### 2012-08-28 b0-llvm.tex
Populating repository

\documentclass{article}

\usepackage[english]{babel}
\usepackage{amssymb,amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{framed}
\usepackage{a4wide}

\title{Translation of B implementations to the LLVM (\textsc{draft})}
\author{} \date{}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\hookleftarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\smallfrown}
\DeclareMathOperator{\isdef}{\blacktriangleright}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\fresh}{\textit{fresh}}
\DeclareMathOperator{\name}{\mathcal{L}()}

\begin{document}

\maketitle

\begin{abstract}
  This document compiles ideas to produce software components to the
  LLVM intermediate format from B implementations. It is a work in
  progress and may be later used as a basis to produce a translator
  from B0 to LLVM.
\end{abstract}

[[[ TODO: PROOF READ ]]]

[[[ SYNTHESIZE TYPES OF PARSE TREE NODES AND THEIR ATTRIBUTES ]]]

\tableofcontents

\section{Introduction}
\label{sec:intro}

The goal of this work is to connect the B method to the LLVM
compilation framework. It is motivated by the fact that many compiling
technologies are implemented in the LLVM framework: optimization,
static analysis, code generation. Providing a connection from the B
method to this framework gives the users of the B method access to
these technologies. Our approach is to define a translation from B0
(the subset of the B language that is used to describe imperative
programs) to the LLVM intermediate format.

\section{Notations and basic definitions}
\label{sec:notations}

The translation is defined by structural recursion on the source
language B0. For each grammatical construct, a code generation
function is defined: $\trad{~}{op}$ for operations, $\trad{~}{inst}$
for instructions, $\trad{~}{exp}$ for expressions and so forth. Each
such function usually produces a text corresponding to bits of LLVM
source code (we use $text$ to denote the type for textual values). It
may have additional parameters and results. In particular, some of
this extra data is a total mapping from B0 identifiers and types to
LLVM text ($dict$ denotes the type of such mapping). Additional
information on those functions will be given below.

We assume that the B0 program has been parsed, and an attributed parse
tree of this program is available. Each node \textsf{n} in the tree
has an attribute \textsf{k} that classifies the node in a syntactic
category. For instance, if \textsf{n.k = BEq}, then \textsf{n}
represents a ``becomes assigned'' instruction. Moreover, for each kind
of node that may appear as an element of a list, there exists an
attribute \textsf{next} that points to the next node in that
list. This is the case for, e.g. instructions (in a block), branches
(in a conditional instruction), names (in a declaration list), etc.
For the last element of that list, the value of this attribute is
$\bot$. Table~\ref{tab:node-attr} summarizes the different kinds of
nodes and their attributes.

\begin{table}
  \begin{center}
    \begin{tabular}[t]{cc}
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Impl} & Implementation & \textsf{sees} : ? \\
        & & \textsf{imports} : ? \\
        & & $\cdots$ \\
        & & \textsf{ops} : Op \\
        \hline
        \textsf{Op} & Operation & \textsf{header}: \textsf{Hd} \\
        & & $\textsf{body} \in \textsf{inst}$ \\
        & & \textsf{next} : \textsf{Op} \\
        \hline
        \textsf{Hd} & Operation header & \textsf{ident} : text \\
        & & \textsf{output} : \textsf{VarDecl} \\
        & & \textsf{input} : \textsf{VarDecl} \\
        \hline
        \textsf{Var} & Variable, parameter & \textsf{ident} : text \\
        & & \textsf{type} : basic type \\
        & & \textsf{next} : \textsf{Var} \\
        \hline
      \end{tabular}
      &
      \begin{tabular}[t]{|l|l|l|}
        \hline
        \textsf{k} & Decription & attributes\\
        \hline
        \hline
        \textsf{Blk} & Block instr. & $\textsf{body} \in \textsf{inst}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{VarDecl} & Variable decl. & \textsf{ident}: \textsf{VarDecl} \\
        & & $\textsf{body} \in \textsf{inst}$ \\
        \hline
        \textsf{If} & If & \textsf{branches}: \textsf{IfBr} \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{IfBr} & If branch & \textsf{cond}: predicate \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & \textsf{next} : \textsf{IfBr} \\
        \hline
        \textsf{BEq} & Becomes equal & \textsf{lhs}: target \\
        & & \textsf{rhs}: expression \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Call} & Call instruction & \textsf{fun}: name \\
        & & \textsf{args}: expression \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{While} & While instr. & \textsf{cond}: predicate \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Assert} & Assert instr. & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{Case} & Case instr. & \textsf{expr}: expression\\
        & & \textsf{branches}: \textsf{CaseBr} \\
        & & $\textsf{next} \in \textsf{Instruction}$ \\
        \hline
        \textsf{CaseBr} & Case branch & \textsf{val}: \textsf{Value} \\
        & & $\textsf{body} \in \textsf{Instr}$ \\
        & & \textsf{next} : \textsf{CaseBr} \\
        \hline
        \textsf{Value} & Value & ??? \\
        & & \textsf{next}: \textsf{Value} \\
        \hline
      \end{tabular} \\
      \\
      \multicolumn{2}{c}{Note: $\textsf{Instr} = \{\textsf{Blk}, \textsf{VarDecl}, \textsf{If}, \textsf{BEq}, \textsf{Call}, \textsf{While}, \textsf{Assert}, \textsf{Case}\}$} \\
    \end{tabular}
    \caption{Structure of the parse tree nodes. For each attribute we give the name of the attribute and
    the type or kind(s) of node that it may be.}
    \label{tab:node-attr}
  \end{center}
\end{table}

Code generation is defined through a set of numbered rules. The
translation rules for some element \textsf{E} of the B0 language are
presented using a pseudo-functional language.  In such rules, the
\textsf{sans serif} and \texttt{typewriter} fonts are used to denote
\textsf{B0} abstract syntax tree nodes and \texttt{LLVM} text
respectively, and $\conc$ denotes string catenation and $\nl$ a new
line. .

Several utility functions are employed throughout those rules.  The
operator $\conc^,$ is the concatenation of two texts separated by a
comma---provided both texts are not empty.
\begin{align}
A \conc^, B \isdef & \textbf{ if } A = \texttt{""} \textbf{ then } B \nonumber \\
& \textbf{ elsif } B = \texttt{""} \textbf{ then } A \nonumber \\
& \textbf{ else } A \conc \texttt{", "} \conc B \nonumber
\end{align}

$\fresh: text \times dict \rightarrow text$ is such that
$\fresh(t, D)$  returns a text $t'$ that is either $t$, if $t$ is not in
the domain of the dictionary $D$, or $t$ followed by the string $n$ of
digits representing the smallest positive number $t \conc n$ does not
belong to the domain of $D$.

We assume that there is an unlimited supply of labels and local variables.
$\name$ returns a text that represents a fresh label or local variable
name for the LLVM currently being produced:
\begin{framed}
\begin{align}
\begin{split}
  &\name : \rightarrow text \\
  &\textbf{let } v : \mathbb{N} = 0 \textbf{ in let } \name = v := v + 1, \textbf{yields } (\textit{text } v)
\end{split}
\end{align}
\end{framed}

\section{A bird's-eye view of the translation}
\label{sec:view}

LLVM programs are structured in \emph{modules}. A LLVM module contains
global \emph{variables}, \emph{functions} and \emph{declarations} of
external entities. The role of variables is to carry some state,
functions may act on that state, and declarations specify the type of
of the external entities.

A B implementation shall then be mapped to a LLVM module.  We also
assume the existence of B library machines that model LLVM variables
that are of a basic type, equipped with operations to read and write
their value.  Other machines, corresponding to more complex types may
also be imported. One may map each section of the B implementation to
the LLVM as summarized in the following table:

\begin{center}
  \begin{tabular}{|c|c|}
    \hline
    B0 & LLVM \\
    \hline
    \hline
    implementation & module \\
    \hline
    imported library machines & global variables \\
    \hline
    other imported machines & other modules \\
    \hline
    operations & functions \\
    \hline
    external references & declarations \\
    \hline
  \end{tabular}
\end{center}

The definition of this translation is structured in four parts:
\begin{description}
\item[control] This part is responsible for describing how B
  operations, and more specifically, B0 instructions, can be mapped to
  LLVM statements.
\item[data] This part is responsible for describing how the data types
  compatible with B0 are translated to LLVM data types.
\item[expressions] This part describes how the evaluation of B0
  expressions and conditions may be codified in LLVM.
\item[modularity] This part is responsible for describing a strategy
  to translate separately the different B modules that may compose a
  B0 implementation to different LLVM source files.
\end{description}

We first define the translation rule for the root of the source
program abstract syntax tree, which shall correspond to a B
implementation. This is a preliminary version where the code
generation is only able to handle standalone B0 implementations, i.e.
the state is defined in the ``concrete variables'' clause. Moreover
the data is restricted to predefined basic types \textsf{INT} and
\textsf{BOOL}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{n.values}{val} \textbf{ and } O_v = \trad{n.vars}{var} \\
  & \quad \textbf{ and } O_{in} = \trad{n.initialization}{ini} \textbf{ and } O_{op} = \trad{n.ops}{ops} \\
  & \quad \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad \trad{i.ops}{ops}  \\
  & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
\end{split}
\end{align}
\end{framed}
Code generation for initialisation and operations is specified in
section~\ref{sec:control}. Code generation for data (type and
state) is defined in section~\ref{sec:data}, and for expression
evaluation in section~\ref{sec:expr}.

\section{Data}
\label{sec:data}

\subsection{Type}
\label{sec:type}
For the time being, the translation only considers two types of
data: integers and booleans. In the case of integers, the translation
is based on the supposition that the B0 definition of \textsf{INT}
is the range $-2^{31}.. 2^{31}-1$:

\begin{framed}
\begin{align}
\begin{split}
  \trad{BOOL}{type} \isdef & \texttt{ "i1"} \\
  \trad{INT}{type} \isdef & \texttt{ "i32"}
\end{split}
\end{align}
\end{framed}

Future version of this document should handle other types of data,
namely integer intervals, enumerated sets, arrays and records (and
any other implementable data type that B0 allows).

\subsection{State}
\label{sec:state}

Concrete variables are mapped to so-called internal variables in the
LLVM program, i.e. private data that is local to the produced
file. The following rule describes the code generation for such
variables:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = internal global "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{type} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

\subsection{Constant values}
\label{sec:values}

Constant valuations are translated to constants in the
LLVM program, as described in the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
  & \textbf{ yields } \\
  & \quad (id \conc \texttt{" = constant "} \conc \trad{v.type}{type} \conc \nl \\
  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
  \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
\end{split}
\end{align}
\end{framed}

\section{Expressions}
\label{sec:expr}

This section presents rules for the generation of LLVM code to
evaluate B0 expressions. These rules are restricted to expressions of
the following basic types: \textsf{BOOL} and \textsf{INT}. Extension
to handle other basic types, arrays and record is left for future
work.

The definition of the code generation is divided as follows:
section~\ref{sec:const} present the base cases (boolean and integer
literals, identifiers), section~\ref{sec:arith} presents code
generation rules for arithmetic expressions, section~\ref{sec:rel} and
section~\ref{sec:cond} is dedicated to relations and conditions.

\subsection{Constants and identifiers}
\label{sec:const}

The following definition provides a mapping from B0 boolean literals
to the corresponding LLVM values:
$$D_{bool} = \{ \textsf{FALSE} \mapsto \texttt{"0"}, \textsf{TRUE} \mapsto \texttt{"1"} \}.$$

Based on this definition, the translation of boolean literals is straightforward:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D_{bool}(\textsf{n.val}), \texttt{"i1"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitBool}
\end{split}
\end{align}
\end{framed}

The translation of integer literals is even simpler, as they have the
same denotation in B0 and LLVM:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, \textsf{n.val}, \texttt{"i32"} \\
  & \textbf{ where } \textsf{n.k} = \textsf{LitInt}
\end{split}
\end{align}
\end{framed}

The last such rule specifies the translation of identifiers. It mainly
consists in a lookup in the translation dictionary, and shall return
also the corresponding LLVM data type:

\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D(\textsf{n}), \trad{n.type}{type} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Ident}
\end{split}
\end{align}
\end{framed}

\subsection{Arithmetic operations}
\label{sec:arith}

The following definition maps B0 binary operators to LLVM binary statements:

$$D_{bari} = \{ + \mapsto \texttt{"add"}, 
- \mapsto \texttt{"sub"}, * \mapsto \texttt{"mul"}, / \mapsto
\texttt{"sdiv"}, \% \mapsto \texttt{"srem"} \}.$$

Based upon this definition, the following rule specifies the
generation of LLVM code responsible for the evaluation of binary
arithmetic expressions (plus the name of the local variable storing
the resulting value, and of its LLVM type):
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } v = \mathcal{V} \textbf{ and } op = D_{bari}(\textsf{n.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad v \conc \texttt{" = i32 "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} \in \dom D_{bari} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Next rule specifies the code generation for the successor and
predecessor operators. As LLVM has no corresponding statement, the
generated code uses addition and subtraction:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 add 1, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{succ} \\
  & \textbf{ assert } O_t = \texttt{i32} \\
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub "} \conc O_l \conc \texttt{", 1"} \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{pred} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

The following rule specifies code generation for the unary
minus. Since LLVM does not have a specific statement for such
operation, the translation uses subtraction:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad v \conc \texttt{" = i32 sub 0, "} \conc O_l \conc \nl), \\
  & \quad v, \texttt{i32} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Arith} \textbf{ and } \textsf{n.op} = \textsf{-} \\
  & \textbf{ assert } O_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

Also, LLVM does not have a predefined operation for the ``power''
operator provided by B0. The corresponding rule introduces a call
to a function, named \texttt{@power} that implements this operator (the LLVM source code
for such function is not presented herein):
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{n.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \mathcal{V} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = call i32 @power(i32, "} \conc O_l \conc \texttt{", i32 "} \conc O'_l \conc \texttt{")"}) \conc \nl), \\
  & \quad l, \texttt{"i32"} \\
  & \textbf{ where } \textsf{r.k} = \textsf{Power} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

\subsection{Relations}
\label{sec:rel}

The LLVM statement \texttt{icmp} compares integers or vectors of integers;
and the \texttt{fcmp} statement compares floats of vectors of floats. The
translation produces \texttt{icmp} statements such as:
\begin{verbatim}
  %5 = icmp slt i32 %2, %4
\end{verbatim}

We first define a mapping from B relational operators to LLVM
keywords:
$$D_{op} = \{ = \mapsto \texttt{"eq"}, 
  \ne \mapsto \texttt{"ne"}, 
  < \mapsto \texttt{"slt"}, 
  \le \mapsto \texttt{"sle"}, 
  > \mapsto \texttt{"sgt"}, 
  \ge \mapsto \texttt{"sge"} \}.$$

\begin{framed}
\begin{align}
\begin{split}
  \trad{r, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
  \trad{r.left, $D$}{exp} \textbf{ and }
  O'_p, O'_l, O'_t = \trad{r.right, $D$}{exp} \\
  & \quad \textbf{ and } l = \mathcal{V} \textbf{ and } op = D_{op}(\textsf{r.op}) \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad (O_p \conc \\
  & \quad O'_p \conc \\
  & \quad l \conc \texttt{" = icmp "} \conc op \conc \texttt{" "} \conc O_t \conc \texttt{" "} \conc O_l \conc^, O'_l \conc \nl), \\
  & \quad l \\
  & \textbf{ where } \textsf{r.k} = \textsf{BinRel} \\
  & \textbf{ assert } O_t = O'_t = \texttt{i32}
\end{split}
\end{align}
\end{framed}

\subsection{Conditions}
\label{sec:cond}

The translation of conditions takes as input a condition node
\textsf{n}, a dictionary $D$, as well as two labels $L_t$ and $L_f$
that correspond to program locations where the execution shall jump
when the condition is evaluted to true or false, respectively.

The first rule defines the translation of an atomic relation:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{cond} : & \textsf{ Node} \times dict \times text \times text \rightarrow text \\
  \trad{n, $D$, $L_t$, $L_f$}{cond} \isdef
  & \textbf{ let } O_p, O_l = \trad{n, $D$}{rel} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad O_p \conc \\
  & \quad \texttt{br i1 } \conc O_l \conc \texttt{" , label \%"} \conc L_t \conc \texttt{", label \%"} \conc L_f \conc \nl  \\
  & \textbf{ where } \textsf{n.k} = \textsf{BinRel}
\end{split}
\end{align}
\end{framed}

Next is the rule for negations:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$, $L_t$, $L_f$}{cond} \isdef
  & \textbf{ yields} \trad{n.arg, $D$, $L_f$, $L_t$}{cond} \\
  & \textbf{ where } \textsf{n.k} = \textsf{Neg}
\end{split}
\end{align}
\end{framed}

The following rule handles the case of conjunctions. Note that it produces code
with a ``short-cut'' when the evaluation of the first argument yields false:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$, $L_t$, $L_f$}{cond} \isdef
  & \textbf{ let } L_i = \name \textbf{ in } \\
  & \textbf{ let } O_l = \trad{n.left, $D$, $L_i$, $L_f$}{cond} \textbf{ and } \trad{n.right, $D$, $L_t$, $L_f$}{cond} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad O_l \conc \\
  & \quad \texttt{; <label>:} \conc L_i \conc \nl \conc \\
  & \quad O_r \\
  & \textbf{ where } \textsf{n.k} = \textsf{Conj}
\end{split}
\end{align}
\end{framed}

The last rule for conditions handles disjunctions. It also produces
code with a short-cut when the first argument evaluates to true:
\begin{framed}
\begin{align}
\begin{split}
  \trad{n, $D$, $L_t$, $L_f$}{cond} \isdef
  & \textbf{ let } L_i = \name \textbf{ in } \\
  & \textbf{ let } O_l = \trad{n.left, $D$, $L_t$, $L_i$}{cond} \textbf{ and } \trad{n.right, $D$, $L_t$, $L_f$}{cond} \textbf{ in } \\
  & \textbf{ yields} \\
  & \quad O_l \conc \\
  & \quad \texttt{; <label>:} \conc L_i \conc \nl \conc \\
  & \quad O_r \\
  & \textbf{ where } \textsf{n.k} = \textsf{Conj}
\end{split}
\end{align}
\end{framed}

\section{Control}
\label{sec:control}

This section addresses how B operations are translated to LLVM
functions. Each operation is translated in turn, and the result of
the translation of individual operations is concatenated to yield the
result of the translation of the operations clause:
\begin{framed}
\begin{align}
  \begin{split}
  \trad{}{ops} : & \textsf{ Node} \rightarrow text \\
  \trad{o}{ops} \isdef 
  & \textbf{ yields } \trad{o}{op} \conc \trad{o.next}{ops}  \\
  & \textbf{ where } \textsf{o} \neq \bot  \textbf{ and } \textsf{o.k} \in \textsf{Op}  \\
  \trad{$\bot$}{ops} \isdef
  & \textbf{ yields } \mty 
  \end{split}
\end{align}
\end{framed}

Three main aspects of the translation are discussed presented, namely
the signature of the operation, information local to the operation,
and the instructions composing the body of the operation. The
translation of the types of local data, as well as that of the
expressions is deferred to sections~\ref{sec:data} and~\ref{sec:expr},
respectively.

One can distinguish three elements in a LLVM function: the signature,
memory allocation for local variables, and blocks of LLVM
statements\footnote{In an attempt to clarify the exposition of our
  ideas, we will call \emph{instructions} the constituents of a B0
  operation, and \emph{statements} those of a LLVM function.}.

To map operations to LLVM functions, one needs to consider the
following:
\begin{itemize}
\item The signature of the function is composed of the name of the
  function, the result type, the name and type of the parameters. The
  signature may carry other information such as linkage, visibility,
  calling convention, etc. The B parser provides this information after
  the type checking phase has been completed.

  As usual in programming languages, LLVM functions may only have one
  result. B0 operations may produce several results. These results are
  also known as output parameters of the operation. The translation of
  output parameters will be function parameters passed by reference,
  i.e. they are pointers. The return type of the LLVM functions
  resulting from the translation of B operations will always be
  \texttt{void}. Translation of operation headers is defined in
  section~\ref{sec:trad-header}

\item The memory to store local variables is allocated on the stack,
  using the statement \verb+alloca+, requiring the type of value that
  need to be stored as well as any memory alignment requirement. The
  Clang compiler places all the stack allocation statements in the
  first block of in the routine. To replicate such behaviour, one
  needs to implement a method to collect the name and type of all the
  local variables used in the operation. Also, the names have to be
  translated to conform to LLVM restrictions. All the data structures
  necessary to implement (straightforwardly) such procedure are
  available in the abstract syntax tree. This part of the translation 
  is defined in section~\ref{sec:trad-alloc}.

\item Finally, the body of the operation has to be transformed to a
  sequence of LLVM blocks of statements. For each kind of instruction
  one (or more) translation function is defined. The details are provided
  in section~\ref{sec:trad-instr}.

\end{itemize}

The translation of an operation is defined by the following rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{op} &: \textsf{ Node} \rightarrow text \\
  \trad{op}{op} \isdef
  & \textbf{ let } O_l = \trad{op}{link} \textbf{ and } O_h, D_h = \trad{op.header}{header} \textbf{ in}  \\
  & \textbf{ let } O_a, D_a = \trad{op.body, $D_h$}{alloc} \textbf{ in}  \\
  & \textbf{ let } O_i = \trad{op.body, $D_a$}{inst} \textbf{ in}  \\
  & \textbf{ yields } \texttt{"declare "} \conc O_l \conc O_h \conc \texttt{"\{"} \conc \nl \conc \\
  & \quad \conc O_a \conc \nl \conc O_i \conc \nl \conc \texttt{"ret void"} \conc \nl \conc \texttt{"\}"} \conc \nl  \\
  & \textbf{ where } \textsf{op} \ne \bot \textbf{ and } \textsf{op.k} = \textsf{Op} 
\end{split}
\end{align}
\end{framed}

The following rule produces the LLVM calling convention keyword
\texttt{internal} that guarantees that the translated B0 local
operations are not visible outside the module:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{link} &: \textsf{ Node} \rightarrow text \\
  \trad{op}{link} \isdef
  & \textbf{ yields } \texttt{"internal "} \\
  & \textbf{ where } \textsf{op} \ne \bot \textbf{ and } \textsf{op.LocOp} = \top \\
  \trad{op}{link} \isdef
  & \textbf{ yields } \texttt{""} \\
  & \textbf{ where } \textsf{op} \ne \bot \textbf{ and } \textsf{op.LocOp} \ne \top
\end{split}
\end{align}
\end{framed}

\subsection{Sub-routines signatures}
\label{sec:trad-header}

The following rules specify the translation of operation headers. It
is defined through four different patterns, depending on the presence
of input and output parameters. The name of the LLVM function is the
name of the B0 operation prefixed by \mbox{@}. The return type is
always \texttt{void} and the parameter list is made of the translation
of the input and output parameters:

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{header} &: Hd \rightarrow text \times dict \times dict \\
    \trad{h}{header} \isdef& \textbf{ let } T_n = \textsf{"@"} \conc \textsf{h.ident} 
    \textbf{ and } T_i, D_i = \trad{h.input}{inp} 
    \textbf{ and } T_o, D_o = \trad{h.output}{out} \textbf{ in}  \\
    & \textbf{ let } T_p = T_i \conc^, T_o \textbf{ in}  \\
    & \textbf{ yields } \texttt{"void "} \conc T_n \conc \texttt{"("} \conc T_p \conc
    \texttt{")"}, D_i \cup D_o, \{ \textsf{h.ident} \mapsto T_n \}  \\
    & \textbf{ where } \textsf{h} \ne \bot \textbf{ and } \textsf{h.k} = \textsf{Hd} 
  \end{split}
\end{align}
\end{framed}

Next are the rules that specify recursively the translation of input
parameters. The first pattern is the base case and the second is the
case for parameter lists with two or more elements. The name of the
parameter is the name of the B0 parameter prefixed with \texttt{\%}.
The type of the parameter is translated through specific rules 
defined in section~\ref{sec:data}:

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{inp} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{i}{inp} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{i.ident} 
    \textbf{ and } T_t = \trad{i.type}{type} 
    \textbf{ and } T, D = \trad{i.next}{inp} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{i} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var}  \\
    \trad{$\bot$}{inp} \isdef& \textbf{ yields } \mty, \emptyset 
  \end{split}
\end{align}
\end{framed}

The following rules describe how output parameters are translated. It
is similar to the rules for input parameters. The only difference is
that parameters need to be of pointer type.

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{out} &: \textsf{ Node } \rightarrow text \times dict \\
    \trad{o}{out} \isdef& \textbf{ let } T_n = \textsf{"\%"} \conc \textsf{o.ident} 
    \textbf{ and } T_t = \trad{o.type}{type} \conc \texttt{"*"}
    \textbf{ and } T, D = \trad{o.next}{out} \textbf{ in}  \\
    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{o} \mapsto T_n \} \cup D  \\
    & \textbf{ where } \textsf{o} \ne \bot \textbf{ and } \textsf{o.k} = \textsf{Var}  \\
    \trad{$\bot$}{out} \isdef& \textbf{ yields } \mty, \emptyset  
  \end{split}
\end{align}
\end{framed}

\subsection{Stack allocation}
\label{sec:trad-alloc}

The generation of stack allocation statements processes the parse tree
in a pre-order depth-first traversal, generating one statement for
each local variable declaration found.

\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{alloc} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
    \trad{i, $D$}{alloc} \isdef 
    & \textbf{ let } O_b, D_b = \trad{n, $D$}{alloc}_{base}  \\
    & \textbf{ and } O_n, D_n = \trad{i.next; $D_b$}{alloc}  \\
    & \textbf{ yields } O_b \conc \nl \conc O_n, D_n  \\
    & \textsf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{Blk}, \textsf{VarDecl}, \textsf{If}, \textsf{IfBr}, \textsf{While}, \textsf{Case}, \textsf{CaseBr} \}  \\
    \trad{i, $D$}{alloc} \isdef 
    & \textbf{ let } O_n, D_n = \trad{i.next; $D$}{alloc}  \\
    & \textbf{ yields } O_n, D_n  \\
    & \textsf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \not\in \{\textsf{Blk}, \textsf{VarDecl}, \textsf{If}, \textsf{IfBr}, \textsf{While}, \textsf{Case}, \textsf{CaseBr} \}  \\
    \trad{$\bot$, $D$}{alloc} \isdef & \textbf{ yields } \mty, D 
  \end{split}
\end{align}
\end{framed}

One must avoid generating name conflicts in the LLVM output. For each
B0 variable \textsf{v} found, a fresh LLVM name \texttt{l} is thus
created. This association is recorded in the $D$ parameters and results.
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{alloc}_{base} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.body, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{Blk}, \textsf{While}, \textsf{IfBr}, \textsf{CaseBr} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ yields } \trad{i.branch, $D$}{alloc}  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{\textsf{If}, \textsf{Case} \}  \\
    \trad{i, $D$}{alloc}_{base} \isdef 
    & \textbf{ let } O_l, D_l = \trad{i.ident, $D$}{local} \textbf{ in}  \\
    & \textbf{ let } O_b, D_b = \trad{i.body, $D_l$}{alloc} \textbf{ in}  \\
    & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} 
  \end{split}
\end{align}
\end{framed}

To illustrate how heap allocation works in LLVM, this is the statement
produced by Clang corresponding to the C declaration
\verb^unsigned short c1^:
\begin{verbatim}
  %c1 = alloca i16
\end{verbatim}
To generate such statement, one needs to create a name
(e.g. \texttt{\%c1}) for the location, and to identify the
corresponding LLVM type (e.g. \texttt{i16}). This is the role of the
last translation function in this section. It produces the desired
source code text and enriches the translation dictionary with the
association between the B0 variable and the created LLVM identifier. It
relies upon an auxiliary translation function $\trad{}{type}$ to be
defined hereafter:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{local} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
  \trad{v, $D$}{local} \isdef & 
  \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v} \mapsto T_v \}$}{local} \textbf{ in}  \\
  & \textbf{ yields } T_v \conc \texttt{" = alloca "} \conc \trad{v.type}{type} \conc \nl \conc T_n, D_n  \\
  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Var}  \\
  \trad{$\bot$, $D$}{local} \isdef & 
  \textbf{ yields } \mty, D 
\end{split}
\end{align}
\end{framed}

Following the allocation statement, the LLVM source code contains the
statements that encodes the instructions in the body of the
operation. It is the role of the next section to present how this
translation can be defined.

\subsection{Control flow}
\label{sec:trad-instr}

The translation function $\trad{}{inst}$ shall produce one labeled,
sequential, block of LLVM statements for each of the following in the
B operation: maximal sequences of instructions that do not contain
branching (i.e. conditionals and loops), and evaluation of branching
condition. In the target code, each block has a unique label that
identifies its first statement. The translator has an unbounded supply
of such labels; recall that a new valid LLVM label may be created with
the expression $\name$.

The translation includes parameters to carry additional information
necessary to the translation. One such parameter is a label for the
next instruction to be translated. If its value is $\bot$ then it is
considered missing. Otherwise, it is a valid LLVM label and it shall
be placed in the LLVM output. An auxiliary definition is helpful to
handle the production of labels:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{inst}_{L} &: Instr \times dict \times text \rightarrow text \\
  \trad{i, $D$, $L$}{inst}_L \isdef 
  & \textbf{yields } L \conc \texttt{:} \trad{i, $D$}{inst}  
\end{split}
\end{align}
\end{framed}

The signature of the translation function for instructions is
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{inst} &: Instr \times dict \rightarrow text
  \end{split}
\end{align}
\end{framed}
In the forthcoming sections, the definition of this function is given
piecewise, according to the kind of instruction being translated. One
special case is the absence of node representing an instruction:
\begin{framed}
  \begin{align}
\begin{split}
    \trad{$\bot$, $D$}{inst} \isdef \textbf{ yields } \texttt{""}
  \end{split}
\end{align}
\end{framed}

\subsubsection{Block}

Block instructions are structuring constructs that have no operational
intent. The translation is thus defined as the concatenation of the
translation of the block body and the translation of the following
statement.

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{let } O_b = \trad{i.body, $D$}{inst} 
  \textbf{ and } O_n = \trad{i.next, $D$}{inst} \textbf{ in} \\
  & \textbf{yields } O_b \conc O_n  \\
  & \textbf{where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Blk} 
\end{split}
\end{align}
\end{framed}

\subsubsection{If}

To translate an if statement \textsf{i}, each branch is translated in
turn. We assume that each branch has a condition, a body, and a
successor. In the case of an ``else'' branch, the condition is $\bot$.
It is the role of $\trad{}{br}$ to produce the translation for such
branches. Note that at the end of the translation of the body of each
branch, a jump to the instruction following the if statement must be
included. Thus a label $L$ is generated and is a parameter of the
branch translation. It will be used to label the statement
\textsf{i.next}:
\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{let } L = \name \textbf{ in} \\
  & \textbf{let } O_b = \trad{i.branch, $D$, $L$}{br} \textbf{ and }
  O_n = \trad{i.next, $D$, $L$}{inst}_L \textbf{ in}  \\
  & \textbf{yields } O_b \conc O_n  \\
  & \textbf{where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{If} 
\end{split}
\end{align}
\end{framed}

Now, to translate the branches of an if instruction, one must first
generate code to evaluate the condition of the branch, followed by an
LLVM conditional branching statement translate, and then translate the
body of the branch. In LLVM, a conditional statement is as follows:
\begin{verbatim}
  br i1 %5, label %6, label %10
\end{verbatim}
where \texttt{\%5} is a local variabel of width one bit (since its
type is \texttt{i1}), storing the value of the condition evaluation,
and \texttt{\%6} and \texttt{\%10} are statement labels.

The first rule consider branches that have non-void condition and
sucessor.  Two labels $L_1$ and $L_2$ are created: $L_1$ is the label
of the branch body, and $L_2$ is the label where the code responsible
for the execution of the remaining branches is located. $O_p$ is the
LLVM code responsible for evaluating the condition, $O_l$ being the
LLVM local variable where the result of the evaluation is stored,
$O_b$ and $O_n$ stand for the LLVM code sections responsible for
executing the branch body and the remaining branches, respectively.
Statement block $O_c$ evaluates the condition, where $O_t$ is the name
of the temporary register storing its value in the LLVM:

\begin{framed}
\begin{align}
\begin{split}
  \trad{b, $D$, $L$}{br} : & \textsf{ Node} \times dict \times text \rightarrow text \\
  \trad{b, $D$, $L$}{br} \isdef 
  & \textbf{ let } L_1 = \name \textbf{ and } L_2 = \name \textbf{ in}  \\
  & \textbf{ and } O_p, O_l = \trad{b.cond, $D$, $L_1$, $L_2$}{cond}  \\
  & \textbf{ and } O_b = \trad{b.body, $D$}{inst} \textbf{ and }
  O_n = \trad{b.next, $D$, $L$}{br} \textbf{ in}  \\
  & \textbf{ yields }  \\
  & \quad O_p \conc \nl \conc  \\
  & \quad \texttt{"br i1 "} \conc O_l \conc \texttt{", label \%"} \conc L_1 \conc \texttt{" , label \%"} \conc L_2 \conc \nl \conc  \\
  & \quad L_1 \conc \texttt{:} \conc \nl \conc  \\
  & \quad O_b \conc  \\
  & \quad \texttt{"br "} \conc L \conc \nl \conc  \\
  & \quad L_2 \conc \texttt{":"} \conc \nl \conc  \\
  & \quad O_n  \\
  & \textbf{ where } \textsf{b} \ne \bot \textbf{ and } \textsf{b.k} = \textsf{IfBr} \textbf{ and } \textsf{b.cond} \ne \bot \textbf{ and } \textsf{b.next} \ne \bot 
\end{split}
\end{align}
\end{framed}

The second case is that of the last branch of the if instruction, when
this branch has a condition (i.e. it is an ``elsif''). It is a simplified
version of the previous rule:
\begin{framed}
\begin{align}
\begin{split}
  \trad{b, $D$, $L$}{br} \isdef 
  & \textbf{ let } L_1 = \name \textbf{ and } O_p, O_l, O_t = \trad{b.cond, D}{exp}  \\
  & \textbf{ and } O_b = \trad{b.body, $D$}{inst} \textbf{ in} \\
  & \textbf{ yields }  \\
  & \quad O_p \conc \nl \conc  \\
  & \quad \texttt{"br i1 "} \conc O_l \conc \texttt{", label \%"} \conc L_1 \conc \texttt{" , label \%"} \conc L \conc \nl \conc  \\
  & \quad L_1 \conc \texttt{:} \conc \nl \conc  \\
  & \quad O_b \conc  \\
  & \quad \texttt{"br "} \conc L \conc \nl \conc  \\
  & \textbf{ where } \textsf{b} \ne \bot \textbf{ and } \textsf{b.k} = \textsf{IfBr} \textbf{ and } \textsf{b.next} = \bot  \textbf{ and } \textsf{b.cond} \ne \bot  \\
  & \textbf{ assert } O_t = \texttt{i1} 
\end{split}
\end{align}
\end{framed}

The last case is even simpler; it applies to ``else'' branches:
\begin{framed}
\begin{align}
\begin{split}
  \trad{b, $D$, $L$}{br} \isdef 
  & \textbf{let } O_b = \trad{b.body, $D$}{inst} \textbf{ in} \\
  & \textbf{yields }  \\
  & \quad O_b \conc  \\
  & \quad \texttt{"br "} \conc L \conc \nl \conc  \\
  & \textbf{where } \textsf{b} \ne \bot \textbf{ and } \textsf{b.k} = \textsf{IfBr} \textbf{ and } \textsf{b.next} = \bot  \textbf{ and } \textsf{b.cond} = \bot 
\end{split}
\end{align}
\end{framed}

\subsubsection{Variable declaration}

The static memory allocation has been produced with the $\trad{}{alloc}$
operator. The rest of the translation consists in translating the body
of the variable declaration:

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{ yields } \trad{i.body, $D$}{inst} \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} 
\end{split}
\end{align}
\end{framed}

\subsubsection{Identity}

If $\textsf{i.k} = \textsf{Id}$, then \textsf{i} has no effect:

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{ yields } \mty \label{trad:id} \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} \in \{ \textsf{Id}, \textsf{Assert} \} 
\end{split}
\end{align}
\end{framed}

\subsubsection{Becomes equal}

In LLVM, assignment is performed with the \texttt{store} statement, e.g.:
\begin{verbatim}
store i16 %17, i16* %c1
\end{verbatim}
where \texttt{\%17} is the local variable storing the assigned value,
\texttt{\%c1} is the local variable being assigned, and \texttt{i16}
is the type of the assigned value.

The translation generates a sequence of statements $O_l$ to evaluate
the left-hand side in local variable $L_l$ , followed by a sequence of
statements $O_r$ to evaluate the right-hand side in the local variable
$L_r$, followed by a single \verb'store' statement to perform the
assignment:

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{ let } O_l, L_l, T_l = \trad{i.lhs, $D$}{lv} \textbf{ and } O_r, L_r, T_r = \trad{i.rhs, $D$}{exp}  \textbf{ in} \\
  & \textbf{ yields }  \\
  & \quad O_l \conc \nl  \\
  & \quad O_r \conc \nl  \\
  & \quad \texttt{"store "} \conc T_r \conc \texttt{" "} \conc L_r \conc \texttt{", "}
  \conc T_l \conc \texttt{" "} \conc L_l \conc \nl  \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{BEq}  \\
  & \textbf{ assert } T_l = T_r \conc \texttt{"*"} 
\end{split}
\end{align}
\end{framed}

$\trad{}{lv}$ is responsible for the translation of assignment
targets, also known as \emph{lvalues} in C. The different kinds of
targets are scalar data, array data, array element and field of a
record. Note that, in the LLVM, arrays are so-called first-class
types, and assignment statements may target directly an
array. Considering the restrictions put on the handled B0 elements,
assignment targets shall only be identifiers:

\begin{framed}
\begin{align}
\begin{split}
  \trad{}{lv} &: Node \times dict \rightarrow text \times text \times text \\
  \trad{n, $D$}{lv} \isdef 
  & \textbf{ let } T = \trad{n.type}{type} \textbf{ in }  \\
  & \textbf{ yields } \texttt{""}, D(\textsf{n.ident}), T \conc \texttt{*} \\
  & \textbf{ where } \textsf{n} \ne \bot \textbf{ and } \textsf{n.k} = \textsf{Ident}
\end{split}
\end{align}
\end{framed}

\subsubsection{While}

For the ``while'' instruction the translation creates three statement
labels: $L_1$ is the start of the block that evaluates the loop condition;
$L_2$ is the start of the block that executes the loop body; $L_3$ labels
the exit of the loop, i.e. the next statement. $O_p$ is the
LLVM code responsible for evaluating the condition, $O_l$ being the
LLVM local variable where the result of the evaluation is stored,
$O_b$ and $O_n$ stand for the LLVM code sections responsible for
executing the loop body and the following statements:

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{ let } L_1 = \name \textbf{ and } L_2 = \name \textbf{ and } L_3 = \name \textbf{ in} \\
  & \textbf{ let } O_p, O_l = \trad{i.cond, $D$, $L_2$, $L_3$}{cond} \textbf{ and } O_b = \trad{i.body, $D$}{inst}  \\
  & \textbf{ and } O_n = \trad{i.next, $D$, $L_3$}{inst} \textbf{ in} \\
  & \textbf{ yields }  \\
  & \quad L_1 \conc \texttt{":"} \conc \nl \conc  \\
  & \quad O_p \conc \nl \conc  \\
  & \quad \texttt{"br i1 "} \conc O_l \conc \texttt{", label \%"} \conc L_2 \conc \texttt{" , label \%"} \conc L_3 \conc \nl \conc  \\
  & \quad L_2 \conc \texttt{:} \conc \nl \conc  \\
  & \quad O_b \conc  \\
  & \quad \texttt{"br "} \conc L_1 \conc \nl \conc  \\
  & \quad O_n  \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{While}  \\
  & \textbf{ assert } O_t = \texttt{"i1"} 
\end{split}
\end{align}
\end{framed}

\subsubsection{Callup}

LLVM has a statement named \texttt{call} for sub-routine calls. For
instance, the following statement has been generated by the Clang
compiler for a call to a C function prototyped as
\verb^void acc(int*, int)^:
\begin{verbatim}
call void @acc(i32* %a, i32 %4)
\end{verbatim}
By convention, each B0 operation are translated to a LLVM with no
return value, and where the input and output parameters of the
operation are translated to parameters passed by value and by
reference, respectively.

Auxiliary definitions $\trad{}{argval}$ and $\trad{}{argref}$ produce
snippets of LLVM code responsible for passing argument by value and
by reference, respectively. The result is a pair of texts, first is
the code for evaluating the arguments, and second is the comma-separated
list of typed arguments:
\begin{framed}
\begin{align}
\begin{split}
  \trad{}{argval} : & \textsf{ Node} \times dict \rightarrow text \times text \times text \\
  \trad{i, $D$}{argval} \isdef 
  & \textbf{ let } O_p, O_v, O_t = \trad{i, D}{exp} \textbf{ and } O'_p, O_r = \trad{i.next, D}{argval} \textbf{ in}  \\
  & \textbf{ yields }  \\
  & \quad O_p \conc O'_p, (O_t \conc \texttt{" "} \conc O_v) \conc^, O_r  \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Call}  \\
  \trad{$\bot$, $D$}{argval} \isdef & \texttt{ ""},\texttt{ ""},\texttt{ ""}  \\
  \trad{}{argref} : & \textsf{ Node} \times dict \rightarrow text \times text \times text \\
  \trad{i, $D$}{argref} \isdef 
  & \textbf{ let } O_p, O_v, O_t = \trad{i, D}{exp} \textbf{ and } O'_p, O_r = \trad{i.next, D}{argref} \textbf{ in}  \\
  & \textbf{ yields }  \\
  & \quad O_p \conc O'_p, (O_t \conc \texttt{"* "} \conc O_v) \conc^, O_r  \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Call}  \\
  \trad{$\bot$, $D$}{argref} \isdef & \texttt{ ""},\texttt{ ""},\texttt{ ""} 
\end{split}
\end{align}
\end{framed}

The result of these auxiliary routines are combined to form the LLVM code
for the whole function call:

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{ let } O_p, O_v = \trad{i.in, D}{argval} \textbf{ and } O'_p, O_r = \trad{i.ou, D}{argref} \textbf{ in} \\
  & \textbf{ yields }  \\
  & \quad O_p \conc O'_p \conc  \\
  & \quad \texttt{"call void "} \conc \trad{i.op, $D$}{exp} \conc \texttt{"("} \conc O_v \conc^, O_r \conc \texttt{")"} \nl  \\
  & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Call} 
\end{split}
\end{align}
\end{framed}

\subsubsection{Case}

This instruction shall be translated to a \verb'switch' statement.
For the sake of simplicity, we assume that case statement without
else branch are changed so that there is such a branch, with a single
identity instruction.

The LLVM switch statement is basically a jump table with the following
concrete syntax (example):
\begin{verbatim}
switch i32 %val, label %otherwise [ i32 0, label %onzero
                                    i32 1, label %onone
                                    i32 2, label %ontwo ]
\end{verbatim}

The translation of a case statement shall thus produce a labeled block
for each branch, and a jump for each value in each branch. We first
provide an auxiliary definition that yields a mapping from branches to
labels.
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{lab} : & \textsf{ Node} \rightarrow \textsf{Node} \rightarrow text \\
    \trad{br}{lab} \isdef 
    & \textbf{ yields } \{ \textsf{br} \mapsto  \name \} \cup
    \trad{br.next}{lab}  \\
    & \textbf{ where } \textsf{br} \ne \bot \textbf{ and } \textsf{br.k} = \textsf{CaseBr}  \\
    \trad{$\bot$}{lab} \isdef & \textbf{ yields } \{\} 
  \end{split}
\end{align}
\end{framed}

Next we provide definitions that yield the jump table for all
but the else branch. First, $\trad{}{jump}$ takes as input a
case branch node, and the mapping to labels and recurses over
all branches and branch values:
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{jump} : & \textsf{ Node} \times (\textsf{Node} \rightarrow text) \rightarrow text \\
    \trad{br, $lab$}{jump} \isdef
    & \textbf{ yields }  \\
    & \quad \trad{br.val, $lab($br$)$}{jump}_v \conc  \\
    & \quad \trad{br.next, $lab$}{jump}  \\
    & \textbf{ where } \textsf{br} \ne \bot \textbf{ and } \textsf{br.k} = \textsf{CaseBr}  \\
    \trad{$\bot$, $lab$}{jump} \isdef
    & \textbf{ yields } \texttt{""} 
  \end{split}
\end{align}
\end{framed}

$\trad{}{jump}_v$ is applies to branch values, and has as second argument a single label $l$:
\begin{framed}
  \begin{align}
\begin{split}
    \trad{}{jump}_v : & \textsf{ Node} \times text \rightarrow text \\
    \trad{val, $l$}{jump}_v \isdef
    & \textbf{ yields }  \\
    & \quad \trad{val.type}{type} \conc \texttt{" "} \conc 
    \trad{val.value}{value} \conc \texttt{", label \%"} \conc l \conc \nl \conc \\
    & \quad \trad{val.next, $l$}{jump}  \\
    & \textbf{ where } \textsf{val} \ne \bot \textbf{ and } \textsf{val.k} = \textsf{Value}  \\
    \trad{$\bot$, $l$}{jump}_v \isdef
    & \textbf{ yields } \texttt{""} 
  \end{split}
\end{align}
\end{framed}

The third and last auxiliary definition $\trad{}{blocks}$ is
responsible for the generation of the statement blocks that correspond
to the branches of a case statement. The parameters are the branch
itself, the variable dictionnary $D$, the branch labeling function
$lab$ and a label $l_f$ where the execution must proceed after
completing the branch:
\begin{framed}
\begin{align}
\begin{split}
  \trad{br, $D$, $lab$, $l_f$}{block} \isdef 
  & \textbf{yields } \\
  & \quad lab(\textsf{br}) \conc \texttt{":"} \conc \nl \conc  \\
  & \quad \trad{br.body, $D$}{inst} \conc  \\
  & \quad \texttt{"br "} \conc l_f \conc \nl \conc  \\
  & \quad \trad{br.next, $D$, $lab$, $l_f$}{block}  \\
  & \textbf{where } \textsf{br} \ne \bot \textbf{ and } \textsf{br.k} = \textsf{CaseBr}  \\
  \trad{$\bot$, $D$, $lab$, $l_f$}{block} \isdef 
  & \textbf{yields } \texttt{""}
\end{split}
\end{align}
\end{framed}

Finally, the following definition details the translation of the case
instruction. First, the tested expression is evaluated, then a switch
statement is emitted, including the jump table, followed by the
translation of the different branches and ended with a label where all
branches eventually jump to.

\begin{framed}
\begin{align}
\begin{split}
  \trad{i, $D$}{inst} \isdef 
  & \textbf{let } lab = \trad{i.branches}{lab} \textbf{ and } p_e, l_e, t_e = \trad{i.exp, D}{exp} \textbf{ in} \\
  & \textbf{let } l_d, l_f = \name, \name \textbf{ and } 
  jumps = \trad{i.branches, $lab$}{jump} \textbf{ and }  \\
  & \quad blocks = \trad{i.branches, $D$, $lab$, $l_f$}{block} \textbf{ in}  \\
  & \textbf{yields }  \\
  & \quad p_e \conc  \\
  & \quad \texttt{"switch "} \conc t_e \conc \texttt{" "} \conc l_e \conc 
  \texttt{ default } \conc l_d \conc \texttt{"["} \conc \nl  \\
  & \quad \quad jumps \conc \texttt{"]"} \conc \nl  \\
  & \quad blocks \conc  \\
  & \quad l_f \conc \texttt{:} \conc \nl  \\
  & \textbf{where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Case}
\end{split}
\end{align}
\end{framed}

As in the case of the if statement, the translation of the following 
instruction initiates a new block, labeled with say \textit{next}. 
Each branch of the case is translated to a block that ends with
an unconditional jump \verb'br label' $\textit{next}$.

% Case_instruction ::=
% ``CASE'' Simple_term ``OF''
% ``EITHER'' Simple_term +'','' ``THEN'' Instruction ( ``OR'' Simple_term +'','' ``THEN'' Instruction )* [ ``ELSE'' Instruction ]
% ``END''
% ``END''

% The CASE substitution is used to define for a program, various
% possible behaviors depending on the value of an expression. Each
% EITHER and OR branch is made up of a non empty list of constants. The
% values of constants in the set of branches must be distinct two by
% two. If the value of the expression belongs to one of the branches,
% then the substitution of this branch is executed. If not, the
% substitution in the ELSE branch is applied, if this latter branch is
% absent, it will by default perform the identity substitution. The
% behavior of this substitution is therefore deterministic and always
% feasible.

%  ; Implement a jump table:
%  switch i32 %val, label %otherwise [ i32 0, label %onzero
%                                      i32 1, label %onone
%                                      i32 2, label %ontwo ]

\subsubsection{Assert}

This instruction has no operational meaning and it needs not be
translated (it might have some effect on the control flow graph, or
may be translated for debugging purposes --- see debugging
intrinsics). Rule~\ref{trad:id} handles this case.

\section{Modularity}

\subsection{The SEES clause}

Suppose a machine $M$ is seen from the implementation being translated
to LLVM. For each constant, set, variable, and operation found in $M$,
the LLVM code must include a corresponding declaration. As an
illustration, consider the following C code:
\begin{verbatim}
extern int f(int, int, int *);
extern char const k1;
extern int k2;
\end{verbatim}
The corresponding LLVM produced by the C compiler is:
\begin{verbatim}
declare i32 @f(i32, i32, i32*)
@k1 = external constant i8
@k2 = external global i32
\end{verbatim}
Also, for the following C++ code:
\begin{verbatim}
class t1;
extern t1 c1;
\end{verbatim}
the compiler produces the following LLVM:
\begin{verbatim}
%class.t1 = type opaque
@c1 = external global %class.t1
\end{verbatim}

\subsection{The PROMOTES clause}

For each promoted operation, a sub-routine definition is generated.
The definition is just a call to the sub-routine corresponding to
the operation being promoted. As an inspiration, let us consider
the following C code:
\begin{verbatim}
extern int f(int);

int promote_f(int p)
{
  return f(p);
}
\end{verbatim}
The compiler produces the following LLVM code:
\begin{verbatim}
define i32 @promote_f(i32 %p) nounwind uwtable ssp {
  %1 = tail call i32 @f(i32 %p) nounwind
  ret i32 %1
}
declare i32 @f(i32)
\end{verbatim}
Note that the function declaration found in the last line shall be
emitted during the processing of the IMPORTS clause for the machine
containing the promoted operation.

\texttt{nounwind}: the called function never returns an unwind or exceptional
control flow.

\section{Completing the definition}

This section is used to document a list of tasks to complete the
definition of LLVM code generation from B0. It is based on the
directory of tests for the C code generation tool ComenC.

The ComenC tests are structured in directories. Each directory tests
the support for some aspect of B0. For each such directory, the
document summarizes the content of each test file and discuss how our
LLVM code generation from B0 needs to be updated.

\subsection{Directory OPERLOC}

This directory is used to test and validate support for ``local operations''.

Local operations are operations that are specified and implemented in
the implementation: they do not refine an operation of the refined
component, they are just a code structuring (factoring) construct. The
specification is located in the "local\_operations" clause, and the
implementation in the "operations clause. Local operations are only
visible to the operations of the machine, and the generated LLVM code
must include corresponding visibility directives.

\subsubsection{Benchmark description}

\begin{description}

\item[operloc\_comments] The implementation has
local operations with comments.

\item[operloc\_empl\_oploc\_ds\_spec] The implementation has three bare
no-op local operations

\item[operloc\_emplacement1] The implementation has three local
operations called in a refined operation.

\item[operloc\_emplacement2] Implementation has three local operations,
two refined operations, two local operations calling the third one.

\item[operloc\_import] Implementation has one local operation, one
imported machine with one operation, the local operations called the
operation of the imported machine

\item[operloc\_ordre1, operloc\_ordre2] Local operations calls other local
operation(s) sometimes appearing further.

\item[operloc\_params] Implementation has local operations with input
and output parameters, and accessing a visible variable

\item[operloc\_promues] The implementation refines an operation by
  promoting an operation of an imported machine, the local operation
  calls the promoted operation. The definition of the local operation
  has a local variable and writes a concrete variable.

\end{description}

\subsubsection{Actions to handle benchmarks}

\begin{enumerate}
\item Emit the keyword \texttt{internal} in the signature of
  operations that appear in the ``local\_operations'' clause.
\item LLVM does not require declaration of sub-routines before their
  use, so code generation shall not worry about ordering operation
  definitions or emitting declarations.
\item Considering promoted operations, it seems that declaring their
  signature is sufficient. For the following C declaration:
\begin{verbatim}
  extern void promop(void);
\end{verbatim}
  the C compiler produces:
\begin{verbatim}
  declare void @promop()
\end{verbatim}
\item Optionally, check which local operations are actually (directly
  or indirectly) referenced by refined operations; only emit code for
  those operations that are used.
\end{enumerate}

\subsection{Directory ORDREDECL}

This directory contains a single benchmark to test modularity constructs:
\begin{description}
\item[test\_ordre] Contains a development with a specification and an
  implementation. The implementation contains imports, extends and
  promotes clauses. Also the benchmark contains declarations of sets
  (abstract and enumerated) and concrete constants (integers and
  intervals).
\end{description}

For each abstract set, a LLVM type is generated. Eventually, if this
abstract set is valued in a B development it will be an integer range.
Since the size of this range remains undefined, there are two
possibilities:

(i) either choose the largest possible integer type available
in the LLVM platform,

(ii) or define it as an opaque type in LLVM (however those seem to be
usually employed for forward declared structure in C or classes
declarations in C++).

\section{Next steps}

Follows a list of tasks to be executed, in no particular order,
and possibly concurrently, to complete this work:
\begin{itemize}

\item Specify code generation for a larger class of data types: arrays,
records, integer intervals, enumerations.

\item Specify modular code generation: sees, imports, extends...

\item Case studies: check that LLVM code is processable by existing
compilation tools.

\item Verify if parse tree produced by BCompiler is close enough to 
  the hypothetic parse tree used in the document. Possibly make adjustements
  to rules to this reality.

\item Validate ideas with manual production of LLVM source files from
simple B0 implementations.

\item Implement.

\end{itemize}

\end{document}

\begin{align}
IMPLEMENTATION & ::= "IMPLEMENTATION" Header \\
& Clause_refines \\
& Clause_implementation* \\
& "END"
\end{align}

Clause_implementation ::= Clause_sees
| Clause_imports
| Clause_promotes
| Clause_extends_B0
| Clause_sets
| Clause_concrete_constants
| Clause_properties
| Clause_values
| Clause_concrete_variables
| Clause_invariant
| Clause_assertions
| Clause_initialization_B0
| Clause_operations_B0

### 2012-08-28 C/obj.c
Adding example C and LLVM files

@@ -1,0 +1,28 @@
+#include <stdbool.h>
+struct impl1 {
+  int v1;
+  int v2;
+  bool v3;
+} * impl1_anonymous;
+
+void op1(struct impl1 * a)
+{
+  a->v1 = a->v1 + 1;
+}
+
+void op2(struct impl1 * a, int * r)
+{
+  * r = a->v1;
+}
+
+struct impl2 {
+  int w1;
+  struct impl1 * w2;
+  struct impl1 * w3;
+} * impl2_anonymous;
+
+void op3(struct impl2 * a)
+{
+  op1(a->w2);
+  op1(a->w3);
+}

### 2012-09-04 b0-llvm.tex
Adding some fix until 5.3 Relations

@@
   \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
-  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{n.values}{val} \textbf{ and } O_v = \trad{n.vars}{var} \\
-  & \quad \textbf{ and } O_{in} = \trad{n.initialization}{ini} \textbf{ and } O_{op} = \trad{n.ops}{ops} \\
+  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{i.values}{val} \textbf{ and } O_v = \trad{i.vars}{var} \\
+  & \quad \textbf{ and } O_{in} = \trad{i.initialization}{ini} \textbf{ and } O_{op} = \trad{i.ops}{ops} \\
   & \quad \textbf{ in} \\
@@
 variables:
+
+% TODO: Qual o valor de n ?  Seria isto: n = L()  (\name) ?
+% Nas duas funções seguintes parece ter uma mistura com
+% relação ao nome das funções:  {var}, {type}, {val} 
 \begin{framed}
@@
   \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
-  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
+  id = \texttt{"@"} \conc \textsf{v.ident}  \textbf{ and }  n = ???? \textbf{ in} \\
   & \textbf{ yields } \\
@@
   \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
-  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
+  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ and }  n = ???? \textbf{ in} \\
   & \textbf{ yields } \\
@@
 \begin{split}
+  \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text \times text \times  text  \\
   \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D_{bool}(\textsf{n.val}), \texttt{"i1"} \\
@@
 \begin{split}
+ \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text  \times text \times  text\\
   \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, \textsf{n.val}, \texttt{"i32"} \\
@@
 \begin{split}
+  \trad{}{exp} : & \textsf{ Node} \times dict \rightarrow text \times text \times  text\\
   \trad{n, $D$}{exp} \isdef & \textbf{ yields } \texttt{""}, D(\textsf{n}), \trad{n.type}{type} \\
@@
 the resulting value, and of its LLVM type):
-\begin{framed}
-\begin{align}
-\begin{split}
+% TODO: Explicar o que significa o O_p,  O_l,   O_t  ?
+\begin{framed}
+\begin{align}
+\begin{split}
+  \trad{}{exp} : & \textsf{ Node} \times dict  \rightarrow text \times text \times  text\\
   \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
@@
   O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
-  & \quad \textbf{ and } v = \mathcal{V} \textbf{ and } op = D_{bari}(\textsf{n.op}) \textbf{ in } \\
+  & \quad \textbf{ and } v = \name \textbf{ and } op = D_{bari}(\textsf{n.op}) \textbf{ in } \\
   & \textbf{ yields} \\
@@
 \begin{split}
+ \trad{}{exp} : & \textsf{ Node} \times dict  \rightarrow text \times text \times  text\\
   \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
-  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
+  \trad{n.arg, $D$}{exp} \textbf{ and } v = \name \textbf{ in } \\
   & \textbf{ yields} \\
@@
   \trad{n, $D$}{exp} \isdef & \textbf{ let } O_p, O_l, O_t =
-  \trad{n.arg, $D$}{exp} \textbf{ and } v = \mathcal{V} \textbf{ in } \\
+  \trad{n.arg, $D$}{exp} \textbf{ and } v = \name \textbf{ in } \\
   & \textbf{ yields} \\
@@
 \begin{split}
+ \trad{}{rel} : & \textsf{ Node} \times dict  \rightarrow text \times text \\
   \trad{n, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
@@
   O'_p, O'_l, O'_t = \trad{n.right, $D$}{exp} \\
-  & \quad \textbf{ and } l = \mathcal{V} \textbf{ in } \\
+  & \quad \textbf{ and } l = \name \textbf{ in } \\
   & \textbf{ yields} \\
@@
 \begin{split}
+ \trad{}{rel} : & \textsf{ Node} \times dict  \rightarrow text \times text \\
   \trad{r, $D$}{rel} \isdef & \textbf{ let } O_p, O_l, O_t =
@@
   O'_p, O'_l, O'_t = \trad{r.right, $D$}{exp} \\
-  & \quad \textbf{ and } l = \mathcal{V} \textbf{ and } op = D_{op}(\textsf{r.op}) \textbf{ in } \\
+  & \quad \textbf{ and } l = \name \textbf{ and } op = D_{op}(\textsf{r.op}) \textbf{ in } \\
   & \textbf{ yields} \\

### 2012-11-13 b0-llvm.tex
- Created the example with the new parameter in functions ( it keeps the state of module).
- Added some elements in the abstract tree.

- Adjusted the dictionary to support rather than (Node -> text), support (text -> text) and domain B keeps identifiers and image identifiers LLVM.

- Fix section 6.1, apparently without using two dictionaries. Fix to use only one.

- Added a parameter in functions to get the reference of the global variables.

@@
         \hline
-        \textsf{Impl} & Implementation & \textsf{sees} : ? \\
-        & & \textsf{imports} : ? \\
+        \textsf{Impl} & Implementation & $\textsf{sees} : \textsf{Identifiers} (?)$\\
+        & & $\textsf{imports} \in \textsf{Identifiers} (?)$ \\
+        & & $\textsf{values} \in \textsf{Identifiers} (?)$ \\
+        & & $\textsf{vars} \in \textsf{Identifiers} (?)$ \\
+        & & $\textsf{initialization} \in Inst (?)$ \\
         & & $\cdots$ \\
@@
         \hline
-        \textsf{Op} & Operation & \textsf{header}: \textsf{Hd} \\
+        \textsf{Op} & Operation & $\textsf{header} \in \textsf{Hd} $\\
         & & $\textsf{body} \in \textsf{inst}$ \\
-        & & \textsf{next} : \textsf{Op} \\
-        \hline
-        \textsf{Hd} & Operation header & \textsf{ident} : text \\
-        & & \textsf{output} : \textsf{VarDecl} \\
-        & & \textsf{input} : \textsf{VarDecl} \\
-        \hline
-        \textsf{Var} & Variable, parameter & \textsf{ident} : text \\
-        & & \textsf{type} : basic type \\
-        & & \textsf{next} : \textsf{Var} \\
+        & & $\textsf{next} \in \textsf{Op} $\\
+        \hline
+        \textsf{Hd} & Operation header & $ \textsf{ident} \in text$ \\
+        & & $\textsf{output} \in \textsf{VarDecl}$ \\
+        & & $\textsf{input} \in \textsf{VarDecl}$ \\
+        \hline
+        \textsf{Var} & Variable, parameter & $ \textsf{ident} \in text $\\
+        & & $\textsf{type} \in basic type $\\
+        & & $\textsf{next} \in \textsf{Var} $\\
         \hline
@@
         \hline
-        \textsf{VarDecl} & Variable decl. & \textsf{ident}: \textsf{VarDecl} \\
+        \textsf{VarDecl} & Variable decl. & $\textsf{ident}\in \textsf{VarDecl} $\\
         & & $\textsf{body} \in \textsf{inst}$ \\
         \hline
-        \textsf{If} & If & \textsf{branches}: \textsf{IfBr} \\
+        \textsf{If} & If & $\textsf{branches} \in \textsf{IfBr}$ \\
         & & $\textsf{next} \in \textsf{Instruction}$ \\
         \hline
-        \textsf{IfBr} & If branch & \textsf{cond}: predicate \\
+        \textsf{IfBr} & If branch & $ \textsf{cond}\in predicate $ \\
         & & $\textsf{body} \in \textsf{Instr}$ \\
-        & & \textsf{next} : \textsf{IfBr} \\
-        \hline
-        \textsf{BEq} & Becomes equal & \textsf{lhs}: target \\
-        & & \textsf{rhs}: expression \\
+        & & $\textsf{next} \in \textsf{IfBr}$ \\
+        \hline
+        \textsf{BEq} & Becomes equal & $\textsf{lhs} \in target $ \\
+        & & $\textsf{rhs} \in expression $ \\
         & & $\textsf{next} \in \textsf{Instruction}$ \\
         \hline
-        \textsf{Call} & Call instruction & \textsf{fun}: name \\
-        & & \textsf{args}: expression \\
+        \textsf{Call} & Call instruction & $\textsf{fun} \in name$ \\
+        & & $\textsf{args} \in expression$ \\
         & & $\textsf{next} \in \textsf{Instruction}$ \\
         \hline
-        \textsf{While} & While instr. & \textsf{cond}: predicate \\
+        \textsf{While} & While instr. & $\textsf{cond} \in predicate $\\
         & & $\textsf{body} \in \textsf{Instr}$ \\
@@
         \hline
-        \textsf{Case} & Case instr. & \textsf{expr}: expression\\
-        & & \textsf{branches}: \textsf{CaseBr} \\
+        \textsf{Case} & Case instr. & $\textsf{expr} \in expression$\\
+        & & $\textsf{branches} \in \textsf{CaseBr}$ \\
         & & $\textsf{next} \in \textsf{Instruction}$ \\
         \hline
-        \textsf{CaseBr} & Case branch & \textsf{val}: \textsf{Value} \\
+        \textsf{CaseBr} & Case branch & $\textsf{val} \in \textsf{Value}$ \\
         & & $\textsf{body} \in \textsf{Instr}$ \\
-        & & \textsf{next} : \textsf{CaseBr} \\
+        & & $\textsf{next} \in \textsf{CaseBr}$ \\
         \hline
@@
 respectively, and $\conc$ denotes string catenation and $\nl$ a new
-line. .
+line.
 
@@
 digits representing the smallest positive number $t \conc n$ does not
-belong to the domain of $D$.
+belong to the domain of $D$. This dictionary is a function that keeps 
+\textbf{B identifiers} in domain and \textbf{LLVM identifiers} in range.
 
@@
   \trad{}{impl} : & \textsf{ Node} \rightarrow text \\
-  \trad{i}{impl} \isdef & \textbf{ let } O_{c} = \trad{i.values}{val} \textbf{ and } O_v = \trad{i.vars}{var} \\
-  & \quad \textbf{ and } O_{in} = \trad{i.initialization}{ini} \textbf{ and } O_{op} = \trad{i.ops}{ops} \\
+  \trad{i}{impl} \isdef & \textbf{ let } Dic = \{ \#ModuleName \mapsto  i.moduleName\} \textbf{ in}\\
+  & \textbf{ let } O_{c}, Dic'  = \trad{i.values, Dic }{val} \textbf{ and } O_v, Dic'' = \trad{i.vars, Dic'}{var} \\
+  & \quad \textbf{ and } O_{in} = \trad{i.initialization, Dic''}{ini} \textbf{ and } O_{op} = \trad{i.ops, Dic''}{ops} \\
   & \quad \textbf{ in} \\
   & \textbf{ yields } \\
-  & \quad \trad{i.ops}{ops}  \\
+  & \quad O_{c} \conc O_{v} \conc O_{in} \conc O_{op}    \\
   & \textbf{ where } \textsf{i.k} = \textsf{Impl} 
@@
 \end{framed}
+
 Code generation for initialisation and operations is specified in
@@
 
-% TODO: Qual o valor de n ?  Seria isto: n = L()  (\name) ?
-% Nas duas funções seguintes parece ter uma mistura com
-% relação ao nome das funções:  {var}, {type}, {val} 
 \begin{framed}
@@
   \trad{}{var} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
-  \trad{v, $D$}{type} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
-  id = \texttt{"@"} \conc \textsf{v.ident}  \textbf{ and }  n = ???? \textbf{ in} \\
+  \trad{v, $D$}{var} \isdef & textListVar = \trad{v,$D$,0}{DicVars}  \textbf{ in}\\
   & \textbf{ yields } \\
-  & \quad (id \conc \texttt{" = internal global "} \conc \trad{v.type}{type} \conc \nl \\
-  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
-  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
-  \trad{$\bot$, $D$}{type} \isdef & \textbf{ yields } \texttt{""}, D
+  & \quad \texttt{" \%struct."}\conc  D(\#ModuleName) \conc \texttt{"\_internalState = type \{"} \\
+  & \quad  \conc textListVar \conc \texttt{"\} "}  \nl , D \\
+  & \textbf{ where } \textsf{v} \ne  \bot \textbf{ and }  \textsf{v.k} = ConcreteVars \\
+  \trad{$\bot$, $D$}{var} \isdef & \textbf{ yields } \texttt{""}, D
+\end{split}
+\end{align}
+\end{framed}
+
+The type list of module variables are declared using this role, where:
+n is the actual node, D is the dictionary and  p is the position of variable.
+
+\begin{framed}
+\begin{align}
+\begin{split}
+  \trad{}{DicVars} : & \textsf{ Node } \times dict \times integer \rightarrow text  \times dict \\
+  \trad{n, $D$, p }{DicVars} \isdef  & \textbf{ let }   type = \trad{n.type}{type}   \textbf{ and }  t', D' =  \trad{n.next,D,p+1} {DicVars}   \textbf{ in}   \\
+  & \textbf{ yields } \\
+  & \quad  type \conc^,  t' , D' \cup \{ \textsf{n.ident} \mapsto (\textit{text }  p) \}  \cup\\
+  & \quad \{ \texttt{"\#type\_"}\conc \textsf{n.ident} \mapsto type \} \\
+  & \textbf{ where } \textsf{n.k} = \textsf{Var}  \\
+\trad{$\bot$, $D$, p }{DicVars} \isdef & \textbf{yields} \texttt{""}, D
 \end{split}
@@
   \trad{}{val} : & \textsf{ Node} \times dict \rightarrow text \times dict \\
-  \trad{v, $D$}{val} \isdef & \textbf{ let } O', D' = \trad{v.next, $D$}{var} \textbf{ and } 
-  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ and }  n = ???? \textbf{ in} \\
+  \trad{v, $D$}{val} \isdef & \textbf{ let } type = \trad{n.type}{type}   \textbf{ and }O', D' = \trad{v.next, $D$}{val} \textbf{ and } 
+  id = \texttt{"@"} \conc \textsf{v.ident} \textbf{ in} \\
   & \textbf{ yields } \\
-  & \quad (id \conc \texttt{" = constant "} \conc \trad{v.type}{type} \conc \nl \\
-  & \quad \quad \conc O'), D' \cup \{ \textsf{n} \mapsto id \} \\
-  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{ConcVar} \\
+  & \quad (id \conc \texttt{" = constant "} \conc type \conc \nl \\
+  & \quad \quad \conc O'), D' \cup \{ \textsf{v.ident} \mapsto id \} \conc  \{ \texttt{"\#type\_"}\conc \textsf{n.ident} \mapsto type \} \\
+  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Values} \\
   \trad{$\bot$, $D$}{val} \isdef & \textbf{ yields } \texttt{""}, D
@@
   \begin{split}
-  \trad{}{ops} : & \textsf{ Node} \rightarrow text \\
-  \trad{o}{ops} \isdef 
-  & \textbf{ yields } \trad{o}{op} \conc \trad{o.next}{ops}  \\
+  \trad{}{ops} : & \textsf{ Node} \times \textsf{ Dict} \rightarrow text \\
+  \trad{o,D}{ops} \isdef 
+  & \textbf{ yields } \trad{o,D}{op} \conc \trad{o.next,D}{ops}  \\
   & \textbf{ where } \textsf{o} \neq \bot  \textbf{ and } \textsf{o.k} \in \textsf{Op}  \\
-  \trad{$\bot$}{ops} \isdef
+  \trad{$\bot$,D}{ops} \isdef
   & \textbf{ yields } \mty 
@@
 \begin{split}
-  \trad{}{op} &: \textsf{ Node} \rightarrow text \\
-  \trad{op}{op} \isdef
-  & \textbf{ let } O_l = \trad{op}{link} \textbf{ and } O_h, D_h = \trad{op.header}{header} \textbf{ in}  \\
+  \trad{}{op} &: \textsf{ Node } \times \textsf{ Dict} \rightarrow text \\
+  \trad{op,d}{op} \isdef
+  & \textbf{ let } O_l = \trad{op}{link} \textbf{ and } O_h, D_h = \trad{op.header,d}{header} \textbf{ in}  \\
   & \textbf{ let } O_a, D_a = \trad{op.body, $D_h$}{alloc} \textbf{ in}  \\
   & \textbf{ let } O_i = \trad{op.body, $D_a$}{inst} \textbf{ in}  \\
-  & \textbf{ yields } \texttt{"declare "} \conc O_l \conc O_h \conc \texttt{"\{"} \conc \nl \conc \\
+  & \textbf{ yields } \texttt{"define "} \conc O_l \conc O_h \conc \texttt{"\{"} \conc \nl \conc \\
   & \quad \conc O_a \conc \nl \conc O_i \conc \nl \conc \texttt{"ret void"} \conc \nl \conc \texttt{"\}"} \conc \nl  \\
@@
 \begin{split}
-    \trad{}{header} &: Hd \rightarrow text \times dict \times dict \\
-    \trad{h}{header} \isdef& \textbf{ let } T_n = \textsf{"@"} \conc \textsf{h.ident} 
-    \textbf{ and } T_i, D_i = \trad{h.input}{inp} 
-    \textbf{ and } T_o, D_o = \trad{h.output}{out} \textbf{ in}  \\
-    & \textbf{ let } T_p = T_i \conc^, T_o \textbf{ in}  \\
-    & \textbf{ yields } \texttt{"void "} \conc T_n \conc \texttt{"("} \conc T_p \conc
-    \texttt{")"}, D_i \cup D_o, \{ \textsf{h.ident} \mapsto T_n \}  \\
+    \trad{}{header} &: Hd \times dict \rightarrow text \times dict \\
+    \trad{h,D}{header} \isdef& \textbf{ let } T_n = \textsf{"@"} \conc \textsf{h.ident}  \textbf{ and } 
+    \textbf{ and } T_i, D_i = \trad{h.input}{inp} \\
+    & \textbf{ and } T_o, D_o = \trad{h.output}{out} \textbf{ in}  \\
+    & \textbf{ let } T_p = \texttt{" \%struct."} \conc D(\#ModuleName) \conc \\
+    & \texttt{"\_internalState* nocapture \%pointerMVars"}  \textbf{ in}  \\
+    & \textbf{ yields } \texttt{"void "} \conc T_n \conc \texttt{"("} \conc T_p \conc^, T_i \conc^, T_o \conc
+    \texttt{")"}, D_i \cup D_o \conc \{ \textsf{h.ident} \mapsto T_n \}  \\
     & \textbf{ where } \textsf{h} \ne \bot \textbf{ and } \textsf{h.k} = \textsf{Hd} 
@@
     \textbf{ and } T, D = \trad{i.next}{inp} \textbf{ in}  \\
-    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{i} \mapsto T_n \} \cup D  \\
+    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{i.ident} \mapsto T_n \} \cup D  \\
     & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var}  \\
@@
     \textbf{ and } T, D = \trad{o.next}{out} \textbf{ in}  \\
-    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{o} \mapsto T_n \} \cup D  \\
+    & \textbf{ yields } T_t \conc \texttt{" "} \conc (T_n \conc^, T), \{ \textsf{o.ident} \mapsto T_n \} \cup D  \\
     & \textbf{ where } \textsf{o} \ne \bot \textbf{ and } \textsf{o.k} = \textsf{Var}  \\
@@
     & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
-    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} 
+    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{VarDecl} \\
+    \trad{i, $D$}{alloc}_{base} \isdef 
+    & \textbf{ let } O_l, D_l = \trad{i.ident, $D$}{global} \textbf{ in}  \\
+    & \textbf{ let } O_b, D_b = \trad{i.body, $D_l$}{alloc} \textbf{ in}  \\
+    & \textbf{ yields } O_l \conc \nl \conc O_b, D_b  \\
+    & \textbf{ where } \textsf{i} \ne \bot \textbf{ and } \textsf{i.k} = \textsf{Var} 
+  \end{split}
+\end{align}
+\end{framed}
+
+This function get internal references to concrete variables from module.
+
+\begin{framed}
+\begin{align}
+\begin{split}
+  \trad{}{global} : & \textsf{ Node } \times dict \rightarrow text \times dict \\
+  \trad{v, $D$}{global} \isdef & 
+  \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
+  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v.ident} \mapsto T_v \}$}{local} \textbf{ in}  \\
+  & \textbf{ yields } T_v \conc \texttt{" = getelementptr inbounds "} \conc\\
+  & \texttt{"\%struct.ModuloName\_internalState* \%pointerMVars, i64 0, i32 "} \conc \\
+  & D(\texttt{"\#type\_"}\conc v.ident) \conc \nl \conc T_n, D_n  \\
+  & \textbf{ where } \textsf{v} \ne \bot \textbf{ and } \textsf{v.k} = \textsf{Var}  \\
+  \trad{$\bot$, $D$}{global} \isdef & 
+  \textbf{ yields } \mty, D 
 \end{split}
@@
   \textbf{ let } T_v = \fresh(\texttt{"\%"}\conc\textsf{v.ident}, D) \textbf{ in}  \\
-  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v} \mapsto T_v \}$}{local} \textbf{ in}  \\
+  & \textbf{ let } T_n, D_n = \trad{v.next, $D \cup \{ \textsf{v.ident} \mapsto T_v \}$}{local} \textbf{ in}  \\
   & \textbf{ yields } T_v \conc \texttt{" = alloca "} \conc \trad{v.type}{type} \conc \nl \conc T_n, D_n  \\
@@
 
+B method supported different types of modularity.
+The implementation modules can use the clauses: SEES, PROMOTES and IMPORT.
+
 \subsection{The SEES clause}
@@
 \end{verbatim}
-Also, for the following C++ code:
-\begin{verbatim}
-class t1;
-extern t1 c1;
-\end{verbatim}
-the compiler produces the following LLVM:
-\begin{verbatim}
-%class.t1 = type opaque
-@c1 = external global %class.t1
-\end{verbatim}
+
+[[TODO: to add and to explain the translation function]]
+
+\subsection{The IMPORT clause}
+
+In B, one module can be imported several times with different
+names creating instances of one module, this feature is called 
+``renaming" in C4B, but it is not supported in version 1.0. Based on
+the following C++ code, the state of each module can be stored in a LLVM
+structure. To represent an reference of module state, one pointer to this
+structure can be used. 
+\begin{verbatim}
+class CRectangle {
+    int x, y;  
+    };
+extern CRectangle vCRectangle;
+extern CRectangle *pCRectangle;
+\end{verbatim}
+The compiler produces the following LLVM:
+\begin{verbatim}
+%struct.CRectangle = type { i32, i32 }
+@pCRectangle = external global %struct.CRectangle* 
+@vCRectangle = external global %struct.CRectangle 
+\end{verbatim}
+
+Now to add support to renaming, all functions contain an first additional
+parameter that allow access data from imported module. This parameter
+is a pointer to structure that store the state of imported module.
+In practice, the translation of module called ``Data" will be based in following
+C code:
+\begin{verbatim}
+long Data__vint;
+struct R_1 { long elem1;long elem2; } Data__vstruct;
+void Data__INITIALISATION(void)
+{
+    Data__vint = 10;
+    Data__vstruct.elem1 = 1;
+    Data__vstruct.elem2 = 2;
+}
+void Data__get_vint( )
+{
+    return vint;
+}
+
+\end{verbatim}
+The result of translation creates a structure ``TStateData" with all
+variables of  module ``Data" to store the state of module, create a pointer
+to this structure and added this pointer as the first parameter in all functions of module.
+The return of a function is passed by last parameter as a pointer.
+\begin{verbatim}
+struct R_1 { long elem1;long elem2; } ;  
+struct TStateData { long vint;long vinterval; R_1 vstruct; } ;
+extern TStateData *pStateData; 
+void Data__INITIALISATION(TStateData * pStateData)
+{
+    pStateData->vint = 10;
+    pStateData->vstruct.elem1 = 1;
+    pStateData->vstruct.elem2 = 2;
+}
+void Data__get_vint(TStateData * pStateData, int * res1 )
+{
+    res1 = &pStateData->vint;
+}
+
+\end{verbatim}
+
+\begin{figure}[ht]
+
+\begin{minipage}{50in}
+\begin{verbatim}
+<how>
+   <to value="make" />
+   <this value="work" />
+</how>
+\end{verbatim}
+\end{minipage}
+
+\begin{tabular}[t]{l}
+\begin{minipage}{50in}
+\begin{verbatim}
+<how>
+   <to value="make" />
+   <this value="work" />
+</how>
+\end{verbatim}
+\end{minipage}
+\end{tabular}
+
+\begin{tabular}[t]{l}
+
+\end{tabular}
+
+\caption{Example.}
+\label{fig:kjdsldkjsa}
+\end{figure}
+
+[[TODO: Explain... ]]
+
+All the above examples are omitting informations about modularity,
+because it becomes the explanation more easy. However, the unique difference
+is when a variable is accessed then is needed add a parameter pointer to keep the state of module
+and to add parameter  to return.
+
+% This clause is supported only in Implementation ( The codes are compiled individually )
+
+The unique additional information is an parameter in functions that represent the 
+This parameter works as object of class.
+
+All the above examples are omitting informations about modularity,
+because it becomes the explanation more easy.
+
+%[[TODO: update the other sections based in this example.]]
+
+% The renaming of imported modules are not supported in C4B (version 1.0 ).
+
+\begin{verbatim}
+
+  // I added this definitions to support renaming
+  struct R_1 { long elem1;long elem2; } ;  // update this declaration to create the type of structures
+  struct TStateData { long vint;long vinterval; R_1 vstruct; } ; // Create this declaration to keep the state
+  extern TStateData vStateData; // Declare the variables
+  extern TStateData *pStateData; // Declare the variables
+
+/* Clause CONCRETE_CONSTANTS */
+/* Basic constants */
+/* Array and record constants */
+extern void Data__INITIALISATION(TStateData * pStateData);
+
+/* Clause OPERATIONS */
+
+extern void Data__set_vint(TStateData * pStateData);
+extern long Data__get_vint(TStateData * pStateData);
+\end{verbatim}
+
+% Why it no use the concept  class and objects ? 
+% Because it is owner compiler uses
 

### 2013-06-05 paper.tex
Files for SBMF 2013 submission

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Control and Modularity}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{May 2013}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\name}{\mathcal{L}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

\begin{document}
\maketitle

\begin{abstract}
  The aim of this work is to lay the foundations of a multi-platform code
  generator for the B method. In particular, this paper presents a translation
  procedure from a large subset of the B language for implementations towards
  LLVM source code. This translation is defined formally as a set of rules
  defined recursively on the abstract syntax for B implementations. It handles
  the following elements of the B language: simple data types, imperative
  instructions and component compositions. 
\end{abstract}

\section{Introduction}

B is a formal refinement-based software design method~\cite{Abrial1996}. It has
a single language encompassing abstract constructs suitable for specification,
as well as classic imperative constructs for computer programming. The starting
point of a B development is a specification, called a \emph{machine}, that is
incrementally refined to an \emph{implementation}, where only imperative-like
constructs may be employed~\cite{Clearsy}. Such implementation may then be
translated to source code in a programming language, say C~\cite{ComenC} or
Ada. All steps in the B method are formally verified using certified theorem
proving technologies. However the translation to some programming language and
its subsequent compilation to the target platform do not benefit from the same
mathematical rigor. In practice, redundancy in the tool chains and execution
platform are employed to increase the level of confidence to the desired levels.

The goal of this work is to connect the B method to the LLVM compilation
framework~\cite{Lattner04LLVM}.  LLVM is a free, open-source project upon which
many compiling contemporary technologies and tools are based. LLVM provides an
intermediate assembly language upon which may be applied techniques such as
optimization, static analysis, code generation, debugging, etc. Providing a
connection from the B method to this framework gives the users of the B method
access to these technologies. Our approach is to define a translation from B0
(the subset of the B language that is used to describe imperative programs) to
the LLVM intermediate format.

Considering the correctness of the proposed translation, we shall rely upon
human inspection for the time being. The standard uniform semantic framework for
B and LLVM that would be necessary to carry out a proof of this correctness does
not exist (yet). A possible approach to mitigate the risk of having an error in
the translation would be to include assertions in the target code. Indeed the
artifacts produced in the B development have many such assertions and it would
be possible to translate at least some of them (B has an assertion instruction
that introduces additional verification conditions in the development). The
produced code could then be (at least partially) verified by testing the absence
of assertion violations.

\paragraph{Overview.} Following this introduction, the paper is organized as
follows. The implementation constructs of the B language, and selected elements
of the LLVM are presented in section~\ref{sec:b-impl}
and~\ref{sec:llvm}. Section~\ref{sec:overview} presents a bird's-eye view that
is then detailed in sections~\ref{sec:module} (for component modularity)
~\ref{sec:data} (for data aspects),~\ref{sec:expr} (expressions) and
instructions~\ref{sec:control}. We conclude and discuss future work
in~\ref{sec:conclusion}.

\section{Implementation Language in the B-method}
\label{sec:b-impl}

This section describes the B implementation constructs that are considered in
the compilation to the LLVM. This description introduces the notations used in
the definition of the translation. The elements of the B language are denoted
using a \B{sans serif} font.

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{imp}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

The initialization and the operations are defined as (sequences of)
instructions. Operations also have a name, inputs and outputs. Possible
instructions are the classic imperative constructs: variable assignment, if and
case conditional, while loop, block with or without local variables, and
operation calls. The if constructs may have several branches, each with a
condition as well as an optional else branch. Called operations may be either
local operations or operations of instances of external modules, in this case
they have an optional prefix identifying the module.

Note that the translation presented in this paper only considers types \B{INT}
(integer) and \B{BOOL} (Boolean). Accordingly, the expression language consists
in arithmetic operations \B{-} (unary and binary), \B{+}, \B{*}, \B{/}, \B{mod},
% \B{**} (exponentiation), 
\B{succ} and \B{pred} and Boolean operations
\B{$\land$}, \B{$\lor$}, \B{$\neg$} as well as relations \B{=}, \B{$\neq$},
\B{$>$}, \B{$<$}, \B{$\leq$} and \B{$\neq$}. Atomic expressions are identifiers,
Boolean constants \B{FALSE} and \B{TRUE}, integer literals, and integer
constants \B{MAXINT} and \B{MININT}. The development through the application of
the B method ensures the absence of overflows and underflows.

\begin{figure}[H]
  \begin{center}
    {\footnotesize
      \frame{
    \begin{tabular}[t]{ccccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{see} & : & seq \B{Name} \\
        \B{imp} & : & seq \B{Impo} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{pre} & : & opt \B{Name} \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}
        \\
      \end{tabular}
      & &
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \\
        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{Name} \\
        \B{rhs} & : & \B{Expr} \\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Name} \\
        \B{pre} & : & opt \B{Name} \\
        \B{inp} & : & seq \B{Expr} \\
        \B{out} & : & seq \B{Name}
        \\
%         \\
%         \hline
%         \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
%         \hline
%         \B{expr} & : & \B{Expr} \\
%         \B{branches} & : & seq \B{CaseBr}
      \end{tabular}
      & &
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
%         \hline
%         \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
%         \hline
%         \B{val} & : & seq \B{Value} \\
%         \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
        \hline
        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred}} \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
        \hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}}
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

\section{Target LLVM Subset}
\label{sec:llvm}

LLVM aims to be a general, language agnostic, intermediate representation for
compilers. The compiler front-end translates the source programming language to
LLVM source code, upon which optimization and other static analysis tasks may be
performed, and the compiler back-end translates to the target platform assembly
language. One key feature of LLVM is that it is a single-static assignment (SSA)
language: a variable may only be assigned once. So for each assignment to a
variable found in the source language, the front-end shall generate a new
temporary variable in the LLVM program. For illustration,
figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
C code appears on the left) with two temporary variables \llvm{\%0} and
\llvm{\%1}.
\begin{figure}
\vspace*{-3em}
  \begin{minipage}[t]{.3\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
      void inc(int * pi) \{\\
      \> *pi += 1; \\
      \}\end{tabbing}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.6\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
      \llvm{define void @inc(i32* \%pi) nounwind \{} \\
      \llvm{entry:}                      \\
      \> \llvm{\%0 = load i32* \%pi}     \\
      \> \llvm{\%1 = add i32 \%0, 1}     \\
      \> \llvm{store i32 \%1, i32* \%pi} \\
      \> \llvm{ret void} \\
      \llvm{\}}
    \end{tabbing}
  \end{minipage}
  \caption{Simple example of a C function and its corresponding LLVM function.
    The first line contains the signature: return type \llvm{void}, the name
    \llvm{@inc} and one parameter named \llvm{\%pi} and typed \llvm{i32*}.  Next
    is the body with a single block, labeled \llvm{entry}, and temporary
    variables \llvm{\%0} and \llvm{\%1}, created in the conversion to SSA. The
    block has four instructions: \llvm{load}, \llvm{add}, \llvm{store} and
    \llvm{ret}. For instance, \llvm{\%1 = add i32 \%0, 1} is an addition
    (\llvm{add}), has result type \llvm{i32} and assigns to
    \llvm{\%1} the sum of variable \llvm{\%0} and integer
    literal \llvm{1}.}
  \label{fig:ex-llvm}
\end{figure}

The introduction to the LLVM that follows is focused in the features
that are relevant for the representation of B
implementations. Figure~\ref{fig:llvm-grammar} presents a grammar of this subset
of the LLVM.

\begin{figure}
  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{struct} \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}
  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
    and \nt{name} correspond respectively to integer types, integer literals,
    labels and names. Choices are separated by \alt and optionally delimited by
    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
    elements of the annotated entity.}
  \label{fig:llvm-grammar}
\end{figure}

LLVM programs are organized into modules, one per translation unit. A module may
contain declarations of external entities (functions, constants) and definitions
of internal items (functions, variables, constants). All data is explicitly
typed. The name and type of external entities must be declared.  All names,
e.g. non-reserved identifiers, shall start with \llvm{@}, when they are global,
or\llvm{\%}, when they are local. For instance, \llvm{@max = external constant
  i32} declares \llvm{@max} as a 32-bit integer constant and \llvm{declare void
  @inc(i32*)} declares \llvm{@inc} as a function having as parameter a pointer
to an integer and \llvm{void} as a return type.

The type system contains the empty type \llvm{void}, an infinite, countable
number of integer types, one for each possible bit width (e.g. \llvm{i8} is the
type for 8-bit integers), and type constructors pointer (monadic \llvm{$\cdot$
  *}) and structure (polyadic \llvm{\{$\cdots$\}}). For instance \llvm{\{ i8*,
  i8, i8 \}} is the type for structures with three fields, the first having as
type pointer to \llvm{i8}. Grammar rule \nt{type\_def} states how types are
named, e.g.: \llvm{\%T1 = type \{i32, i32\}} and \llvm{\%T2 = type \{\%T1 *,
  \%T1 *\}}. Note that there is no \llvm{void *} type in the LLVM, instead
pointer values are of integer types.

Local entities are constants, variables or functions. An example of constant
definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
and value. The elements of a variable definition are the name, type and code
generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
information for target code generation: linkage type (the variable is merged
with other of the same name), the scope (global) and initialization of the
variable (zeroed is mandatory for common linkage). LLVM defines other attributes
for these aspects, but they are not needed in this work and their presentation
is omitted.  Function definitions are composed of the signature and body. The
signature contains the return type, name, parameters, and attributes for target
code generation. Attribute \llvm{nounwind} states that the function never
generate exceptions. The body is a sequence of blocks of instructions in
single-static assignment (SSA) form, i.e. are such that each variable may be
assigned exactly once.

The grammar rule \nt{inst} describes the different kind of
instructions. Instruction \llvm{alloca} allocates a memory block the size of the
given type from the stack segment. This and the following classes of
instructions assign the resulting value to a fresh variable. Arithmetic
operations are binary, and include signed quotient and reminder. Arithmetic
comparisons compare two values according to the first operand and return a 1-bit
integer value.  Instruction \llvm{call} invokes the given function with the
given arguments and assigns the result to a fresh variable, which is of the
given type.  In general, \llvm{getelementptr} gets the address of an element in
an aggregate object through indexing. In the target LLVM subset, aggregate
objects are structures, the first index has type \llvm{i32} and value \llvm{0}
to select the first structure at the given location \nt{exp}, and the second
index selects a field in that structure.  Instruction \llvm{load} assigns to a
fresh variable \nt{name} the contents of a memory address of type \nt{type}
specified by \nt{exp} (e.g. in fig.~\ref{fig:ex-llvm}).  Instruction
\llvm{store} writes a value to memory address (again, see
fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two forms.  In the first
form, it directs conditionally the control flow to one of two blocks.  If the
given expression evaluates to \llvm{1}, control goes to the first block,
otherwise to the second. In the second form, the control jumps unconditionally
to the given block.
% Instruction \llvm{switch} directs the
% control flow to one of several blocks. The given expression is evaluated and
% control goes to the first branch with the matching value, or to the default
% branch (appearing first in the instruction).  
Finally, instruction \llvm{ret}
ends the current function call, optionally returning a value.

The expression language is limited to names (local and global), integer literals
and selection of an element in a structure.

\section{Translation: a Bird's-Eye View}
\label{sec:overview}

The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
module, say \llvm{M}. A B implementation \B{M} has several sections, each
translated as follows:
\begin{enumerate}
\item Visibility clauses provide access to definitions found in external
  B components and are translated to external declarations in LLVM.
\item Constants and value definitions are translated to LLVM constants. To avoid
  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
\item To encode the implementation state, the translation produces a
  structure type, named \llvm{@M\$state\$}, and an instance of that type, named
  \llvm{@M\$self\$}. All the variables and instances of imported machines
  are encoded as elements of that structure.  Also, for each operation of each
  imported machine, an external function declaration is produced.
\item Each operation \B{op} of the implementation is translated into the
  definition of a function named \llvm{@M\$op}; also the translation produces a
  special function \llvm{@M\$init\$} responsible for the initialization.

  The parameter list of the functions thus produced contain one item for each
  input and output of the corresponding operation, and one item of type
  \llvm{@M\$state\$*}, to pass the address of the structure encoding the
  state of the implementation and its modification by the callee. The return
  type is always \llvm{void}.
\end{enumerate}
This approach assumes that information on external B components used in \B{M} is
available during translation. To fulfill this hypothesis, the translation
produces an ``interface'' artifact containing the information that the
translation of an importation of \B{M} in a client machine
requires. Figure~\ref{fig:skel} summarizes the translation principles by
providing the structure of both the interface and the LLVM implementation thus
produced.

\begin{figure}
  \begin{center}
    (a) Structure of the interface artifact produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{(references to imported components)} \\
      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \llvm{declare void @M\$init\$(@M\$state\$*)} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{declare void @M\$op(@M\$state\$*, \ListOf{\nt{type}})} \\
    \end{tabbing}
    (b) Structure of the LLVM module produced from a B implementation \B{M}:
    \begin{tabbing}
      foo \= foo \= \kill
      \textit{for each imported component} \B{Q} \textit{generate} \\
      \> \llvm{@Q\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \> \llvm{declare void @Q\$init\$(@Q\$state\$*)} \\
      \> \textit{for each constant} \B{cst} \textit{generate} \\
      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
      \> \textit{for each operation} \B{op} \textit{generate} \\
      \> \> \llvm{declare void @Q\$op(@Q\$state\$*, \ListOf{\nt{type}})} \\
      \textit{for each constant} \B{cst} \textit{generate} \\
      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
      \llvm{@M\$self\$ = common global @M\$state\$ zeroinitializer} \\
      \llvm{define void @M\$init\$(@M\$state\$* \%self\$) \{} \\
      \> \llvm{\ListOf{\nt{block}}} \\
      \> \llvm{exit: ret void} \\
      \llvm{\}} \\
      \textit{for each operation} \B{op} \textit{generate} \\
      \> \llvm{define void @M\$op(@M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
      \> \> \llvm{\ListOf{\nt{block}}} \\
      \> \> \llvm{exit: ret void} \\
      \> \llvm{\}}
    \end{tabbing}
  \end{center}
  \caption{Skeleton of the artifacts produced by the translation.}
  \label{fig:skel}
\end{figure}

The detailed presentation of the translation of the different elements of a B
implementation is divided into:
\begin{description}
\item[Modularity] constructs to combine B components are described in
  sec.~\ref{sec:module}.
\item[Data] translation is briefly described sec.~\ref{sec:data}, including
  the encoding of B data types.
\item[Expressions] and conditions are codified as LLVM instruction blocks, as
  described in section~\ref{sec:expr}.
\item[Control] flow, instructions are translated to LLVM instructions as
  discussed in sec.~\ref{sec:control}, which also presents the encoding
  of B operations and initialization.
\end{description}

Code generation is defined with a set of rules defined by structural
recursion. For each grammatical construct, a code generation function is
defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name. Also $\state(\B{M})
\isdef \llvm{"@\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
state of component \B{M}. We assume an unlimited supply of labels and local
variables: let $\name$ return a fresh LLVM label or local variable name.

\section{Translation: Modularity Aspects}
\label{sec:module}

The translation of a B module produces two artifacts: an interface and an (LLVM)
implementation. In the following, we describe how they are produced.

\subsection{Interface}

The definition of the interface $\intf(\B{M})$ of a component \B{M} has four
key elements, as shown in figure~\ref{fig:skel} (b):
\begin{enumerate}
\item $\intf(\B{M})_u$ is the list of used components: $\intf(\B{M})_u \isdef \B{M.imp}$.

\item $\intf(\B{M})_t$ is a textual description of the type representing the
  state of the component, i.e. a structure with one element per imported
  component and variable:
\begin{small}
$$\intf(\B{M})_t \isdef \llvm{"struct \{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
\end{small}
where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.

\item $\intf(\B{M})_c$ is the definition of the component constants. It maps
the constant names to their type:
$\intf(\B{M})_c \isdef \trad{\B{M.const}}{\ListOf{\B{Cons}}}$ and
$\trad{\B{c}}{\B{Cons}} \isdef \B{c} \mapsto \trad{\B{c.type}}{\B{Type}}$.

\item $\intf(\B{M})_o$ represents the operation declarations. It maps
operation names to the list of input parameter types and output parameter
types:
$\intf(\B{M})_o \isdef \trad{\B{M.op}}{\ListOf{\B{Oper}}}_t$ and
$\trad{\B{o}}{\B{Oper}}_t \isdef \B{o} \mapsto ( \trad{\B{o.inp}}{\ListOf{inp}} \sep \trad{\B{o.out}}{\ListOf{out}} )$, where
$\trad{\B{i}}{inp} \isdef \trad{\B{i.type}}{\B{Type}}$ and
$\trad{\B{o}}{out} \isdef \llvm{"\PH{\trad{\B{o.type}}{\B{Type}}}*"}$.
\end{enumerate}

\subsection{Implementation}

The actual LLVM source code is generated following the template shown in
figure~\ref{fig:skel} (a), as follows:
\begin{align*}
  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
  \quad \PH{\trad{\B{i}}{State}}} \\
& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
& \llvm{\PH{\trad{\B{i.init}}{Init}} 
  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
\end{align*}
We shall now define each rule referenced in this template.
\begin{enumerate}
\item For the type definitions corresponding to the state of the imported
  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
    type \PH{\intf(\B{i})_t}\nl"}$.
\item For the declarations of the initialization functions in the imported
  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
\item For the definitions of the imported constants: $\trad{\B{i}, \intf}{ICons}
  \isdef \trad{$\intf(\B{i})_c$}{\ListOf{Cons}}$, where
  \noindent $\trad{$\B{c} \mapsto t)$}{Cons} \isdef \llvm{"\PH{\Global(\B{c})} =
    external constant \PH{t}\nl"}$.
\item For the declarations of the imported operations: $\trad{\B{i},
    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
\item The translation of constant definitions is presented in section~\ref{sec:data}.
\item For the definition of the type for component states: \\
\noindent$\trad{\B{M}}{State} \isdef
\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
\item For the initialization, the definition of a function is generated. It
  first initializes the imported modules and then the current module (see
  section~\ref{sec:control}).
\item For each operation, a function definition is generated (see
  section~\ref{sec:control}).
\end{enumerate}
Let $\idx$ be a function that maps each imported machine and state variable to
its position (starting from zero).
\section{Translation: Data}
\label{sec:data}

The translation presented in this paper is currently limited to integers and
booleans. In the case of integers, the translation is based on the supposition
that the B0 definition of \B{INT} is the range $-2^{31}.. 2^{31}-1$. We have
$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}$ and $\trad{INT}{\B{Type}} \isdef
\llvm{i32}$.

B constants are translated to constants in the LLVM program, as described in the
following rule, where parameter \B{c} denote the constant, $M$ the name of the
module and $\Global$ is the mapping between source and target identifiers. Two
results are produced: the LLVM source code and an updated identifier mapping:
\begin{small}
\begin{align*}
  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
\end{align*}
\end{small}
The definition of the translation for expressions is described next in
section~\ref{sec:expr}.

\section{Translation: Expressions
  \label{sec:expr}}

To evaluate a composed expression using the LLVM, one must first generate a
sequence of instructions to evaluate each argument, the resulting value being
stored in a temporary variable, and second generate an instance of the LLVM
instruction corresponding to the expression argument.  Translation of
expressions is defined as functions that take as input some class of expressions
and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
sequence of instructions required in the evaluation (e.g.  for the
sub-expressions), $v$ is the LLVM expression representing the resulting value,
and $t$ represents the LLVM type of the expression. In the case of composed
predicates, though, the evaluation is different as boolean operations need to be
implemented with conditional and unconditional jumps.

\paragraph{Atomic expressions} have a simple translation. In the case of
literals, one only needs to issue the corresponding LLVM literal:
$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
$\B{l.type} = \B{INT}$.

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

\paragraph{Comparisons} are formed by the application of a relational operator
to two expressions. Mapping $\opmap$ associates B operators to corresponding
LLVM keywords:
$$\opmap = \{ = \mapsto \llvm{"eq"}, 
  \ne \mapsto \llvm{"ne"}, 
  < \mapsto \llvm{"slt"}, 
  \le \mapsto \llvm{"sle"}, 
  > \mapsto \llvm{"sgt"}, 
  \ge \mapsto \llvm{"sge"} \}.
$$

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \AND v = \name \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar structure and interpretation
as comparisons and their presentation is mostly omitted. Let us present the
translation of the unary \B{succ} operator though:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

\paragraph{Predicates}
The translation of conditions takes as input a condition node \B{n}, as well as
two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
execution shall jump when the condition is evaluted to true or false,
respectively.  The first rule defines the translation of an atomic relation:
\begin{align*}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
\end{align*}

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

\section{Translation: Control Flow and Instructions
  \label{sec:control}}

This section addresses how operations in B implementations are translated to
LLVM functions. It also discusses initialization, which might be seen as a
special kind of operation. For operations, the presentation of the translation
is divided into three steps: the signature, the local data, and the body
instructions.

The signature of the function is composed of the name of the function, the
result type, the name and type of the parameters. The signature may carry other
information such as linkage, visibility, calling convention, etc. The B parser
provides this information after the type checking phase has been completed.  On
the one hand LLVM functions may only have one result, on the other hand B0
operations may produce several results. The translation of output parameters
will be function parameters passed by reference, i.e. they are pointers. For
uniformity of treatment, the return type of the produced LLVM functions is
\llvm{void}. Translation of operation headers is defined in
sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
the stack, using the \llvm{alloca} instruction. We describe a function in
sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
type of its local variables used in the operation. These names are translated to
conform to LLVM restrictions. The body of the operation has to be transformed to
a sequence of LLVM blocks of statements. For each kind of instruction one (or
more) translation function is defined. The details are provided in
section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
combined as follows:
\begin{align*}
\begin{split}
  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
  & \llvm{"define void \PH{h} \{\nl} \\
  & \quad \llvm{entry:\nl} \\
  & \quad \quad \PH{a} \\
  & \quad \quad \PH{i} \\
  & \quad \llvm{exit: ret void\nl} \\
  & \llvm{\}\nl"}
\end{split}
\end{align*}

Initialization of a B implementation shares many characteristics with operations
but needs to deal with initialization of imported modules. It is discussed in
sec.~\ref{sec:trad-init}.

\subsection{Sub-routines signatures}
\label{sec:trad-header}

The following rule specifies the translation of operation headers. Its
parameters are the operation identifier, inputs, outputs, the name of the
machine it belongs to, and the results are the LLVM function definition
header: \\
\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$

Next are the rules defining the translation of individual input and output
parameters. They have local scope and they are translated by prepending
\llvm{\%}:
\begin{align*}
  \trad{i}{inp} \isdef 
  \llvm{"\PH{\trad{i.type}{\B{Type}}} \PH{\local(\B{i})}"} \quad \quad
  \trad{o}{out} \isdef 
  \llvm{"\PH{\trad{o.type}{\B{Type}}}* \PH{\local(\B{o})}"}
\end{align*}

\subsection{Stack allocation}
\label{sec:trad-alloc}

B operations may have local variables, in which case the translator needs to
issue LLVM stack allocation instructions. Such instructions have two arguments:
the type of values to store, and a temporary variable that is assigned the
address (see figure~\ref{fig:llvm-grammar} for the concrete syntax). The
generation of stack allocation code visits the parse tree of the operation body
in a pre-order depth-first traversal, generating one statement for each local
variable declaration found. The detailed definition of this traversal is omitted
for space reasons. For each B0 variable \B{v} found, a LLVM name is thus
created.
\begin{align*}
  \trad{v}{alloc} \isdef 
  \llvm{"\PH{\local(\B{v})} = alloca \PH{\trad{v.type}{\B{Type}}} \nl"}
\end{align*}

After the stack allocations, the LLVM source code contains the statements that
encodes the instructions in the body of the operation, as described in the
following section.

\subsection{Translation of instructions}
\label{sec:trad-instr}

Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
of B instructions \B{il} and a label $\ell$ and produces the source code of a
LLVM block with $\ell$ as exit point. It is defined as follows:

\begin{align*}
\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
\end{align*}

Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
instructions. Its definition is specialized for each class of instructions as
detailed in the rest of this section. Note that a new block is created after
each \B{If} and \B{While} instruction.

\paragraph{While instructions} are translated in two blocks of instructions:
block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
body:
\begin{align*}
\begin{split}
\trad{w, $\ell$}{\B{While}} \isdef 
  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
  & \llvm{"\PH{\ell_1} : \PH{c}} \\
  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
\end{split}
\end{align*}

\paragraph{If instructions} are sequences of possibly conditional blocks of
instructions, i.e. branches. After a branch is executed, control must go
to the next block $\ell$.
\begin{align*}
  \trad{i, $\ell$}{\B{If}}_L \isdef \trad{i.branches, $\ell$}{\ListOf{\B{IfBr}}}
\end{align*}
To translate conditional branches, one must first generate code to evaluate the
condition of the branch, followed by an LLVM conditional branching statement,
translate the body of the branch, and add an unconditional branch to the block
following the corresponding \B{If} statement.  The translation for branches has
as arguments a branch \B{b}, the remaining branches \B{bl}, the lexicon, and the
label $\ell$ of the block following the \B{If} statement they belong to. The
following definition applies for the case of conditional branches:
\begin{align*}
\begin{split}
  \trad{\B{b bl}, $\ell$}{\B{IfBr}} \isdef & \LET \ell_1 = \name \AND \ell_2 = \name \IN \\
  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}} \\
  & \llvm{\PH{\ell_2} : \PH{\trad{bl, $\ell$}{\ListOf{\B{IfBr}}}}"} \\
\end{split}
\end{align*}
The following rule handles the case of the last branch, which might have a condition or not:
\begin{align*}
\begin{split}
  \trad{\B{b}, $\ell$}{\B{IfBr}} \isdef & \IF \B{b.cond} = \text{None} \THEN
  \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
  & \ELSE \LET \ell_1 = \name \IN \\
  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}}
\end{split}
\end{align*}

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.
\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef 
  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
%   & l \conc r \conc 
%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%   \conc t \conc \llvm{"* "} \conc p \conc \nl
\end{split}
\end{align*}
The assignment target may be either local to the operation or a state
variable. In the former case, it can be referenced by the corresponding LLVM
identifier. In the latter case, it is represented as an element of the state
structure $\Sigma$, and its address must be calculated and assigned to a new temporary
variable.
\begin{align*}
\lefteqn{\trad{n}{lv} \isdef} \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
& \quad \ELSE \LET v = \name \AND t = \trad{n.type}{\B{Type}} \IN \\
& \quad \quad \llvm{"\PH{v} = load \PH{t}* getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
\end{align*}

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generated must include instructions to evaluate the
operation arguments, the address of the variables receiving the results as well
as an instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated, or an instance of an imported
machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"@\PH{c.pre}\$state\$"} \AND \\
  & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
\end{align*}

\paragraph{Block instructions} are structuring constructs that have no
operational intent; their translation is that of their body: $\trad{i,
  $\ell$}{\B{Blk}} \isdef \trad{i.body, $\ell$}{\ListOf{\B{Inst}}}$.

\subsection{Initialization}
\label{sec:trad-init}

Initialization is done is two steps: first all imported modules are initialized,
second the initialization clause is executed.
\begin{align*}
  \trad{i}{Init} \isdef
  & \llvm{"define @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
  & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
  & \quad\llvm{\PH{\trad{i.init, "exit"}{\B{Inst}}_L}} \\
  & \quad\llvm{exit: ret void \nl} \\
  & \llvm{\}"}
\end{align*}
The initialization of an imported module is performed by a call to the
corresponding initialization function:
\begin{align*}
  \trad{m}{InitImp} \isdef
  & \LET \ell = \name \IN \\
  & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}* \PH{\ell})"}
\end{align*}

\section{Conclusion and future work}
\label{sec:conclusion}

This paper presents the formal specification of a translation from a large
subset of the B implementation language to LLVM, a modern compiler internal
language.  Even though this is still a work in progress, the definition is
self-contained and has a large enough scope to be applied to B implementations
where the data belongs to basic types. This specification is a blueprint for a
code synthesis tool for B that is currently being implemented. This tool will be
distributed as an extension to Atelier-B under an open-source license.

The next step is to extend the scope to the full B implementation language. This
entails that the translation must be defined for aggregate data types and
adapted to handle some syntactic sugar. We are also planning for producing a
LLVM output with debugging information. Such output would be indeed very helpful
to provide feedback to the user when applying testing to validate the produced
code.

To prove the verification of the correctness of the translation, we would have
to define the semantics of both LLVM and B in a unified framework. One possible
starting point is Vellvm~\cite{vellvm}, a framework to reason about the
correctness of LLVM programs and transformations. We would then have to extend
the framework to include B implementation language. Another possible approach
would be to translate verification conditions from the B development artifacts
as assertinons in the generated LLVM code. The compiled program would include
checks that such assertions hold while executing.

\bibliographystyle{plain}
\bibliography{b2llvm}

\end{document}

### 2013-07-19 PAPERS/2013-SBMF/paper_short.tex
Organizing to SBMF short

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{A Proposed of Translation of B Implementations to the LLVM}
\author{David D\'{e}harbe \inst{1} \and Val\'{e}rio Medeiros Jr.\inst{2}}
\institute{Federal University of Rio Grande do Norte, UFRN, Brazil\\
\and 
Federal Institute of Education, Science and Technology  of Rio Grande do Norte, IFRN, Brazil}
\date{May 2013}

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\text{\Return}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\name}{\mathcal{L}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

\begin{document}
\maketitle

\begin{abstract}
  The aim of this work is to lay the foundations of a multi-platform code
  generator for the B method. In particular, this paper presents a translation
  procedure from a large subset of the B language for implementations towards
  LLVM source code. This translation is defined formally as a set of rules
  defined recursively on the abstract syntax for B implementations. It handles
  the following elements of the B language: simple data types, imperative
  instructions and component compositions. 
\end{abstract}

\section{Introduction}

B is a formal refinement-based software design method~\cite{Abrial1996}. It has
a single language encompassing abstract constructs suitable for specification,
as well as classic imperative constructs for computer programming. The starting
point of a B development is a specification, called a \emph{machine}, that is
incrementally refined to an \emph{implementation}, where only imperative-like
constructs may be employed~\cite{Clearsy}. Such implementation may then be
translated to source code in a programming language, say C~\cite{ComenC} or
Ada. All steps in the B method are formally verified using certified theorem
proving technologies. However the translation to some programming language and
its subsequent compilation to the target platform do not benefit from the same
mathematical rigor. In practice, redundancy in the tool chains and execution
platform are employed to increase the level of confidence to the desired levels.

The goal of this work is to connect the B method to the LLVM compilation
framework~\cite{Lattner04LLVM}.  LLVM is a free, open-source project upon which
many compiling contemporary technologies and tools are based. LLVM provides an
intermediate assembly language upon which may be applied techniques such as
optimization, static analysis, code generation, debugging, etc. Providing a
connection from the B method to this framework gives the users of the B method
access to these technologies. Our approach is to define a translation from B0
(the subset of the B language that is used to describe imperative programs) to
the LLVM intermediate format.

Considering the correctness of the proposed translation, we shall rely upon
human inspection for the time being. The standard uniform semantic framework for
B and LLVM that would be necessary to carry out a proof of this correctness does
not exist (yet). A possible approach to mitigate the risk of having an error in
the translation would be to include assertions in the target code. Indeed the
artifacts produced in the B development have many such assertions and it would
be possible to translate at least some of them (B has an assertion instruction
that introduces additional verification conditions in the development). The
produced code could then be (at least partially) verified by testing the absence
of assertion violations.

\paragraph{Overview.} Following this introduction, the paper is organized as
follows. The implementation constructs of the B language, and selected elements
of the LLVM are presented in section~\ref{sec:b-impl}
and~\ref{sec:llvm}. Section~\ref{sec:overview} presents a bird's-eye view that
is then detailed in sections~\ref{sec:module} (for component modularity)
~\ref{sec:data} (for data aspects),~\ref{sec:expr} (expressions) and
instructions~\ref{sec:control}. We conclude and discuss future work
in~\ref{sec:conclusion}.

\section{Implementation Language in the B-method}
\label{sec:b-impl}

This section describes the B implementation constructs that are considered in
the compilation to the LLVM. This description introduces the notations used in
the definition of the translation. The elements of the B language are denoted
using a \B{sans serif} font.

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{imp}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

The initialization and the operations are defined as (sequences of)
instructions. Operations also have a name, inputs and outputs. Possible
instructions are the classic imperative constructs: variable assignment, if and
case conditional, while loop, block with or without local variables, and
operation calls. The if constructs may have several branches, each with a
condition as well as an optional else branch. Called operations may be either
local operations or operations of instances of external modules, in this case
they have an optional prefix identifying the module.

Note that the translation presented in this paper only considers types \B{INT}
(integer) and \B{BOOL} (Boolean). Accordingly, the expression language consists
in arithmetic operations \B{-} (unary and binary), \B{+}, \B{*}, \B{/}, \B{mod},
% \B{**} (exponentiation), 
\B{succ} and \B{pred} and Boolean operations
\B{$\land$}, \B{$\lor$}, \B{$\neg$} as well as relations \B{=}, \B{$\neq$},
\B{$>$}, \B{$<$}, \B{$\leq$} and \B{$\neq$}. Atomic expressions are identifiers,
Boolean constants \B{FALSE} and \B{TRUE}, integer literals, and integer
constants \B{MAXINT} and \B{MININT}. The development through the application of
the B method ensures the absence of overflows and underflows.
%TODO optionally to remove unsed elements in this table
\begin{figure}[H]
  \begin{center}
    {\footnotesize
      \frame{
    \begin{tabular}[t]{ccccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{see} & : & seq \B{Name} \\
        \B{imp} & : & seq \B{Impo} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{pre} & : & opt \B{Name} \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}
        \\
      \end{tabular}
      & &
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \\
        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{Name} \\
        \B{rhs} & : & \B{Expr} \\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Name} \\
        \B{pre} & : & opt \B{Name} \\
        \B{inp} & : & seq \B{Expr} \\
        \B{out} & : & seq \B{Name}
        \\
%         \\
%         \hline
%         \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
%         \hline
%         \B{expr} & : & \B{Expr} \\
%         \B{branches} & : & seq \B{CaseBr}
      \end{tabular}
      & &
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
%         \hline
%         \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
%         \hline
%         \B{val} & : & seq \B{Value} \\
%         \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
        \hline
        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred}} \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
        \hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}}
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

\section{Target LLVM Subset}
\label{sec:llvm}

LLVM aims to be a general, language agnostic, intermediate representation for
compilers. The compiler front-end translates the source programming language to
LLVM source code, upon which optimization and other static analysis tasks may be
performed, and the compiler back-end translates to the target platform assembly
language. One key feature of LLVM is that it is a single-static assignment (SSA)
language: a variable may only be assigned once. So for each assignment to a
variable found in the source language, the front-end shall generate a new
temporary variable in the LLVM program. For illustration,
figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
C code appears on the left) with two temporary variables \llvm{\%0} and
\llvm{\%1}.
\begin{figure}
\vspace*{-2.4em}
  \begin{minipage}[t]{.3\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%      \B{VARIABLES}\\
%      \> y\\
      inc(x) =\\
      \B{PRE} x : INT \B{THEN}   \\
      \> 	y = x+y\\
      \B{END} \\
      \end{tabbing}
  \end{minipage}
  \:\vline\:
  \begin{minipage}[t]{.2\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 int y;\\      
      void inc(int x) \{\\
      \> 	y = x+y; \\
      \}\end{tabbing}
  \end{minipage}
	  \:\vline\:
  \begin{minipage}[t]{.6\textwidth}
    \begin{tabbing}
      foo \= \kill  \\
%	 \llvm{@y = global i32 0 } \\      
      \llvm{define void @inc(i32 \%x) nounwind \{} \\
      \llvm{entry:}                      \\
      \> \llvm{\%0 = load i32* @y, align 4}     \\
      \> \llvm{\%1 = add nsw i32 \%0, \%x}     \\
      \> \llvm{store i32 \%1, i32* @y, align 4} \\
      \> \llvm{ret void} \\
      \llvm{\}}
    \end{tabbing}

  \end{minipage}
  \caption{Simple example of a B function and its corresponding  C function and LLVM function.
    It is not a real translation, but it help to understand the translation %TODO to update the variables
    The first line contains the signature: return type \llvm{void}, the name
    \llvm{@inc} and one parameter named \llvm{\%pi} and typed \llvm{i32*}.  Next
    is the body with a single block, labeled \llvm{entry}, and temporary
    variables \llvm{\%0} and \llvm{\%1}, created in the conversion to SSA. The
    block has four instructions: \llvm{load}, \llvm{add}, \llvm{store} and
    \llvm{ret}. For instance, \llvm{\%1 = add i32 \%0, 1} is an addition
    (\llvm{add}), has result type \llvm{i32} and assigns to
    \llvm{\%1} the sum of variable \llvm{\%0} and integer
    literal \llvm{1}.}
  \label{fig:ex-llvm}
\end{figure}

%The introduction to the LLVM that follows is focused in the features
%that are relevant for the representation of B
%implementations. Figure~\ref{fig:llvm-grammar} presents a grammar of this subset
%of the LLVM. 
%
%\begin{figure}
%  \begin{center}
%    \begin{tabular}{rcl}
%      \nt{module} & ::= & \ListOf{\nt{item}} \\
%      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
%      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
%      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
%      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
%      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{struct} \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
%      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
%      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
%      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
%      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
%      \nt{param} & ::= & \nt{type} \nt{name} \\
%      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
%      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
%      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
%      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
%      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
%      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
%      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
%      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
%      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
%      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
%%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
%      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
%      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
%      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
%      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
%      \nt{arg} & ::= & \nt{type} \nt{exp}
%    \end{tabular}
%  \end{center}
%  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
%    and \nt{name} correspond respectively to integer types, integer literals,
%    labels and names. Choices are separated by \alt and optionally delimited by
%    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
%    elements of the annotated entity.}
%  \label{fig:llvm-grammar}
%\end{figure}
%
%LLVM programs are organized into modules, one per translation unit. A module may
%contain declarations of external entities (functions, constants) and definitions
%of internal items (functions, variables, constants). All data is explicitly
%typed. The name and type of external entities must be declared.  All names,
%e.g. non-reserved identifiers, shall start with \llvm{@}, when they are global,
%or\llvm{\%}, when they are local. For instance, \llvm{@max = external constant
%  i32} declares \llvm{@max} as a 32-bit integer constant and \llvm{declare void
%  @inc(i32*)} declares \llvm{@inc} as a function having as parameter a pointer
%to an integer and \llvm{void} as a return type.
%
%The type system contains the empty type \llvm{void}, an infinite, countable
%number of integer types, one for each possible bit width (e.g. \llvm{i8} is the
%type for 8-bit integers), and type constructors pointer (monadic \llvm{$\cdot$
%  *}) and structure (polyadic \llvm{\{$\cdots$\}}). For instance \llvm{\{ i8*,
%  i8, i8 \}} is the type for structures with three fields, the first having as
%type pointer to \llvm{i8}. Grammar rule \nt{type\_def} states how types are
%named, e.g.: \llvm{\%T1 = type \{i32, i32\}} and \llvm{\%T2 = type \{\%T1 *,
%  \%T1 *\}}. Note that there is no \llvm{void *} type in the LLVM, instead
%pointer values are of integer types. 
%
%Local entities are constants, variables or functions. An example of constant
%definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
%and value. The elements of a variable definition are the name, type and code
%generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
%Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
%information for target code generation: linkage type (the variable is merged
%with other of the same name), the scope (global) and initialization of the
%variable (zeroed is mandatory for common linkage). LLVM defines other attributes
%for these aspects, but they are not needed in this work and their presentation
%is omitted.  Function definitions are composed of the signature and body. The
%signature contains the return type, name, parameters, and attributes for target
%code generation. Attribute \llvm{nounwind} states that the function never
%generate exceptions. The body is a sequence of blocks of instructions in
%single-static assignment (SSA) form, i.e. are such that each variable may be
%assigned exactly once.
%
%The grammar rule \nt{inst} describes the different kind of
%instructions. Instruction \llvm{alloca} allocates a memory block the size of the
%given type from the stack segment. This and the following classes of
%instructions assign the resulting value to a fresh variable. Arithmetic
%operations are binary, and include signed quotient and reminder. Arithmetic
%comparisons compare two values according to the first operand and return a 1-bit
%integer value.  Instruction \llvm{call} invokes the given function with the
%given arguments and assigns the result to a fresh variable, which is of the
%given type.  In general, \llvm{getelementptr} gets the address of an element in
%an aggregate object through indexing. In the target LLVM subset, aggregate
%objects are structures, the first index has type \llvm{i32} and value \llvm{0}
%to select the first structure at the given location \nt{exp}, and the second
%index selects a field in that structure.  Instruction \llvm{load} assigns to a
%fresh variable \nt{name} the contents of a memory address of type \nt{type}
%specified by \nt{exp} (e.g. in fig.~\ref{fig:ex-llvm}).  Instruction
%\llvm{store} writes a value to memory address (again, see
%fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two forms.  In the first
%form, it directs conditionally the control flow to one of two blocks.  If the
%given expression evaluates to \llvm{1}, control goes to the first block,
%otherwise to the second. In the second form, the control jumps unconditionally
%to the given block.
%% Instruction \llvm{switch} directs the
%% control flow to one of several blocks. The given expression is evaluated and
%% control goes to the first branch with the matching value, or to the default
%% branch (appearing first in the instruction).  
%Finally, instruction \llvm{ret}
%ends the current function call, optionally returning a value.
%
%The expression language is limited to names (local and global), integer literals
%and selection of an element in a structure.
%
%\section{Translation: a Bird's-Eye View}
%\label{sec:overview}
%
%The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
%module, say \llvm{M}. A B implementation \B{M} has several sections, each
%translated as follows:
%\begin{enumerate}
%\item Visibility clauses provide access to definitions found in external
%  B components and are translated to external declarations in LLVM.
%\item Constants and value definitions are translated to LLVM constants. To avoid
%  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
%  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
%\item To encode the implementation state, the translation produces a
%  structure type, named \llvm{@M\$state\$}, and an instance of that type, named
%  \llvm{@M\$self\$}. All the variables and instances of imported machines
%  are encoded as elements of that structure.  Also, for each operation of each
%  imported machine, an external function declaration is produced.
%\item Each operation \B{op} of the implementation is translated into the
%  definition of a function named \llvm{@M\$op}; also the translation produces a
%  special function \llvm{@M\$init\$} responsible for the initialization.
%
%  The parameter list of the functions thus produced contain one item for each
%  input and output of the corresponding operation, and one item of type
%  \llvm{@M\$state\$*}, to pass the address of the structure encoding the
%  state of the implementation and its modification by the callee. The return
%  type is always \llvm{void}.
%\end{enumerate}
%This approach assumes that information on external B components used in \B{M} is
%available during translation. To fulfill this hypothesis, the translation
%produces an ``interface'' artifact containing the information that the
%translation of an importation of \B{M} in a client machine
%requires. Figure~\ref{fig:skel} summarizes the translation principles by
%providing the structure of both the interface and the LLVM implementation thus
%produced.
%
%\begin{figure}
%  \begin{center}
%    (a) Structure of the interface artifact produced from a B implementation \B{M}:
%    \begin{tabbing}
%      foo \= foo \= \kill
%      \textit{(references to imported components)} \\
%      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
%      \llvm{declare void @M\$init\$(@M\$state\$*)} \\
%      \textit{for each constant} \B{cst} \textit{generate} \\
%      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
%      \textit{for each operation} \B{op} \textit{generate} \\
%      \> \llvm{declare void @M\$op(@M\$state\$*, \ListOf{\nt{type}})} \\
%    \end{tabbing}
%    (b) Structure of the LLVM module produced from a B implementation \B{M}:
%    \begin{tabbing}
%      foo \= foo \= \kill
%      \textit{for each imported component} \B{Q} \textit{generate} \\
%      \> \llvm{@Q\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
%      \> \llvm{declare void @Q\$init\$(@Q\$state\$*)} \\
%      \> \textit{for each constant} \B{cst} \textit{generate} \\
%      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
%      \> \textit{for each operation} \B{op} \textit{generate} \\
%      \> \> \llvm{declare void @Q\$op(@Q\$state\$*, \ListOf{\nt{type}})} \\
%      \textit{for each constant} \B{cst} \textit{generate} \\
%      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
%      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
%      \llvm{@M\$self\$ = common global @M\$state\$ zeroinitializer} \\
%      \llvm{define void @M\$init\$(@M\$state\$* \%self\$) \{} \\
%      \> \llvm{\ListOf{\nt{block}}} \\
%      \> \llvm{exit: ret void} \\
%      \llvm{\}} \\
%      \textit{for each operation} \B{op} \textit{generate} \\
%      \> \llvm{define void @M\$op(@M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
%      \> \> \llvm{\ListOf{\nt{block}}} \\
%      \> \> \llvm{exit: ret void} \\
%      \> \llvm{\}}
%    \end{tabbing}
%  \end{center}
%  \caption{Skeleton of the artifacts produced by the translation.}
%  \label{fig:skel}
%\end{figure}
%
The detailed presentation of the translation of the different elements of a B
implementation is divided into:
\begin{description} %TODO remove the references and to organize it.
\item[Modularity] constructs to combine B components are described in
  sec.~\ref{sec:module}.
\item[Data] translation is briefly described sec.~\ref{sec:data}, including
  the encoding of B data types.
\item[Expressions] and conditions are codified as LLVM instruction blocks, as
  described in section~\ref{sec:expr}.
\item[Control] flow, instructions are translated to LLVM instructions as
  discussed in sec.~\ref{sec:control}, which also presents the encoding
  of B operations and initialization.
\end{description}

Code generation is defined with a set of rules defined by structural
recursion. For each grammatical construct, a code generation function is
defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name. Also $\state(\B{M})
\isdef \llvm{"@\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
state of component \B{M}. We assume an unlimited supply of labels and local
variables: let $\name$ return a fresh LLVM label or local variable name.
%
%\section{Translation: Modularity Aspects}
%\label{sec:module}
%
%The translation of a B module produces two artifacts: an interface and an (LLVM)
%implementation. In the following, we describe how they are produced.
%
%\subsection{Interface}
%
%The definition of the interface $\intf(\B{M})$ of a component \B{M} has four
%key elements, as shown in figure~\ref{fig:skel} (b):
%\begin{enumerate}
%\item $\intf(\B{M})_u$ is the list of used components: $\intf(\B{M})_u \isdef \B{M.imp}$.
%
%\item $\intf(\B{M})_t$ is a textual description of the type representing the
%  state of the component, i.e. a structure with one element per imported
%  component and variable:
%\begin{small}
%$$\intf(\B{M})_t \isdef \llvm{"struct \{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
%\end{small}
%where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
%$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.
%
%\item $\intf(\B{M})_c$ is the definition of the component constants. It maps
%the constant names to their type:
%$\intf(\B{M})_c \isdef \trad{\B{M.const}}{\ListOf{\B{Cons}}}$ and
%$\trad{\B{c}}{\B{Cons}} \isdef \B{c} \mapsto \trad{\B{c.type}}{\B{Type}}$.
%
%\item $\intf(\B{M})_o$ represents the operation declarations. It maps
%operation names to the list of input parameter types and output parameter
%types:
%$\intf(\B{M})_o \isdef \trad{\B{M.op}}{\ListOf{\B{Oper}}}_t$ and
%$\trad{\B{o}}{\B{Oper}}_t \isdef \B{o} \mapsto ( \trad{\B{o.inp}}{\ListOf{inp}} \sep \trad{\B{o.out}}{\ListOf{out}} )$, where
%$\trad{\B{i}}{inp} \isdef \trad{\B{i.type}}{\B{Type}}$ and
%$\trad{\B{o}}{out} \isdef \llvm{"\PH{\trad{\B{o.type}}{\B{Type}}}*"}$.
%\end{enumerate}
%
%\subsection{Implementation}
%
%The actual LLVM source code is generated following the template shown in
%figure~\ref{fig:skel} (a), as follows:
%\begin{align*}
%  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
%  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
%  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
%  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
%& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
%  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
%  \quad \PH{\trad{\B{i}}{State}}} \\
%& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
%& \llvm{\PH{\trad{\B{i.init}}{Init}} 
%  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
%\end{align*}
%We shall now define each rule referenced in this template.
%\begin{enumerate}
%\item For the type definitions corresponding to the state of the imported
%  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
%    type \PH{\intf(\B{i})_t}\nl"}$.
%\item For the declarations of the initialization functions in the imported
%  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
%    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
%\item For the definitions of the imported constants: $\trad{\B{i}, \intf}{ICons}
%  \isdef \trad{$\intf(\B{i})_c$}{\ListOf{Cons}}$, where
%  \noindent $\trad{$\B{c} \mapsto t)$}{Cons} \isdef \llvm{"\PH{\Global(\B{c})} =
%    external constant \PH{t}\nl"}$.
%\item For the declarations of the imported operations: $\trad{\B{i},
%    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
%  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
%    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
%\item The translation of constant definitions is presented in section~\ref{sec:data}.
%\item For the definition of the type for component states: \\
%\noindent$\trad{\B{M}}{State} \isdef
%\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
%\item For the initialization, the definition of a function is generated. It
%  first initializes the imported modules and then the current module (see
%  section~\ref{sec:control}).
%\item For each operation, a function definition is generated (see
%  section~\ref{sec:control}).
%\end{enumerate}
%Let $\idx$ be a function that maps each imported machine and state variable to
%its position (starting from zero).
%\section{Translation: Data}
%\label{sec:data}
%
%The translation presented in this paper is currently limited to integers and
%booleans. In the case of integers, the translation is based on the supposition
%that the B0 definition of \B{INT} is the range $-2^{31}.. 2^{31}-1$. We have
%$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}$ and $\trad{INT}{\B{Type}} \isdef
%\llvm{i32}$.
%
%B constants are translated to constants in the LLVM program, as described in the
%following rule, where parameter \B{c} denote the constant, $M$ the name of the
%module and $\Global$ is the mapping between source and target identifiers. Two
%results are produced: the LLVM source code and an updated identifier mapping:
%\begin{small}
%\begin{align*}
%  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
%\end{align*}
%\end{small}
%The definition of the translation for expressions is described next in
%section~\ref{sec:expr}.
%
\section{Translation: Expressions
  \label{sec:expr}}

To evaluate a composed expression using the LLVM, one must first generate a
sequence of instructions to evaluate each argument, the resulting value being
stored in a temporary variable, and second generate an instance of the LLVM
instruction corresponding to the expression argument.  Translation of
expressions is defined as functions that take as input some class of expressions
and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
sequence of instructions required in the evaluation (e.g.  for the
sub-expressions), $v$ is the LLVM expression representing the resulting value,
and $t$ represents the LLVM type of the expression. In the case of composed
predicates, though, the evaluation is different as boolean operations need to be
implemented with conditional and unconditional jumps.

\paragraph{Atomic expressions} have a simple translation. In the case of
literals, one only needs to issue the corresponding LLVM literal:
$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
$\B{l.type} = \B{INT}$.

For identifiers, three cases are possible: global constants, local variables,
and module state variables. For state variables, the translation issues
statements to assign the address of the corresponding element in the state
structure to pointer variable $p$ and then dereference $p$ to assign $v$:
\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\end{align*}

\paragraph{Comparisons} are formed by the application of a relational operator
to two expressions. Mapping $\opmap$ associates B operators to corresponding
LLVM keywords:
$$\opmap = \{ = \mapsto \llvm{"eq"}, 
  \ne \mapsto \llvm{"ne"}, 
  < \mapsto \llvm{"slt"}, 
  \le \mapsto \llvm{"sle"}, 
  > \mapsto \llvm{"sgt"}, 
  \ge \mapsto \llvm{"sge"} \}.
$$

The code responsible for evaluating a comparison \B{r} is defined by the
following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
in the fresh variable $v$.
\begin{align*}
\lefteqn{\trad{r}{\B{Comp}} \isdef
  \LET
  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
& \AND v = \name \IN \\
& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
\end{align*}

\paragraph{Arithmetic operations} have a similar structure and interpretation
as comparisons and their presentation is mostly omitted. Let us present the
translation of the unary \B{succ} operator though:
\begin{align*}
  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
\end{align*}

\paragraph{Predicates}
The translation of conditions takes as input a condition node \B{n}, as well as
two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
execution shall jump when the condition is evaluted to true or false,
respectively.  The first rule defines the translation of an atomic relation:
\begin{align*}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
\end{align*}

\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
following rule handles the case of conjunctions. Note that it produces code with
a ``short-cut'' when the evaluation of the first argument yields false:
\begin{align*}
\begin{split}
  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
\end{split}
\end{align*}
The definition of the translation for disjunctions is similar and is omitted.

%\section{Translation: Control Flow and Instructions
%  \label{sec:control}}
%
%This section addresses how operations in B implementations are translated to
%LLVM functions. It also discusses initialization, which might be seen as a
%special kind of operation. For operations, the presentation of the translation
%is divided into three steps: the signature, the local data, and the body
%instructions.
%
%The signature of the function is composed of the name of the function, the
%result type, the name and type of the parameters. The signature may carry other
%information such as linkage, visibility, calling convention, etc. The B parser
%provides this information after the type checking phase has been completed.  On
%the one hand LLVM functions may only have one result, on the other hand B0
%operations may produce several results. The translation of output parameters
%will be function parameters passed by reference, i.e. they are pointers. For
%uniformity of treatment, the return type of the produced LLVM functions is
%\llvm{void}. Translation of operation headers is defined in
%sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
%the stack, using the \llvm{alloca} instruction. We describe a function in
%sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
%type of its local variables used in the operation. These names are translated to
%conform to LLVM restrictions. The body of the operation has to be transformed to
%a sequence of LLVM blocks of statements. For each kind of instruction one (or
%more) translation function is defined. The details are provided in
%section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
%combined as follows:
%\begin{align*}
%\begin{split}
%  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
%  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
%  & \llvm{"define void \PH{h} \{\nl} \\
%  & \quad \llvm{entry:\nl} \\
%  & \quad \quad \PH{a} \\
%  & \quad \quad \PH{i} \\
%  & \quad \llvm{exit: ret void\nl} \\
%  & \llvm{\}\nl"}
%\end{split}
%\end{align*}
%
%Initialization of a B implementation shares many characteristics with operations
%but needs to deal with initialization of imported modules. It is discussed in
%sec.~\ref{sec:trad-init}.
%
%\subsection{Sub-routines signatures}
%\label{sec:trad-header}
%
%The following rule specifies the translation of operation headers. Its
%parameters are the operation identifier, inputs, outputs, the name of the
%machine it belongs to, and the results are the LLVM function definition
%header: \\
%\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
%
%Next are the rules defining the translation of individual input and output
%parameters. They have local scope and they are translated by prepending
%\llvm{\%}:
%\begin{align*}
%  \trad{i}{inp} \isdef 
%  \llvm{"\PH{\trad{i.type}{\B{Type}}} \PH{\local(\B{i})}"} \quad \quad
%  \trad{o}{out} \isdef 
%  \llvm{"\PH{\trad{o.type}{\B{Type}}}* \PH{\local(\B{o})}"}
%\end{align*}
%
%\subsection{Stack allocation}
%\label{sec:trad-alloc}
%
%B operations may have local variables, in which case the translator needs to
%issue LLVM stack allocation instructions. Such instructions have two arguments:
%the type of values to store, and a temporary variable that is assigned the
%address (see figure~\ref{fig:llvm-grammar} for the concrete syntax). The
%generation of stack allocation code visits the parse tree of the operation body
%in a pre-order depth-first traversal, generating one statement for each local
%variable declaration found. The detailed definition of this traversal is omitted
%for space reasons. For each B0 variable \B{v} found, a LLVM name is thus
%created.
%\begin{align*}
%  \trad{v}{alloc} \isdef 
%  \llvm{"\PH{\local(\B{v})} = alloca \PH{\trad{v.type}{\B{Type}}} \nl"}
%\end{align*}
%
%After the stack allocations, the LLVM source code contains the statements that
%encodes the instructions in the body of the operation, as described in the
%following section.
%
%\subsection{Translation of instructions}
%\label{sec:trad-instr}
%
%Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
%of B instructions \B{il} and a label $\ell$ and produces the source code of a
%LLVM block with $\ell$ as exit point. It is defined as follows:
%
%\begin{align*}
%\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
%& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
%& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
%& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
%\end{align*}
%
%Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
%instructions. Its definition is specialized for each class of instructions as
%detailed in the rest of this section. Note that a new block is created after
%each \B{If} and \B{While} instruction.
%
%\paragraph{While instructions} are translated in two blocks of instructions:
%block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
%body:
%\begin{align*}
%\begin{split}
%\trad{w, $\ell$}{\B{While}} \isdef 
%  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
%  & \llvm{"\PH{\ell_1} : \PH{c}} \\
%  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
%  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
%\end{split}
%\end{align*}
%
%\paragraph{If instructions} are sequences of possibly conditional blocks of
%instructions, i.e. branches. After a branch is executed, control must go
%to the next block $\ell$.
%\begin{align*}
%  \trad{i, $\ell$}{\B{If}}_L \isdef \trad{i.branches, $\ell$}{\ListOf{\B{IfBr}}}
%\end{align*}
%To translate conditional branches, one must first generate code to evaluate the
%condition of the branch, followed by an LLVM conditional branching statement,
%translate the body of the branch, and add an unconditional branch to the block
%following the corresponding \B{If} statement.  The translation for branches has
%as arguments a branch \B{b}, the remaining branches \B{bl}, the lexicon, and the
%label $\ell$ of the block following the \B{If} statement they belong to. The
%following definition applies for the case of conditional branches:
%\begin{align*}
%\begin{split}
%  \trad{\B{b bl}, $\ell$}{\B{IfBr}} \isdef & \LET \ell_1 = \name \AND \ell_2 = \name \IN \\
%  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
%  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}} \\
%  & \llvm{\PH{\ell_2} : \PH{\trad{bl, $\ell$}{\ListOf{\B{IfBr}}}}"} \\
%\end{split}
%\end{align*}
%The following rule handles the case of the last branch, which might have a condition or not:
%\begin{align*}
%\begin{split}
%  \trad{\B{b}, $\ell$}{\B{IfBr}} \isdef & \IF \B{b.cond} = \text{None} \THEN
%  \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
%  & \ELSE \LET \ell_1 = \name \IN \\
%  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
%  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}}
%\end{split}
%\end{align*}
%
%\paragraph{Becomes equal instructions} must evaluate the target and the source
%of the assignment, and then copy the result of the latter in the former, using
%the LLVM instruction \llvm{store}. The evaluation of the target is defined with
%$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
%variable $p$ containing the assigned location. The evaluation of the source
%yields the corresponding code $r$, the variable $v$ holding the result, and its
%type $t$.
%\begin{align*}
%\begin{split}
%  \trad{a}{\B{Beq}} \isdef 
%  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
%  & \llvm{"\PH{l}} \\
%  & \llvm{\PH{r}} \\
%  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
%%   & l \conc r \conc 
%%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%%   \conc t \conc \llvm{"* "} \conc p \conc \nl
%\end{split}
%\end{align*}
%The assignment target may be either local to the operation or a state
%variable. In the former case, it can be referenced by the corresponding LLVM
%identifier. In the latter case, it is represented as an element of the state
%structure $\Sigma$, and its address must be calculated and assigned to a new temporary
%variable.
%\begin{align*}
%\lefteqn{\trad{n}{lv} \isdef} \\
%& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
%& \quad \ELSE \LET v = \name \AND t = \trad{n.type}{\B{Type}} \IN \\
%& \quad \quad \llvm{"\PH{v} = load \PH{t}* getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
%& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
%\end{align*}
%
%\paragraph{Call up} instructions apply B operations with some values and obtain
%results. The LLVM code generated must include instructions to evaluate the
%operation arguments, the address of the variables receiving the results as well
%as an instance of the LLVM \llvm{call} instruction. In addition, the proposed
%translation includes as argument the address of a structure storing the state of
%the component corresponding to this operation. Note that this component may be
%either the implementation being translated, or an instance of an imported
%machine.
%\begin{align*}
%  \trad{c}{\B{Call}} \isdef 
%  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
%  \trad{c.out}{\ListOf{Out}} \AND \\
%  & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
%  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"@\PH{c.pre}\$state\$"} \AND \\
%  & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
%  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
%  & \IN \\
%  & \quad \llvm{"\PH{p_i}} \\
%  & \quad \llvm{\PH{p_o}} \\
%  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
%\end{align*}
%The translation of input and output arguments produces two code snippets. The
%first corresponds to the LLVM instruction sequence $p$ responsible for
%evaluating the argument, and the second to the type $t$ and name $v$ of the
%variable storing the resulting value:
%\begin{align*}
%  \trad{i}{Inp} \isdef 
%  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
%  \trad{o}{Out} \isdef 
%  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
%\end{align*}
%
%\paragraph{Block instructions} are structuring constructs that have no
%operational intent; their translation is that of their body: $\trad{i,
%  $\ell$}{\B{Blk}} \isdef \trad{i.body, $\ell$}{\ListOf{\B{Inst}}}$.
%
%
%\subsection{Initialization}
%\label{sec:trad-init}
%
%Initialization is done is two steps: first all imported modules are initialized,
%second the initialization clause is executed.
%\begin{align*}
%  \trad{i}{Init} \isdef
%  & \llvm{"define @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
%  & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
%  & \quad\llvm{\PH{\trad{i.init, "exit"}{\B{Inst}}_L}} \\
%  & \quad\llvm{exit: ret void \nl} \\
%  & \llvm{\}"}
%\end{align*}
%The initialization of an imported module is performed by a call to the
%corresponding initialization function:
%\begin{align*}
%  \trad{m}{InitImp} \isdef
%  & \LET \ell = \name \IN \\
%  & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
%  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}* \PH{\ell})"}
%\end{align*}

\section{Conclusion and future work}
\label{sec:conclusion}

This paper presents the formal specification of a translation from a large
subset of the B implementation language to LLVM, a modern compiler internal
language.  Even though this is still a work in progress, the definition is
self-contained and has a large enough scope to be applied to B implementations
where the data belongs to basic types. This specification is a blueprint for a
code synthesis tool for B that is currently being implemented. This tool will be
distributed as an extension to Atelier-B under an open-source license.

The next step is to extend the scope to the full B implementation language. This
entails that the translation must be defined for aggregate data types and
adapted to handle some syntactic sugar. We are also planning for producing a
LLVM output with debugging information. Such output would be indeed very helpful
to provide feedback to the user when applying testing to validate the produced
code.

To prove the verification of the correctness of the translation, we would have
to define the semantics of both LLVM and B in a unified framework. One possible
starting point is Vellvm~\cite{vellvm}, a framework to reason about the
correctness of LLVM programs and transformations. We would then have to extend
the framework to include B implementation language. Another possible approach
would be to translate verification conditions from the B development artifacts
as assertinons in the generated LLVM code. The compiled program would include
checks that such assertions hold while executing.

\bibliographystyle{plain}
\bibliography{b2llvm}

\end{document}

### 2013-07-19 PAPERS/2013-SBMF/paper_short.tex
no message

@@
 \pagestyle{headings}
-\title{A Proposed of Translation of B Implementations to the LLVM}
+\title{A Proposed Translation of B Implementations to the LLVM}
 \author{David D\'{e}harbe \inst{1} \and Val\'{e}rio Medeiros Jr.\inst{2}}
@@
   \caption{Simple example of a B function and its corresponding  C function and LLVM function.
-    It is not a real translation, but it help to understand the translation %TODO to update the variables
+    It is not a real translation, but it helps to understand the translation. %TODO to update the variables
     The first line contains the signature: return type \llvm{void}, the name

### 2013-07-20 PAPERS/2013-SBMF/paper_short.tex
no message

@@
       \B{PRE} x : INT \B{THEN}   \\
-      \> 	y = x+y\\
+      \> 	y := x+y\\
       \B{END} \\
@@
 %
-\section{Translation: Expressions
-  \label{sec:expr}}
-
-To evaluate a composed expression using the LLVM, one must first generate a
-sequence of instructions to evaluate each argument, the resulting value being
-stored in a temporary variable, and second generate an instance of the LLVM
-instruction corresponding to the expression argument.  Translation of
-expressions is defined as functions that take as input some class of expressions
-and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
-sequence of instructions required in the evaluation (e.g.  for the
-sub-expressions), $v$ is the LLVM expression representing the resulting value,
-and $t$ represents the LLVM type of the expression. In the case of composed
-predicates, though, the evaluation is different as boolean operations need to be
-implemented with conditional and unconditional jumps.
-
-\paragraph{Atomic expressions} have a simple translation. In the case of
-literals, one only needs to issue the corresponding LLVM literal:
-$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
-$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
-$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
-$\B{l.type} = \B{INT}$.
-
-For identifiers, three cases are possible: global constants, local variables,
-and module state variables. For state variables, the translation issues
-statements to assign the address of the corresponding element in the state
-structure to pointer variable $p$ and then dereference $p$ to assign $v$:
-\begin{align*}
-\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
-& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
-& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
-\llvm{""} \sep \local(\B{n}) \sep t \\
-& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
-& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
-& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
-\end{align*}
-
-\paragraph{Comparisons} are formed by the application of a relational operator
-to two expressions. Mapping $\opmap$ associates B operators to corresponding
-LLVM keywords:
-$$\opmap = \{ = \mapsto \llvm{"eq"}, 
-  \ne \mapsto \llvm{"ne"}, 
-  < \mapsto \llvm{"slt"}, 
-  \le \mapsto \llvm{"sle"}, 
-  > \mapsto \llvm{"sgt"}, 
-  \ge \mapsto \llvm{"sge"} \}.
-$$
-
-The code responsible for evaluating a comparison \B{r} is defined by the
-following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
-an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
-the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
-in the fresh variable $v$.
-\begin{align*}
-\lefteqn{\trad{r}{\B{Comp}} \isdef
-  \LET
-  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
-  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
-& \AND v = \name \IN \\
-& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
-\end{align*}
-
-\paragraph{Arithmetic operations} have a similar structure and interpretation
-as comparisons and their presentation is mostly omitted. Let us present the
-translation of the unary \B{succ} operator though:
-\begin{align*}
-  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
-  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
-  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
-\end{align*}
-
-\paragraph{Predicates}
-The translation of conditions takes as input a condition node \B{n}, as well as
-two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
-execution shall jump when the condition is evaluted to true or false,
-respectively.  The first rule defines the translation of an atomic relation:
-\begin{align*}
-  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
-  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
-  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
-\end{align*}
-
-\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
-  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
-following rule handles the case of conjunctions. Note that it produces code with
-a ``short-cut'' when the evaluation of the first argument yields false:
+
+%\section{Translation: Expressions
+%  \label{sec:expr}}
+%
+%To evaluate a composed expression using the LLVM, one must first generate a
+%sequence of instructions to evaluate each argument, the resulting value being
+%stored in a temporary variable, and second generate an instance of the LLVM
+%instruction corresponding to the expression argument.  Translation of
+%expressions is defined as functions that take as input some class of expressions
+%and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
+%sequence of instructions required in the evaluation (e.g.  for the
+%sub-expressions), $v$ is the LLVM expression representing the resulting value,
+%and $t$ represents the LLVM type of the expression. In the case of composed
+%predicates, though, the evaluation is different as boolean operations need to be
+%implemented with conditional and unconditional jumps.
+%
+%\paragraph{Atomic expressions} have a simple translation. In the case of
+%literals, one only needs to issue the corresponding LLVM literal:
+%$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
+%$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
+%$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
+%$\B{l.type} = \B{INT}$.
+%
+%For identifiers, three cases are possible: global constants, local variables,
+%and module state variables. For state variables, the translation issues
+%statements to assign the address of the corresponding element in the state
+%structure to pointer variable $p$ and then dereference $p$ to assign $v$:
+%\begin{align*}
+%\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
+%& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
+%& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
+%\llvm{""} \sep \local(\B{n}) \sep t \\
+%& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
+%& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
+%& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
+%\end{align*}
+%
+%\paragraph{Comparisons} are formed by the application of a relational operator
+%to two expressions. Mapping $\opmap$ associates B operators to corresponding
+%LLVM keywords:
+%$$\opmap = \{ = \mapsto \llvm{"eq"}, 
+%  \ne \mapsto \llvm{"ne"}, 
+%  < \mapsto \llvm{"slt"}, 
+%  \le \mapsto \llvm{"sle"}, 
+%  > \mapsto \llvm{"sgt"}, 
+%  \ge \mapsto \llvm{"sge"} \}.
+%$$
+%
+%The code responsible for evaluating a comparison \B{r} is defined by the
+%following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
+%an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
+%the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
+%in the fresh variable $v$.
+%\begin{align*}
+%\lefteqn{\trad{r}{\B{Comp}} \isdef
+%  \LET
+%  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
+%  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
+%& \AND v = \name \IN \\
+%& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
+%\end{align*}
+%   
+%\paragraph{Arithmetic operations} have a similar structure and interpretation
+%as comparisons and their presentation is mostly omitted. Let us present the
+%translation of the unary \B{succ} operator though:
+%\begin{align*}
+%  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
+%  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
+%  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
+%\end{align*}
+%
+%\paragraph{Predicates}
+%The translation of conditions takes as input a condition node \B{n}, as well as
+%two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
+%execution shall jump when the condition is evaluted to true or false,
+%respectively.  The first rule defines the translation of an atomic relation:
+%\begin{align*}
+%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
+%  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
+%  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
+%\end{align*}
+%
+%\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
+%  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
+%following rule handles the case of conjunctions. Note that it produces code with
+%a ``short-cut'' when the evaluation of the first argument yields false:
+%\begin{align*}
+%\begin{split}
+%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
+%  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
+%  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
+%  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
+%\end{split}
+%\end{align*}
+%The definition of the translation for disjunctions is similar and is omitted.
+
+\section{Translation: Control Flow and Instructions
+  \label{sec:control}}
+
+This section addresses how operations in B implementations are translated to
+LLVM functions. It also discusses initialization, which might be seen as a
+special kind of operation. For operations, the presentation of the translation
+is divided into three steps: the signature, the local data, and the body
+instructions.
+
+The signature of the function is composed of the name of the function, the
+result type, the name and type of the parameters. The signature may carry other
+information such as linkage, visibility, calling convention, etc. The B parser
+provides this information after the type checking phase has been completed.  On
+the one hand LLVM functions may only have one result, on the other hand B0
+operations may produce several results. The translation of output parameters
+will be function parameters passed by reference, i.e. they are pointers. For
+uniformity of treatment, the return type of the produced LLVM functions is
+\llvm{void}. Translation of operation headers is defined in
+sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
+the stack, using the \llvm{alloca} instruction. We describe a function in
+sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
+type of its local variables used in the operation. These names are translated to
+conform to LLVM restrictions. The body of the operation has to be transformed to
+a sequence of LLVM blocks of statements. For each kind of instruction one (or
+more) translation function is defined. The details are provided in
+section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
+combined as follows:
 \begin{align*}
 \begin{split}
-  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
-  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
-  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
-  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
+  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
+  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
+  & \llvm{"define void \PH{h} \{\nl} \\
+  & \quad \llvm{entry:\nl} \\
+  & \quad \quad \PH{a} \\
+  & \quad \quad \PH{i} \\
+  & \quad \llvm{exit: ret void\nl} \\
+  & \llvm{\}\nl"}
 \end{split}
 \end{align*}
-The definition of the translation for disjunctions is similar and is omitted.
-
-%\section{Translation: Control Flow and Instructions
-%  \label{sec:control}}
-%
-%This section addresses how operations in B implementations are translated to
-%LLVM functions. It also discusses initialization, which might be seen as a
-%special kind of operation. For operations, the presentation of the translation
-%is divided into three steps: the signature, the local data, and the body
-%instructions.
-%
-%The signature of the function is composed of the name of the function, the
-%result type, the name and type of the parameters. The signature may carry other
-%information such as linkage, visibility, calling convention, etc. The B parser
-%provides this information after the type checking phase has been completed.  On
-%the one hand LLVM functions may only have one result, on the other hand B0
-%operations may produce several results. The translation of output parameters
-%will be function parameters passed by reference, i.e. they are pointers. For
-%uniformity of treatment, the return type of the produced LLVM functions is
-%\llvm{void}. Translation of operation headers is defined in
-%sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
-%the stack, using the \llvm{alloca} instruction. We describe a function in
-%sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
-%type of its local variables used in the operation. These names are translated to
-%conform to LLVM restrictions. The body of the operation has to be transformed to
-%a sequence of LLVM blocks of statements. For each kind of instruction one (or
-%more) translation function is defined. The details are provided in
-%section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
-%combined as follows:
-%\begin{align*}
-%\begin{split}
-%  \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
-%  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
-%  & \llvm{"define void \PH{h} \{\nl} \\
-%  & \quad \llvm{entry:\nl} \\
-%  & \quad \quad \PH{a} \\
-%  & \quad \quad \PH{i} \\
-%  & \quad \llvm{exit: ret void\nl} \\
-%  & \llvm{\}\nl"}
-%\end{split}
-%\end{align*}
-%
-%Initialization of a B implementation shares many characteristics with operations
-%but needs to deal with initialization of imported modules. It is discussed in
-%sec.~\ref{sec:trad-init}.
-%
+
+Initialization of a B implementation shares many characteristics with operations
+but needs to deal with initialization of imported modules. It is discussed in
+sec.~\ref{sec:trad-init}.
+
 %\subsection{Sub-routines signatures}
@@
 %
-%\subsection{Translation of instructions}
-%\label{sec:trad-instr}
-%
-%Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
-%of B instructions \B{il} and a label $\ell$ and produces the source code of a
-%LLVM block with $\ell$ as exit point. It is defined as follows:
-%
-%\begin{align*}
-%\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
-%& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
-%& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
-%& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
-%\end{align*}
-%
-%Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
-%instructions. Its definition is specialized for each class of instructions as
-%detailed in the rest of this section. Note that a new block is created after
-%each \B{If} and \B{While} instruction.
-%
-%\paragraph{While instructions} are translated in two blocks of instructions:
-%block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
-%body:
-%\begin{align*}
-%\begin{split}
-%\trad{w, $\ell$}{\B{While}} \isdef 
-%  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
-%  & \llvm{"\PH{\ell_1} : \PH{c}} \\
-%  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
-%  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
-%\end{split}
-%\end{align*}
-%
+\subsection{Translation of instructions}
+\label{sec:trad-instr}
+
+Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
+of B instructions \B{il} and a label $\ell$ and produces the source code of a
+LLVM block with $\ell$ as exit point. It is defined as follows:
+
+\begin{align*}
+\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
+& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
+& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
+& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
+\end{align*}
+
+Eventually, $\trad{}{\B{Inst}}$ defines the translation for individual
+instructions. Its definition is specialized for each class of instructions as
+detailed in the rest of this section. Note that a new block is created after
+each \B{If} and \B{While} instruction.
+
+\paragraph{While instructions} are translated in two blocks of instructions:
+block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
+body:
+\begin{align*}
+\begin{split}
+\trad{w, $\ell$}{\B{While}} \isdef 
+  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
+  & \llvm{"\PH{\ell_1} : \PH{c}} \\
+  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
+  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
+\end{split}
+\end{align*}
+
 %\paragraph{If instructions} are sequences of possibly conditional blocks of
@@
 %
-%\paragraph{Becomes equal instructions} must evaluate the target and the source
-%of the assignment, and then copy the result of the latter in the former, using
-%the LLVM instruction \llvm{store}. The evaluation of the target is defined with
-%$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
-%variable $p$ containing the assigned location. The evaluation of the source
-%yields the corresponding code $r$, the variable $v$ holding the result, and its
-%type $t$.
-%\begin{align*}
-%\begin{split}
-%  \trad{a}{\B{Beq}} \isdef 
-%  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
-%  & \llvm{"\PH{l}} \\
-%  & \llvm{\PH{r}} \\
-%  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
-%%   & l \conc r \conc 
-%%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
-%%   \conc t \conc \llvm{"* "} \conc p \conc \nl
-%\end{split}
-%\end{align*}
+\paragraph{Becomes equal instructions} must evaluate the target and the source
+of the assignment, and then copy the result of the latter in the former, using
+the LLVM instruction \llvm{store}. The evaluation of the target is defined with
+$\trad{}{lv}$, which yields the corresponding LLVM source code $l$ and the
+variable $p$ containing the assigned location. The evaluation of the source
+yields the corresponding code $r$, the variable $v$ holding the result, and its
+type $t$.
+\begin{align*}
+\begin{split}
+  \trad{a}{\B{Beq}} \isdef 
+  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
+  & \llvm{"\PH{l}} \\
+  & \llvm{\PH{r}} \\
+  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
+%   & l \conc r \conc 
+%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
+%   \conc t \conc \llvm{"* "} \conc p \conc \nl
+\end{split}
+\end{align*}
 %The assignment target may be either local to the operation or a state

### 2013-07-20 PAPERS/2013-SBMF/paper_short.tex
almost there

@@
 The aim of this work is to lay the foundations of a multi-platform code
-  generator for the B method. In particular, this paper presents a translation
-  procedure from a large subset of the B language for implementations towards
-  LLVM source code. This translation is defined formally as a set of rules
-  defined recursively on the abstract syntax for B implementations. It handles
-  the following elements of the B language: simple data types, imperative
-  instructions and component compositions. 
+generator for the B method. In particular, this paper presents a proposed
+translation  procedure from a subset of the B language for implementations
+towards LLVM source code. This translation is defined formally as a set of
+rules defined recursively on the abstract syntax for B implementations. It 
+presents a overview of translation and a simplified example.
 \end{abstract}
@@
 follows. The implementation constructs of the B language, and selected elements
-of the LLVM are presented in section~\ref{sec:b-impl}
-and~\ref{sec:llvm}. Section~\ref{sec:overview} presents a bird's-eye view that
-is then detailed in sections~\ref{sec:module} (for component modularity)
-~\ref{sec:data} (for data aspects),~\ref{sec:expr} (expressions) and
-instructions~\ref{sec:control}. We conclude and discuss future work
-in~\ref{sec:conclusion}.
+of the LLVM are presented in section~\ref{sec:b-impl} and~\ref{sec:llvm}.
+Section~\ref{sec:control} presents the main rules to translate the operations and
+instructions. We conclude and discuss future work in~\ref{sec:conclusion}.
 
@@
 containing definitions used in the implementation; \B{const}, a set of constant
-names, their type and their values; \B{imp}, a sequence of optionally
-prefixed B machines corresponding to external modules instantiated in the
-implementation; \B{var}, a set of variable names, their type and possibly
+names, their type and their values; \B{imp},%a sequence of optionally prefixed
+%B machines corresponding to external modules instantiated in the
+%implementation;
+\B{var}, a set of variable names, their type and possibly
 additional functional restrictions forming the rest of the state; \B{init}, a
@@
 in arithmetic operations \B{-} (unary and binary), \B{+}, \B{*}, \B{/}, \B{mod},
-% \B{**} (exponentiation), 
 \B{succ} and \B{pred} and Boolean operations
@@
 the B method ensures the absence of overflows and underflows.
-%TODO optionally to remove unsed elements in this table
+
 \begin{figure}[H]
@@
         \B{see} & : & seq \B{Name} \\
-        \B{imp} & : & seq \B{Impo} \\
+%        \B{imp} & : & seq \B{Impo} \\
         \B{const} & : & seq \B{Cons} \\
@@
         \hline
-        \multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{pre} & : & opt \B{Name} \\
-        \hline
+        %\multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\
+        %\hline
+        %\B{id} & : & \B{Name} \\
+        %\B{pre} & : & opt \B{Name} \\
+        %\hline
         \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
@@
         \B{type} & : & \B{Type} \\
-        \B{scope} & : & \B{Impl} \alt \B{Oper}
-        \\
+        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
+        \hline
+        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{inp} & : & seq \B{Vari} \\
+        \B{out} & : & seq \B{Vari} \\
+        \B{body} & : & \B{Inst}	\\
       \end{tabular}
@@
       \begin{tabular}[t]{rcl}
-        \hline
-        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{inp} & : & seq \B{Vari} \\
-        \B{out} & : & seq \B{Vari} \\
-        \B{body} & : & \B{Inst}
-        \\
         \hline
@@
         \B{out} & : & seq \B{Name}
+        \\
+	\hline
+	\multicolumn{3}{|c|}{While loop: \B{While}} \\
+        \hline
+        \B{cond} & : & \B{Pred} \\
+        \B{body} & : & \B{Inst}
         \\
@@
       \begin{tabular}[t]{rcl}
-        \hline
-        \multicolumn{3}{|c|}{While loop: \B{While}} \\
-        \hline
-        \B{cond} & : & \B{Pred} \\
-        \B{body} & : & \B{Inst}
-        \\
         \hline
@@
       Sequence and optional attributes are denoted ``seq'' and ``opt'',
-      respectively}
+      respectively.}
     \label{tab:node-attr}

### 2013-07-20 PAPERS/2013-SBMF/paper_short.tex
I believe that it is finished, but I need show to David.

@@
 components of a B implementation are: \B{id}, the name identifying the
-implementation within the project; \B{see}, a set of references to other modules
-containing definitions used in the implementation; \B{const}, a set of constant
-names, their type and their values; \B{imp},%a sequence of optionally prefixed
-%B machines corresponding to external modules instantiated in the
-%implementation;
-\B{var}, a set of variable names, their type and possibly
-additional functional restrictions forming the rest of the state; \B{init}, a
-sequence of instructions executed upon initialization of the implementation;
-\B{op}, a set of operations, each being an algorithmic description of the
-different functionalities provided by the implementation.
+implementation within the project; \B{const}, a set of constant
+names, their type and their values; \B{var}, a set of variable names, their
+type and possibly additional functional restrictions forming the rest of the
+state; \B{init}, a sequence of instructions executed upon initialization of
+the implementation; \B{op}, a set of operations, each being an algorithmic 
+description of the different functionalities provided by the implementation.
 
-The initialization and the operations are defined as (sequences of)
-instructions. Operations also have a name, inputs and outputs. Possible
-instructions are the classic imperative constructs: variable assignment, if and
-case conditional, while loop, block with or without local variables, and
-operation calls. The if constructs may have several branches, each with a
-condition as well as an optional else branch. Called operations may be either
-local operations or operations of instances of external modules, in this case
-they have an optional prefix identifying the module.
+The operations are defined as (sequences of) instructions. Operations also
+have a name, inputs and outputs. Possible instructions are the classic
+imperative constructs: variable assignment, if and case conditional, while loop,
+block with or without local variables, and operation calls. The if constructs
+may have several branches, each with a condition as well as an optional else
+branch.
 
@@
         \B{id} & : & \B{Name} \\
-        \B{see} & : & seq \B{Name} \\
+%        \B{see} & : & seq \B{Name} \\
 %        \B{imp} & : & seq \B{Impo} \\
@@
     \caption{Abstract syntax structure of a B implementation.  For each abstract
-      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
-      branch in a case instruction, and the different attributes of the class),
-      or a list of the possible sub-classes (e.g. \B{Inst} for the different
-      kinds of instructions).
-      Sequence and optional attributes are denoted ``seq'' and ``opt'',
-      respectively.}
+      syntax element, we give the name of the class  or a list of the possible sub-classes.
+      Sequence and optional attributes are denoted ``seq'' and ``opt'', respectively.}
     \label{tab:node-attr}
@@
 compilers. The compiler front-end translates the source programming language to
-LLVM source code, upon which optimization and other static analysis tasks may be
-performed, and the compiler back-end translates to the target platform assembly
-language. One key feature of LLVM is that it is a single-static assignment (SSA)
-language: a variable may only be assigned once. So for each assignment to a
-variable found in the source language, the front-end shall generate a new
-temporary variable in the LLVM program. For illustration,
-figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
-C code appears on the left) with two temporary variables \llvm{\%0} and
-\llvm{\%1}.
+LLVM source code, upon which optimization and other static analysis tasks may
+be performed, and the compiler back-end translates to the target platform
+assembly language. One key feature of LLVM is that it is a single-static
+assignment (SSA) language: a variable may only be assigned once. So for each
+assignment to a variable found in the source language, the front-end shall
+generate a new temporary variable in the LLVM program. For illustration,
+figure~\ref{fig:ex-llvm} presents a B specification, C program and a LLVM
+program with two temporary variables \llvm{\%0} and \llvm{\%1}.
 \begin{figure}
-\vspace*{-2.4em}
+\vspace*{-2.3em}
   \begin{minipage}[t]{.3\textwidth}
@@
   \caption{Simple example of a B function and its corresponding  C function and LLVM function.
-    It is not a real translation, but it helps to understand the translation. %TODO to update the variables
-    The first line contains the signature: return type \llvm{void}, the name
-    \llvm{@inc} and one parameter named \llvm{\%pi} and typed \llvm{i32*}.  Next
+    It is simplified and not a faithful translation, but it helps to understand the translation. 
+    The first line in LLVM code contains the signature: return type \llvm{void}, the name
+    \llvm{@inc} and one parameter named \llvm{\%x} and typed \llvm{i32}.  Next
     is the body with a single block, labeled \llvm{entry}, and temporary
@@
     block has four instructions: \llvm{load}, \llvm{add}, \llvm{store} and
-    \llvm{ret}. For instance, \llvm{\%1 = add i32 \%0, 1} is an addition
-    (\llvm{add}), has result type \llvm{i32} and assigns to
-    \llvm{\%1} the sum of variable \llvm{\%0} and integer
-    literal \llvm{1}.}
+    \llvm{ret}. For instance, \llvm{\%1 = add nsw i32 \%0, \%x} is an addition
+    (\llvm{add}), has result type \llvm{i32} and assigns to \llvm{\%1} the sum
+    of variable \llvm{\%0} and  variable \llvm{\%x}, the result is stored in
+     \llvm{@y}.}
   \label{fig:ex-llvm}
@@
 
-%The introduction to the LLVM that follows is focused in the features
-%that are relevant for the representation of B
-%implementations. Figure~\ref{fig:llvm-grammar} presents a grammar of this subset
-%of the LLVM. 
-%
-%\begin{figure}
-%  \begin{center}
-%    \begin{tabular}{rcl}
-%      \nt{module} & ::= & \ListOf{\nt{item}} \\
-%      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def} 
-%      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
-%      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
-%      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
-%      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{struct} \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
-%      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
-%      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
-%      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
-%      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
-%      \nt{param} & ::= & \nt{type} \nt{name} \\
-%      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
-%      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
-%      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
-%      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
-%      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
-%      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
-%      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
-%      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
-%      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
-%      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
-%%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
-%      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
-%      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
-%      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
-%      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
-%      \nt{arg} & ::= & \nt{type} \nt{exp}
-%    \end{tabular}
-%  \end{center}
-%  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
-%    and \nt{name} correspond respectively to integer types, integer literals,
-%    labels and names. Choices are separated by \alt and optionally delimited by
-%    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
-%    elements of the annotated entity.}
-%  \label{fig:llvm-grammar}
-%\end{figure}
-%
-%LLVM programs are organized into modules, one per translation unit. A module may
-%contain declarations of external entities (functions, constants) and definitions
-%of internal items (functions, variables, constants). All data is explicitly
-%typed. The name and type of external entities must be declared.  All names,
-%e.g. non-reserved identifiers, shall start with \llvm{@}, when they are global,
-%or\llvm{\%}, when they are local. For instance, \llvm{@max = external constant
-%  i32} declares \llvm{@max} as a 32-bit integer constant and \llvm{declare void
-%  @inc(i32*)} declares \llvm{@inc} as a function having as parameter a pointer
-%to an integer and \llvm{void} as a return type.
-%
-%The type system contains the empty type \llvm{void}, an infinite, countable
-%number of integer types, one for each possible bit width (e.g. \llvm{i8} is the
-%type for 8-bit integers), and type constructors pointer (monadic \llvm{$\cdot$
-%  *}) and structure (polyadic \llvm{\{$\cdots$\}}). For instance \llvm{\{ i8*,
-%  i8, i8 \}} is the type for structures with three fields, the first having as
-%type pointer to \llvm{i8}. Grammar rule \nt{type\_def} states how types are
-%named, e.g.: \llvm{\%T1 = type \{i32, i32\}} and \llvm{\%T2 = type \{\%T1 *,
-%  \%T1 *\}}. Note that there is no \llvm{void *} type in the LLVM, instead
-%pointer values are of integer types. 
-%
-%Local entities are constants, variables or functions. An example of constant
-%definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
-%and value. The elements of a variable definition are the name, type and code
-%generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
-%Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
-%information for target code generation: linkage type (the variable is merged
-%with other of the same name), the scope (global) and initialization of the
-%variable (zeroed is mandatory for common linkage). LLVM defines other attributes
-%for these aspects, but they are not needed in this work and their presentation
-%is omitted.  Function definitions are composed of the signature and body. The
-%signature contains the return type, name, parameters, and attributes for target
-%code generation. Attribute \llvm{nounwind} states that the function never
-%generate exceptions. The body is a sequence of blocks of instructions in
-%single-static assignment (SSA) form, i.e. are such that each variable may be
-%assigned exactly once.
-%
-%The grammar rule \nt{inst} describes the different kind of
-%instructions. Instruction \llvm{alloca} allocates a memory block the size of the
-%given type from the stack segment. This and the following classes of
-%instructions assign the resulting value to a fresh variable. Arithmetic
-%operations are binary, and include signed quotient and reminder. Arithmetic
-%comparisons compare two values according to the first operand and return a 1-bit
-%integer value.  Instruction \llvm{call} invokes the given function with the
-%given arguments and assigns the result to a fresh variable, which is of the
-%given type.  In general, \llvm{getelementptr} gets the address of an element in
-%an aggregate object through indexing. In the target LLVM subset, aggregate
-%objects are structures, the first index has type \llvm{i32} and value \llvm{0}
-%to select the first structure at the given location \nt{exp}, and the second
-%index selects a field in that structure.  Instruction \llvm{load} assigns to a
-%fresh variable \nt{name} the contents of a memory address of type \nt{type}
-%specified by \nt{exp} (e.g. in fig.~\ref{fig:ex-llvm}).  Instruction
-%\llvm{store} writes a value to memory address (again, see
-%fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two forms.  In the first
-%form, it directs conditionally the control flow to one of two blocks.  If the
-%given expression evaluates to \llvm{1}, control goes to the first block,
-%otherwise to the second. In the second form, the control jumps unconditionally
-%to the given block.
-%% Instruction \llvm{switch} directs the
-%% control flow to one of several blocks. The given expression is evaluated and
-%% control goes to the first branch with the matching value, or to the default
-%% branch (appearing first in the instruction).  
-%Finally, instruction \llvm{ret}
-%ends the current function call, optionally returning a value.
-%
-%The expression language is limited to names (local and global), integer literals
-%and selection of an element in a structure.
-%
-%\section{Translation: a Bird's-Eye View}
-%\label{sec:overview}
-%
-%The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
-%module, say \llvm{M}. A B implementation \B{M} has several sections, each
-%translated as follows:
-%\begin{enumerate}
-%\item Visibility clauses provide access to definitions found in external
-%  B components and are translated to external declarations in LLVM.
-%\item Constants and value definitions are translated to LLVM constants. To avoid
-%  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
-%  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
-%\item To encode the implementation state, the translation produces a
-%  structure type, named \llvm{@M\$state\$}, and an instance of that type, named
-%  \llvm{@M\$self\$}. All the variables and instances of imported machines
-%  are encoded as elements of that structure.  Also, for each operation of each
-%  imported machine, an external function declaration is produced.
-%\item Each operation \B{op} of the implementation is translated into the
-%  definition of a function named \llvm{@M\$op}; also the translation produces a
-%  special function \llvm{@M\$init\$} responsible for the initialization.
-%
-%  The parameter list of the functions thus produced contain one item for each
-%  input and output of the corresponding operation, and one item of type
-%  \llvm{@M\$state\$*}, to pass the address of the structure encoding the
-%  state of the implementation and its modification by the callee. The return
-%  type is always \llvm{void}.
-%\end{enumerate}
-%This approach assumes that information on external B components used in \B{M} is
-%available during translation. To fulfill this hypothesis, the translation
-%produces an ``interface'' artifact containing the information that the
-%translation of an importation of \B{M} in a client machine
-%requires. Figure~\ref{fig:skel} summarizes the translation principles by
-%providing the structure of both the interface and the LLVM implementation thus
-%produced.
-%
-%\begin{figure}
-%  \begin{center}
-%    (a) Structure of the interface artifact produced from a B implementation \B{M}:
-%    \begin{tabbing}
-%      foo \= foo \= \kill
-%      \textit{(references to imported components)} \\
-%      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-%      \llvm{declare void @M\$init\$(@M\$state\$*)} \\
-%      \textit{for each constant} \B{cst} \textit{generate} \\
-%      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
-%      \textit{for each operation} \B{op} \textit{generate} \\
-%      \> \llvm{declare void @M\$op(@M\$state\$*, \ListOf{\nt{type}})} \\
-%    \end{tabbing}
-%    (b) Structure of the LLVM module produced from a B implementation \B{M}:
-%    \begin{tabbing}
-%      foo \= foo \= \kill
-%      \textit{for each imported component} \B{Q} \textit{generate} \\
-%      \> \llvm{@Q\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-%      \> \llvm{declare void @Q\$init\$(@Q\$state\$*)} \\
-%      \> \textit{for each constant} \B{cst} \textit{generate} \\
-%      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
-%      \> \textit{for each operation} \B{op} \textit{generate} \\
-%      \> \> \llvm{declare void @Q\$op(@Q\$state\$*, \ListOf{\nt{type}})} \\
-%      \textit{for each constant} \B{cst} \textit{generate} \\
-%      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
-%      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-%      \llvm{@M\$self\$ = common global @M\$state\$ zeroinitializer} \\
-%      \llvm{define void @M\$init\$(@M\$state\$* \%self\$) \{} \\
-%      \> \llvm{\ListOf{\nt{block}}} \\
-%      \> \llvm{exit: ret void} \\
-%      \llvm{\}} \\
-%      \textit{for each operation} \B{op} \textit{generate} \\
-%      \> \llvm{define void @M\$op(@M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
-%      \> \> \llvm{\ListOf{\nt{block}}} \\
-%      \> \> \llvm{exit: ret void} \\
-%      \> \llvm{\}}
-%    \end{tabbing}
-%  \end{center}
-%  \caption{Skeleton of the artifacts produced by the translation.}
-%  \label{fig:skel}
-%\end{figure}
-%
-The detailed presentation of the translation of the different elements of a B
-implementation is divided into:
-\begin{description} %TODO remove the references and to organize it.
-\item[Modularity] constructs to combine B components are described in
-  sec.~\ref{sec:module}.
-\item[Data] translation is briefly described sec.~\ref{sec:data}, including
-  the encoding of B data types.
-\item[Expressions] and conditions are codified as LLVM instruction blocks, as
-  described in section~\ref{sec:expr}.
-\item[Control] flow, instructions are translated to LLVM instructions as
-  discussed in sec.~\ref{sec:control}, which also presents the encoding
-  of B operations and initialization.
-\end{description}
+Operations and instructions are translated to LLVM instructions as discussed in
+sec.~\ref{sec:control}. For reason of reduced space, only the operations and
+instructions will be presented with more details.
 
@@
 element \B{n} of the abstract syntax tree returns the corresponding LLVM local
-name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
-\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
-LLVM global name, composed by special characters (to avoid name conflicts), the
-name of the component where it is defined and its own name. Also $\state(\B{M})
-\isdef \llvm{"@\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
-state of component \B{M}. We assume an unlimited supply of labels and local
-variables: let $\name$ return a fresh LLVM label or local variable name.
-%
-%\section{Translation: Modularity Aspects}
-%\label{sec:module}
-%
-%The translation of a B module produces two artifacts: an interface and an (LLVM)
-%implementation. In the following, we describe how they are produced.
-%
-%\subsection{Interface}
-%
-%The definition of the interface $\intf(\B{M})$ of a component \B{M} has four
-%key elements, as shown in figure~\ref{fig:skel} (b):
-%\begin{enumerate}
-%\item $\intf(\B{M})_u$ is the list of used components: $\intf(\B{M})_u \isdef \B{M.imp}$.
-%
-%\item $\intf(\B{M})_t$ is a textual description of the type representing the
-%  state of the component, i.e. a structure with one element per imported
-%  component and variable:
-%\begin{small}
-%$$\intf(\B{M})_t \isdef \llvm{"struct \{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
-%\end{small}
-%where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
-%$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.
-%
-%\item $\intf(\B{M})_c$ is the definition of the component constants. It maps
-%the constant names to their type:
-%$\intf(\B{M})_c \isdef \trad{\B{M.const}}{\ListOf{\B{Cons}}}$ and
-%$\trad{\B{c}}{\B{Cons}} \isdef \B{c} \mapsto \trad{\B{c.type}}{\B{Type}}$.
-%
-%\item $\intf(\B{M})_o$ represents the operation declarations. It maps
-%operation names to the list of input parameter types and output parameter
-%types:
-%$\intf(\B{M})_o \isdef \trad{\B{M.op}}{\ListOf{\B{Oper}}}_t$ and
-%$\trad{\B{o}}{\B{Oper}}_t \isdef \B{o} \mapsto ( \trad{\B{o.inp}}{\ListOf{inp}} \sep \trad{\B{o.out}}{\ListOf{out}} )$, where
-%$\trad{\B{i}}{inp} \isdef \trad{\B{i.type}}{\B{Type}}$ and
-%$\trad{\B{o}}{out} \isdef \llvm{"\PH{\trad{\B{o.type}}{\B{Type}}}*"}$.
-%\end{enumerate}
-%
-%\subsection{Implementation}
-%
-%The actual LLVM source code is generated following the template shown in
-%figure~\ref{fig:skel} (a), as follows:
-%\begin{align*}
-%  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
-%  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
-%  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
-%  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
-%& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
-%  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
-%  \quad \PH{\trad{\B{i}}{State}}} \\
-%& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
-%& \llvm{\PH{\trad{\B{i.init}}{Init}} 
-%  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
-%\end{align*}
-%We shall now define each rule referenced in this template.
-%\begin{enumerate}
-%\item For the type definitions corresponding to the state of the imported
-%  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
-%    type \PH{\intf(\B{i})_t}\nl"}$.
-%\item For the declarations of the initialization functions in the imported
-%  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
-%    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
-%\item For the definitions of the imported constants: $\trad{\B{i}, \intf}{ICons}
-%  \isdef \trad{$\intf(\B{i})_c$}{\ListOf{Cons}}$, where
-%  \noindent $\trad{$\B{c} \mapsto t)$}{Cons} \isdef \llvm{"\PH{\Global(\B{c})} =
-%    external constant \PH{t}\nl"}$.
-%\item For the declarations of the imported operations: $\trad{\B{i},
-%    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
-%  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
-%    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
-%\item The translation of constant definitions is presented in section~\ref{sec:data}.
-%\item For the definition of the type for component states: \\
-%\noindent$\trad{\B{M}}{State} \isdef
-%\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
-%\item For the initialization, the definition of a function is generated. It
-%  first initializes the imported modules and then the current module (see
-%  section~\ref{sec:control}).
-%\item For each operation, a function definition is generated (see
-%  section~\ref{sec:control}).
-%\end{enumerate}
-%Let $\idx$ be a function that maps each imported machine and state variable to
-%its position (starting from zero).
-%\section{Translation: Data}
-%\label{sec:data}
-%
-%The translation presented in this paper is currently limited to integers and
-%booleans. In the case of integers, the translation is based on the supposition
-%that the B0 definition of \B{INT} is the range $-2^{31}.. 2^{31}-1$. We have
-%$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}$ and $\trad{INT}{\B{Type}} \isdef
-%\llvm{i32}$.
-%
-%B constants are translated to constants in the LLVM program, as described in the
-%following rule, where parameter \B{c} denote the constant, $M$ the name of the
-%module and $\Global$ is the mapping between source and target identifiers. Two
-%results are produced: the LLVM source code and an updated identifier mapping:
-%\begin{small}
-%\begin{align*}
-%  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
-%\end{align*}
-%\end{small}
-%The definition of the translation for expressions is described next in
-%section~\ref{sec:expr}.
-%
+name, obtained by prepending character \% to its identifier. We assume an 
+unlimited supply of labels and local variables: let $\name$ return a fresh LLVM
+label or local variable name.
 
-%\section{Translation: Expressions
-%  \label{sec:expr}}
-%
-%To evaluate a composed expression using the LLVM, one must first generate a
-%sequence of instructions to evaluate each argument, the resulting value being
-%stored in a temporary variable, and second generate an instance of the LLVM
-%instruction corresponding to the expression argument.  Translation of
-%expressions is defined as functions that take as input some class of expressions
-%and produce a triple of strings $p \sep v \sep t$, where $p$ is a possibly empty
-%sequence of instructions required in the evaluation (e.g.  for the
-%sub-expressions), $v$ is the LLVM expression representing the resulting value,
-%and $t$ represents the LLVM type of the expression. In the case of composed
-%predicates, though, the evaluation is different as boolean operations need to be
-%implemented with conditional and unconditional jumps.
-%
-%\paragraph{Atomic expressions} have a simple translation. In the case of
-%literals, one only needs to issue the corresponding LLVM literal:
-%$\trad{TRUE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"1"} \sep \llvm{"i1"}$,
-%$\trad{FALSE}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"0"} \sep \llvm{"i1"}$ and
-%$\trad{l}{\B{Lit}} \isdef \llvm{""} \sep \llvm{"\PH{l}"} \sep \llvm{"i32"}$ when
-%$\B{l.type} = \B{INT}$.
-%
-%For identifiers, three cases are possible: global constants, local variables,
-%and module state variables. For state variables, the translation issues
-%statements to assign the address of the corresponding element in the state
-%structure to pointer variable $p$ and then dereference $p$ to assign $v$:
-%\begin{align*}
-%\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
-%& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
-%& \ELSIF \B{n} \mbox{ is a local variable } \THEN 
-%\llvm{""} \sep \local(\B{n}) \sep t \\
-%& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
-%& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
-%& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
-%\end{align*}
-%
-%\paragraph{Comparisons} are formed by the application of a relational operator
-%to two expressions. Mapping $\opmap$ associates B operators to corresponding
-%LLVM keywords:
-%$$\opmap = \{ = \mapsto \llvm{"eq"}, 
-%  \ne \mapsto \llvm{"ne"}, 
-%  < \mapsto \llvm{"slt"}, 
-%  \le \mapsto \llvm{"sle"}, 
-%  > \mapsto \llvm{"sgt"}, 
-%  \ge \mapsto \llvm{"sge"} \}.
-%$$
-%
-%The code responsible for evaluating a comparison \B{r} is defined by the
-%following rule. First both arguments are evaluated in $p_1$ and $p_2$, and then
-%an instance of the LLVM comparison instruction \llvm{icmp} is used to compare
-%the temporary variables $v_1$ and $v_2$, of type $t_1 = t_2$, storing the result
-%in the fresh variable $v$.
-%\begin{align*}
-%\lefteqn{\trad{r}{\B{Comp}} \isdef
-%  \LET
-%  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
-%  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
-%& \AND v = \name \IN \\
-%& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
-%\end{align*}
-%   
-%\paragraph{Arithmetic operations} have a similar structure and interpretation
-%as comparisons and their presentation is mostly omitted. Let us present the
-%translation of the unary \B{succ} operator though:
-%\begin{align*}
-%  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
-%  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
-%  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
-%\end{align*}
-%
-%\paragraph{Predicates}
-%The translation of conditions takes as input a condition node \B{n}, as well as
-%two labels $\ell_t$ and $\ell_f$ that correspond to program locations where the
-%execution shall jump when the condition is evaluted to true or false,
-%respectively.  The first rule defines the translation of an atomic relation:
-%\begin{align*}
-%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{Form}} \isdef
-%  \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
-%  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
-%\end{align*}
-%
-%\noindent The translation for negation is simply: $\trad{not f, $\ell_t$,
-%  $\ell_f$}{\B{not}} \isdef \trad{f, $\ell_f$, $\ell_t$}{\B{Form}}$.  The
-%following rule handles the case of conjunctions. Note that it produces code with
-%a ``short-cut'' when the evaluation of the first argument yields false:
-%\begin{align*}
-%\begin{split}
-%  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
-%  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
-%  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
-%  & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
-%\end{split}
-%\end{align*}
-%The definition of the translation for disjunctions is similar and is omitted.
-
-\section{Translation: Control Flow and Instructions
+\section{Translation: Operations
 \label{sec:control}}
@@
 This section addresses how operations in B implementations are translated to
-LLVM functions. It also discusses initialization, which might be seen as a
-special kind of operation. For operations, the presentation of the translation
+LLVM functions. For operations, the presentation of the translation
 is divided into three steps: the signature, the local data, and the body
@@
 uniformity of treatment, the return type of the produced LLVM functions is
-\llvm{void}. Translation of operation headers is defined in
-sec.~\ref{sec:trad-header}. The memory to store local variables is allocated on
-the stack, using the \llvm{alloca} instruction. We describe a function in
-sec.~\ref{sec:trad-alloc} that, applied to a B operation, returns the name and
-type of its local variables used in the operation. These names are translated to
-conform to LLVM restrictions. The body of the operation has to be transformed to
-a sequence of LLVM blocks of statements. For each kind of instruction one (or
-more) translation function is defined. The details are provided in
-section~\ref{sec:trad-instr}.  The results of the translation of each aspect are
-combined as follows:
+\llvm{void}.
+
+The \trad{}{sig} rule specifies the translation of operation headers and
+\trad{}{alloc} rule defines the LLVM stack allocation.  The results of the
+translation of each aspect are combined as follows:
 \begin{align*}
@@
   & \quad \llvm{entry:\nl} \\
-  & \quad \quad \PH{a} \\
-  & \quad \quad \PH{i} \\
+  & \quad \quad \PH{a}  \quad \quad \PH{i}  \\
+%  & \quad \quad \PH{i} \\
   & \quad \llvm{exit: ret void\nl} \\
@@
 
-Initialization of a B implementation shares many characteristics with operations
-but needs to deal with initialization of imported modules. It is discussed in
-sec.~\ref{sec:trad-init}.
-
-%\subsection{Sub-routines signatures}
-%\label{sec:trad-header}
-%
-%The following rule specifies the translation of operation headers. Its
-%parameters are the operation identifier, inputs, outputs, the name of the
-%machine it belongs to, and the results are the LLVM function definition
-%header: \\
-%\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
-%
-%Next are the rules defining the translation of individual input and output
-%parameters. They have local scope and they are translated by prepending
-%\llvm{\%}:
-%\begin{align*}
-%  \trad{i}{inp} \isdef 
-%  \llvm{"\PH{\trad{i.type}{\B{Type}}} \PH{\local(\B{i})}"} \quad \quad
-%  \trad{o}{out} \isdef 
-%  \llvm{"\PH{\trad{o.type}{\B{Type}}}* \PH{\local(\B{o})}"}
-%\end{align*}
-%
-%\subsection{Stack allocation}
-%\label{sec:trad-alloc}
-%
-%B operations may have local variables, in which case the translator needs to
-%issue LLVM stack allocation instructions. Such instructions have two arguments:
-%the type of values to store, and a temporary variable that is assigned the
-%address (see figure~\ref{fig:llvm-grammar} for the concrete syntax). The
-%generation of stack allocation code visits the parse tree of the operation body
-%in a pre-order depth-first traversal, generating one statement for each local
-%variable declaration found. The detailed definition of this traversal is omitted
-%for space reasons. For each B0 variable \B{v} found, a LLVM name is thus
-%created.
-%\begin{align*}
-%  \trad{v}{alloc} \isdef 
-%  \llvm{"\PH{\local(\B{v})} = alloca \PH{\trad{v.type}{\B{Type}}} \nl"}
-%\end{align*}
-%
-%After the stack allocations, the LLVM source code contains the statements that
-%encodes the instructions in the body of the operation, as described in the
-%following section.
-%
 \subsection{Translation of instructions}
@@
 
-%\paragraph{If instructions} are sequences of possibly conditional blocks of
-%instructions, i.e. branches. After a branch is executed, control must go
-%to the next block $\ell$.
-%\begin{align*}
-%  \trad{i, $\ell$}{\B{If}}_L \isdef \trad{i.branches, $\ell$}{\ListOf{\B{IfBr}}}
-%\end{align*}
-%To translate conditional branches, one must first generate code to evaluate the
-%condition of the branch, followed by an LLVM conditional branching statement,
-%translate the body of the branch, and add an unconditional branch to the block
-%following the corresponding \B{If} statement.  The translation for branches has
-%as arguments a branch \B{b}, the remaining branches \B{bl}, the lexicon, and the
-%label $\ell$ of the block following the \B{If} statement they belong to. The
-%following definition applies for the case of conditional branches:
-%\begin{align*}
-%\begin{split}
-%  \trad{\B{b bl}, $\ell$}{\B{IfBr}} \isdef & \LET \ell_1 = \name \AND \ell_2 = \name \IN \\
-%  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
-%  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}} \\
-%  & \llvm{\PH{\ell_2} : \PH{\trad{bl, $\ell$}{\ListOf{\B{IfBr}}}}"} \\
-%\end{split}
-%\end{align*}
-%The following rule handles the case of the last branch, which might have a condition or not:
-%\begin{align*}
-%\begin{split}
-%  \trad{\B{b}, $\ell$}{\B{IfBr}} \isdef & \IF \B{b.cond} = \text{None} \THEN
-%  \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
-%  & \ELSE \LET \ell_1 = \name \IN \\
-%  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
-%  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}}
-%\end{split}
-%\end{align*}
-%
 \paragraph{Becomes equal instructions} must evaluate the target and the source
@@
 \end{align*}
-%The assignment target may be either local to the operation or a state
-%variable. In the former case, it can be referenced by the corresponding LLVM
-%identifier. In the latter case, it is represented as an element of the state
-%structure $\Sigma$, and its address must be calculated and assigned to a new temporary
-%variable.
-%\begin{align*}
-%\lefteqn{\trad{n}{lv} \isdef} \\
-%& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
-%& \quad \ELSE \LET v = \name \AND t = \trad{n.type}{\B{Type}} \IN \\
-%& \quad \quad \llvm{"\PH{v} = load \PH{t}* getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
-%& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
-%\end{align*}
-%
-%\paragraph{Call up} instructions apply B operations with some values and obtain
-%results. The LLVM code generated must include instructions to evaluate the
-%operation arguments, the address of the variables receiving the results as well
-%as an instance of the LLVM \llvm{call} instruction. In addition, the proposed
-%translation includes as argument the address of a structure storing the state of
-%the component corresponding to this operation. Note that this component may be
-%either the implementation being translated, or an instance of an imported
-%machine.
-%\begin{align*}
-%  \trad{c}{\B{Call}} \isdef 
-%  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
-%  \trad{c.out}{\ListOf{Out}} \AND \\
-%  & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
-%  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"@\PH{c.pre}\$state\$"} \AND \\
-%  & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
-%  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
-%  & \IN \\
-%  & \quad \llvm{"\PH{p_i}} \\
-%  & \quad \llvm{\PH{p_o}} \\
-%  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
-%\end{align*}
-%The translation of input and output arguments produces two code snippets. The
-%first corresponds to the LLVM instruction sequence $p$ responsible for
-%evaluating the argument, and the second to the type $t$ and name $v$ of the
-%variable storing the resulting value:
-%\begin{align*}
-%  \trad{i}{Inp} \isdef 
-%  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
-%  \trad{o}{Out} \isdef 
-%  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
-%\end{align*}
-%
-%\paragraph{Block instructions} are structuring constructs that have no
-%operational intent; their translation is that of their body: $\trad{i,
-%  $\ell$}{\B{Blk}} \isdef \trad{i.body, $\ell$}{\ListOf{\B{Inst}}}$.
-%
-%
-%\subsection{Initialization}
-%\label{sec:trad-init}
-%
-%Initialization is done is two steps: first all imported modules are initialized,
-%second the initialization clause is executed.
-%\begin{align*}
-%  \trad{i}{Init} \isdef
-%  & \llvm{"define @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
-%  & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
-%  & \quad\llvm{\PH{\trad{i.init, "exit"}{\B{Inst}}_L}} \\
-%  & \quad\llvm{exit: ret void \nl} \\
-%  & \llvm{\}"}
-%\end{align*}
-%The initialization of an imported module is performed by a call to the
-%corresponding initialization function:
-%\begin{align*}
-%  \trad{m}{InitImp} \isdef
-%  & \LET \ell = \name \IN \\
-%  & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
-%  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}* \PH{\ell})"}
-%\end{align*}
 

### 2013-08-22 PROTO/example-01.tex
Created file translation.tex to include all translation rules, and the description of the validation. Added an example used for manual translation. Created a prototype implementation of the translation in Python. Added a pythonesque representation of the example suitable for processing by the prototype.

\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\pagestyle{headings}
\title{Translation of B Implementations to the LLVM: Example 01}
\author{David Déharbe \and Valério Medeiros Jr} 
\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
  Federal University of Rio Grande do Norte \\
  Natal (Brazil)}
\date{August 2013}

\usepackage{xcolor}
\usepackage{listings}
\usepackage{textcomp}

\lstset{
    inputencoding=utf8,
%    backgroundcolor=\color{white},
    tabsize=4,
    rulecolor=,
    upquote=true,
%    aboveskip={1.5\baselineskip},
    columns=fixed,
    showstringspaces=false,
    extendedchars=true,
    breaklines=true,
    prebreak = \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
    frame=single,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    basicstyle=\scriptsize\ttfamily,
    identifierstyle=\ttfamily,
    keywordstyle=\ttfamily\color[rgb]{0,0,1},
    commentstyle=\ttfamily\color[rgb]{0.133,0.545,0.133},
    stringstyle=\ttfamily\color[rgb]{0.627,0.126,0.941},
}

\makeatletter
\lstdefinelanguage{llvm}{
  morecomment = [l]{;},
  morestring=[b]", 
  sensitive = true,
  classoffset=0,
  morekeywords={
    define, declare, global, constant,
    internal, external, private,
    linkonce, linkonce_odr, weak, weak_odr, appending,
    common, extern_weak,
    thread_local, dllimport, dllexport,
    hidden, protected, default,
    except, deplibs,
    volatile, fastcc, coldcc, cc, ccc,
    x86_stdcallcc, x86_fastcallcc,
    ptx_kernel, ptx_device,
    signext, zeroext, inreg, sret, nounwind, noreturn,
    nocapture, byval, nest, readnone, readonly, noalias, uwtable,
    inlinehint, noinline, alwaysinline, optsize, ssp, sspreq,
    noredzone, noimplicitfloat, naked, alignstack,
    module, asm, align, tail, to,
    addrspace, section, alias, sideeffect, c, gc,
    target, datalayout, triple,
    blockaddress
  },
  classoffset=1, keywordstyle=\color{purple},
  morekeywords={
    fadd, sub, fsub, mul, fmul,
    sdiv, udiv, fdiv, srem, urem, frem,
    and, or, xor,
    icmp, fcmp,
    eq, ne, ugt, uge, ult, ule, sgt, sge, slt, sle,
    oeq, ogt, oge, olt, ole, one, ord, ueq, ugt, uge,
    ult, ule, une, uno,
    nuw, nsw, exact, inbounds,
    phi, call, select, shl, lshr, ashr, va_arg,
    trunc, zext, sext,
    fptrunc, fpext, fptoui, fptosi, uitofp, sitofp,
    ptrtoint, inttoptr, bitcast,
    ret, br, indirectbr, switch, invoke, unwind, unreachable,
    malloc, alloca, free, load, store, getelementptr,
    extractelement, insertelement, shufflevector,
    extractvalue, insertvalue,
  },
  alsoletter={\%},
  keywordsprefix={\%},
}
\makeatother

\makeatletter
\lstdefinelanguage{b}{
  morecomment = [l]{;},
  morestring=[b]", 
  sensitive = true,
  classoffset=0,
  morekeywords={
    IMPLEMENTATION, REFINES, CONCRETE_VARIABLES, INVARIANT, INITIALISATION, 
    BEGIN, END, IF, THEN, ELSE
  },
  classoffset=1, keywordstyle=\color{purple},
  morekeywords={
    FALSE, TRUE
  },
  alsoletter={\%,\&},
}
\makeatother

\begin{document}
\maketitle

\lstinputlisting[language=b]{example-01.bimp}

The following is the translation, where only the data part, initialization
and operation \verb'zero' have been taken into account. This file compiles
with \verb'llc-mp-3.1' (LLVM static compiler).

\lstinputlisting[language=llvm]{example-01.llvm}

\end{document}

### 2013-08-22 translation.tex
Created file translation.tex to include all translation rules, and the description of the validation. Added an example used for manual translation. Created a prototype implementation of the translation in Python. Added a pythonesque representation of the example suitable for processing by the prototype.

--- 2013-06-05 paper.tex
@@
 \usepackage[fleqn]{amsmath}
-\usepackage[colorlinks=true]{hyperref}
+\usepackage[colorlinks=false]{hyperref}
 \usepackage{xspace,framed,here,keystroke}
@@
 \DeclareMathOperator{\dom}{\mbox{dom}}
-\DeclareMathOperator{\name}{\mathcal{L}()}
+\DeclareMathOperator{\lbl}{\mathcal{L}()}
+\DeclareMathOperator{\variable}{\mathcal{V}()}
 \newcommand{\llvm}[1]{\texttt{#1}}
@@
 
+\newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}
+
 \begin{document}
@@
         \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & \B{Inst} \\
+        \B{init} & : & seq \B{Inst} \\
         \B{op} & : & seq \B{Oper} \\
@@
         \hline
-        \B{lhs} & : & \B{Name} \\
+        \B{lhs} & : & \B{Vari} \\
         \B{rhs} & : & \B{Expr} \\
@@
 \end{figure}
+\DD{Include \B{Vari} as one of the alternative for an \B{Expr}.}
+\DD{Restore \B{Case} and \B{CaseBr}.}
 
@@
       \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
-      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{struct} \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
+      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
       \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
@@
 \item To encode the implementation state, the translation produces a
-  structure type, named \llvm{@M\$state\$}, and an instance of that type, named
+  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
   \llvm{@M\$self\$}. All the variables and instances of imported machines
@@
   input and output of the corresponding operation, and one item of type
-  \llvm{@M\$state\$*}, to pass the address of the structure encoding the
+  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
   state of the implementation and its modification by the callee. The return
@@
       \textit{(references to imported components)} \\
-      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-      \llvm{declare void @M\$init\$(@M\$state\$*)} \\
+      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
+      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
       \textit{for each constant} \B{cst} \textit{generate} \\
@@
       \textit{for each operation} \B{op} \textit{generate} \\
-      \> \llvm{declare void @M\$op(@M\$state\$*, \ListOf{\nt{type}})} \\
+      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
     \end{tabbing}
@@
       \textit{for each imported component} \B{Q} \textit{generate} \\
-      \> \llvm{@Q\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-      \> \llvm{declare void @Q\$init\$(@Q\$state\$*)} \\
+      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
+      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
       \> \textit{for each constant} \B{cst} \textit{generate} \\
@@
       \> \textit{for each operation} \B{op} \textit{generate} \\
-      \> \> \llvm{declare void @Q\$op(@Q\$state\$*, \ListOf{\nt{type}})} \\
+      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
       \textit{for each constant} \B{cst} \textit{generate} \\
       \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
-      \llvm{@M\$state\$ = type struct \{ \ListOf{\nt{type}} \}} \\
-      \llvm{@M\$self\$ = common global @M\$state\$ zeroinitializer} \\
-      \llvm{define void @M\$init\$(@M\$state\$* \%self\$) \{} \\
+      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
+      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
+      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
       \> \llvm{\ListOf{\nt{block}}} \\
@@
       \textit{for each operation} \B{op} \textit{generate} \\
-      \> \llvm{define void @M\$op(@M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
+      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
       \> \> \llvm{\ListOf{\nt{block}}} \\
@@
 name of the component where it is defined and its own name. Also $\state(\B{M})
-\isdef \llvm{"@\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
+\isdef \llvm{"\%\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
 state of component \B{M}. We assume an unlimited supply of labels and local
-variables: let $\name$ return a fresh LLVM label or local variable name.
+variables: we use $\lbl$ to return a fresh LLVM label and $\variable$ for a
+fresh local variable name.
 
@@
 \begin{small}
-$$\intf(\B{M})_t \isdef \llvm{"struct \{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
+$$\intf(\B{M})_t \isdef \llvm{"\{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
 \end{small}
@@
 \end{small}
+
 The definition of the translation for expressions is described next in
@@
 
+To facilitate the defintiion of rules that apply to similar cases, we introduce
+mapping $\opmap$, that associates B operators to corresponding LLVM keywords:
+$$\opmap = \{ 
+\begin{array}[t]{l}
+  = \mapsto \llvm{"eq"}, 
+  \ne \mapsto \llvm{"ne"}, 
+  < \mapsto \llvm{"slt"}, 
+  \le \mapsto \llvm{"sle"}, 
+  > \mapsto \llvm{"sgt"}, 
+  \ge \mapsto \llvm{"sge"}, \\
+  + \mapsto \llvm{"add"} \}.
+  \end{array}
+$$
+
 \paragraph{Atomic expressions} have a simple translation. In the case of
@@
 \llvm{""} \sep \local(\B{n}) \sep t \\
-& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \name \AND v = \name \IN \\
+& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
 & \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
@@
 \paragraph{Comparisons} are formed by the application of a relational operator
-to two expressions. Mapping $\opmap$ associates B operators to corresponding
-LLVM keywords:
-$$\opmap = \{ = \mapsto \llvm{"eq"}, 
-  \ne \mapsto \llvm{"ne"}, 
-  < \mapsto \llvm{"slt"}, 
-  \le \mapsto \llvm{"sle"}, 
-  > \mapsto \llvm{"sgt"}, 
-  \ge \mapsto \llvm{"sge"} \}.
-$$
+to two expressions.
 
@@
   p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
-& \AND v = \name \IN \\
+& \IN \LET v = \variable \IN \\
 & \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
@@
 
-\paragraph{Arithmetic operations} have a similar structure and interpretation
-as comparisons and their presentation is mostly omitted. Let us present the
-translation of the unary \B{succ} operator though:
+\paragraph{Arithmetic operations}\DD{Actually some arithmetic operations are not binary and different translation rules shall be written.} have a similar 
+structure and interpretation as comparisons. Here is the rule for binary
+arithmetic operations:
+\begin{align*}
+\lefteqn{\trad{r}{\B{Term}} \isdef
+  \LET
+  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
+  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
+& \IN \LET v = \variable \IN \\
+& \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = i32 \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v \sep \llvm{i32}
+\end{align*}
+
+Next is the translation of the unary \B{succ} operator:
 \begin{align*}
   \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
-  \trad{n.arg}{\B{expr}} \AND w = \name \IN \\
-  & \llvm{"\PH{p} \quad \PH{w} = i32 add 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
+  \trad{n.arg}{\B{expr}} \IN \LET w = \variable \IN \\
+  & \llvm{"\PH{p} \quad \PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
 \end{align*}
@@
 \begin{split}
-  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \name \IN} \\
+  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \lbl \IN} \\
   & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
@@
 of B instructions \B{il} and a label $\ell$ and produces the source code of a
-LLVM block with $\ell$ as exit point. It is defined as follows:
-
-\begin{align*}
-\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br \PH{\ell} \nl"} \\
-& \ELSE \LET i, il' = il  \AND p = \trad{il, $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
-& \quad \IF i \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \name \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
+LLVM block with $\ell$ as exit point. It is defined as follows: \DD{Rewrite so
+  that the temporary variable names are generated in the correct order.}
+
+\begin{align*}
+\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
+& \ELSE \LET \B{i}, \B{il'} = \B{il}  \AND p = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
+& \quad \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
 & \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
@@
 \trad{w, $\ell$}{\B{While}} \isdef 
-  & \LET \ell_1 = \name \AND \ell_2 = \name \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
+  & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
   & \llvm{"\PH{\ell_1} : \PH{c}} \\
@@
 \begin{split}
-  \trad{\B{b bl}, $\ell$}{\B{IfBr}} \isdef & \LET \ell_1 = \name \AND \ell_2 = \name \IN \\
+  \trad{\B{b bl}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \IN \\
   & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
@@
 \begin{split}
-  \trad{\B{b}, $\ell$}{\B{IfBr}} \isdef & \IF \B{b.cond} = \text{None} \THEN
+  \trad{\B{b}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \IF \B{b.cond} = \text{None} \THEN
   \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
-  & \ELSE \LET \ell_1 = \name \IN \\
+  & \ELSE \LET \ell_1 = \lbl \IN \\
   & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
@@
 & \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
-& \quad \ELSE \LET v = \name \AND t = \trad{n.type}{\B{Type}} \IN \\
-& \quad \quad \llvm{"\PH{v} = load \PH{t}* getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \ELSE \LET v = \variable \AND t = \trad{n.type}{\B{Type}} \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
 & \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
@@
   & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
-  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"@\PH{c.pre}\$state\$"} \AND \\
+  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"\%\PH{c.pre}\$state\$"} \AND \\
   & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
-  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
+  & \quad \quad \llvm{"getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
   & \IN \\
@@
   \trad{i}{Init} \isdef
-  & \llvm{"define @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
+  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
   & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
-  & \quad\llvm{\PH{\trad{i.init, "exit"}{\B{Inst}}_L}} \\
+  & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
   & \quad\llvm{exit: ret void \nl} \\
@@
 \begin{align*}
-  \trad{m}{InitImp} \isdef
-  & \LET \ell = \name \IN \\
+  \lefteqn{\trad{m}{InitImp} \isdef } \\
+  & \LET \ell = \lbl \IN \\
   & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
-  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}* \PH{\ell})"}
-\end{align*}
-
+  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}*
+    \PH{\ell})"}
+\end{align*}
+
+\section{Validation}
+\label{sec:validation}
+
+Before implementing this translation, we want validate the rules that define it.
+For this, we follow a pragmatic approach, based on testing and simulation of the
+generated LLVM artifacts, and using the existing LLVM tool set.
+
+First, we define a sample set of small B implementation modules, that contain
+all the B constructs handled in the translation. For each such module, the rules
+are applied manually, yielding a LLVM module. This LLVM module is then amended
+with extra code that have two purposes:
+\begin{itemize}
+\item a standalone executable may be generated from it;
+\item the executable creates an instance of the B implementation, applies the
+  initialization code, and offers to the user the choice to interactively invoke
+  all the operations of the implementation; after initialization and each
+  operation application the state of the implementation is printed. 
+\end{itemize}
+
+The LLVM tools that are employed are:
+\begin{description}
+\item[llvm-as] The LLVM assembler generates LLVM bitcode, invoked, e.g., as
+  follows:
+\begin{verbatim}
+llvm-as-mp-3.1 example-01.llvm -o example-01.bc
+\end{verbatim}
+\item[lli] The LLVM bitcode interpreter, which may be invoked as follows:
+\begin{verbatim}
+lli-mp-3.1 example-01.bc
+\end{verbatim}
+\end{description}
 \section{Conclusion and future work}

### 2013-08-22 PROTO/counter_i.py

@@ -1,0 +1,39 @@
+value = { "kind": "Vari",
+          "id":"value",
+          "type": "INT",
+          "scope": "Impl" }
+
+error = { "kind": "Vari",
+          "id":"error",
+          "type": "BOOL",
+          "scope": "Impl" }
+
+expr1 = { "kind": "IntegerLit",
+          "value": "0"}
+
+inst1 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": expr1}
+
+expr2 = { "kind": "BooleanLit",
+          "value": "FALSE"}
+
+inst2 = { "kind": "Beq",
+          "lhs": error,
+          "rhs": expr2}
+
+zero = {}
+inc = {}
+dec = {}
+valid = {}
+get = {}
+set = {}
+
+counter_i = { "kind": "Impl",
+              "id": "counter_i",
+              "concrete_variables": [ value, error ],
+              "initialisation": [ inst1, inst2 ],
+              "operations": [ zero, inc, dec, valid, get, set ] }
+
+value["root"] = counter_i
+error["root"] = counter_i

### 2013-08-22 PROTO/example_01.py

@@ -1,0 +1,87 @@
+value = { "kind": "Vari",
+          "id":"value",
+          "type": "INT",
+          "scope": "Impl" }
+
+error = { "kind": "Vari",
+          "id":"error",
+          "type": "BOOL",
+          "scope": "Impl" }
+
+zero = { "kind": "IntegerLit",
+          "value": "0"}
+
+false = { "kind": "BooleanLit",
+          "value": "FALSE"}
+
+max_int = { "kind": "IntegerLit",
+          "value": "2147483647"}
+
+one = { "kind": "IntegerLit",
+          "value": "1"}
+
+true = { "kind": "BooleanLit",
+          "value": "TRUE"}
+
+term1 = { "kind": "Term",
+          "op": "+",
+          "args" : [value, one] }
+
+comp1 = { "kind": "Comp",
+          "op": "<",
+          "arg1": value,
+          "arg2": max_int }
+
+inst1 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": zero}
+
+inst2 = { "kind": "Beq",
+          "lhs": error,
+          "rhs": false}
+
+inst3 = { "kind": "Blk",
+          "body": [inst1, inst2] }
+
+inst4 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": term1}
+
+inst5 = { "kind": "Beq",
+          "lhs": error,
+          "rhs": true}
+
+branch1 = { "kind": "IfBr",
+            "cond": comp1,
+            "body": inst4 }
+
+branch2 = { "kind": "IfBr",
+            "body": inst5 }
+
+inst6 = { "kind": "If",
+          "branches": [branch1, branch2]}
+
+zero = { "kind": "Oper",
+         "id": "zero",
+         "inp": [],
+         "out": [],
+         "body": inst3 }
+inc = { "kind": "Oper",
+        "id": "inc",
+        "inp": [],
+        "out": [],
+        "body": inst6 }
+
+dec = {}
+valid = {}
+get = {}
+set = {}
+
+counter_i = { "kind": "Impl",
+              "id": "counter_i",
+              "concrete_variables": [ value, error ],
+              "initialisation": [ inst1, inst2 ],
+              "operations": [ zero, inc ] }
+
+for n in [value, error, zero, inc]:
+    n["root"] = counter_i

### 2013-08-22 PROTO/translate.py

@@ -1,0 +1,338 @@
+
+### LLVM IDENTIFIER GENERATION ###
+
+# the code in this section is responsible for generating LLVM names for
+# local variables and labels
+
+llvm_local_var_counter = 0
+llvm_label_counter = 0
+def new_llvm_local_var():
+    global llvm_local_var_counter
+    result = "%"+str(llvm_local_var_counter)
+    llvm_local_var_counter += 1
+    return result
+
+def new_llvm_label():
+    global llvm_label_counter
+    result = "label"+str(llvm_label_counter)
+    llvm_label_counter += 1
+    return result
+
+def reset_llvm_names():
+    global llvm_local_var_counter, llvm_label_counter
+    llvm_local_var_counter = 0
+    llvm_label_counter = 0
+
+def llvm_op(str):
+    if str == "=":
+        return "eq"
+    elif str == "!=":
+        return "ne"
+    elif str == "<":
+        return "slt"
+    elif str == "<=":
+        return "sle"
+    elif str == ">":
+        return "sgt"
+    elif str == ">=":
+        return "sge"
+    elif str == "+":
+        return "add"
+    else:
+        print("error: operator " + str + " not translated")
+        return ""
+        
+### MISC ###
+
+#
+# Function check_kind is used to assert that the arguments of translation
+# function arguments are in the correct syntactic category
+#
+def check_kind(n, s):
+    assert(n["kind"] in s)
+
+nl = "\n"
+sp = " "
+tb = "  "
+
+### TYPE TRANSLATION ###
+
+#
+# This function is responsible for translation B0 type names to LLVM types
+#
+def translate_type(t):
+    assert(t == "INT" or t == "BOOL")
+    if t == "INT":
+        return "i32"
+    else:
+        return "i1"
+
+#
+# Function responsible for producing the name of the type corresponding
+# to the state of B0 implementation n
+#
+def state_name(n):
+    check_kind(n, {"Impl"})
+    return "%"+n["id"]+"$state$"
+
+#
+# This function is responsible for building the LLVM type expression for
+# the type corresponding to the state of B0 implementation n.
+#
+def state_expression(n):
+    check_kind(n, {"Impl"})
+    result = "{"
+    for idx in range(len(n["concrete_variables"])):
+        if idx != 0:
+            result = result + ","
[…]

### 2013-08-22 PAPERS/2013-SBMF/paper_short.tex

@@
 \pagestyle{headings}
-\title{A Proposed Translation of B Implementations to the LLVM}
+\title{Proposal: Translation of B Implementations to LLVM-IR}
 \author{David D\'{e}harbe \inst{1} \and Val\'{e}rio Medeiros Jr.\inst{2}}
@@
 rules defined recursively on the abstract syntax for B implementations. It 
-presents a overview of translation and a simplified example.
+presents an overview of translation and a simplified example.
 \end{abstract}
@@
 Operations and instructions are translated to LLVM instructions as discussed in
-sec.~\ref{sec:control}. For reason of reduced space, only the operations and
+sec.~\ref{sec:control}. For a reason of reduced space, only the operations and
 instructions will be presented with more details.
@@
 
-This paper presents the formal specification of a translation from a large
-subset of the B implementation language to LLVM, a modern compiler internal
-language.  Even though this is still a work in progress, the definition is
-self-contained and has a large enough scope to be applied to B implementations
-where the data belongs to basic types. This specification is a blueprint for a
-code synthesis tool for B that is currently being implemented. This tool will be
-distributed as an extension to Atelier-B under an open-source license.
+This paper presents the formal specification of a translation from a subset of
+the B implementation language to LLVM, a modern compiler internal language.
+Even though this is still a work in progress, the definition is self-contained
+and has a large enough scope to be applied to B implementations where the data
+belongs to basic types. This specification is a blueprint for a code synthesis
+tool for B that is currently being implemented. This tool will be distributed
+as an extension to Atelier-B under an open-source license.
 

### 2013-08-22 PROTO/counter_i.py
Fixed small problems in translation prototype. Added one operation in example_01. Added a README file.

@@ -1,39 +1,0 @@
-value = { "kind": "Vari",
-          "id":"value",
-          "type": "INT",
-          "scope": "Impl" }
-
-error = { "kind": "Vari",
-          "id":"error",
-          "type": "BOOL",
-          "scope": "Impl" }
-
-expr1 = { "kind": "IntegerLit",
-          "value": "0"}
-
-inst1 = { "kind": "Beq",
-          "lhs": value,
-          "rhs": expr1}
-
-expr2 = { "kind": "BooleanLit",
-          "value": "FALSE"}
-
-inst2 = { "kind": "Beq",
-          "lhs": error,
-          "rhs": expr2}
-
-zero = {}
-inc = {}
-dec = {}
-valid = {}
-get = {}
-set = {}
-
-counter_i = { "kind": "Impl",
-              "id": "counter_i",
-              "concrete_variables": [ value, error ],
-              "initialisation": [ inst1, inst2 ],
-              "operations": [ zero, inc, dec, valid, get, set ] }
-
-value["root"] = counter_i
-error["root"] = counter_i

### 2013-08-22 PROTO/example_01.py

@@ -27,10 +27,19 @@ term1 = { "kind": "Term",
           "op": "+",
           "args" : [value, one] }
 
+term2 = { "kind": "Term",
+          "op": "-",
+          "args" : [value, one] }
+
 comp1 = { "kind": "Comp",
           "op": "<",
           "arg1": value,
           "arg2": max_int }
+
+comp2 = { "kind": "Comp",
+          "op": "<",
+          "arg1": zero,
+          "arg2": value }
 
 inst1 = { "kind": "Beq",
           "lhs": value,
@@ -61,6 +70,20 @@ branch2 = { "kind": "IfBr",
 inst6 = { "kind": "If",
           "branches": [branch1, branch2]}
 
+inst7 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": term2}
+
+branch3 = { "kind": "IfBr",
+            "cond": comp2,
+            "body": inst7 }
+
+branch4 = { "kind": "IfBr",
+            "body": inst5 }
+
+inst8 = { "kind": "If",
+          "branches": [branch3, branch4]}
+
 zero = { "kind": "Oper",
          "id": "zero",
          "inp": [],
@@ -72,7 +95,11 @@ inc = { "kind": "Oper",
         "out": [],
         "body": inst6 }
 
-dec = {}
+dec = { "kind": "Oper",
+        "id": "dec",
+        "inp": [],
+        "out": [],
+        "body": inst8 }
 valid = {}
 get = {}
 set = {}
@@ -81,7 +108,7 @@ counter_i = { "kind": "Impl",
               "id": "counter_i",
               "concrete_variables": [ value, error ],
               "initialisation": [ inst1, inst2 ],
-              "operations": [ zero, inc ] }
+              "operations": [ zero, inc, dec ] }
 
-for n in [value, error, zero, inc]:
+for n in [value, error, zero, inc, dec]:
     n["root"] = counter_i

### 2013-08-22 PROTO/translate.py

@@ -1,3 +1,12 @@
+###
+#
+# Prototype for a LLVM-IR generator for B
+#
+# Caveat: This is the first Python program of the author, and he is aware
+# that an object oriented (or, rather, a class oriented) solution could be
+# appropriate.
+#
+###
 
 ### LLVM IDENTIFIER GENERATION ###
 
@@ -23,7 +32,21 @@ def reset_llvm_names():
     llvm_local_var_counter = 0
     llvm_label_counter = 0
 
+### LLVM names for B operators ###
+
 def llvm_op(str):
+    '''
+    Input:
+        - str: the name of an operator in B0
+    Output: The name of the corresponding LLVM operator
+    Example:
+        >>> llvm_op("=") == eq
+        True
+        >>> llvm_op("+") == add
+        True
+    Note: An error message is printed and the empty string
+    is returned if the translation has not been defined.
+    '''
     if str == "=":
         return "eq"
     elif str == "!=":
@@ -38,6 +61,8 @@ def llvm_op(str):
         return "sge"
     elif str == "+":
         return "add"
+    elif str == "-":
+        return "sub"
     else:
         print("error: operator " + str + " not translated")
         return ""
@@ -49,6 +74,15 @@ def llvm_op(str):
 # function arguments are in the correct syntactic category
 #
 def check_kind(n, s):
+    '''
+    Input:
+       - n: represents a B0 syntactic entity
+       - s: a set of syntactic entity class names
+    Output: None.
+    Description: Checks that the class of n is one in s.
+    Example:
+       >>> check_kind(n, {"IntegerLit, BooleanLit"})
+    '''
     assert(n["kind"] in s)
 
 nl = "\n"
@@ -131,7 +165,7 @@ def translate_name(n):
         p = new_llvm_local_var()
         v = new_llvm_local_var()
         text = ""
-        text += tb + p + " = getelementptr %self$, i32 0, i32 " + str() + nl
+        text += tb + p + " = getelementptr " + state_name(n["root"]) + "*" + sp + "%self$, i32 0, i32 " + str(variable_position(n)) + nl
         text += tb + v + " = load " + t + "* " + p + nl
         return (text, v, t)
     else:
@@ -147,7 +181,7 @@ def translate_term(n):
     v = new_llvm_local_var()
[…]

### 2013-08-23 translation.tex
Updated the python encoding of example-01 with all the operations. Updated the translation to handle all of example_01, and verified that it is compilable by llvm-as, and that the interpretation with lli behaves as expected. Small changes in the textual description of the translation.

@@
 \end{description}
+
+Once a rule has been manually validated, we implement it in a prototype tool
+written in Python. Unfortunately the tool does not take as input the B0 source
+file, but instead a collection of Python definitions corresponding to the
+syntactic entities present in the parse tree. For instance, for an integer
+concrete variable named \B{value}, we would have the following Python entity:
+\begin{verbatim}
+value = { "kind": "Vari",
+          "id": "value",
+          "type": "INT",
+          "scope": "impl",
+          "root": counter_i}
+\end{verbatim}
+where \verb'counter_i' would be the Python entity representing the parse tree
+root for the B implementation. Again the output generated by the prototype
+is compiled and simulated using off-the-shelf LLVM tools \verb'llvm-as' and
+\verb'lli'.
+
 \section{Conclusion and future work}

### 2013-08-23 PROTO/example_01.py

@@ -8,6 +8,21 @@ error = { "kind": "Vari",
           "type": "BOOL",
           "scope": "Impl" }
 
+res = { "kind": "Vari",
+        "id": "res",
+        "type": "BOOL",
+        "scope": "Oper" }
+
+res2 = { "kind": "Vari",
+        "id": "res",
+        "type": "INT",
+        "scope": "Oper" }
+
+avalue = { "kind": "Vari",
+           "id":"avalue",
+           "type": "INT",
+           "scope": "Oper" }
+
 zero = { "kind": "IntegerLit",
           "value": "0"}
 
@@ -84,6 +99,40 @@ branch4 = { "kind": "IfBr",
 inst8 = { "kind": "If",
           "branches": [branch3, branch4]}
 
+inst9 = { "kind": "Beq",
+          "lhs": res,
+          "rhs": false }
+
+inst10 = { "kind": "Beq",
+           "lhs": res,
+           "rhs": true }
+
+comp3 = { "kind": "Comp",
+          "op": "=",
+          "arg1": error,
+          "arg2": true }
+
+branch5 = { "kind": "IfBr",
+            "cond": comp3,
+            "body": inst9 }
+
+branch6 = { "kind": "IfBr",
+            "body": inst10 }
+
+inst11 = { "kind": "If",
+           "branches": [branch5, branch6] }
+
+inst12 = { "kind": "Blk",
+           "body": [inst11] }
+
+inst13 = { "kind": "Beq",
+           "lhs": res2,
+           "rhs": value }
+
+inst14 = { "kind": "Beq",
+           "lhs": value,
+           "rhs": avalue }
+
 zero = { "kind": "Oper",
          "id": "zero",
          "inp": [],
@@ -100,15 +149,30 @@ dec = { "kind": "Oper",
         "inp": [],
         "out": [],
         "body": inst8 }
-valid = {}
-get = {}
-set = {}
+
+valid = { "kind": "Oper",
+          "id": "valid",
+          "inp": [],
+          "out": [res],
+          "body": inst12 }
+
+get = { "kind": "Oper",
+        "id": "get",
+        "inp": [],
+        "out": [res2],
+        "body": inst13 }
+
+set = { "kind": "Oper",
+        "id": "set",
+        "inp": [avalue],
+        "out": [],
+        "body": inst14 }
 
 counter_i = { "kind": "Impl",
               "id": "counter_i",
               "concrete_variables": [ value, error ],
[…]

### 2013-08-23 PROTO/translate.py

@@ -168,6 +168,8 @@ def translate_name(n):
         text += tb + p + " = getelementptr " + state_name(n["root"]) + "*" + sp + "%self$, i32 0, i32 " + str(variable_position(n)) + nl
         text += tb + v + " = load " + t + "* " + p + nl
         return (text, v, t)
+    elif n["scope"] == "Oper":
+        return ("", "%"+n["id"], t)
     else:
         print("error: name translation not supported")
         return ("", "", "")
@@ -206,7 +208,7 @@ def translate_comp(n, lbl1, lbl2):
     v = new_llvm_local_var()
     return (p1 +
             p2 +
-            tb + v + " = icmp " + llvm_op(n["op"]) + sp + t1 + sp + ", " + v2 + nl), v
+            tb + v + " = icmp " + llvm_op(n["op"]) + sp + t1 + sp + v1 + ", " + v2 + nl), v
 
 def translate_pred(n, lbl1, lbl2):
     if n["kind"] == "Comp":
@@ -223,8 +225,14 @@ def translate_lv(n):
     global tb, nl
     check_kind(n, {"Vari"})
     t = n["type"]
-    v = new_llvm_local_var()
-    return (tb + v + " = getelementptr " + state_name(n["root"])+ "* %self$, i32 0, i32 " + str(variable_position(n)) + nl, v)
+    if n["scope"] == "Impl":
+        v = new_llvm_local_var()
+        return (tb + v + " = getelementptr " + state_name(n["root"])+ "* %self$, i32 0, i32 " + str(variable_position(n)) + nl, v)
+    elif n["scope"] == "Oper":
+        return ("", "%"+n["id"])
+    else:
+        print("unknown scope for variable " + v["id"])
+        return ("", "UNKNOWN")
 
 def translate_beq(n):
     global tb, sp, nl
@@ -316,29 +324,37 @@ def translate_inst_list_label(l, lbl):
         p2 = translate_inst_list_label(l2, lbl)
         return p1 + p2
 
+### TRANSLATION OF OPERATIONS AND INITIALISATION
+
 def translate_init(i):
     global tb, nl
     check_kind(i, {"Impl"})
     reset_llvm_names()
-    result = "define void @"+i["id"]+"$init$() {" + nl
-    result = result + translate_inst_list_label(i["initialisation"], "exit")
-    result = result + "exit:" + nl
-    result = result + tb + "ret void" + nl
-    result = result + "}" + nl
+    result = "define void @"+i["id"]+"$init$"
[…]

### 2013-08-23 translation.tex
Added a new example with operators pred and succ and a scalar constant. Updated translate.py and translation.tex to reflect the changes necessary to process this example and simulate the generated LLVM code.

@@
         \hline
-        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred}} \\
+        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Vari}} \\
         \hline
@@
 \end{figure}
-\DD{Include \B{Vari} as one of the alternative for an \B{Expr}.}
 \DD{Restore \B{Case} and \B{CaseBr}.}
@@
   B components and are translated to external declarations in LLVM.
-\item Constants and value definitions are translated to LLVM constants. To avoid
-  name conflicts, a B constant named \B{cst} is translated to a LLVM constant
-  named \llvm{@M\$cst} (the rules for name construction avoid possible conflicts).
+\item Constants and value definitions are translated to their value in the
+  LLVM platform.
 \item To encode the implementation state, the translation produces a
@@
       \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
-      \textit{for each constant} \B{cst} \textit{generate} \\
-      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
       \textit{for each operation} \B{op} \textit{generate} \\
@@
       \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
-      \> \textit{for each constant} \B{cst} \textit{generate} \\
-      \> \> \llvm{@Q\$cst = external constant \nt{itype}} \\
       \> \textit{for each operation} \B{op} \textit{generate} \\
       \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
-      \textit{for each constant} \B{cst} \textit{generate} \\
-      \> \llvm{@M\$cst = constant \nt{itype} \nt{iliteral}} \\
       \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
@@
 
-\item $\intf(\B{M})_c$ is the definition of the component constants. It maps
-the constant names to their type:
-$\intf(\B{M})_c \isdef \trad{\B{M.const}}{\ListOf{\B{Cons}}}$ and
-$\trad{\B{c}}{\B{Cons}} \isdef \B{c} \mapsto \trad{\B{c.type}}{\B{Type}}$.
-
 \item $\intf(\B{M})_o$ represents the operation declarations. It maps
@@
     @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
-\item For the definitions of the imported constants: $\trad{\B{i}, \intf}{ICons}
-  \isdef \trad{$\intf(\B{i})_c$}{\ListOf{Cons}}$, where
-  \noindent $\trad{$\B{c} \mapsto t)$}{Cons} \isdef \llvm{"\PH{\Global(\B{c})} =
-    external constant \PH{t}\nl"}$.
 \item For the declarations of the imported operations: $\trad{\B{i},
@@
 
-B constants are translated to constants in the LLVM program, as described in the
-following rule, where parameter \B{c} denote the constant, $M$ the name of the
-module and $\Global$ is the mapping between source and target identifiers. Two
-results are produced: the LLVM source code and an updated identifier mapping:
-\begin{small}
-\begin{align*}
-  \trad{\B{c}}{\B{Cons}} \isdef \llvm{"\PH{\Global(\B{c})} = constant \PH{\trad{\B{c.type}}{\B{Type}}} \PH{\trad{\B{c.val}}{\B{Expr}}} \nl"}
-\end{align*}
-\end{small}
-
-The definition of the translation for expressions is described next in
-section~\ref{sec:expr}.
-
-\section{Translation: Expressions
-  \label{sec:expr}}
+B constants shall be translated to constants in the LLVM program. The
+translation of scalar constants consists in substituting them with their value.
+\DD{The translation of non-scalar constants will be defined in a next version of
+  this document.}  The definition of the translation for expressions is
+described next in section~\ref{sec:expr}.
+
+\section{Translation: Expressions}
+\label{sec:expr}
 
@@
 \lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
-& \IF \B{n} \mbox{ is a constant } \THEN \llvm{""} \sep \Global(\B{n}) \sep t \\
+& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
 & \ELSIF \B{n} \mbox{ is a local variable } \THEN 
@@
 & \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
-& \quad \llvm{"\PH{p} = getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
+& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
 & \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
@@
 \begin{align*}
-  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t =
-  \trad{n.arg}{\B{expr}} \IN \LET w = \variable \IN \\
-  & \llvm{"\PH{p} \quad \PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
+  \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t = \trad{n.arg}{\B{expr}} \IN  \\
+  & \quad \LET w = \variable \IN \\
+  & \quad \quad \llvm{"\PH{p}} \\
+  & \quad \quad \llvm{\PH{w} = add i32 1, \PH{v} \nl"} \sep w \sep \llvm{i32}
 \end{align*}
@@
 
+One issue in the translation of instructions is the disruption of the flow graph
+intro a linear sequence of instructions and branches. This requires the creation
+of instruction labels. As a consequence, our definition of the translation of
+instructions has two kinds of rules, those that take a label as parameter, and
+those that do not. Such label parameter indicates where the control should go
+after the corresponding execution is executed.
+
+\begin{align*}
+\trad{\B{i}, $\ell$}{\B{Inst}}_L \isdef
+& \IF \B{i} \mbox{ is \B{If}} \THEN \trad{\B{i}, $\ell$}{\B{If}}_L \\
+& \ELSIF \B{i} \mbox{ is \B{While}} \THEN \trad{\B{i}, $\ell$}{\B{While}}_L \\
+& \ELSIF \B{i} \mbox{ is \B{Blk}} \THEN \trad{\B{i.body}, $\ell$}{\ListOf{\B{Inst}}}_L \\
+& \ELSE \LET p = \trad{\B{i}}{\B{Inst}} \IN \\
+& \quad \llvm{"\PH{p}} \\
+& \quad \llvm{ br label \%\PH{\ell} \nl"}
+\end{align*}
+
 Function $\trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L$ takes as input a sequence
 of B instructions \B{il} and a label $\ell$ and produces the source code of a
-LLVM block with $\ell$ as exit point. It is defined as follows: \DD{Rewrite so
-  that the temporary variable names are generated in the correct order.}
+LLVM block with $\ell$ as exit point. It is defined as follows:
 
@@
 \trad{\B{il}, $\ell$}{\ListOf{\B{Inst}}}_L \isdef & \IF \B{il} \mbox{ is empty} \THEN \llvm{"br label \%\PH{\ell} \nl"} \\
-& \ELSE \LET \B{i}, \B{il'} = \B{il}  \AND p = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \IN \\
-& \quad \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}} \PH{\ell'}: \PH{p}"} \\
-& \quad \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"}
+& \ELSE \\
+& \LET \B{i}, \B{il'} = \B{il}  \IN \\
+& \quad \quad \LET p_1 = 
+\begin{array}[t]{l}
+  \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \\
+  \quad \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
+  \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"} \IN
+\end{array} \\
+& \quad \quad \IN \\
+& \quad \quad \quad \LET p_2 = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \\
+& \quad \quad \quad \quad \llvm{"\PH{p_1} \PH{p_2}"}
 \end{align*}
@@
 
+\begin{center}
+\newcommand{\no}[0]{--}
+\newcommand{\yes}[0]{$\circ$}
+\begin{tabular}{|l|c|c|}
+\hline
+Example nº & 01 & 02 \\
+\hline
+variables & \yes & \yes \\
+constants & \no & \yes \\
+op. input & \yes & \no \\
+op. output & \yes & \no \\
+\hline
+\multicolumn{3}{|c|}{instructions} \\
+\hline
+block & \yes & \no \\
+local var & \no & \no \\
+identity & \no & \no \\
+becomes equal & \yes & \yes \\
+operation call & \no & \no \\
+condition & \yes & \no \\
+case & \no & \no \\
+assertion & \no & \no \\
+sequence & \yes & \yes \\
+while & \no & \no \\
+\hline
+\multicolumn{3}{|c|}{expressions} \\
+\hline
+succ & \no & \yes \\
+pred & \no & \yes \\
++ & \yes & \no \\
+- & \yes & \no \\
+\hline
+\multicolumn{3}{|c|}{formulas} \\
+\hline
+and & \no & \no \\
+or & \no & \no \\
+not & \no & \no \\
+comparison & \yes & \yes \\
+\hline
+\end{tabular}
+\end{center}
+
 \section{Conclusion and future work}

### 2013-08-23 PROTO/example_02.py

@@ -1,0 +1,133 @@
+value = { "kind": "Vari",
+          "id":"value",
+          "type": "INT",
+          "scope": "Impl" }
+
+error = { "kind": "Vari",
+          "id":"error",
+          "type": "BOOL",
+          "scope": "Impl" }
+
+ilit1 = { "kind": "IntegerLit",
+          "value": "3"}
+
+max = { "kind": "Cons",
+        "id": "MAX",
+        "type": "INT",
+        "value" : ilit1 }
+
+res = { "kind": "Vari",
+        "id": "res",
+        "type": "BOOL",
+        "scope": "Oper" }
+
+res2 = { "kind": "Vari",
+        "id": "res",
+        "type": "INT",
+        "scope": "Oper" }
+
+avalue = { "kind": "Vari",
+           "id":"avalue",
+           "type": "INT",
+           "scope": "Oper" }
+
+zero = { "kind": "IntegerLit",
+          "value": "0"}
+
+false = { "kind": "BooleanLit",
+          "value": "FALSE"}
+
+one = { "kind": "IntegerLit",
+          "value": "1"}
+
+true = { "kind": "BooleanLit",
+          "value": "TRUE"}
+
+term1 = { "kind": "Term",
+          "op": "succ",
+          "args" : [value] }
+
+term2 = { "kind": "Term",
+          "op": "pred",
+          "args" : [value] }
+
+comp1 = { "kind": "Comp",
+          "op": "<",
+          "arg1": value,
+          "arg2": max }
+
+comp2 = { "kind": "Comp",
+          "op": "<",
+          "arg1": zero,
+          "arg2": value }
+
+inst1 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": zero}
+
+inst2 = { "kind": "Beq",
+          "lhs": error,
+          "rhs": false}
+
+inst3 = { "kind": "Blk",
+          "body": [inst1, inst2] }
+
+inst4 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": term1}
+
+inst5 = { "kind": "Beq",
+          "lhs": error,
+          "rhs": true}
+
+branch1 = { "kind": "IfBr",
+            "cond": comp1,
+            "body": inst4 }
+
+branch2 = { "kind": "IfBr",
+            "body": inst5 }
+
+inst6 = { "kind": "If",
+          "branches": [branch1, branch2]}
+
+inst7 = { "kind": "Beq",
+          "lhs": value,
+          "rhs": term2}
+
[…]

### 2013-08-23 PROTO/translate.py

@@ -109,6 +109,8 @@ def state_name(n):
     check_kind(n, {"Impl"})
     return "%"+n["id"]+"$state$"
 
+def global_name(n):
+    return "@" + n["root"]["id"] + "$" + n["id"]
 #
 # This function is responsible for building the LLVM type expression for
 # the type corresponding to the state of B0 implementation n.
@@ -161,32 +163,56 @@ def translate_booleanlit(n):
 def translate_name(n):
     check_kind(n, {"Vari"})
     t = translate_type(n["type"])
-    if n["scope"] == "Impl":
+    if n["scope"] == "Oper":
+        return ("", "%"+n["id"], t)
+    elif n["scope"] == "Impl":
         p = new_llvm_local_var()
         v = new_llvm_local_var()
         text = ""
         text += tb + p + " = getelementptr " + state_name(n["root"]) + "*" + sp + "%self$, i32 0, i32 " + str(variable_position(n)) + nl
         text += tb + v + " = load " + t + "* " + p + nl
         return (text, v, t)
-    elif n["scope"] == "Oper":
-        return ("", "%"+n["id"], t)
     else:
         print("error: name translation not supported")
         return ("", "", "")
         
+def translate_unary(n):
+    check_kind(n, {"Term"})
+    assert (n["op"] in {"succ", "pred"})
+    if n["op"] == "succ":
+        p,v,t = translate_expression(n["args"][0])
+        w = new_llvm_local_var()
+        text = ""
+        text += p
+        text += tb + w + " = add i32 1," + sp + v + nl
+        return (text, w, "i32")
+    elif n["op"] == "pred":
+        p,v,t = translate_expression(n["args"][0])
+        w = new_llvm_local_var()
+        text = ""
+        text += p
+        text += tb + w + " = sub i32 " + v + ", 1" + nl
+        return (text, w, "i32")
+    else:
+        print("error: unary operator translation not supported")
+        return ("", "", "")
+
 def translate_term(n):
     global tb, sp, nl
     check_kind(n, {"Term"})
-    assert(len(n["args"]) == 2)
-    p1,v1,t1 = translate_expression(n["args"][0])
-    p2,v2,t2 = translate_expression(n["args"][1])
-    v = new_llvm_local_var()
-    return (p1 +
-            p2 +
-            tb + v + " = " + llvm_op(n["op"]) + sp + t1 + sp + v1 + ", " + v2 + nl), v, t1
[…]

### 2013-08-28 translation.tex
Added an example to validate translation rules for local variables. Updated translation prototype to handle local variables. Updated translation definition to reflect corrections.

@@
 \newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|}
-\hline
-Example nº & 01 & 02 \\
-\hline
-variables & \yes & \yes \\
-constants & \no & \yes \\
-op. input & \yes & \no \\
-op. output & \yes & \no \\
-\hline
-\multicolumn{3}{|c|}{instructions} \\
-\hline
-block & \yes & \no \\
-local var & \no & \no \\
-identity & \no & \no \\
-becomes equal & \yes & \yes \\
-operation call & \no & \no \\
-condition & \yes & \no \\
-case & \no & \no \\
-assertion & \no & \no \\
-sequence & \yes & \yes \\
-while & \no & \no \\
-\hline
-\multicolumn{3}{|c|}{expressions} \\
-\hline
-succ & \no & \yes \\
-pred & \no & \yes \\
-+ & \yes & \no \\
-- & \yes & \no \\
-\hline
-\multicolumn{3}{|c|}{formulas} \\
-\hline
-and & \no & \no \\
-or & \no & \no \\
-not & \no & \no \\
-comparison & \yes & \yes \\
+\begin{tabular}{|l|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 \\
+\hline
+variables & \yes & \yes & \yes \\
+constants & \no & \yes & \no \\
+op. input & \yes & \no & \yes \\
+op. output & \yes & \no & \yes \\
+\hline
+\multicolumn{4}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes \\
+local var & \no & \no & \yes \\
+identity & \no & \no & \yes \\
+becomes equal & \yes & \yes & \yes \\
+operation call & \no & \no & \no\\
+condition & \yes & \no & \no\\
+case & \no & \no & \no\\
+assertion & \no & \no & \no\\
+sequence & \yes & \yes & \yes\\
+while & \no & \no & \no\\
+\hline
+\multicolumn{4}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes \\
+bool lit & \yes & \yes & \no \\
+vars & \yes & \yes & \yes \\
+loc. vars & \no & \no & \yes \\
+const & \no & \yes & \no \\ 
+succ & \no & \yes & \no \\
+pred & \no & \yes & \no \\
++ & \yes & \no & \no \\
+- & \yes & \no & \no \\
+\hline
+\multicolumn{4}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no \\
+or & \no & \no & \no \\
+not & \no & \no & \no \\
+$<$ & \yes & \yes & \no \\
 \hline

### 2013-08-28 PROTO/example_02.py

@@ -1,3 +1,36 @@
+
+# IMPLEMENTATION
+#    swap_i
+# REFINES
+#    swap
+# CONCRETE_VARIABLES
+#    v1 ,  v2
+# INVARIANT
+#     v1 : INT & v2 : INT
+# INITIALISATION
+#    v1 := 0; v2 := 1
+# OPERATIONS
+#    step =
+#    VAR tmp IN
+#        BEGIN
+#            tmp := v1;
+#            v1 := v2;
+#            v2 := tmp
+#        END
+#    END;
+#    set ( av1 , av2 ) =
+#    BEGIN
+#        v1 := av1;
+#        v2 := av2
+#    END
+#    ;
+#    r1 , r2 <-- get =
+#    BEGIN
+#        r1 := v1;
+#        r2 := v2
+#    END
+# END
+
 value = { "kind": "Vari",
           "id":"value",
           "type": "INT",

### 2013-08-28 PROTO/example_03.py

@@ -1,0 +1,146 @@
+# IMPLEMENTATION
+#    swap_i
+# REFINES
+#    swap
+# CONCRETE_VARIABLES
+#    v1 ,  v2
+# INVARIANT
+#     v1 : INT & v2 : INT
+# INITIALISATION
+#    v1 := 0; v2 := 1
+# OPERATIONS
+#    step =
+#    VAR tmp IN
+#        BEGIN
+#            tmp := v1;
+#            v1 := v2;
+#            v2 := tmp
+#        END
+#    END;
+#    set ( av1 , av2 ) =
+#    BEGIN
+#        v1 := av1;
+#        v2 := av2
+#    END
+#    ;
+#    r1 , r2 <-- get =
+#    BEGIN
+#        r1 := v1;
+#        r2 := v2
+#    END
+# END
+
+v1 = { "kind": "Vari",
+       "id":"v1",
+       "type": "INT",
+       "scope": "Impl" }
+
+v2 = { "kind": "Vari",
+       "id":"v2",
+       "type": "INT",
+       "scope": "Impl" }
+
+ilit0 = { "kind": "IntegerLit",
+          "value": "0"}
+
+ilit1 = { "kind": "IntegerLit",
+          "value": "1"}
+
+r1 = { "kind": "Vari",
+       "id": "r1",
+       "type": "INT",
+       "scope": "Oper" }
+
+r2 = { "kind": "Vari",
+       "id": "r2",
+       "type": "INT",
+       "scope": "Oper" }
+
+av1 = { "kind": "Vari",
+        "id":"av1",
+        "type": "INT",
+        "scope": "Oper" }
+
+av2 = { "kind": "Vari",
+        "id":"av2",
+        "type": "INT",
+        "scope": "Oper" }
+
+tmp = { "kind": "Vari",
+        "id":"tmp",
+        "type": "INT",
+        "scope": "Local" }
+
+
+inst1 = { "kind": "Beq",
+          "lhs": v1,
+          "rhs": ilit0}
+
+inst2 = { "kind": "Beq",
+          "lhs": v2,
+          "rhs": ilit1}
+
+inst3 = { "kind": "Beq",
+          "lhs": tmp,
+          "rhs": v1}
+
+inst4 = { "kind": "Beq",
+          "lhs": v1,
+          "rhs": v2}
+
+inst5 = { "kind": "Beq",
+          "lhs": v2,
+          "rhs": tmp}
+
+inst6 = { "kind": "VarD",
+          "vars": [tmp],
[…]

### 2013-08-28 PROTO/translate.py

@@ -163,7 +163,13 @@ def translate_booleanlit(n):
 def translate_name(n):
     check_kind(n, {"Vari"})
     t = translate_type(n["type"])
-    if n["scope"] == "Oper":
+    if n["scope"] == "Local":
+        v1 = "%"+n["id"]
+        v2 = new_llvm_local_var()
+        t = translate_type(n["type"])
+        text = tb+ v2 + " = load" + sp + t + "*" + sp + v1 + nl
+        return (text, v2, t)
+    elif n["scope"] == "Oper":
         return ("", "%"+n["id"], t)
     elif n["scope"] == "Impl":
         p = new_llvm_local_var()
@@ -256,7 +262,7 @@ def translate_lv(n):
     if n["scope"] == "Impl":
         v = new_llvm_local_var()
         return (tb + v + " = getelementptr " + state_name(n["root"])+ "* %self$, i32 0, i32 " + str(variable_position(n)) + nl, v)
-    elif n["scope"] == "Oper":
+    elif n["scope"] in {"Oper", "Local"}:
         return ("", "%"+n["id"])
     else:
         print("unknown scope for variable " + v["id"])
@@ -301,14 +307,16 @@ def translate_if(n, lbl):
     return translate_if_br(n["branches"], lbl)
 
 def translate_inst(n):
-    check_kind(n, {"Beq"})
+    check_kind(n, {"Beq", "VarD"})
     if n["kind"] == "Beq":
         return translate_beq(n)
+    elif n["kind"] == "VarD":
+        return translate_inst_list(n)
     else:
         return ""
 
 def translate_inst_label(n, lbl):
-    check_kind(n, {"Blk", "If", "While", "Beq"})
+    check_kind(n, {"Blk", "If", "Beq", "VarD"})
     if n["kind"] == "Blk":
         return translate_inst_list_label(n["body"], lbl)
     elif n["kind"] == "If":
@@ -318,6 +326,10 @@ def translate_inst_label(n, lbl):
         result += translate_beq(n)
         result += tb + "br label %" + lbl + nl
         return result
+    elif n["kind"] == "VarD":
+        return translate_inst_list_label(n["body"], lbl)
+    else:
+        print("error: instruction type unknown")
 
 def translate_inst_list(l):
     if len(l) == 0:
@@ -352,21 +364,41 @@ def translate_inst_list_label(l, lbl):
         p2 = translate_inst_list_label(l2, lbl)
         return p1 + p2
 
+### TRANSLATION OF STACK VARIABLE ALLOCATION
+
[…]

### 2013-08-28 translation.tex
Added a B implementation example to validate the translation of WHILE instructions. Added a LLVM scaffolding to instantiate the B implementation and call its operations. Implemented the translation for WHILE instructions. Created a Python module to ease constructing the representation of the abstract syntax trees of B implementation. Updated the definition of the translation to reflect necessary corrections. Updated the contents of the README file and removed useless files from the PROTO directory.

--- 2013-08-23 translation.tex
@@
   \textbf{ let } p \sep v = \trad{n}{\B{Comp}} \IN} \\
-  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \PH{\ell_t}, label \PH{\ell_f} \nl"}
+  & \llvm{"\PH{p} \quad br i1 \PH{v}, label \%\PH{\ell_t}, label \%\PH{\ell_f} \nl"}
 \end{align*}
@@
   & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
-  & \llvm{"\PH{\ell_1} : \PH{c}} \\
-  & \llvm{br i1 \PH{v}, label \PH{\ell_2}, label \PH{\ell} \nl} \\
+  & \llvm{"br label \%\PH{\ell_1} \nl} \\
+  & \llvm{\PH{\ell_1} : \PH{c}} \\
+  & \llvm{br i1 \PH{v}, label \%\PH{\ell_2}, label \%\PH{\ell} \nl} \\
   & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
@@
 \newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|}
-\hline
-Example nº & 01 & 02 \\
-\hline
-variables & \yes & \yes \\
-constants & \no & \yes \\
-op. input & \yes & \no \\
-op. output & \yes & \no \\
-\hline
-\multicolumn{3}{|c|}{instructions} \\
-\hline
-block & \yes & \no \\
-local var & \no & \no \\
-identity & \no & \no \\
-becomes equal & \yes & \yes \\
-operation call & \no & \no \\
-condition & \yes & \no \\
-case & \no & \no \\
-assertion & \no & \no \\
-sequence & \yes & \yes \\
-while & \no & \no \\
-\hline
-\multicolumn{3}{|c|}{expressions} \\
-\hline
-succ & \no & \yes \\
-pred & \no & \yes \\
-+ & \yes & \no \\
-- & \yes & \no \\
-\hline
-\multicolumn{3}{|c|}{formulas} \\
-\hline
-and & \no & \no \\
-or & \no & \no \\
-not & \no & \no \\
-comparison & \yes & \yes \\
+\begin{tabular}{|l|c|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 & 04\\
+\hline
+variables & \yes & \yes & \yes & \yes \\
+constants & \no & \yes & \no & \no \\
+op. input & \yes & \no & \yes & \no \\
+op. output & \yes & \no & \yes & \no \\
+\hline
+\multicolumn{5}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes & \yes\\
+local var & \no & \no & \yes & \yes \\
+identity & \no & \no & \yes & \no \\
+becomes equal & \yes & \yes & \yes & \yes \\
+operation call & \no & \no & \no & \no \\
+condition & \yes & \no & \no & \no \\
+case & \no & \no & \no & \no \\
+assertion & \no & \no & \no & \no\\
+sequence & \yes & \yes & \yes & \yes \\
+while & \no & \no & \no & \yes \\
+\hline
+\multicolumn{5}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes & \yes \\
+bool lit & \yes & \yes & \no & \yes \\
+vars & \yes & \yes & \yes & \yes \\
+loc. vars & \no & \no & \yes & \yes \\
+const & \no & \yes & \no & \no \\ 
+succ & \no & \yes & \no & \yes \\
+pred & \no & \yes & \no & \yes \\
++ & \yes & \no & \no & \yes \\
+- & \yes & \no & \no & \no\\
+\hline
+\multicolumn{5}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no & \no \\
+or & \no & \no & \no & \no \\
+not & \no & \no & \no & \no \\
+$<$ & \yes & \yes & \no & \no\\
 \hline

### 2013-08-28 PROTO/bimp.py

@@ -1,0 +1,75 @@
+### TYPES ###
+
+INT = "INT"
+BOOL = "BOOL"
+
+### TERMINALS ###
+
+def make_var(name, type, scope):
+    return { "kind": "Vari", "id": name, "type": type, "scope": scope }
+
+def make_imp_var(name, type):
+    return make_var(name, type, "Impl")
+
+def make_loc_var(name, type):
+    return make_var(name, type, "Local")
+
+def make_intlit(value):
+    return { "kind": "IntegerLit", "value": str(value)}
+
+zero = make_intlit(0)
+one = make_intlit(1)
+
+def make_boollit(value):
+    return { "kind": "BooleanLit",
+             "value": value}
+false = make_boollit("FALSE")
+true = make_boollit("TRUE")
+
+### COMPOSED EXPRESSIONS ###
+
+def make_term(op, args):
+    return { "kind": "Term", "op": op, "args" : args }
+
+def make_succ(term):
+    return make_term("succ", [term])
+
+def make_pred(term):
+    return make_term("pred", [term])
+
+def make_sum(term1, term2):
+    return make_term("+", [term1, term2])
+
+def make_comp(op, arg1, arg2):
+    return { "kind": "Comp", "op": op, "arg1": arg1, "arg2": arg2 }
+
+def make_lt(term1, term2):
+    return make_comp("<", term1, term2)
+
+### INSTRUCTIONS ###
+
+def make_beq(lhs, rhs):
+    return { "kind": "Beq", "lhs": lhs, "rhs": rhs}
+
+def make_blk(body):
+    return { "kind": "Blk", "body": body }
+
+def make_var_decl(vars, body):
+    return { "kind": "VarD", "vars": vars, "body": body }
+
+def make_while(cond, body):
+    return { "kind": "While", "cond": cond, "body": body }
+
+### COMPONENT ###
+
+def make_oper(id, inp, out, body):
+    return { "kind": "Oper", "id": id, "inp": inp, "out": out, "body": body }
+
+def make_implementation(id, vars, consts, init, ops):
+    root = { "kind": "Impl", "id": id,
+             "concrete_variables": vars, "concrete_constants": consts,
+             "initialisation": init, "operations": ops }
+    for node in root["concrete_variables"]+root["operations"]:
+        node["root"] = root
+    return root
+

### 2013-08-28 PROTO/example_01.py

@@ -168,7 +168,7 @@ set = { "kind": "Oper",
         "out": [],
         "body": inst14 }
 
-counter_i = { "kind": "Impl",
+root = { "kind": "Impl",
               "id": "counter_i",
               "concrete_variables": [ value, error ],
               "initialisation": [ inst1, inst2 ],

### 2013-08-28 PROTO/example_04.py

@@ -1,0 +1,106 @@
+# IMPLEMENTATION
+#    mult_i
+# REFINES
+#    mult
+# CONCRETE_VARIABLES
+#    op1 , op2 , res , ok
+# INVARIANT
+#     op1: INT & op2: INT & res: INT & ok: BOOL
+# INITIALISATION
+#    op1 := 0;
+#    op2 := 0;
+#    res := 0;
+#    ok := TRUE
+# OPERATIONS
+#    inc1 =
+#    BEGIN
+#        op1 := succ(op1);
+#        ok := FALSE
+#    END;
+#    dec1 =
+#    BEGIN
+#        op1 := pred(op1);
+#        ok := FALSE
+#    END;
+#    inc2 =
+#    BEGIN
+#        op2 := succ(op2);
+#        ok := FALSE
+#    END;
+#    dec2 =
+#    BEGIN
+#        op2 := pred(op2);
+#        ok := FALSE
+#    END;
+#    calc =
+#    VAR acc, idx IN
+#        acc := 0;
+#        idx := 0;
+#        WHILE idx < op2 DO
+#            idx := succ(idx);
+#            acc := acc + op1
+#        INVARIANT
+#            idx: 0..op2 &
+#            acc = op1 * idx
+#        VARIANT
+#            op2 - idx
+#        END;
+#        res := acc;
+#        ok := TRUE
+#    END
+# END
+
+from bimp import *
+
+op1 = make_imp_var("op1", INT)
+op2 = make_imp_var("op2", INT)
+res = make_imp_var("res", INT)
+ok = make_imp_var("ok", BOOL)
+
+inst1 = make_beq(op1, zero)
+inst2 = make_beq(op2, zero)
+inst3 = make_beq(res, zero)
+inst4 = make_beq(ok, true)
+
+term1 = make_succ(op1)
+inst5 = make_beq(op1, term1)
+inst6 = make_beq(ok, false)
+inst7 = make_blk([inst5, inst6])
+
+inc1 = make_oper("inc1", [], [], inst7)
+
+term2 = make_pred(op1)
+inst8 = make_beq(op1, term2)
+inst9 = make_blk([inst8, inst6])
+
+dec1 = make_oper("dec1", [], [], inst9)
+
+term3 = make_succ(op2)
+inst10 = make_beq(op2, term3)
+inst11 = make_blk([inst10, inst6])
+
+inc2 = make_oper("inc2", [], [], inst11)
+
[…]

### 2013-08-28 PROTO/translate.py

@@ -16,18 +16,35 @@
 llvm_local_var_counter = 0
 llvm_label_counter = 0
 def new_llvm_local_var():
+    '''
+    Input: None
+    Output: A string representing a LLVM local variable. This string
+    is composed by the prefix "%" followed by a number. Function
+    reset_llvm_names is responsible for zeroing that number.
+    '''
     global llvm_local_var_counter
     result = "%"+str(llvm_local_var_counter)
     llvm_local_var_counter += 1
     return result
 
 def new_llvm_label():
+    '''
+    Input: None
+    Output: A string representing a LLVM instruction label. This string
+    is composed by the prefix "label" followed by a number. Function
+    reset_llvm_names is responsible for zeroing that number.
+    '''
     global llvm_label_counter
     result = "label"+str(llvm_label_counter)
     llvm_label_counter += 1
     return result
 
 def reset_llvm_names():
+    '''
+    Output: None
+    The role of this function is to zero the counters used to build
+    label and local variable identifiers.
+    '''
     global llvm_local_var_counter, llvm_label_counter
     llvm_local_var_counter = 0
     llvm_label_counter = 0
@@ -234,7 +251,7 @@ def translate_expression(n):
 
 ### TRANSLATION OF CONDITIONS ###
 
-def translate_comp(n, lbl1, lbl2):
+def translate_comp(n):
     global tb, sp, nl
     check_kind(n, {"Comp"})
     p1,v1,t1 = translate_expression(n["arg1"])
@@ -244,13 +261,20 @@ def translate_comp(n, lbl1, lbl2):
             p2 +
             tb + v + " = icmp " + llvm_op(n["op"]) + sp + t1 + sp + v1 + ", " + v2 + nl), v
 
-def translate_pred(n, lbl1, lbl2):
+def translate_pred(n):
     if n["kind"] == "Comp":
-        text, v = translate_comp(n, lbl1, lbl2)
+        return translate_comp(n)
+    else:
+        print("not implemented translation for such predicate")
+        return ("", "")
+
+def translate_form(n, lbl1, lbl2):
+    if n["kind"] == "Comp":
+        text, v = translate_comp(n)
         text += tb + "br i1" + sp + v + sp + ", label %" + lbl1 + ", label %" + lbl2 + nl
         return text
     else:
-        print("not implemented translate_pred for formulas.")
[…]

### 2013-08-28 translation.tex
Added an B implementation example with boolean formulas and operators (and, or, not). Added LLVM scaffolding to instantiate implementation and call its operations. Created a Python data structure representing the abstract syntax tree of that B implementation. Updated the Python module with utility functions to build such data structure. Renamed the scaffolding files so that they better match the corresponding example. Implemented translation of boolean operators. Updated the coverage of the B implementation constructs in the documentation of the translation definition.

@@
 
-\begin{center}
-\newcommand{\no}[0]{--}
-\newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|c|c|}
-\hline
-Example nº & 01 & 02 & 03 & 04\\
-\hline
-variables & \yes & \yes & \yes & \yes \\
-constants & \no & \yes & \no & \no \\
-op. input & \yes & \no & \yes & \no \\
-op. output & \yes & \no & \yes & \no \\
-\hline
-\multicolumn{5}{|c|}{instructions} \\
-\hline
-block & \yes & \no & \yes & \yes\\
-local var & \no & \no & \yes & \yes \\
-identity & \no & \no & \yes & \no \\
-becomes equal & \yes & \yes & \yes & \yes \\
-operation call & \no & \no & \no & \no \\
-condition & \yes & \no & \no & \no \\
-case & \no & \no & \no & \no \\
-assertion & \no & \no & \no & \no\\
-sequence & \yes & \yes & \yes & \yes \\
-while & \no & \no & \no & \yes \\
-\hline
-\multicolumn{5}{|c|}{expressions} \\
-\hline
-int lit & \yes & \yes & \yes & \yes \\
-bool lit & \yes & \yes & \no & \yes \\
-vars & \yes & \yes & \yes & \yes \\
-loc. vars & \no & \no & \yes & \yes \\
-const & \no & \yes & \no & \no \\ 
-succ & \no & \yes & \no & \yes \\
-pred & \no & \yes & \no & \yes \\
-+ & \yes & \no & \no & \yes \\
-- & \yes & \no & \no & \no\\
-\hline
-\multicolumn{5}{|c|}{formulas} \\
-\hline
-and & \no & \no & \no & \no \\
-or & \no & \no & \no & \no \\
-not & \no & \no & \no & \no \\
-$<$ & \yes & \yes & \no & \no\\
-\hline
-\end{tabular}
-\end{center}
-
 \section{Conclusion and future work}
@@
 
+\newpage
+\appendix
+
+\section{Test summary}
+
+For each test file, indicates the B0 constructs found in the implementation.
+\begin{center}
+\newcommand{\no}[0]{--}
+\newcommand{\yes}[0]{$\circ$}
+\begin{tabular}{|l|c|c|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 & 04 & 05\\
+\hline
+variables & \yes & \yes & \yes & \yes & \yes\\
+constants & \no & \yes & \no & \no & \yes\\
+op. input & \yes & \no & \yes & \no & \no \\
+op. output & \yes & \no & \yes & \no & \no \\
+\hline
+\multicolumn{6}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes & \yes & \no\\
+local var & \no & \no & \yes & \yes & \no \\
+identity & \no & \no & \yes & \no & \no \\
+becomes equal & \yes & \yes & \yes & \yes & \yes\\
+operation call & \no & \no & \no & \no & \no \\
+condition & \yes & \no & \no & \no & \yes\\
+case & \no & \no & \no & \no & \no \\
+assertion & \no & \no & \no & \no & \no \\
+sequence & \yes & \yes & \yes & \yes & \no \\
+while & \no & \no & \no & \yes & \no \\
+\hline
+\multicolumn{6}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes & \yes & \yes\\
+bool lit & \yes & \yes & \no & \yes & \yes\\
+vars & \yes & \yes & \yes & \yes & \no \\
+loc. vars & \no & \no & \yes & \yes & \no \\
+const & \no & \yes & \no & \no & \yes\\ 
+succ & \no & \yes & \no & \yes & \yes\\
+pred & \no & \yes & \no & \yes & \yes\\
++ & \yes & \no & \no & \yes & \no \\
+- & \yes & \no & \no & \no & \no\\
+\hline
+\multicolumn{6}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no & \no & \yes\\
+or & \no & \no & \no & \no & \yes\\
+not & \no & \no & \no & \no & \yes\\
+$<$ & \yes & \yes & \no & \no & \no\\
+$>$ & \no & \no & \no & \no & \yes \\
+\hline
+\end{tabular}
+\end{center}
+
 \end{document}

### 2013-08-28 PROTO/bimp.py

@@ -5,6 +5,10 @@ BOOL = "BOOL"
 
 ### TERMINALS ###
 
+def make_const(id, type, value):
+    return { "kind": "Cons", "id": id, "type": type, "value" : value }
+
+
 def make_var(name, type, scope):
     return { "kind": "Vari", "id": name, "type": type, "scope": scope }
 
@@ -13,6 +17,9 @@ def make_imp_var(name, type):
 
 def make_loc_var(name, type):
     return make_var(name, type, "Local")
+
+def make_arg_var(name, type):
+    return make_var(name, type, "Oper")
 
 def make_intlit(value):
     return { "kind": "IntegerLit", "value": str(value)}
@@ -46,6 +53,21 @@ def make_comp(op, arg1, arg2):
 def make_lt(term1, term2):
     return make_comp("<", term1, term2)
 
+def make_gt(term1, term2):
+    return make_comp(">", term1, term2)
+
+def make_form(op, args):
+    return { "kind": "Form", "op": op, "args" : args }
+
+def make_and(arg1, arg2):
+    return make_form("and", [arg1, arg2])
+
+def make_or(arg1, arg2):
+    return make_form("or", [arg1, arg2])
+
+def make_not(arg):
+    return make_form("not", [arg])
+
 ### INSTRUCTIONS ###
 
 def make_beq(lhs, rhs):
@@ -56,6 +78,12 @@ def make_blk(body):
 
 def make_var_decl(vars, body):
     return { "kind": "VarD", "vars": vars, "body": body }
+
+def make_if_br(cond, body):
+    return { "kind": "IfBr", "cond": cond, "body": body}
+
+def make_if(branches):
+    return { "kind": "If", "branches": branches }
 
 def make_while(cond, body):
     return { "kind": "While", "cond": cond, "body": body }

### 2013-08-28 PROTO/example_05.py

@@ -1,0 +1,93 @@
+# IMPLEMENTATION
+#    mult_i
+# REFINES
+#    mult
+# CONCRETE_CONSTANTS
+#    THR
+# VALUES
+#    THR = 2
+# CONCRETE_VARIABLES
+#    op1 ,
+#    op2
+# INVARIANT
+#     op1: INT & op2: INT
+# INITIALISATION
+#    op1 := 0;
+#    op2 := 0
+# OPERATIONS
+#    inc1 = op1 := succ(op1);
+#    dec1 = op1 := pred(op1);
+#    inc2 = op2 := succ(op2);
+#    dec2 = op2 := pred(op2);
+#    IF op1 > THR and op2 > THR THEN
+#      res := TRUE
+#    ELSE
+#      res := FALSE
+#    END;
+#    res <-- some =
+#    IF op1 > THR or op2 > THR THEN
+#      res := TRUE
+#    ELSE
+#      res := FALSE
+#    END;
+#    res <-- none =
+#    IF not(op1 > THR) and not (op2 > THR) THEN
+#      res := TRUE
+#    ELSE
+#      res := FALSE
+#    END
+# END
+
+from bimp import *
+
+THR = make_const("THR", INT, make_intlit(2))
+
+op1 = make_imp_var("op1", INT)
+op2 = make_imp_var("op2", INT)
+val = make_imp_var("val", BOOL)
+
+inst1 = make_beq(op1, zero)
+inst2 = make_beq(op2, zero)
+
+term1 = make_succ(op1)
+inst3 = make_beq(op1, term1)
+
+inc1 = make_oper("inc1", [], [], inst3)
+
+term2 = make_pred(op1)
+inst4 = make_beq(op1, term2)
+
+dec1 = make_oper("dec1", [], [], inst4)
+
+term3 = make_succ(op2)
+inst5 = make_beq(op2, term3)
+
+inc2 = make_oper("inc2", [], [], inst5)
+
+term4 = make_pred(op2)
+inst6 = make_beq(op2, term4)
+
+dec2 = make_oper("dec2", [], [], inst6)
+
+cond1 = make_gt(op1, THR)
+cond2 = make_gt(op2, THR)
+form1 = make_and(cond1, cond2)
+form2 = make_or(cond1, cond2)
+form3 = make_and(make_not(cond1), make_not(cond2))
+inst7 = make_beq(val, true)
+inst8 = make_beq(val, false)
[…]

### 2013-08-28 PROTO/translate.py

@@ -268,11 +268,57 @@ def translate_pred(n):
         print("not implemented translation for such predicate")
         return ("", "")
 
+def translate_and(n, lbl1, lbl2):
+    check_kind(n, {"Form"})
+    assert(n["op"] == "and")
+    assert(len(n["args"]) == 2)
+    arg1 = n["args"][0]
+    arg2 = n["args"][1]
+    lbl = new_llvm_label()
+    p1 = translate_form(arg1, lbl, lbl2)
+    p2 = translate_form(arg2, lbl1, lbl2)
+    result = ""
+    result += p1
+    result += lbl + ":" + nl
+    result += p2
+    return result
+
+def translate_or(n, lbl1, lbl2):
+    check_kind(n, {"Form"})
+    assert(n["op"] == "or")
+    assert(len(n["args"]) == 2)
+    arg1 = n["args"][0]
+    arg2 = n["args"][1]
+    lbl = new_llvm_label()
+    p1 = translate_form(arg1, lbl1, lbl)
+    p2 = translate_form(arg2, lbl1, lbl2)
+    result = ""
+    result += p1
+    result += lbl + ":" + nl
+    result += p2
+    return result
+
+def translate_not(n, lbl1, lbl2):
+    check_kind(n, {"Form"})
+    assert(n["op"] == "not")
+    return translate_form(n["args"][0], lbl2, lbl1)
+
 def translate_form(n, lbl1, lbl2):
+    check_kind(n, {"Comp", "Form"})
     if n["kind"] == "Comp":
         text, v = translate_comp(n)
         text += tb + "br i1" + sp + v + sp + ", label %" + lbl1 + ", label %" + lbl2 + nl
         return text
+    elif n["kind"] == "Form":
+        if n["op"] == "and":
+            return translate_and(n, lbl1, lbl2)
+        elif n["op"] == "or":
+            return translate_or(n, lbl1, lbl2)
+        elif n["op"] == "not":
+            return translate_not(n, lbl1, lbl2)
+        else:
+            print("error: unrecognized formula")
+            return ""
     else:
         print("not implemented translate_form for formulas.")
         return ""
@@ -316,7 +362,7 @@ def translate_if_br(lbr, lbl):
         result += lbl_2 + ":\n"
         result += translate_if_br(lbr2, lbl)
     else:
-        if "cond" not in br.keys():
+        if "cond" not in br.keys() or br["cond"] == None:
             result = translate_inst_label(br["body"], lbl)
         else:
             lbl_1 = new_llvm_label()

### 2013-08-29 translation.tex
Restored specification of the translation for the CASE instruction. Created a B implementation with a CASE statement. Created a Python data structure representing that B implementation. Updated the Python module bimp to handle the new constructs (case, multiplication). Created a LLVM scaffolding to instantiate the B implementation and call its operations. Implemented the translation of the CASE statement to LLVM-IR.

--- 2013-08-28 translation.tex
@@
         \\
-      \end{tabular}
-      & &
-      \begin{tabular}[t]{rcl}
         \hline
@@
         \\
+      \end{tabular}
+      & &
+      \begin{tabular}[t]{rcl}
         \hline
@@
         \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
-        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \\
+        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \alt \B{Skip} \\
         \hline
@@
         \\
-%         \\
-%         \hline
-%         \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
-%         \hline
-%         \B{expr} & : & \B{Expr} \\
-%         \B{branches} & : & seq \B{CaseBr}
+        \hline
+        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
+        \hline
+        \B{expr} & : & \B{Expr} \\
+        \B{branches} & : & seq \B{CaseBr}
+        \\
+        \hline
+        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
+        \hline
+        \B{val} & : & seq \B{Value} \\
+        \B{body} & : & \B{Inst} \\
       \end{tabular}
@@
         \B{body} & : & \B{Inst} \\
-%         \hline
-%         \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
-%         \hline
-%         \B{val} & : & seq \B{Value} \\
-%         \B{body} & : & \B{Inst} \\
         \hline
@@
 \DD{Restore \B{Case} and \B{CaseBr}.}
+\DD{Specify \B{Skip}.}
 
@@
 
+\paragraph{Case instructions} are formed by an expression, and a list of case
+branches, each of which has a list of values and an instruction block.
+Optionally, there might be a default branch, and we consider that, in this case,
+the list of values is empty.
+
+The translation to the LLVM IR uses its switch statement. It is composed of a
+variable, a default jump and a jump table. The jump table is a sequence of pairs
+formed by a value, and an instruction label. In the following, \llvm{\%val} is
+the variable, \llvm{\%otherwise} is the default label, and the jump table has
+three entries. The first corresponds to the case where the value of \llvm{\%val}
+is \llvm{0}, in which case the execution jumps to the label \llvm{onzero}.
+\begin{verbatim}
+switch i32 %val, label %otherwise [ i32 0, label %onzero
+                                    i32 1, label %onone
+                                    i32 2, label %ontwo ]
+\end{verbatim}
+
+The translation of the case statement first consists in producing the code $p$
+to evaluate the expression tested in the statement. When that code completes
+execution, the value is stored in variable $v$ of type $t$. Then the jump table
+$j$ and the code $b$ of the different branches are generated. This generation
+takes also as parameters labels $\ell$ and $\ell_o$.
+\begin{align*}
+  \trad{v, $\ell$}{\B{Case}} \isdef & \LET p \sep v \sep t = \trad{v.expr}{\B{Expr}} \IN \\
+  & \LET \ell_o = \lbl \IN \\
+  & \LET j, b = \trad{v.branches, $\ell_o$, $\ell$}{\ListOf{CaseBr}} \IN \\
+  & \quad \llvm{"\PH{p}} \\
+  & \quad \llvm{switch \PH{t} \PH{v}, label \%\PH{\ell_o} [ \PH{j} ] \PH{b}"}
+\end{align*}
+
+The following rule defines the composition of the translation of the different
+branches of a case statement. Again, this translation has two parts: bits of the
+jump table $j$, and instruction blocks $b$. If a case branch has several values,
+then several entries need to be created in the jump table. This is the role
+of the translation function $\trad{}{\B{ListOf{CaseVal}}}$.
+\begin{align*}
+  \trad{b bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \isdef 
+  & \LET \ell = \lbl \IN \\
+  & \LET j, b = \trad{bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \IN \\
+  & \quad \llvm{"\PH{\trad{b.values, $\ell$}{\ListOf{\B{CaseVal}}}} \PH{j}"}
+  \sep \\
+  & \quad
+  \llvm{"\PH{\ell}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L} \PH{b}"}
+\end{align*}
+This rule handles the last branch of a case statement, and needs to test if
+that branch is a default branch or not. If not, then one needs to be artificially
+created (as this is mandatory in the LLVM-IR switch statement):
+\begin{align*}
+  \trad{b, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \isdef 
+  & \IF \B{b.values} = \text{None} \THEN \\
+  & \quad \llvm{""} \sep \llvm{"\PH{\ell_o}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L}"} \\
+  & \ELSE \LET \ell_b = \lbl \IN \\
+  & \quad \trad{b.values, $\ell_b$}{\ListOf{\B{CaseVal}}} \sep \\
+  & \quad \llvm{"\PH{\ell_b}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L}} \\
+  & \quad \quad \llvm{\PH{\ell_o}: \nl br label \% \PH{\ell_e}}
+\end{align*}
+
+The following rules specifies that the values of a case branch are translated in
+order, their translations being concatenated to form the result. The parameter
+$\ell$ is the label of the instruction block containing the translation of the
+branch body:
+\begin{align*}
+  \trad{v, $\ell$}{\ListOf{\B{CaseVal}}} \isdef & \trad{v, $\ell$}{\B{CaseVal}} \\
+  \trad{v vl, $\ell$}{\ListOf{\B{CaseVal}}} \isdef & 
+  \llvm{"\PH{\trad{v, $\ell$}{\B{CaseVal}}} \PH{\trad{vl, $\ell$}{\ListOf{\B{CaseVal}}}}"}
+\end{align*}
+
+This rule defines how a single value of a case branch is translated, namely
+as an entry in the jump table of a switch statement:
+\begin{align*}
+  \trad{v, $\ell$}{\B{CaseVal}} \isdef & \LET p \sep v \sep t = \trad{v}{\B{Expr}} \IN \\
+  & \quad \llvm{"\PH{t} \PH{v}, label \%\PH{\ell} \nl"}
+\end{align*}
+
 \paragraph{Call up} instructions apply B operations with some values and obtain
@@
 
+\newpage
+\appendix
+
+\section{Test summary}
+
+For each test file, indicates the B0 constructs found in the implementation.
+\begin{center}
+\newcommand{\no}[0]{--}
+\newcommand{\yes}[0]{$\circ$}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 & 04 & 05 & 06 \\
+\hline
+variables & \yes & \yes & \yes & \yes & \yes & \yes\\
+constants & \no & \yes & \no & \no & \yes & \no \\
+op. input & \yes & \no & \yes & \no & \no & \no \\
+op. output & \yes & \no & \yes & \no & \no & \no \\
+\hline
+\multicolumn{7}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes & \yes & \no & \yes\\
+local var & \no & \no & \yes & \yes & \no & \yes \\
+identity & \no & \no & \yes & \no & \no & \no \\
+becomes equal & \yes & \yes & \yes & \yes & \yes & \yes\\
+operation call & \no & \no & \no & \no & \no & \no \\
+condition & \yes & \no & \no & \no & \yes & \no \\
+case & \no & \no & \no & \no & \no & \yes \\
+assertion & \no & \no & \no & \no & \no & \no \\
+sequence & \yes & \yes & \yes & \yes & \no & \yes \\
+while & \no & \no & \no & \yes & \no & \no\\
+\hline
+\multicolumn{7}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes & \yes & \yes & \yes\\
+bool lit & \yes & \yes & \no & \yes & \yes & \no \\
+vars & \yes & \yes & \yes & \yes & \no & \yes \\
+loc. vars & \no & \no & \yes & \yes & \no & \yes \\
+const & \no & \yes & \no & \no & \yes & \no \\ 
+succ & \no & \yes & \no & \yes & \yes & \no \\
+pred & \no & \yes & \no & \yes & \yes & \no \\
++ & \yes & \no & \no & \yes & \no & \yes \\
+- & \yes & \no & \no & \no & \no & \no \\
+* & \no & \no & \no & \no & \no & \yes \\
+\hline
+\multicolumn{7}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no & \no & \yes & \no \\
+or & \no & \no & \no & \no & \yes & \no \\
+not & \no & \no & \no & \no & \yes & \no \\
+$<$ & \yes & \yes & \no & \no & \no & \no \\
+$>$ & \no & \no & \no & \no & \yes & \no \\
+\hline
+\end{tabular}
+\end{center}
+
 \end{document}

### 2013-08-29 PROTO/bimp.py

@@ -47,6 +47,9 @@ def make_pred(term):
 def make_sum(term1, term2):
     return make_term("+", [term1, term2])
 
+def make_prod(term1, term2):
+    return make_term("*", [term1, term2])
+
 def make_comp(op, arg1, arg2):
     return { "kind": "Comp", "op": op, "arg1": arg1, "arg2": arg2 }
 
@@ -84,6 +87,12 @@ def make_if_br(cond, body):
 
 def make_if(branches):
     return { "kind": "If", "branches": branches }
+
+def make_case_br(values, body):
+    return { "kind": "CaseBr", "val": values, "body": body }
+
+def make_case(expression, branches):
+    return { "kind": "Case", "expr": expression, "branches": branches }
 
 def make_while(cond, body):
     return { "kind": "While", "cond": cond, "body": body }

### 2013-08-29 PROTO/example_06.py

@@ -1,0 +1,81 @@
+# IMPLEMENTATION
+#     switch_i
+# REFINES
+#     switch
+# CONCRETE_VARIABLES
+#     v1 ,
+#     v2
+# INVARIANT
+#     v1 : INT & v2 : INT &
+#     v1 : 0..4 & v2 : 0..1 &
+#     ((v1 = 0 & v2 = 0) or
+#         (v1 = 1 & v2 = 0) or
+#         (v1 = 1 & v2 = 1) or
+#         (v1 = 2 & v2 = 1) or
+#         (v1 = 4 & v2 = 1))
+# INITIALISATION
+#     v1 := 0;
+#     v2 := 0
+    
+# OPERATIONS
+#     step =
+#     VAR sum IN
+#         sum := v1 + v2;
+#         CASE sum OF
+#             EITHER 0 THEN
+#                 v1 := v1 + 1
+#             OR 1 THEN
+#                 v2 := v1 + 1
+#             OR 2, 3 THEN
+#                 v1 := v1 * 2
+#             ELSE
+#                 BEGIN
+#                     v1 := 0;
+#                     v2 := 0
+#                 END
+#             END
+#         END
+#     END
+# END
+
+from bimp import *
+
+v1 = make_imp_var("v1", INT)
+v2 = make_imp_var("v2", INT)
+
+one = make_intlit(1)
+two = make_intlit(2)
+three = make_intlit(3)
+
+inst1 = make_beq(v1, zero)
+inst2 = make_beq(v2, zero)
+
+sum = make_loc_var("sum", INT)
+term1 = make_sum(v1, v2)
+inst3 = make_beq(sum, term1)
+
+term2 = make_sum(v1, one)
+inst4 = make_beq(v1, term2)
+br1 = make_case_br([zero], inst4)
+
+term3 = make_sum(v2, one)
+inst5 = make_beq(v2, term3)
+br2 = make_case_br([one], inst5)
+
+term4 = make_prod(v1, two)
+inst6 = make_beq(v1, term4)
+br3 = make_case_br([two,three], inst6)
+
+inst7 = make_blk([inst1, inst2])
+br4 = make_case_br([], inst7)
+
+inst8 = make_case(sum, [br1, br2, br3, br4])
+
+inst9 = make_var_decl([sum], [inst3, inst8])
+step = make_oper("step", [], [], inst9)
+
+vars = [v1, v2]
+consts = []
+init = [inst1, inst2]
+ops = [step]
[…]

### 2013-08-29 PROTO/translate.py

@@ -80,6 +80,8 @@ def llvm_op(str):
         return "add"
     elif str == "-":
         return "sub"
+    elif str == "*":
+        return "mul"
     else:
         print("error: operator " + str + " not translated")
         return ""
@@ -105,6 +107,7 @@ def check_kind(n, s):
 nl = "\n"
 sp = " "
 tb = "  "
+tb2 = tb*2
 
 ### TYPE TRANSLATION ###
 
@@ -392,6 +395,53 @@ def translate_while(n, lbl):
     result += body
     return result
 
+def translate_case_val_list(vl, lbl):
+    global tb2, sp, nl
+    result = ""
+    for v in vl:
+        p, v, t = translate_expression(v)
+        result += tb2 + t + sp + v + ", label %" + lbl + nl
+    return result
+
+def translate_case_branch_list(bl, lblo, lble):
+    br = bl[0]
+    bl2 = bl[1:]
+    j, b = "", ""
+    if bl2 == []:
+        if "val" not in br.keys() or br["val"] == [] or br["val"] == None:
+            b += tb + lblo + ":" + nl
+            b += translate_inst_label(br["body"], lble)
+        else:
+            lbl = new_llvm_label()
+            j += translate_case_val_list(br["val"], lbl)
+            b += tb + lbl + ":" + nl
+            b += translate_inst_label(br["body"], lble)
+            b += tb + lblo + ":" + nl
+            b += tb + "branch label %" + lble
+    else:
+        lbl = new_llvm_label()
+        j += translate_case_val_list(br["val"], lbl)
+        b += lbl + ":" + nl
+        b += translate_inst_label(br["body"], lble)
+        j2, b2 = translate_case_branch_list(bl2, lblo, lble)
+        j += j2
+        b += b2
+    return j, b
+
+def translate_case(n, lbl):
+    global nl, tb, sp
+    check_kind(n, {"Case"})
+    p,v,t = translate_expression(n["expr"])
+    lblo = new_llvm_label()
+    j, b = translate_case_branch_list(n["branches"], lblo, lbl)
+    result = ""
+    result += p
+    result += tb + "switch " + t + sp + v + ", label %" + lblo + "[" + nl
+    result += j
+    result += tb + "]" + nl
+    result += b
+    return result
+
 def translate_inst(n):
     check_kind(n, {"Beq", "VarD"})
     if n["kind"] == "Beq":
[…]

### 2013-09-04 translation.tex
Updated the specification of the translation to reflect evolution in handling modular design (more precisely: import clauses in implementations).

@@
 \newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
-\newcommand{\state}[0]{\ensuremath{\sf\Sigma}}
+\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
+\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
+\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
 \newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
@@
 containing definitions used in the implementation; \B{const}, a set of constant
-names, their type and their values; \B{imp}, a sequence of optionally
+names, their type and their values; \B{impo}, a sequence of optionally
 prefixed B machines corresponding to external modules instantiated in the
@@
         \B{id} & : & \B{Name} \\
-        \B{see} & : & seq \B{Name} \\
-        \B{imp} & : & seq \B{Impo} \\
+        \B{see} & : & seq \B{Mach} \\
+        \B{impo} & : & seq \B{Impo} \\
         \B{const} & : & seq \B{Cons} \\
@@
         \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} \\
+        \B{op} & : & seq \B{Oper} 
+        \\
         \hline
@@
         \hline
+        \B{mach} & : & \B{Mach} \\
+        \B{pre} & : & opt \B{Name} 
+        \\
+        \hline
+        \multicolumn{3}{|c|}{Machine: \B{Mach}} \\
+        \hline
         \B{id} & : & \B{Name} \\
-        \B{pre} & : & opt \B{Name} \\
+        \B{impl} & : & \B{Impl} \\
         \hline
@@
         \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
-        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \alt \B{Skip} \\
+        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
         \hline
@@
         \B{branches} & : & seq \B{CaseBr}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
-        \hline
-        \B{val} & : & seq \B{Value} \\
-        \B{body} & : & \B{Inst} \\
       \end{tabular}
@@
       \begin{tabular}[t]{rcl}
+        \hline
+        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
+        \hline
+        \B{val} & : & seq \B{Value} \\
+        \B{body} & : & \B{Inst}
+        \\
         \hline
@@
 \end{figure}
-\DD{Restore \B{Case} and \B{CaseBr}.}
 \DD{Specify \B{Skip}.}
@@
       & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
-      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
+      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
       & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
@@
 
+To conclude, we make no assumption on the existence of a library to obtain
+resources managed by the operating system, such as dynamic memory allocation. In
+the absence of such resource, all data must be allocated either statically, or
+on the current stack frame, by using the \llvm{alloca} instruction.
+
 \section{Translation: a Bird's-Eye View}
@@
 
-  The parameter list of the functions thus produced contain one item for each
+  The parameter list of the functions thus produced contains one item for each
   input and output of the corresponding operation, and one item of type
@@
 \end{enumerate}
-This approach assumes that information on external B components used in \B{M} is
-available during translation. To fulfill this hypothesis, the translation
-produces an ``interface'' artifact containing the information that the
-translation of an importation of \B{M} in a client machine
-requires. Figure~\ref{fig:skel} summarizes the translation principles by
-providing the structure of both the interface and the LLVM implementation thus
-produced.
+The translation is designed to support modular B development. In a B
+implementation, one may reuse existing components, by instantiating (the B term
+is \emph{importing} so-called library machines. Such library machines have their
+own development and implementation. Borrowing terminology from the
+object-oriented world, one may view library machines as classes, and the B
+implementation as an object that contains instances of those classes. In
+practice, different from most OO languages, such instances are not references,
+but the actual objects. We consider this necessary as dynamic memory allocation
+is precluded.
+
+To translate the instantiations of library machines in an implementation, the
+translator needs to have access to the following information on those machines
+and their respective implementation:
+\begin{itemize}
+\item for each library machine \B{M}, the corresponding implementation, denoted
+  \B{M.impl}.
+\item for such implementation, the type of the state and the signature of the
+  operations.
+\end{itemize}
+Note that, when translating the implementation of a library machine, no instance
+shall be generated. It is in the implementation of the main design, that we call
+the \emph{top-level} component, that such instances are created.
+
+Figure~\ref{fig:skel} summarizes the translation principles by providing the
+structure of the LLVM implementation thus produced.
 
@@
   \begin{center}
-    (a) Structure of the interface artifact produced from a B implementation \B{M}:
-    \begin{tabbing}
-      foo \= foo \= \kill
-      \textit{(references to imported components)} \\
-      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
-      \llvm{declare void @M\$init\$(\%M\$state\$*)} \\
-      \textit{for each operation} \B{op} \textit{generate} \\
-      \> \llvm{declare void @M\$op(\%M\$state\$*, \ListOf{\nt{type}})} \\
-    \end{tabbing}
-    (b) Structure of the LLVM module produced from a B implementation \B{M}:
     \begin{tabbing}
@@
       \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
-      \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer} \\
       \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
@@
       \> \> \llvm{exit: ret void} \\
-      \> \llvm{\}}
+      \> \llvm{\}} \\
+      \textit{if this is the top-level component, generate} \\
+      \> \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer}
     \end{tabbing}
   \end{center}
-  \caption{Skeleton of the artifacts produced by the translation.}
+  \caption{Structure of the LLVM module produced from a B implementation \B{M}.}
   \label{fig:skel}
@@
 
-Code generation is defined with a set of rules defined by structural
-recursion. For each grammatical construct, a code generation function is
-defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
+\paragraph{Notations.} Code generation is defined with a set of rules defined by
+structural recursion. For each grammatical construct, a code generation function
+is defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
 instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
@@
 LLVM global name, composed by special characters (to avoid name conflicts), the
-name of the component where it is defined and its own name. Also $\state(\B{M})
-\isdef \llvm{"\%\PH{\B{M.id}}\$state\$"}$ yields the name of the type for the
-state of component \B{M}. We assume an unlimited supply of labels and local
-variables: we use $\lbl$ to return a fresh LLVM label and $\variable$ for a
-fresh local variable name.
+name of the component where it is defined and its own name.  Also, we define
+three auxiliary functions responsible to produce such identifiers of,
+respectively, the type representing the state of a B module \B{n}, its
+initialization function, and the object carrying its state:
+\begin{align*}
+\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
+\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
+\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
+\end{align*}
+We assume an unlimited supply of labels and local variables: we use $\lbl$ to
+return a fresh LLVM label and $\variable$ for a fresh local variable name.
 
@@
 
-The translation of a B module produces two artifacts: an interface and an (LLVM)
-implementation. In the following, we describe how they are produced.
-
-\subsection{Interface}
-
-The definition of the interface $\intf(\B{M})$ of a component \B{M} has four
-key elements, as shown in figure~\ref{fig:skel} (b):
-\begin{enumerate}
-\item $\intf(\B{M})_u$ is the list of used components: $\intf(\B{M})_u \isdef \B{M.imp}$.
-
-\item $\intf(\B{M})_t$ is a textual description of the type representing the
-  state of the component, i.e. a structure with one element per imported
-  component and variable:
-\begin{small}
-$$\intf(\B{M})_t \isdef \llvm{"\{\PH{\trad{\B{M.imp}}{\ListOf{\B{Impo}}}_t} \PH{\trad{\B{M.var}}{\ListOf{\B{Vari}}}_t} \}"},$$
-\end{small}
-where $\trad{\B{i}}{\B{Impo}}_t \isdef \llvm{\state(\B{i})}$ and
-$\trad{\B{v}}{\B{Vari}}_t \isdef \trad{\B{v.type}}{\B{Type}}$.
-
-\item $\intf(\B{M})_o$ represents the operation declarations. It maps
-operation names to the list of input parameter types and output parameter
-types:
-$\intf(\B{M})_o \isdef \trad{\B{M.op}}{\ListOf{\B{Oper}}}_t$ and
-$\trad{\B{o}}{\B{Oper}}_t \isdef \B{o} \mapsto ( \trad{\B{o.inp}}{\ListOf{inp}} \sep \trad{\B{o.out}}{\ListOf{out}} )$, where
-$\trad{\B{i}}{inp} \isdef \trad{\B{i.type}}{\B{Type}}$ and
-$\trad{\B{o}}{out} \isdef \llvm{"\PH{\trad{\B{o.type}}{\B{Type}}}*"}$.
-\end{enumerate}
-
-\subsection{Implementation}
-
-The actual LLVM source code is generated following the template shown in
-figure~\ref{fig:skel} (a), as follows:
-\begin{align*}
-  \lefteqn{\trad{\B{i}, \intf}{\B{Impl}} \isdef
-  \llvm{"\PH{\trad{\B{i.imp}, \intf}{\ListOf{IType}}}} 
-  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{IInit}}}
-  \quad \PH{\trad{\B{i.imp}, \intf}{\ListOf{ICons}}}} \\
-& \llvm{\PH{\trad{\B{i.imp}, \intf}{\ListOf{IOper}}}
-  \quad \PH{\trad{\B{i.id}, \B{i.const}}{\ListOf{\B{Cons}}}}
-  \quad \PH{\trad{\B{i}}{State}}} \\
-& \llvm{@\PH{\B{i.id}}\$self\$ = common global \PH{\state(\B{i.id})} zeroinitializer \nl} \\
-& \llvm{\PH{\trad{\B{i.init}}{Init}} 
-  \quad \PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
-\end{align*}
-We shall now define each rule referenced in this template.
-\begin{enumerate}
-\item For the type definitions corresponding to the state of the imported
-  components: $\trad{i, \intf}{IType} \isdef \llvm{"\PH{\state(\B{i})} =
-    type \PH{\intf(\B{i})_t}\nl"}$.
-\item For the declarations of the initialization functions in the imported
-  components: $\trad{\B{i}, \intf}{Init} \isdef \llvm{"declare void
-    @\PH{\B{i.id}}\$init\$(\PH{\state(\B{i.id})}*)\nl"}$.
-\item For the declarations of the imported operations: $\trad{\B{i},
-    \intf}{IOper} = \trad{\B{i.id}, $\intf(\B{i})_o$}{\ListOf{Oper}}$ where
-  $\trad{\B{Q}, $\B{op} \mapsto (i, o)$}{Oper} = \llvm{"declare void
-    \PH{\Global(\B{op})}(\PH{\state(\B{Q})}*,\PH{i},\PH{o})"}$.
-\item The translation of constant definitions is presented in section~\ref{sec:data}.
-\item For the definition of the type for component states: \\
-\noindent$\trad{\B{M}}{State} \isdef
-\llvm{"\PH{\state(\B{M})} = type \PH{\intf(\B{M})_t}\nl"}$.
-\item For the initialization, the definition of a function is generated. It
-  first initializes the imported modules and then the current module (see
-  section~\ref{sec:control}).
-\item For each operation, a function definition is generated (see
-  section~\ref{sec:control}).
-\end{enumerate}
-Let $\idx$ be a function that maps each imported machine and state variable to
-its position (starting from zero).
+The translation of a B implementation to LLVM, following the template sketched
+in Figure~\ref{fig:skel}, is specified by the following rule, where \B{n} is the
+root of the implementation representation and $\top^?$ a boolean indicating if
+\B{n} is the implementation of a top-level component or of a library machine:
+\begin{align*}
+  \lefteqn{\trad{\B{n}, $\top^?$}{\B{Impl}} \isdef } \\
+  & \llvm{"\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{State}}}_{def}}} & \text{defines the type of imported modules} \\
+  & \llvm{\state(\B{n}) = type \PH{\trad{n}{State}_{expr}}} & \text{defines the type of current module} \\
+  & \llvm{\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{Bhv}}}_{decl}}} & \text{declares of imported modules operations} \\
+  & \llvm{\PH{\trad{\B{n.init}}{\B{Init}}}} & \text{defines the initialization function} \\
+  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}}} & \text{defines functions implementing operations}\\
+  & \llvm{\PH{\trad{\B{n}, $\top^?$}{\B{Inst}}}"} & \text{creates instance (conditional)}
+\end{align*}
+
+To handle the importation of a machine, one needs to produce the definition of
+the type for the state of its implementation. Since this implementation may reference
+other modules, the definition is recursive (parameter $A$ is an accumulator to avoid
+duplicating such definitions):
+\begin{align*}
+  \trad{\B{n}, $A$}{\B{State}}_{def} \isdef 
+  & \IF \B{n} \not\in A \THEN \\
+  & \quad \LET \B{m} = \B{n.mach} \IN \\
+  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
+  & \quad \quad \quad \llvm{"\PH{\trad{\B{i.impo}, $A \cup \{ \B{n} \}$}{\ListOf{\B{State}}}_{def}}} \\
+  & \quad \quad \quad \llvm{\PH{\state(\B{m})} = type \PH{\trad{i}{State}_{expr}} \nl"}
+\end{align*}
+
+The following rules specify how the type representing the state of an
+implementation is generated. It is a structure type, with one element per
+imported machine, and one element per variable. The definition of such type is
+just the enumeration of the type of its elements (the individual elements are
+referenced, not by name, but by position inside the structure):
+\begin{align*}
+  \trad{\B{n}}{\B{State}}_{expr} \isdef & \llvm{"\{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{expr}},  \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{expr}} \}"} \\
+  \trad{\B{n}}{\B{Impo}}_{expr} \isdef & \state(\B{n.mach}) \\
+  \trad{\B{n}}{\B{Vari}}_{expr} \isdef & \trad{\B{n.type}}{\B{Type}}
+\end{align*}
+
+The code generation for the functions implementing the initialization and
+operations of imported machine is specified by the following rules. Not that
+this definition needs not be recursive, as only the operations of a directly
+imported machine may be called in an implementation. Again, parameter $A$ plays
+the role of an accumulator to avoid (harmless, but useless) duplicate
+declarations.
+\begin{align*}
+  \trad{\B{n}, $A$}{\B{Bhv}}_{decl} \isdef 
+  & \IF \B{n} \not\in A \THEN \\
+  & \quad \LET \B{m} = \B{n.mach} \IN \\
+  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
+  & \quad \quad \quad \llvm{declare void \PH{\init(\B{m})}(\PH{\state(\B{m})} *) \nl} \\
+  & \quad \quad \quad \llvm{\PH{\trad{\B{m.op}}{\ListOf{\B{Oper}}}_{decl}}"} \\
+  \trad{\B{n}}{\B{Oper}}_{decl} \isdef &
+  \llvm{"declare void \PH{\Global(\B{n})}(\PH{\state(\B{n.root})}, \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}}, \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}})"}\\
+  \trad{\B{n}}{\B{inp}}_{decl} \isdef & \trad{\B{n.type}}{\B{Type}} \\
+  \trad{\B{n}}{\B{out}}_{decl} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
+\end{align*}
+
+The last rule is responsible for producing the code creating the instance when the
+implementation is that of the top-level component of a design:
+\begin{align*}
+\lefteqn{  \trad{\B{n}, $\top^?$}{\B{Inst}} \isdef } \\
+  & \IF \top^? \THEN \llvm{"\PH{\self(\B{n})} = common global \PH{\state(\B{n})} zeroinitializer \nl"} \\
+  & \ELSE \llvm{""}
+\end{align*}
+
 \section{Translation: Data}
@@
 \hline
+\hline
+\multicolumn{7}{|c|}{clauses} \\
+\hline
 variables & \yes & \yes & \yes & \yes & \yes & \yes\\
@@
 \hline
+\multicolumn{7}{|c|}{types} \\
+\hline
+INT & \yes & \yes & \yes & \yes & \yes & \yes \\
+BOOL & \yes & \yes & \no & \no & \yes & \no \\
+\hline
 \multicolumn{7}{|c|}{expressions} \\

### 2013-09-05 translation.tex
Completed a new version of the translation definition updating support for "IMPORT" clauses and "CALL" operations. Updated Python module bimp.py to build abstract syntax tree for B implementations. Registered a new example of B implementation with a single imported machine, not prefixed. Created the Python data structure corresponding to that example. obs. 1: The implementation of the translation for import clauses is not completed. obs. 2: The construction Python data structures of existing examples needs to be synchronized with respect to bimp.py.

@@
         \hline
-        \B{op} & : & \B{Name} \\
-        \B{pre} & : & opt \B{Name} \\
+        \B{op} & : & \B{Oper} \\
         \B{inp} & : & seq \B{Expr} \\
-        \B{out} & : & seq \B{Name}
+        \B{out} & : & seq \B{Name} \\
+        \B{inst} & : & opt \B{Impo}
         \\
@@
 \end{figure}
-\DD{Specify \B{Skip}.}
 
@@
   LLVM platform.
-\item To encode the implementation state, the translation produces a
-  structure type, named \llvm{\%M\$state\$}, and an instance of that type, named
-  \llvm{@M\$self\$}. All the variables and instances of imported machines
-  are encoded as elements of that structure.  Also, for each operation of each
-  imported machine, an external function declaration is produced.
+\item To encode the implementation state, the translation produces a structure
+  type, named \llvm{\%M\$state\$}, and an instance of that type, named
+  \llvm{@M\$self\$}. All the variables and instances of imported machines are
+  encoded as elements of that structure. These elements are referenced by their
+  position in the structure within the LLVM code. Also, for each operation of
+  each imported machine, an external function declaration is produced.
 \item Each operation \B{op} of the implementation is translated into the
@@
 name of the component where it is defined and its own name.  Also, we define
-three auxiliary functions responsible to produce such identifiers of,
-respectively, the type representing the state of a B module \B{n}, its
-initialization function, and the object carrying its state:
+three auxiliary functions responsible to produce such identifiers:
 \begin{align*}
@@
 return a fresh LLVM label and $\variable$ for a fresh local variable name.
+
+Finally, we need a function that, given a concrete variable, or an imported
+machine, returns the position of the corresponding element in the structure
+representing the state of the implementation. The position of the elements in
+the structure is the same as the position of corresponding B construct in the
+source code:
+\begin{align*}
+  \idx(\B{n}) & \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
+\B{n.root.vari})
+\end{align*}
+where $\mathit{index}(e, s)$ is a function that returns the position of $e$ in
+sequence $s$ and $\conc$ is the concatenation operator for sequences.
 
@@
 & \quad \quad \IN \\
-& \quad \quad \quad \LET p_2 = \trad{il', $\ell$}{\ListOf{\B{Inst}}}_L \\
-& \quad \quad \quad \quad \llvm{"\PH{p_1} \PH{p_2}"}
+& \quad \quad \quad \llvm{"\PH{p_1}} \\
+& \quad \quad \quad \llvm{\PH{\trad{il', $\ell$}{\ListOf{\B{Inst}}}_L}"}
 \end{align*}
@@
 each \B{If} and \B{While} instruction.
-
-\paragraph{While instructions} are translated in two blocks of instructions:
-block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
-body:
-\begin{align*}
-\begin{split}
-\trad{w, $\ell$}{\B{While}} \isdef 
-  & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
-  & \llvm{"br label \%\PH{\ell_1} \nl} \\
-  & \llvm{\PH{\ell_1} : \PH{c}} \\
-  & \llvm{br i1 \PH{v}, label \%\PH{\ell_2}, label \%\PH{\ell} \nl} \\
-  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
-\end{split}
-\end{align*}
-
-\paragraph{If instructions} are sequences of possibly conditional blocks of
-instructions, i.e. branches. After a branch is executed, control must go
-to the next block $\ell$.
-\begin{align*}
-  \trad{i, $\ell$}{\B{If}}_L \isdef \trad{i.branches, $\ell$}{\ListOf{\B{IfBr}}}
-\end{align*}
-To translate conditional branches, one must first generate code to evaluate the
-condition of the branch, followed by an LLVM conditional branching statement,
-translate the body of the branch, and add an unconditional branch to the block
-following the corresponding \B{If} statement.  The translation for branches has
-as arguments a branch \B{b}, the remaining branches \B{bl}, the lexicon, and the
-label $\ell$ of the block following the \B{If} statement they belong to. The
-following definition applies for the case of conditional branches:
-\begin{align*}
-\begin{split}
-  \trad{\B{b bl}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \IN \\
-  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
-  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}} \\
-  & \llvm{\PH{\ell_2} : \PH{\trad{bl, $\ell$}{\ListOf{\B{IfBr}}}}"} \\
-\end{split}
-\end{align*}
-The following rule handles the case of the last branch, which might have a condition or not:
-\begin{align*}
-\begin{split}
-  \trad{\B{b}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \IF \B{b.cond} = \text{None} \THEN
-  \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
-  & \ELSE \LET \ell_1 = \lbl \IN \\
-  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
-  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}}
-\end{split}
-\end{align*}
 
@@
 & \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
+\end{align*}
+
+\paragraph{Call up} instructions apply B operations with some values and obtain
+results. The LLVM code generator produces instructions to evaluate the operation
+arguments, the address of the variables receiving the results as well as an
+instance of the LLVM \llvm{call} instruction. In addition, the proposed
+translation includes as argument the address of a structure storing the state of
+the component corresponding to this operation. Note that this component may be
+either the implementation being translated (in the case of a so-called local
+operation), or an instance of an imported machine.
+\begin{align*}
+  \trad{c}{\B{Call}} \isdef 
+  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
+  \trad{c.out}{\ListOf{Out}} \AND \\
+  & \quad id = \Global(\B{c}) \AND ty = \state(\B{c.root}) \AND \\
+  & \quad val = \IF \B{c.inst} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
+  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})}"}\\
+  & \IN \\
+  & \quad \llvm{"\PH{p_i}} \\
+  & \quad \llvm{\PH{p_o}} \\
+  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
+\end{align*}
+The translation of input and output arguments produces two code snippets. The
+first corresponds to the LLVM instruction sequence $p$ responsible for
+evaluating the argument, and the second to the type $t$ and name $v$ of the
+variable storing the resulting value:
+\begin{align*}
+  \trad{i}{Inp} \isdef 
+  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
+  \trad{o}{Out} \isdef 
+  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
+\end{align*}
+
+\paragraph{Identity} does nothing and no code is generated:
+\begin{align*}
+\begin{split}
+  \trad{n}{\B{Skip}} \isdef & \llvm{""}
+\end{split}
+\end{align*}
+
+\paragraph{Block instructions} are structuring constructs that have no
+operational intent; their translation is that of their body: $\trad{i,
+  $\ell$}{\B{Blk}} \isdef \trad{i.body, $\ell$}{\ListOf{\B{Inst}}}$.
+
+\paragraph{If instructions} are sequences of possibly conditional blocks of
+instructions, i.e. branches. After a branch is executed, control must go
+to the next block $\ell$.
+\begin{align*}
+  \trad{i, $\ell$}{\B{If}}_L \isdef \trad{i.branches, $\ell$}{\ListOf{\B{IfBr}}}
+\end{align*}
+To translate conditional branches, one must first generate code to evaluate the
+condition of the branch, followed by an LLVM conditional branching statement,
+translate the body of the branch, and add an unconditional branch to the block
+following the corresponding \B{If} statement.  The translation for branches has
+as arguments a branch \B{b}, the remaining branches \B{bl}, the lexicon, and the
+label $\ell$ of the block following the \B{If} statement they belong to. The
+following definition applies for the case of conditional branches:
+\begin{align*}
+\begin{split}
+  \trad{\B{b bl}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \IN \\
+  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
+  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}} \\
+  & \llvm{\PH{\ell_2} : \PH{\trad{bl, $\ell$}{\ListOf{\B{IfBr}}}}"} \\
+\end{split}
+\end{align*}
+The following rule handles the case of the last branch, which might have a condition or not:
+\begin{align*}
+\begin{split}
+  \trad{\B{b}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \IF \B{b.cond} = \text{None} \THEN
+  \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
+  & \ELSE \LET \ell_1 = \lbl \IN \\
+  & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
+  & \llvm{\PH{\ell_1} : \PH{\trad{b.body, $\ell$}{\ListOf{\B{Inst}}}_L}}
+\end{split}
 \end{align*}
@@
 
-\paragraph{Call up} instructions apply B operations with some values and obtain
-results. The LLVM code generated must include instructions to evaluate the
-operation arguments, the address of the variables receiving the results as well
-as an instance of the LLVM \llvm{call} instruction. In addition, the proposed
-translation includes as argument the address of a structure storing the state of
-the component corresponding to this operation. Note that this component may be
-either the implementation being translated, or an instance of an imported
-machine.
-\begin{align*}
-  \trad{c}{\B{Call}} \isdef 
-  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
-  \trad{c.out}{\ListOf{Out}} \AND \\
-  & \quad id = \IF \B{c.pre} = \text{None} \THEN \Global(\B{c}) \ELSE \llvm{"@\PH{c.pre}\$\PH{c.id}"} \AND \\
-  & \quad ty = \IF \B{c.pre} = \text{None} \THEN \state(\B{c.root}) \ELSE \llvm{"\%\PH{c.pre}\$state\$"} \AND \\
-  & \quad val = \IF \B{c.pre} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
-  & \quad \quad \llvm{"getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{c.pre})}"}\\
-  & \IN \\
-  & \quad \llvm{"\PH{p_i}} \\
-  & \quad \llvm{\PH{p_o}} \\
-  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
-\end{align*}
-The translation of input and output arguments produces two code snippets. The
-first corresponds to the LLVM instruction sequence $p$ responsible for
-evaluating the argument, and the second to the type $t$ and name $v$ of the
-variable storing the resulting value:
-\begin{align*}
-  \trad{i}{Inp} \isdef 
-  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
-  \trad{o}{Out} \isdef 
-  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
-\end{align*}
-
-\paragraph{Block instructions} are structuring constructs that have no
-operational intent; their translation is that of their body: $\trad{i,
-  $\ell$}{\B{Blk}} \isdef \trad{i.body, $\ell$}{\ListOf{\B{Inst}}}$.
+\paragraph{While instructions} are translated in two blocks of instructions:
+block $\ell_1$ evaluates the loop condition and block $\ell_2$ executes the loop
+body:
+\begin{align*}
+\begin{split}
+\trad{w, $\ell$}{\B{While}} \isdef 
+  & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
+  & \llvm{"br label \%\PH{\ell_1} \nl} \\
+  & \llvm{\PH{\ell_1} : \PH{c}} \\
+  & \llvm{br i1 \PH{v}, label \%\PH{\ell_2}, label \%\PH{\ell} \nl} \\
+  & \llvm{\PH{\ell_2} : \PH{\trad{i.body, $\ell_1$}{\ListOf{\B{Inst}}}_L}}
+\end{split}
+\end{align*}
 

### 2013-09-05 PROTO/bimp.py

@@ -22,16 +22,27 @@ def make_arg_var(name, type):
     return make_var(name, type, "Oper")
 
 def make_intlit(value):
-    return { "kind": "IntegerLit", "value": str(value)}
-
-zero = make_intlit(0)
-one = make_intlit(1)
+    '''
+    - Input: an integer value
+    Creates an integer literal.
+    '''
+    return { "kind": "IntegerLit", 
+             "value": str(value)}
 
 def make_boollit(value):
+    '''
+    - Input: "FALSE" or "TRUE"
+    Creates a boolean literal.
+    '''
     return { "kind": "BooleanLit",
              "value": value}
-false = make_boollit("FALSE")
-true = make_boollit("TRUE")
+
+FALSE = make_boollit("FALSE")
+TRUE = make_boollit("TRUE")
+
+ZERO = make_intlit(0)
+ONE = make_intlit(1)
+MAXINT = make_intlit(2147483647)
 
 ### COMPOSED EXPRESSIONS ###
 
@@ -46,6 +57,9 @@ def make_pred(term):
 
 def make_sum(term1, term2):
     return make_term("+", [term1, term2])
+
+def make_diff(term1, term2):
+    return make_term("-", [term1, term2])
 
 def make_prod(term1, term2):
     return make_term("*", [term1, term2])
@@ -97,16 +111,46 @@ def make_case(expression, branches):
 def make_while(cond, body):
     return { "kind": "While", "cond": cond, "body": body }
 
+def make_call(op, inp, out, inst=None):
+    '''
+    - Input:
+    op: an operation
+    inp: a sequence of inputs
+    out: a sequence of outputs
+    inst: an element of an imported clause (optional)
+    - Output:
+    an operation call
+    '''
+    return { "kind": "Call", "inp": inp, "out": out, 
+             "inst": inst }
+
 ### COMPONENT ###
 
 def make_oper(id, inp, out, body):
     return { "kind": "Oper", "id": id, "inp": inp, "out": out, "body": body }
 
-def make_implementation(id, vars, consts, init, ops):
+def make_import(mach, prefix = None):
+    '''
+    - Input: 
+    mach: a machine
+    prefix: a string prefix (optional).
+    - Output:
+    Creates an element of an import clause.
+    '''
+    assert(mach["kind"] in { "Mach" })
+    return { "kind": "Impo", "mach": mach, "pre": prefix }
+
+def make_implementation(id, imports, consts, vars, init, ops):
     root = { "kind": "Impl", "id": id,
[…]

### 2013-09-05 PROTO/example_07.py

@@ -1,0 +1,136 @@
+# MACHINE counter
+# ...
+# END
+#
+# IMPLEMENTATION counter_i
+# REFINES counter
+# CONCRETE_VARIABLES value, error
+# INVARIANT
+#     value: INT & error : BOOL & (error = overflow)
+# INITIALISATION
+#     value := 0;                 inst0
+#     error := FALSE              inst1
+# OPERATIONS
+#     zero =
+#     BEGIN                       inst6
+#         value := 0;             inst0
+#         error := FALSE          inst1
+#     END;
+#     inc =
+#     IF value < MAXINT           inst4
+#     THEN                        ifbr0
+#         value := value + 1      inst2
+#     ELSE                        ifbr1
+#         error := TRUE           inst3
+#     END;
+#     res <-- get = 
+#         res := value            inst5
+# END
+
+import bimp
+
+value = bimp.make_imp_var("value", bimp.INT)
+error = bimp.make_imp_var("error", bimp.BOOL)
+res = bimp.make_arg_var("res", bimp.INT)
+
+inst0 = bimp.make_beq(value, bimp.ZERO)
+inst1 = bimp.make_beq(error, bimp.FALSE)
+
+inst2 = bimp.make_beq(value, bimp.make_sum(value, bimp.ONE))
+inst3 = bimp.make_beq(error, bimp.TRUE)
+
+res1 = bimp.make_arg_var("res", bimp.INT)
+ifbr0 = bimp.make_if_br(bimp.make_lt(value, bimp.MAXINT), inst2)
+ifbr1 = bimp.make_if_br(None, inst3)
+inst4 = bimp.make_if([ifbr0, ifbr1])
+inst5 = bimp.make_beq(res1, value)
+inst6 = bimp.make_blk([inst0, inst1])
+
+zero = bimp.make_oper("zero", [], [], inst6)
+inc = bimp.make_oper("inc", [], [], inst4)
+get = bimp.make_oper("get", [], [res1], inst5)
+
+counter_i = bimp.make_implementation("counter_i", 
+                                     [],
+                                     [value, error], 
+                                     [], 
+                                     [], 
+                                     [zero, inc, get])
+counter = bimp.make_machine("counter", counter_i)
+
+# IMPLEMENTATION wd_i
+# REFINES wd
+# VALUES timeout=50
+# IMPORTS counter
+# INITIALISATION
[…]

### 2013-09-06 translation.tex
Updated examples so that they are built with the latest version of bimp.py Added an example to test translation of an implementation with an import clause without prefix. Implemented translation of import clauses, and call instructions. Updated the specification of the translation.

@@
 \newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
-\newcommand{\nl}[0]{\text{\Return}}
+\newcommand{\nl}[0]{\ensuremath{\downarrow}}
 \newcommand{\mty}[0]{\texttt{""}}
@@
   \trad{a}{\B{Beq}} \isdef 
-  & \LET l \sep p = \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
+  & \LET l \sep p \sep t'= \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
   & \llvm{"\PH{l}} \\
   & \llvm{\PH{r}} \\
-  & \llvm{store \PH{t} \PH{v}, \PH{t}* \PH{p} \nl"}
+  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
 %   & l \conc r \conc 
@@
 \lefteqn{\trad{n}{lv} \isdef} \\
-& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""}, \local(\B{n}) \\
-& \quad \ELSE \LET v = \variable \AND t = \trad{n.type}{\B{Type}} \IN \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
+& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
+& \quad \ELSE \LET v = \variable \IN \\
 & \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
-& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v
+& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
 \end{align*}
@@
   \trad{c.out}{\ListOf{Out}} \AND \\
-  & \quad id = \Global(\B{c}) \AND ty = \state(\B{c.root}) \AND \\
-  & \quad val = \IF \B{c.inst} = \text{None} \THEN \llvm{"\%self\$"} \ELSE \\
-  & \quad \quad \llvm{"getelementptr \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})}"}\\
+  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
+  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
+  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
+  & \quad \quad \ELSE \LET l = \lbl \IN \\
+  & \quad \quad \quad l \sep \\
+  & \quad \quad \quad \llvm{"\PH{l} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
   & \IN \\
@@
   & \quad \llvm{\PH{p_o}} \\
+  & \quad \llvm{\PH{p}} \\
   & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
@@
   \trad{o}{Out} \isdef 
-  & \LET p \sep v \sep t = \trad{o}{\B{Expr}} \IN p \sep \llvm{"\PH{t}* \PH{v}"}
+  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
 \end{align*}
@@
 \newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|c|c|c|c|}
-\hline
-Example nº & 01 & 02 & 03 & 04 & 05 & 06 \\
-\hline
-\hline
-\multicolumn{7}{|c|}{clauses} \\
-\hline
-variables & \yes & \yes & \yes & \yes & \yes & \yes\\
-constants & \no & \yes & \no & \no & \yes & \no \\
-op. input & \yes & \no & \yes & \no & \no & \no \\
-op. output & \yes & \no & \yes & \no & \no & \no \\
-\hline
-\multicolumn{7}{|c|}{instructions} \\
-\hline
-block & \yes & \no & \yes & \yes & \no & \yes\\
-local var & \no & \no & \yes & \yes & \no & \yes \\
-identity & \no & \no & \yes & \no & \no & \no \\
-becomes equal & \yes & \yes & \yes & \yes & \yes & \yes\\
-operation call & \no & \no & \no & \no & \no & \no \\
-condition & \yes & \no & \no & \no & \yes & \no \\
-case & \no & \no & \no & \no & \no & \yes \\
-assertion & \no & \no & \no & \no & \no & \no \\
-sequence & \yes & \yes & \yes & \yes & \no & \yes \\
-while & \no & \no & \no & \yes & \no & \no\\
-\hline
-\multicolumn{7}{|c|}{types} \\
-\hline
-INT & \yes & \yes & \yes & \yes & \yes & \yes \\
-BOOL & \yes & \yes & \no & \no & \yes & \no \\
-\hline
-\multicolumn{7}{|c|}{expressions} \\
-\hline
-int lit & \yes & \yes & \yes & \yes & \yes & \yes\\
-bool lit & \yes & \yes & \no & \yes & \yes & \no \\
-vars & \yes & \yes & \yes & \yes & \no & \yes \\
-loc. vars & \no & \no & \yes & \yes & \no & \yes \\
-const & \no & \yes & \no & \no & \yes & \no \\ 
-succ & \no & \yes & \no & \yes & \yes & \no \\
-pred & \no & \yes & \no & \yes & \yes & \no \\
-+ & \yes & \no & \no & \yes & \no & \yes \\
-- & \yes & \no & \no & \no & \no & \no \\
-* & \no & \no & \no & \no & \no & \yes \\
-\hline
-\multicolumn{7}{|c|}{formulas} \\
-\hline
-and & \no & \no & \no & \no & \yes & \no \\
-or & \no & \no & \no & \no & \yes & \no \\
-not & \no & \no & \no & \no & \yes & \no \\
-$<$ & \yes & \yes & \no & \no & \no & \no \\
-$>$ & \no & \no & \no & \no & \yes & \no \\
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 \\
+\hline
+\hline
+\multicolumn{8}{|c|}{clauses} \\
+\hline
+import w.o prefix & \no & \no & \no & \no & \no & \no & \yes \\
+import w prefix & \no & \no & \no & \no & \no & \no & \no \\
+variables & \yes & \yes & \yes & \yes & \yes & \yes & \no \\
+constants & \no & \yes & \no & \no & \yes & \no & \yes \\
+op. input & \yes & \no & \yes & \no & \no & \no & \no \\
+op. output & \yes & \no & \yes & \no & \no & \no & \yes \\
+\hline
+\multicolumn{8}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes & \yes & \no & \yes & \no \\
+local var & \no & \no & \yes & \yes & \no & \yes & \yes \\
+identity & \no & \no & \yes & \no & \no & \no & \no \\
+becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+operation call & \no & \no & \no & \no & \no & \no & \yes \\
+condition & \yes & \no & \no & \no & \yes & \no & \yes \\
+case & \no & \no & \no & \no & \no & \yes & \no \\
+assertion & \no & \no & \no & \no & \no & \no & \no \\
+sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes \\
+while & \no & \no & \no & \yes & \no & \no & \yes \\
+\hline
+\multicolumn{8}{|c|}{types} \\
+\hline
+INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+BOOL & \yes & \yes & \no & \no & \yes & \no & \yes \\
+\hline
+\multicolumn{8}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes \\
+vars & \yes & \yes & \yes & \yes & \no & \yes & \no \\
+loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes \\
+const & \no & \yes & \no & \no & \yes & \no & \yes \\ 
+succ & \no & \yes & \no & \yes & \yes & \no & \no \\
+pred & \no & \yes & \no & \yes & \yes & \no & \no \\
++ & \yes & \no & \no & \yes & \no & \yes & \yes \\
+- & \yes & \no & \no & \no & \no & \no & \yes \\
+* & \no & \no & \no & \no & \no & \yes & \no \\
+\hline
+\multicolumn{8}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no & \no & \yes & \no & \no \\
+or & \no & \no & \no & \no & \yes & \no & \no \\
+not & \no & \no & \no & \no & \yes & \no & \no \\
+$<$ & \yes & \yes & \no & \no & \no & \no & \yes \\
+$>$ & \no & \no & \no & \no & \yes & \no & \yes \\
 \hline

### 2013-09-06 PROTO/bimp.py

@@ -73,6 +73,9 @@ def make_lt(term1, term2):
 def make_gt(term1, term2):
     return make_comp(">", term1, term2)
 
+def make_eq(term1, term2):
+    return make_comp("=", term1, term2)
+
 def make_form(op, args):
     return { "kind": "Form", "op": op, "args" : args }
 
@@ -86,6 +89,9 @@ def make_not(arg):
     return make_form("not", [arg])
 
 ### INSTRUCTIONS ###
+
+def make_skip():
+    return { "kind": "Skip" }
 
 def make_beq(lhs, rhs):
     return { "kind": "Beq", "lhs": lhs, "rhs": rhs}
@@ -121,7 +127,7 @@ def make_call(op, inp, out, inst=None):
     - Output:
     an operation call
     '''
-    return { "kind": "Call", "inp": inp, "out": out, 
+    return { "kind": "Call", "op": op, "inp": inp, "out": out, 
              "inst": inst }
 
 ### COMPONENT ###
@@ -146,7 +152,7 @@ def make_implementation(id, imports, consts, vars, init, ops):
              "concrete_constants": consts,
              "concrete_variables": vars, 
              "initialisation": init, "operations": ops }
-    for node in root["concrete_variables"]+root["operations"]:
+    for node in imports + vars + ops:
         node["root"] = root
     return root
 

### 2013-09-06 PROTO/example_01.py

@@ -1,178 +1,89 @@
-value = { "kind": "Vari",
-          "id":"value",
-          "type": "INT",
-          "scope": "Impl" }
+# IMPLEMENTATION counter_i
+# CONCRETE_VARIABLES value, error
+# INITIALISATION
+#   BEGIN
+#     value := 0;
+#     error := FALSE
+#   END
+# OPERATIONS
+# zero =
+#   BEGIN
+#     value := 0;
+#     error := FALSE
+#   END
+# inc =
+#   IF value < MAXINT
+#   THEN
+#     value := value + 1
+#   ELSE
+#     error := TRUE
+#   END;
+# dec =
+#   IF 0 < value
+#   THEN
+#     value := value - 1
+#   ELSE
+#     error := TRUE
+#   END;
+# res <-- valid =
+#   BEGIN
+#     IF error = TRUE
+#     THEN
+#       res := FALSE
+#     ELSE
+#       res := TRUE
+#     END
+#   END;
+# res <-- get =
+#   res := value;
+# set(avalue) =
+#   value := avalue
+# END
 
-error = { "kind": "Vari",
-          "id":"error",
-          "type": "BOOL",
-          "scope": "Impl" }
+import bimp
 
-res = { "kind": "Vari",
-        "id": "res",
-        "type": "BOOL",
-        "scope": "Oper" }
+value = bimp.make_imp_var("value", bimp.INT)
+error = bimp.make_imp_var("error", bimp.INT)
+res1 = bimp.make_arg_var("res", bimp.BOOL)
+res2 = bimp.make_arg_var("res", bimp.INT)
+avalue = bimp.make_arg_var("avalue", bimp.INT)
 
-res2 = { "kind": "Vari",
-        "id": "res",
-        "type": "INT",
-        "scope": "Oper" }
+inst1 = bimp.make_beq(value, bimp.ZERO)
+inst2 = bimp.make_beq(error, bimp.FALSE)
+inst3 = bimp.make_blk([inst1, inst2])
+zero = bimp.make_oper("zero", [], [], inst3)
 
-avalue = { "kind": "Vari",
-           "id":"avalue",
-           "type": "INT",
-           "scope": "Oper" }
+inst4 = bimp.make_beq(value, bimp.make_sum(value, bimp.ONE))
+inst5 = bimp.make_beq(error, bimp.TRUE)
[…]

### 2013-09-06 PROTO/example_02.py

@@ -1,166 +1,64 @@
-
-# IMPLEMENTATION
-#    swap_i
-# REFINES
-#    swap
-# CONCRETE_VARIABLES
-#    v1 ,  v2
-# INVARIANT
-#     v1 : INT & v2 : INT
+# IMPLEMENTATION counter_i
+# REFINES...
+# CONCRETE_CONSTANTS MAX
+# PROPRTIES MAX : INT & MAX = 3
+# CONCRETE_VARIABLES value, error
 # INITIALISATION
-#    v1 := 0; v2 := 1
+#   BEGIN
+#     value := 0;
+#     error := FALSE
+#   END
 # OPERATIONS
-#    step =
-#    VAR tmp IN
-#        BEGIN
-#            tmp := v1;
-#            v1 := v2;
-#            v2 := tmp
-#        END
-#    END;
-#    set ( av1 , av2 ) =
-#    BEGIN
-#        v1 := av1;
-#        v2 := av2
-#    END
-#    ;
-#    r1 , r2 <-- get =
-#    BEGIN
-#        r1 := v1;
-#        r2 := v2
-#    END
+# zero =
+#   BEGIN
+#     value := 0;
+#     error := FALSE
+#   END
+# inc =
+#   IF value < MAX
+#   THEN
+#     value := succ value
+#   ELSE
+#     error := TRUE
+#   END;
+# END
+# dec =
+#   IF 0 < value
+#   THEN
+#     value := pred value
+#   ELSE
+#     error := TRUE
+#   END;
 # END
 
-value = { "kind": "Vari",
-          "id":"value",
-          "type": "INT",
-          "scope": "Impl" }
+import bimp
 
-error = { "kind": "Vari",
-          "id":"error",
-          "type": "BOOL",
-          "scope": "Impl" }
+value = bimp.make_imp_var("value", bimp.INT)
+error = bimp.make_imp_var("error", bimp.INT)
+three = bimp.make_intlit(3)
+max = bimp.make_const("MAX", bimp.INT, three)
 
-ilit1 = { "kind": "IntegerLit",
-          "value": "3"}
+comp1 = bimp.make_lt(value, max)
+comp2 = bimp.make_lt(bimp.ZERO, value)
 
-max = { "kind": "Cons",
-        "id": "MAX",
-        "type": "INT",
-        "value" : ilit1 }
-
-res = { "kind": "Vari",
-        "id": "res",
-        "type": "BOOL",
[…]

### 2013-09-06 PROTO/example_03.py

@@ -30,117 +30,33 @@
 #    END
 # END
 
-v1 = { "kind": "Vari",
-       "id":"v1",
-       "type": "INT",
-       "scope": "Impl" }
+import bimp
 
-v2 = { "kind": "Vari",
-       "id":"v2",
-       "type": "INT",
-       "scope": "Impl" }
-
-ilit0 = { "kind": "IntegerLit",
-          "value": "0"}
-
-ilit1 = { "kind": "IntegerLit",
-          "value": "1"}
-
-r1 = { "kind": "Vari",
-       "id": "r1",
-       "type": "INT",
-       "scope": "Oper" }
-
-r2 = { "kind": "Vari",
-       "id": "r2",
-       "type": "INT",
-       "scope": "Oper" }
-
-av1 = { "kind": "Vari",
-        "id":"av1",
-        "type": "INT",
-        "scope": "Oper" }
-
-av2 = { "kind": "Vari",
-        "id":"av2",
-        "type": "INT",
-        "scope": "Oper" }
-
-tmp = { "kind": "Vari",
-        "id":"tmp",
-        "type": "INT",
-        "scope": "Local" }
-
-
-inst1 = { "kind": "Beq",
-          "lhs": v1,
-          "rhs": ilit0}
-
-inst2 = { "kind": "Beq",
-          "lhs": v2,
-          "rhs": ilit1}
-
-inst3 = { "kind": "Beq",
-          "lhs": tmp,
-          "rhs": v1}
-
-inst4 = { "kind": "Beq",
-          "lhs": v1,
-          "rhs": v2}
-
-inst5 = { "kind": "Beq",
-          "lhs": v2,
-          "rhs": tmp}
-
-inst6 = { "kind": "VarD",
-          "vars": [tmp],
-          "body": [inst3, inst4, inst5] }
-
-inst7 = { "kind": "Beq",
-          "lhs": v1,
-          "rhs": av1 }
-
-inst8 = { "kind": "Beq",
-          "lhs": v2,
-          "rhs": av2 }
-
-inst9 = { "kind": "Blk",
-          "body": [inst7, inst8] }
-
-inst10 = { "kind": "Beq",
-           "lhs": r1,
-           "rhs": v1 }
-
-inst11 = { "kind": "Beq",
-           "lhs": r2,
-           "rhs": v2 }
-
-inst12 = { "kind": "Blk",
[…]

### 2013-09-06 PROTO/example_04.py

@@ -50,57 +50,59 @@
 #    END
 # END
 
-from bimp import *
+import bimp
 
-op1 = make_imp_var("op1", INT)
-op2 = make_imp_var("op2", INT)
-res = make_imp_var("res", INT)
-ok = make_imp_var("ok", BOOL)
+op1 = bimp.make_imp_var("op1", bimp.INT)
+op2 = bimp.make_imp_var("op2", bimp.INT)
+res = bimp.make_imp_var("res", bimp.INT)
+ok = bimp.make_imp_var("ok", bimp.BOOL)
 
-inst1 = make_beq(op1, zero)
-inst2 = make_beq(op2, zero)
-inst3 = make_beq(res, zero)
-inst4 = make_beq(ok, true)
+inst1 = bimp.make_beq(op1, bimp.ZERO)
+inst2 = bimp.make_beq(op2, bimp.ZERO)
+inst3 = bimp.make_beq(res, bimp.ZERO)
+inst4 = bimp.make_beq(ok, bimp.TRUE)
 
-term1 = make_succ(op1)
-inst5 = make_beq(op1, term1)
-inst6 = make_beq(ok, false)
-inst7 = make_blk([inst5, inst6])
+term1 = bimp.make_succ(op1)
+inst5 = bimp.make_beq(op1, term1)
+inst6 = bimp.make_beq(ok, bimp.FALSE)
+inst7 = bimp.make_blk([inst5, inst6])
 
-inc1 = make_oper("inc1", [], [], inst7)
+inc1 = bimp.make_oper("inc1", [], [], inst7)
 
-term2 = make_pred(op1)
-inst8 = make_beq(op1, term2)
-inst9 = make_blk([inst8, inst6])
+term2 = bimp.make_pred(op1)
+inst8 = bimp.make_beq(op1, term2)
+inst9 = bimp.make_blk([inst8, inst6])
 
-dec1 = make_oper("dec1", [], [], inst9)
+dec1 = bimp.make_oper("dec1", [], [], inst9)
 
-term3 = make_succ(op2)
-inst10 = make_beq(op2, term3)
-inst11 = make_blk([inst10, inst6])
+term3 = bimp.make_succ(op2)
+inst10 = bimp.make_beq(op2, term3)
+inst11 = bimp.make_blk([inst10, inst6])
 
-inc2 = make_oper("inc2", [], [], inst11)
+inc2 = bimp.make_oper("inc2", [], [], inst11)
 
-term4 = make_pred(op2)
-inst12 = make_beq(op2, term4)
[…]

### 2013-09-06 PROTO/example_05.py

@@ -38,56 +38,57 @@
 #    END
 # END
 
-from bimp import *
+import bimp
 
-THR = make_const("THR", INT, make_intlit(2))
+THR = bimp.make_const("THR", bimp.INT, bimp.make_intlit(2))
 
-op1 = make_imp_var("op1", INT)
-op2 = make_imp_var("op2", INT)
-val = make_imp_var("val", BOOL)
+op1 = bimp.make_imp_var("op1", bimp.INT)
+op2 = bimp.make_imp_var("op2", bimp.INT)
+val = bimp.make_imp_var("val", bimp.BOOL)
 
-inst1 = make_beq(op1, zero)
-inst2 = make_beq(op2, zero)
+inst1 = bimp.make_beq(op1, bimp.ZERO)
+inst2 = bimp.make_beq(op2, bimp.ZERO)
 
-term1 = make_succ(op1)
-inst3 = make_beq(op1, term1)
+term1 = bimp.make_succ(op1)
+inst3 = bimp.make_beq(op1, term1)
 
-inc1 = make_oper("inc1", [], [], inst3)
+inc1 = bimp.make_oper("inc1", [], [], inst3)
 
-term2 = make_pred(op1)
-inst4 = make_beq(op1, term2)
+term2 = bimp.make_pred(op1)
+inst4 = bimp.make_beq(op1, term2)
 
-dec1 = make_oper("dec1", [], [], inst4)
+dec1 = bimp.make_oper("dec1", [], [], inst4)
 
-term3 = make_succ(op2)
-inst5 = make_beq(op2, term3)
+term3 = bimp.make_succ(op2)
+inst5 = bimp.make_beq(op2, term3)
 
-inc2 = make_oper("inc2", [], [], inst5)
+inc2 = bimp.make_oper("inc2", [], [], inst5)
 
-term4 = make_pred(op2)
-inst6 = make_beq(op2, term4)
+term4 = bimp.make_pred(op2)
+inst6 = bimp.make_beq(op2, term4)
 
-dec2 = make_oper("dec2", [], [], inst6)
+dec2 = bimp.make_oper("dec2", [], [], inst6)
 
-cond1 = make_gt(op1, THR)
-cond2 = make_gt(op2, THR)
-form1 = make_and(cond1, cond2)
-form2 = make_or(cond1, cond2)
-form3 = make_and(make_not(cond1), make_not(cond2))
-inst7 = make_beq(val, true)
-inst8 = make_beq(val, false)
[…]

### 2013-09-06 PROTO/example_06.py

@@ -38,44 +38,44 @@
 #     END
 # END
 
-from bimp import *
+import bimp
 
-v1 = make_imp_var("v1", INT)
-v2 = make_imp_var("v2", INT)
+v1 = bimp.make_imp_var("v1", bimp.INT)
+v2 = bimp.make_imp_var("v2", bimp.INT)
 
-one = make_intlit(1)
-two = make_intlit(2)
-three = make_intlit(3)
+two = bimp.make_intlit(2)
+three = bimp.make_intlit(3)
 
-inst1 = make_beq(v1, zero)
-inst2 = make_beq(v2, zero)
+inst1 = bimp.make_beq(v1, bimp.ZERO)
+inst2 = bimp.make_beq(v2, bimp.ZERO)
 
-sum = make_loc_var("sum", INT)
-term1 = make_sum(v1, v2)
-inst3 = make_beq(sum, term1)
+sum = bimp.make_loc_var("sum", bimp.INT)
+term1 = bimp.make_sum(v1, v2)
+inst3 = bimp.make_beq(sum, term1)
 
-term2 = make_sum(v1, one)
-inst4 = make_beq(v1, term2)
-br1 = make_case_br([zero], inst4)
+term2 = bimp.make_sum(v1, bimp.ONE)
+inst4 = bimp.make_beq(v1, term2)
+br1 = bimp.make_case_br([bimp.ZERO], inst4)
 
-term3 = make_sum(v2, one)
-inst5 = make_beq(v2, term3)
-br2 = make_case_br([one], inst5)
+term3 = bimp.make_sum(v2, bimp.ONE)
+inst5 = bimp.make_beq(v2, term3)
+br2 = bimp.make_case_br([bimp.ONE], inst5)
 
-term4 = make_prod(v1, two)
-inst6 = make_beq(v1, term4)
-br3 = make_case_br([two,three], inst6)
+term4 = bimp.make_prod(v1, two)
+inst6 = bimp.make_beq(v1, term4)
+br3 = bimp.make_case_br([two,three], inst6)
 
-inst7 = make_blk([inst1, inst2])
-br4 = make_case_br([], inst7)
+inst7 = bimp.make_blk([inst1, inst2])
+br4 = bimp.make_case_br([], inst7)
 
-inst8 = make_case(sum, [br1, br2, br3, br4])
+inst8 = bimp.make_case(sum, [br1, br2, br3, br4])
 
-inst9 = make_var_decl([sum], [inst3, inst8])
-step = make_oper("step", [], [], inst9)
[…]

### 2013-09-06 PROTO/example_07.py

@@ -50,12 +50,13 @@ zero = bimp.make_oper("zero", [], [], inst6)
 inc = bimp.make_oper("inc", [], [], inst4)
 get = bimp.make_oper("get", [], [res1], inst5)
 
-counter_i = bimp.make_implementation("counter_i", 
-                                     [],
-                                     [value, error], 
-                                     [], 
-                                     [], 
-                                     [zero, inc, get])
+imports = []
+consts = []
+vars = [value, error]
+init = [inst0, inst1]
+ops = [zero, inc, get]
+
+counter_i = bimp.make_implementation("counter_i", imports, consts, vars, init, ops)
 counter = bimp.make_machine("counter", counter_i)
 
 # IMPLEMENTATION wd_i

### 2013-09-06 PROTO/translate.py

@@ -49,6 +49,47 @@ def reset_llvm_names():
     llvm_local_var_counter = 0
     llvm_label_counter = 0
 
+def instance_name(n):
+    '''
+    - Input:
+      n: A node representing a B implementation
+    - Output:
+      A string for the name of the LLVM variable representing the instance 
+      of that implementation.
+    '''
+    check_kind(n, {"Impl"})
+    return "@"+n["id"]+"$self$"
+
+def state_t_name(n):
+    '''
+    - Input:
+      n: A node representing a B implementation
+    - Output:
+      A string for the name of the LLVM type representing the state of
+      the implementation n.
+    '''
+    check_kind(n, {"Impl"})
+    return "%"+n["id"]+"$state$"
+
+def global_name(n):
+    '''
+    - Input:
+      n: A node representing a B construct
+    - Output:
+      A string for the name of the LLVM construct representing n.
+    '''
+    return "@" + n["root"]["id"] + "$" + n["id"]
+
+def init_name(n):
+    '''
+    - Input:
+      n : a node representing a B implementation
+    - Output:
+    String with the name of the LLVM function encoding the initialization
+    of that implementation.
+    '''
+    return "@"+n["id"]+"$init$"
+
 ### LLVM names for B operators ###
 
 def llvm_op(str):
@@ -82,6 +123,10 @@ def llvm_op(str):
         return "sub"
     elif str == "*":
         return "mul"
+    elif str == "/":
+        return "sdiv"
+    elif str == "mod":
+        return "srem"
     else:
         print("error: operator " + str + " not translated")
         return ""
@@ -109,6 +154,24 @@ sp = " "
 tb = "  "
 tb2 = tb*2
 
+def state_position(n):
+    '''
+    Input:
+      - n: represents a B state variable or imported machine
+    Output:
+      - position of n in the list of imported machines and state variables
+    An error message is printed and the value 0 is returned if n is
+    not found.
+    '''
+    root = n["root"]
+    result = 0
+    for n2 in root["imports"] + root["concrete_variables"]:
+        if n2 == n:
+            return result
+        result += 1
+    print("error: position of imported machine or variable not found in implementation")
+    return 0
+
 ### TYPE TRANSLATION ###
 
 #
@@ -122,27 +185,22 @@ def translate_type(t):
         return "i1"
 
[…]

### 2013-09-07 translation.tex
Added to translation.tex a reminder to fix generation of initialization function (call to initialization of imported modules). Added B source exerpts for example07. Added B and Python data structure for example08 (multiple imports, with prefix, and state variables).

@@
 
-Initialization is done is two steps: first all imported modules are initialized,
-second the initialization clause is executed.
+Initialization is done is two steps: first all imported modules are
+initialized,\DD{Verify that Python implementation includes call to such
+  initialisation functions.}  second the initialization clause is executed.
 \begin{align*}

### 2013-09-07 PROTO/example_08.py

@@ -1,0 +1,205 @@
+# MACHINE counter
+# ...
+# END
+#
+# IMPLEMENTATION counter_i
+# REFINES counter
+# CONCRETE_VARIABLES value, error
+# INVARIANT
+#     value: INT & error : BOOL & (error = overflow)
+# INITIALISATION
+#     value := 0;                 inst0
+#     error := FALSE              inst1
+# OPERATIONS
+#     zero =
+#     BEGIN                       inst6
+#         value := 0;             inst0
+#         error := FALSE          inst1
+#     END;
+#     inc =
+#     IF value < MAXINT           inst4
+#     THEN                        ifbr0
+#         value := value + 1      inst2
+#     ELSE                        ifbr1
+#         error := TRUE           inst3
+#     END;
+#     res <-- get = 
+#         res := value            inst5
+# END
+
+import bimp
+
+value = bimp.make_imp_var("value", bimp.INT)
+error = bimp.make_imp_var("error", bimp.BOOL)
+res = bimp.make_arg_var("res", bimp.INT)
+
+inst0 = bimp.make_beq(value, bimp.ZERO)
+inst1 = bimp.make_beq(error, bimp.FALSE)
+
+inst2 = bimp.make_beq(value, bimp.make_sum(value, bimp.ONE))
+inst3 = bimp.make_beq(error, bimp.TRUE)
+
+res1 = bimp.make_arg_var("res", bimp.INT)
+ifbr0 = bimp.make_if_br(bimp.make_lt(value, bimp.MAXINT), inst2)
+ifbr1 = bimp.make_if_br(None, inst3)
+inst4 = bimp.make_if([ifbr0, ifbr1])
+inst5 = bimp.make_beq(res1, value)
+inst6 = bimp.make_blk([inst0, inst1])
+
+zero = bimp.make_oper("zero", [], [], inst6)
+inc = bimp.make_oper("inc", [], [], inst4)
+get = bimp.make_oper("get", [], [res1], inst5)
+
+imports = []
+consts = []
+vars = [value, error]
+init = [inst0, inst1]
+ops = [zero, inc, get]
+
+counter_i = bimp.make_implementation("counter_i", imports, consts, vars, init, ops)
+counter = bimp.make_machine("counter", counter_i)
+
+# IMPLEMENTATION timer_i
+# IMPORTS hh.counter, mm.counter, ss.counter
+
+imp_hh = bimp.make_import(counter, "hh")
+imp_mm = bimp.make_import(counter, "mm")
[…]

### 2013-09-09 PROTO/example_01.py
Fixed regression in example_01.py. Fixed stupid errors in example_08.py. Fixed regression in translate.py.

@@ -43,7 +43,7 @@
 import bimp
 
 value = bimp.make_imp_var("value", bimp.INT)
-error = bimp.make_imp_var("error", bimp.INT)
+error = bimp.make_imp_var("error", bimp.BOOL)
 res1 = bimp.make_arg_var("res", bimp.BOOL)
 res2 = bimp.make_arg_var("res", bimp.INT)
 avalue = bimp.make_arg_var("avalue", bimp.INT)

### 2013-09-09 PROTO/example_08.py

@@ -83,13 +83,13 @@ timer_init = [inst7, inst8]
 # OPERATIONS
 #     tick =
 #     IF ( is_running = TRUE ) & ( overflow = FALSE )
-cond1 = make_and(make_eq(is_running, bimp.TRUE),
-                 make_eq(overflow, bimp.FALSE))
+cond1 = bimp.make_and(bimp.make_eq(is_running, bimp.TRUE),
+                      bimp.make_eq(overflow, bimp.FALSE))
 #     THEN
 #         VAR hours, minutes, seconds IN
-hours = make_loc_var("hours", bimp.INT)
-minutes = make_loc_var("minutes", bimp.INT)
-seconds = make_loc_var("seconds", bimp.INT)
+hours = bimp.make_loc_var("hours", bimp.INT)
+minutes = bimp.make_loc_var("minutes", bimp.INT)
+seconds = bimp.make_loc_var("seconds", bimp.INT)
 #             hours <-- hh.get;
 #             minutes <-- mm.get;
 #             seconds <-- ss.get;
@@ -97,41 +97,41 @@ inst9 = bimp.make_call(get, [], [hours], imp_hh)
 inst10 = bimp.make_call(get, [], [minutes], imp_mm)
 inst11 = bimp.make_call(get, [], [seconds], imp_ss)
 #             IF (seconds < 59) THEN
-fiftynine = make_intlit(59)
-cond2 = make_lt(seconds, fiftynine)
+fiftynine = bimp.make_intlit(59)
+cond2 = bimp.make_lt(seconds, fiftynine)
 #                 ss.inc
-inst12 = make_call(inc, [], [], imp_ss)
-if_br0 = make_if_br(cond2, inst12)
+inst12 = bimp.make_call(inc, [], [], imp_ss)
+if_br0 = bimp.make_if_br(cond2, inst12)
 #             ELSE
 #                 BEGIN
 #                     ss.zero;
-inst13 = make_call(zero, [], [], imp_ss)
+inst13 = bimp.make_call(zero, [], [], imp_ss)
 #                     IF (minutes < 59) THEN
-cond3 = make_lt(minutes, fiftynine)
+cond3 = bimp.make_lt(minutes, fiftynine)
 #                         mm.inc
-inst14 = make_call(inc, [], [], imp_mm)
-if_br1 = make_if_br(cond3, inst14)
+inst14 = bimp.make_call(inc, [], [], imp_mm)
[…]

### 2013-09-09 PROTO/translate.py

@@ -534,13 +534,13 @@ def translate_call(n):
     result += pi
     result += po
     id = global_name(operation)
-    ty = state_name(operation["root"])
+    ty = state_t_name(operation["root"])
     if n["inst"] == None:
         val = "%self$"
     else:
         val = new_llvm_local_var()
         result += (tb + val + " = getelementptr " + 
-                   state_name(n["inst"]["root"]) + "* " +
+                   state_t_name(n["inst"]["root"]) + "* " +
                    "%self$, i32 0, i32 " + 
                    str(state_position(n["inst"])) + nl)
     args = [ty + sp + val] + il + ol
@@ -861,7 +861,6 @@ def translate_implementation(i, toplevel):
     check_kind(i, {"Impl"})
     result = ""
     result += translate_type_def_import_list(i)
-    result += state_t_name(i)+ " = type "+ translate_type_expr_impl(i)+ nl
     result += translate_op_decl_import_list(i)
     result += translate_constants(i)
     result += translate_state(i)

### 2013-09-09 translation.tex
Fixed issue #1. Tested code generation with multiple imports of the same machine.

--- 2013-09-06 translation.tex
@@
 \B{op}, a set of operations, each being an algorithmic description of the
-different functionalities provided by the implementation.
+different functionalities provided by the implementation.\DD{Include the
+possibility that the import is a machine without implementation. Instead it
+has a list of typed, implementable, variables.}
 
@@
   & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
-  & \quad \quad \ELSE \LET l = \lbl \IN \\
-  & \quad \quad \quad l \sep \\
-  & \quad \quad \quad \llvm{"\PH{l} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
+  & \quad \quad \ELSE \LET v = \variable \IN \\
+  & \quad \quad \quad v \sep \\
+  & \quad \quad \quad \llvm{"\PH{v} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
   & \IN \\
@@
 
-Initialization is done is two steps: first all imported modules are initialized,
-second the initialization clause is executed.
+Initialization is done is two steps: first all imported modules are
+initialized,  second the initialization clause is executed.
 \begin{align*}
@@
   & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
-  & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
+  & \quad\llvm{\PH{\trad{i, i.impl}{\ListOf{InitImp}}}} \\
   & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
@@
 \begin{align*}
-  \lefteqn{\trad{m}{InitImp} \isdef } \\
-  & \LET \ell = \lbl \IN \\
-  & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
-  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}*
-    \PH{\ell})"}
+  \lefteqn{\trad{\B{impl}, \B{impo}}{InitImp} \isdef} \\
+  & \LET v = \variable \AND \B{m} = \B{impo.mach} \IN \\
+  & \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{impl.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{impo})}\nl}\\
+  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m.impl})}*
+    \PH{\ell}) \nl"}
 \end{align*}
@@
 \newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-\hline
-Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 \\
-\hline
-\hline
-\multicolumn{8}{|c|}{clauses} \\
-\hline
-import w.o prefix & \no & \no & \no & \no & \no & \no & \yes \\
-import w prefix & \no & \no & \no & \no & \no & \no & \no \\
-variables & \yes & \yes & \yes & \yes & \yes & \yes & \no \\
-constants & \no & \yes & \no & \no & \yes & \no & \yes \\
-op. input & \yes & \no & \yes & \no & \no & \no & \no \\
-op. output & \yes & \no & \yes & \no & \no & \no & \yes \\
-\hline
-\multicolumn{8}{|c|}{instructions} \\
-\hline
-block & \yes & \no & \yes & \yes & \no & \yes & \no \\
-local var & \no & \no & \yes & \yes & \no & \yes & \yes \\
-identity & \no & \no & \yes & \no & \no & \no & \no \\
-becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-operation call & \no & \no & \no & \no & \no & \no & \yes \\
-condition & \yes & \no & \no & \no & \yes & \no & \yes \\
-case & \no & \no & \no & \no & \no & \yes & \no \\
-assertion & \no & \no & \no & \no & \no & \no & \no \\
-sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes \\
-while & \no & \no & \no & \yes & \no & \no & \yes \\
-\hline
-\multicolumn{8}{|c|}{types} \\
-\hline
-INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-BOOL & \yes & \yes & \no & \no & \yes & \no & \yes \\
-\hline
-\multicolumn{8}{|c|}{expressions} \\
-\hline
-int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes \\
-vars & \yes & \yes & \yes & \yes & \no & \yes & \no \\
-loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes \\
-const & \no & \yes & \no & \no & \yes & \no & \yes \\ 
-succ & \no & \yes & \no & \yes & \yes & \no & \no \\
-pred & \no & \yes & \no & \yes & \yes & \no & \no \\
-+ & \yes & \no & \no & \yes & \no & \yes & \yes \\
-- & \yes & \no & \no & \no & \no & \no & \yes \\
-* & \no & \no & \no & \no & \no & \yes & \no \\
-\hline
-\multicolumn{8}{|c|}{formulas} \\
-\hline
-and & \no & \no & \no & \no & \yes & \no & \no \\
-or & \no & \no & \no & \no & \yes & \no & \no \\
-not & \no & \no & \no & \no & \yes & \no & \no \\
-$<$ & \yes & \yes & \no & \no & \no & \no & \yes \\
-$>$ & \no & \no & \no & \no & \yes & \no & \yes \\
+\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
+\hline
+Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 & 08 \\
+\hline
+\hline
+\multicolumn{9}{|c|}{clauses} \\
+\hline
+import w.o prefix & \no & \no & \no & \no & \no & \no & \yes & \no \\
+import w prefix & \no & \no & \no & \no & \no & \no & \no & \yes \\
+variables & \yes & \yes & \yes & \yes & \yes & \yes & \no & \yes \\
+constants & \no & \yes & \no & \no & \yes & \no & \yes & \yes \\
+op. input & \yes & \no & \yes & \no & \no & \no & \no & \no \\
+op. output & \yes & \no & \yes & \no & \no & \no & \yes & \yes \\
+\hline
+\multicolumn{9}{|c|}{instructions} \\
+\hline
+block & \yes & \no & \yes & \yes & \no & \yes & \no & \yes \\
+local var & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
+identity & \no & \no & \yes & \no & \no & \no & \no & \no \\
+becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+operation call & \no & \no & \no & \no & \no & \no & \yes & \yes \\
+condition & \yes & \no & \no & \no & \yes & \no & \yes & \yes \\
+case & \no & \no & \no & \no & \no & \yes & \no & \no \\
+assertion & \no & \no & \no & \no & \no & \no & \no & \no \\
+sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes & \yes \\
+while & \no & \no & \no & \yes & \no & \no & \yes & \no \\
+\hline
+\multicolumn{9}{|c|}{types} \\
+\hline
+INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+BOOL & \yes & \yes & \no & \no & \yes & \no & \yes & \yes \\
+\hline
+\multicolumn{9}{|c|}{expressions} \\
+\hline
+int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
+bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes & \yes \\
+vars & \yes & \yes & \yes & \yes & \no & \yes & \no & \yes \\
+loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
+const & \no & \yes & \no & \no & \yes & \no & \yes & \yes\\ 
+succ & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
+pred & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
++ & \yes & \no & \no & \yes & \no & \yes & \yes & \no \\
+- & \yes & \no & \no & \no & \no & \no & \yes & \no \\
+* & \no & \no & \no & \no & \no & \yes & \no & \no \\
+\hline
+\multicolumn{9}{|c|}{formulas} \\
+\hline
+and & \no & \no & \no & \no & \yes & \no & \no & \no \\
+or & \no & \no & \no & \no & \yes & \no & \no & \no \\
+not & \no & \no & \no & \no & \yes & \no & \no & \no \\
+$<$ & \yes & \yes & \no & \no & \no & \no & \yes & \yes \\
+$>$ & \no & \no & \no & \no & \yes & \no & \yes & \no \\
 \hline

### 2013-09-09 PROTO/example_08.py

@@ -197,7 +197,7 @@ has_overflown = bimp.make_oper("has_overflown", [], [answer], inst37)
 # END
 
 timer_ops = [tick, reset, start, stop, elapsed, has_overflown]
-wd_i = bimp.make_implementation("wd_i", 
+timer_i = bimp.make_implementation("timer_i", 
                                 timer_imports, 
                                 timer_consts, 
                                 timer_vars, 

### 2013-09-09 PROTO/translate.py

@@ -192,8 +192,8 @@ def state_expression(n):
     check_kind(n, {"Impl", "Mach"})
     result = "{"
     components = []
-    for imp in n["imports"]:
-        components += [state_t_name(imp)]
+    for impo in n["imports"]:
+        components += [state_t_name(impo["mach"]["impl"])]
     for var in n["concrete_variables"]:
         components += [translate_type(var["type"])]
     for i in range(len(components)):
@@ -665,7 +665,37 @@ def translate_signature(n):
     result += ")"
     return result
 
+def translate_import_init_list(n):
+    '''
+    Input:
+      - n: root node of a B implementation AST
+    Output:
+      LLVM source code consisting of the calls to the initialization functions
+      of the instances of machines imported by n.
+    '''
+    global nl, tb
+    check_kind(n, {"Impl"})
+    imports = n["imports"]
+    idx = 0
+    result = ""
+    for impo in imports:
+        mach = impo["mach"]
+        lbl = new_llvm_local_var()
+        m = impo["mach"]
+        result += (tb + lbl + " = getelementptr " + 
+                   state_t_name(n) + "%self$, i32 0, i32 " + str(idx) + nl)
+        result += (tb + "call void " + init_name(m) + 
+                   "(" + state_t_name(m["impl"]) + "* " + lbl + ")" + nl)
+        idx += 1
+    return result
+    
 def translate_init(i):
+    '''
+    Input:
+      - i: root node of a B implementation AST
+    Output:
+    LLVM implementation of the initialization clause of i (a LLVM function).
+    '''
     global tb, nl, sp
     check_kind(i, {"Impl"})
     reset_llvm_names()
@@ -673,6 +703,7 @@ def translate_init(i):
     result += "("+state_t_name(i)+"* %self$) {" + nl
     result += "entry:" + nl
     result += translate_alloc_inst_list(i["initialisation"])
+    result += translate_import_init_list(i)
     result += translate_inst_list_label(i["initialisation"], "exit")
     result += "exit:" + nl
     result += tb + "ret void" + nl

### 2013-09-16 PROTO/loadbxml.py
Yet another work in progress version of the Python loader of BXML files. Able to load very basic operations, substitutions and expressions.

@@ -50,16 +50,16 @@ def sym_table_add(table, id, pyt):
 #
 
 def get(node):
-    return val.get("ident")
+    return node.get("ident")
 
 def value(node):
-    return val.get("value")
+    return node.get("value")
 
 def operator(node):
-    return val.get("operator")
+    return node.get("operator")
 
 def name(node):
-    return val.get("name")
+    return node.get("name")
 
 ###
 # general-purpose accumulators
@@ -190,38 +190,42 @@ def load_nary_predicate(n, symbols):
 def load_boolean_expression(n, symbols):
     assert n.tag == "Boolean_Expression"
     if n.tag == "Binary_Predicate":
-        load_binary_predicate(n, symbols)
+        return load_binary_predicate(n, symbols)
     elif n.tag == "Expression_Comparison":
-        load_expression_comparison(n, symbols)
+        return load_expression_comparison(n, symbols)
     elif n.tag == "Unary_Predicate":
-        load_unary_predicate(n, symbols)
+        return load_unary_predicate(n, symbols)
     elif n.tag == "Nary_Predicate":
-        load_nary_predicate(n, symbols)
-    elif n.tag in { "Quantified_predicat", "Set" }
+        return load_nary_predicate(n, symbols)
+    elif n.tag in {"Quantified_Predicate", "Set"}:
         error("unexpected boolean expression" + n.tag)
+        return None
     else:
         error("unknown boolean expression " + n.tag)
+        return None
 
 def load_expression(n, symbols):
     if n.tag == "Binary_Expression":
-        load_binary_expression(n, symbols)
+        return load_binary_expression(n, symbols)
     elif n.tag == "Nary_Expression":
-        load_nary_expression(n, symbols)
+        return load_nary_expression(n, symbols)
     elif n.tag == "Unary_Expression":
-        load_unary_expression(n, symbols)
+        return load_unary_expression(n, symbols)
     elif n.tag == "Boolean_Litteral":
-        load_boolean_literal(n, symbols)
+        return load_boolean_literal(n, symbols)
     elif n.tag == "Integer_Litteral":
-        load_integer_literal(n, symbols)
+        return load_integer_literal(n, symbols)
     elif n.tag == "Identifier":
-        load_identifier(n, symbols)
+        return load_identifier(n, symbols)
     elif n.tag == "Boolean_Expression":
-        load_boolean_expression(n, symbols)
[…]

### 2013-09-23 translation.tex
Yet another work in progress version of the Python loader of BXML files. Able to load very basic operations, substitutions and expressions.
Profound changes in the text laying out the principles for composing translation of different modules related by import clauses. The specification of the translation has not been updated accordingly and is deprecated.

@@
 \usepackage{xspace,framed,here,keystroke}
+\usepackage{graphicx}
 \pagestyle{headings}
@@
         \B{id} & : & \B{Name} \\
-        \B{see} & : & seq \B{Mach} \\
-        \B{impo} & : & seq \B{Impo} \\
+        \B{impo} & : & seq \B{Module} \\
         \B{const} & : & seq \B{Cons} \\
@@
         \hline
-        \multicolumn{3}{|c|}{Imported machine: \B{Impo}} \\
-        \hline
-        \B{mach} & : & \B{Mach} \\
+        \multicolumn{3}{|c|}{Modules: \B{Module}} \\
+        \hline
+        \B{mach} & : & \B{MachRef} \\
         \B{pre} & : & opt \B{Name} 
@@
         \hline
-        \multicolumn{3}{|c|}{Machine: \B{Mach}} \\
+        \multicolumn{3}{|c|}{Machine: \B{MachRef}} \\
+        \hline
+        \B{spec} & : & \B{MachInterface} \\
+        \B{impl} & : & opt \B{Impl} \\
+        \hline
+        \multicolumn{3}{|c|}{Machine interface: \B{MachInterface}} \\
         \hline
         \B{id} & : & \B{Name} \\
-        \B{impl} & : & \B{Impl} \\
+        \B{const} & : & seq \B{Cons} \\
+        \B{var} & : & seq \B{Vari} \\
+        \B{init} & : & seq \B{Inst} \\
+        \B{op} & : & seq \B{Oper} 
+        \\
         \hline
@@
 for these aspects, but they are not needed in this work and their presentation
-is omitted.  Function definitions are composed of the signature and body. The
-signature contains the return type, name, parameters, and attributes for target
-code generation. Attribute \llvm{nounwind} states that the function never
-generate exceptions. The body is a sequence of blocks of instructions in
-single-static assignment (SSA) form, i.e. are such that each variable may be
-assigned exactly once.
-
-The grammar rule \nt{inst} describes the different kind of
-instructions. Instruction \llvm{alloca} allocates a memory block the size of the
-given type from the stack segment. This and the following classes of
-instructions assign the resulting value to a fresh variable. Arithmetic
-operations are binary, and include signed quotient and reminder. Arithmetic
-comparisons compare two values according to the first operand and return a 1-bit
-integer value.  Instruction \llvm{call} invokes the given function with the
-given arguments and assigns the result to a fresh variable, which is of the
-given type.  In general, \llvm{getelementptr} gets the address of an element in
-an aggregate object through indexing. In the target LLVM subset, aggregate
-objects are structures, the first index has type \llvm{i32} and value \llvm{0}
-to select the first structure at the given location \nt{exp}, and the second
-index selects a field in that structure.  Instruction \llvm{load} assigns to a
-fresh variable \nt{name} the contents of a memory address of type \nt{type}
-specified by \nt{exp} (e.g. in fig.~\ref{fig:ex-llvm}).  Instruction
-\llvm{store} writes a value to memory address (again, see
-fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two forms.  In the first
-form, it directs conditionally the control flow to one of two blocks.  If the
-given expression evaluates to \llvm{1}, control goes to the first block,
-otherwise to the second. In the second form, the control jumps unconditionally
-to the given block.
+is omitted. For each such definition, a memory block is statically allocated to
+stores the variable value. Function definitions are composed of the signature
+and body. The signature contains the return type, name, parameters, and
+attributes for target code generation. Attribute \llvm{nounwind} states that the
+function never generate exceptions. The body is a sequence of blocks of
+instructions in single-static assignment (SSA) form, i.e. are such that each
+variable may be assigned exactly once.
+
+The grammar rule \nt{inst} describes the different kind of instructions. All
+instructions producing a value assign it to a fresh variable.  Instruction
+\llvm{alloca} allocates a memory block the size of the given type on the stack
+segment. This memory is automatically freed when the current frame is popped
+from the stack. The LLVM does not include instructions for allocating from the
+heap. Arithmetic operations are binary, and include signed quotient and
+reminder. Arithmetic comparisons compare two values according to the first
+operand and return a 1-bit integer value.  Instruction \llvm{call} invokes the
+given function with the given arguments and assigns the result to a fresh
+variable, which is of the given type.  In general, \llvm{getelementptr} gets the
+address of an element in an aggregate object through indexing. In the target
+LLVM subset, aggregate objects are structures, the first index has type
+\llvm{i32} and value \llvm{0} to select the first structure at the given
+location \nt{exp}, and the second index selects a field in that structure.
+Instruction \llvm{load} assigns to a fresh variable \nt{name} the contents of a
+memory address of type \nt{type} specified by \nt{exp} (e.g. in
+fig.~\ref{fig:ex-llvm}).  Instruction \llvm{store} writes a value to memory
+address (again, see fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two
+forms.  In the first form, it directs conditionally the control flow to one of
+two blocks.  If the given expression evaluates to \llvm{1}, control goes to the
+first block, otherwise to the second. In the second form, the control jumps
+unconditionally to the given block.
 % Instruction \llvm{switch} directs the
@@
 
-\section{Translation: a Bird's-Eye View}
+\section{Translation: a global picture}
 \label{sec:overview}
 
-The translation unit is the B implementation, say \B{M}, and is mapped to a LLVM
-module, say \llvm{M}. A B implementation \B{M} has several sections, each
-translated as follows:
+In addition to producing LLVM code that is faithful to the design developed in
+B, our code generator is designed to satisfy the following two requirements:
 \begin{enumerate}
-\item Visibility clauses provide access to definitions found in external
-  B components and are translated to external declarations in LLVM.
-\item Constants and value definitions are translated to their value in the
-  LLVM platform.
-\item To encode the implementation state, the translation produces a structure
-  type, named \llvm{\%M\$state\$}, and an instance of that type, named
-  \llvm{@M\$self\$}. All the variables and instances of imported machines are
-  encoded as elements of that structure. These elements are referenced by their
-  position in the structure within the LLVM code. Also, for each operation of
-  each imported machine, an external function declaration is produced.
-\item Each operation \B{op} of the implementation is translated into the
-  definition of a function named \llvm{@M\$op}; also the translation produces a
-  special function \llvm{@M\$init\$} responsible for the initialization.
-
-  The parameter list of the functions thus produced contains one item for each
-  input and output of the corresponding operation, and one item of type
-  \llvm{\%M\$state\$*}, to pass the address of the structure encoding the
-  state of the implementation and its modification by the callee. The return
-  type is always \llvm{void}.
+\item Memory is allocated statically, save for the function call frame stack: no
+  dynamic memory allocation library is required. Many safety critical systems
+  preclude use of this programming facility.
+\item The code generator features separate compilation: an internal change in a
+  component should only require generating new LLVM code for that component, and
+  not for the other components that depend on it. This condition is important
+  for large, distributed projects.
 \end{enumerate}
-The translation is designed to support modular B development. In a B
-implementation, one may reuse existing components, by instantiating (the B term
-is \emph{importing} so-called library machines. Such library machines have their
-own development and implementation. Borrowing terminology from the
-object-oriented world, one may view library machines as classes, and the B
-implementation as an object that contains instances of those classes. In
-practice, different from most OO languages, such instances are not references,
-but the actual objects. We consider this necessary as dynamic memory allocation
-is precluded.
-
-To translate the instantiations of library machines in an implementation, the
-translator needs to have access to the following information on those machines
-and their respective implementation:
-\begin{itemize}
-\item for each library machine \B{M}, the corresponding implementation, denoted
-  \B{M.impl}.
-\item for such implementation, the type of the state and the signature of the
-  operations.
-\end{itemize}
-Note that, when translating the implementation of a library machine, no instance
-shall be generated. It is in the implementation of the main design, that we call
-the \emph{top-level} component, that such instances are created.
-
-Figure~\ref{fig:skel} summarizes the translation principles by providing the
-structure of the LLVM implementation thus produced.
+
+In Atelier-B, software development is organized in libraries of modules. A
+module has a specification, called a machine. A machine may be:
+\begin{description}
+\item[basic] In that case its implementation is carried out outside the realm of
+  the B method. A possible application is a set of related I/O routines. Even
+  though no LLVM code needs to be generated to implement the basic machines, the
+  code generator needs still to produce LLVM declarations of the entities
+  (e.g., constants and operations) that are supplied by those machines.
+\item[developed] In that case, this module also contains a series of
+  refinements, eventually reaching an implementable version, from which LLVM
+  code may be generated.
+\end{description}
+Amongst the several constructs of the B notation supporting modular design, the
+\emph{import} relation is related to separate compilation. At the implementation
+level, one module may import several instances of a machine to construct its
+internal data structures. Such imported machines may be either basic, or
+developed. In the latter case, the implementation of the imported machines may
+in turn import other machine instances and there is no pre-established limit to
+such chain of imports. Hence, the import relation between module instances forms
+a tree, where the root is an implementation and the descendants of a node are
+the instances imported from the module in that node.
 
 \begin{figure}
+\begin{center}
+\includegraphics[width=.8\textwidth]{b-project.pdf}
+\caption{Structure of the imports relation in a B project}
+\label{fig:structure}
+\end{center}
+\end{figure}
+Figure~\ref{fig:structure}
+illustrates this with the structure of a B project where an implementation
+\B{I0} of a specification \B{M0} imports one unnamed instance of a basic machine
+\B{M2} and two instances named \B{A} and \B{B} of a developed machine \B{M1}. In
+turn its implementation \B{I1} itself imports two instances \B{C} and \B{D} of a
+developed machine \B{M3}, implemented as \B{I3} (which has no imports itself).
+The generation of LLVM code for the project shall allocate memory for one
+instance of \B{I0}, two instances of \B{I2} and four instances of \B{I3}.
+
+At that point, let us remark that, when generating code for a given module, we
+need to consider the following two cases:
+\begin{enumerate}
+\item The code generation aims at producing a component, that will further be
+  instantiated by importations. We call this the \emph{COMP} mode for code
+  generation. This case is further split into two sub-cases:
+  \begin{enumerate}
+  \item The module is basic, and only declarations need to be emitted.
+  \item The module is developed, and declarations and definitions are produced.
+  \end{enumerate}
+\item The generated code corresponds to a full project. Given the root module,
+  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
+\end{enumerate}
+One key issue in the code generation is how memory shall be allocated to store
+the state of the modules. To satisfy our first initial requirement (no dynamic
+memory allocation), this allocation shall be static, that is to say the
+generator must include the code that creates the module instances when the
+project starts execution. Note that this cannot be done in \emph{COMP} mode as,
+at that point, we cannot assume any knowledge on the number of instances the
+module being compiled we there be at run time. Consequently, the responsibility
+for the static allocation of instances is assigned to the \emph{PROJ} mode,
+which shall also perform the proper bindings between the actual and formal
+instances in the generated code.
+
+To encode the data space of a module, the translation produces a structure type,
+named \llvm{\%M\$state\$}, such that the elements of the structure represent the
+variables and the instances of imported modules. To satisfy our second initial
+requirement (separate compilation), the instances cannot be part of the
+structure itself, instead references to those instances (encoded as pointers)
+are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
+To encode behavior of a module, each operation \B{op} is translated into the
+definition of a function named \llvm{@M\$op}The parameter list of the functions
+thus produced contains one item for each input and output of the corresponding
+operation. There is also one parameter of type \llvm{\%M\$ref\$}, to pass the
+address of the structure encoding instance associated with that operation. Also
+the translation produces a special function \llvm{@M\$init\$} responsible for
+the initialization. The parameters of this function are the addresses of the
+instances found in the import tree of the module (including the module itself).
+
+Our approach consists in dividing the translation into three sections, and to
+combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
+a given module \B{M}, the sections are the following:
+\begin{description}
+\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
+  the data space of the module. In case the data space of a module is empty,
+  this section is empty and the type is not defined:
+  \begin{tabbing}
+    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
+  \end{tabbing}
+\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data}, if
+  this has been defined. 
+  \begin{tabbing}
+    \llvm{\%M\$ref\$ = type \%M\$state\$*}
+  \end{tabbing}
+  The interface also contains declarations of LLVM
+  functions: one implementing the initialisation of the module, and one for each
+  operation of the module. If the data space of the module is empty, no function
+  definition corresponding to the initialisation is emitted
+  \begin{tabbing}
+    foo \= \kill
+    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
+    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
+  \end{tabbing}
+\item[implementation:] Defines the functions declared in the interface. This
+  section is produced only for the developed modules.
+  \begin{tabbing}
+    foo \= \kill
+    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
+    \> \llvm{\ListOf{\nt{block}}} \\
+    \> \llvm{exit: ret void} \\
+    \llvm{\}}
+  \end{tabbing}
+\end{description}
+Assuming these three sections are available, the \emph{COMP\/} translation
+assembles them as follows:
 \begin{center}
@@
     foo \= foo \= \kill
-      \textit{for each imported component} \B{Q} \textit{generate} \\
-      \> \llvm{\%Q\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
-      \> \llvm{declare void @Q\$init\$(\%Q\$state\$*)} \\
-      \> \textit{for each operation} \B{op} \textit{generate} \\
-      \> \> \llvm{declare void @Q\$op(\%Q\$state\$*, \ListOf{\nt{type}})} \\
-      \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
-      \llvm{define void @M\$init\$(\%M\$state\$* \%self\$) \{} \\
-      \> \llvm{\ListOf{\nt{block}}} \\
-      \> \llvm{exit: ret void} \\
-      \llvm{\}} \\
-      \textit{for each operation} \B{op} \textit{generate} \\
-      \> \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
-      \> \> \llvm{\ListOf{\nt{block}}} \\
+    \textit{for each imported module} \B{Q} \textit{generate} \\
+    \> \textit{generate a type declaration for the state space of} \B{Q} \\
+    \> \llvm{\%Q\$state\$ = type opaque} \\
+    \> \textit{include the interface section of }\B{Q} \\
+    \textit{include the typedef section of} \B{M} \\
+    \textit{include the implementation section of} \B{M}
+  \end{tabbing}
+\end{center}
+For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
+\begin{center}
+  \begin{tabbing}
+    foo \= foo \= \kill
+    \textit{for each transitively imported component} \B{Q} \\
+    \> \textit{include the typedef section of \B{Q}} \\
+    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
+    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
+    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\
+
+    \textit{for each imported component} \B{Q} \\
+    \> \textit{include the interface section of} \B{Q} \\
+    \textit{define a function} \llvm{\%\$init\$} \textit{with a call to the } \\
+    \> \textit{initialization function of} \B{M} \textit{with the proper bindings} \\
+    \> \llvm{define void @\$init\$(void) \{} \\
+    \> \> \llvm{call void @M\$init\$(@M, \{ \ListOf{instances} \}) \{} \\
     \> \> \llvm{exit: ret void} \\
     \> \llvm{\}} \\
-      \textit{if this is the top-level component, generate} \\
-      \> \llvm{@M\$self\$ = common global \%M\$state\$ zeroinitializer}
+    \> 
   \end{tabbing}
 \end{center}
-  \caption{Structure of the LLVM module produced from a B implementation \B{M}.}
-  \label{fig:skel}
-\end{figure}
 
@@
 \label{sec:module}
+
+[[[ THIS SECTION IS DEPRECATED ]]]
 
@@
 
+[[[ DEPRECATED ]]]
+
 Initialization is done is two steps: first all imported modules are

### 2013-09-25 translation.tex
Remodeled the definition of code generation for imports.

@@
 \newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
+\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
+\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
+\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
+\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
+\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
+\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
 \newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
 \newcommand{\state}[0]{\ensuremath{\sf\Theta}}
+\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
 \newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
@@
         \hline
-        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{impo} & : & seq \B{Module} \\
-        \B{const} & : & seq \B{Cons} \\
-        \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} 
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Modules: \B{Module}} \\
-        \hline
-        \B{mach} & : & \B{MachRef} \\
-        \B{pre} & : & opt \B{Name} 
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Machine: \B{MachRef}} \\
-        \hline
-        \B{spec} & : & \B{MachInterface} \\
-        \B{impl} & : & opt \B{Impl} \\
-        \hline
-        \multicolumn{3}{|c|}{Machine interface: \B{MachInterface}} \\
+        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
         \hline
@@
         \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} 
+        \B{op} & : & seq \B{Oper} \\
+        \B{impl} & : & opt \B{Impl} \\
+        \hline
+        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{impo} & : & seq \B{Import} \\
+        \B{const} & : & seq \B{Cons} \\
+        \B{var} & : & seq \B{Vari} \\
+        \B{init} & : & seq \B{Inst} \\
+        \B{op} & : & seq \B{Oper} \\
+        \hline
+        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
+        \hline
+        \B{mach} & : & \B{Machine} \\
+        \B{pre} & : & opt \B{Name} 
         \\
@@
 \item[basic] In that case its implementation is carried out outside the realm of
-  the B method. A possible application is a set of related I/O routines. Even
-  though no LLVM code needs to be generated to implement the basic machines, the
-  code generator needs still to produce LLVM declarations of the entities
-  (e.g., constants and operations) that are supplied by those machines.
+  the B method. Even though no LLVM code needs to be generated to implement the
+  basic machines, the code generator still needs to produce LLVM declarations of
+  the entities (e.g., constants and operations) that are supplied by those
+  machines.
 \item[developed] In that case, this module also contains a series of
@@
 \end{description}
+We distinguish basic machines from developed machines in the abstract syntax by
+testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
+the machine is basic, otherwise it is the root of the corresponding
+implementation.
+
 Amongst the several constructs of the B notation supporting modular design, the
@@
 \end{figure}
-Figure~\ref{fig:structure}
-illustrates this with the structure of a B project where an implementation
-\B{I0} of a specification \B{M0} imports one unnamed instance of a basic machine
-\B{M2} and two instances named \B{A} and \B{B} of a developed machine \B{M1}. In
-turn its implementation \B{I1} itself imports two instances \B{C} and \B{D} of a
-developed machine \B{M3}, implemented as \B{I3} (which has no imports itself).
-The generation of LLVM code for the project shall allocate memory for one
-instance of \B{I0}, two instances of \B{I2} and four instances of \B{I3}.
+Figure~\ref{fig:structure} illustrates this with the structure of a B project
+where an implementation \B{I0} of a specification \B{M0} imports one unnamed
+instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
+developed machine \B{M2}. Its implementation \B{I2} in turn imports two
+instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
+(which has no imports itself).  The generation of LLVM code for the project
+shall produce the representation for one instance of \B{I0}, two instances of
+\B{I2} and four instances of \B{I3}.
 
@@
 at that point, we cannot assume any knowledge on the number of instances the
-module being compiled we there be at run time. Consequently, the responsibility
-for the static allocation of instances is assigned to the \emph{PROJ} mode,
-which shall also perform the proper bindings between the actual and formal
-instances in the generated code.
+module being compiled will there be at run time. Consequently, the
+responsibility for the static allocation of instances is assigned to the
+\emph{PROJ} mode, which shall also perform the proper bindings between the
+actual and formal instances in the generated code.
 
@@
 To encode behavior of a module, each operation \B{op} is translated into the
-definition of a function named \llvm{@M\$op}The parameter list of the functions
-thus produced contains one item for each input and output of the corresponding
-operation. There is also one parameter of type \llvm{\%M\$ref\$}, to pass the
-address of the structure encoding instance associated with that operation. Also
-the translation produces a special function \llvm{@M\$init\$} responsible for
-the initialization. The parameters of this function are the addresses of the
-instances found in the import tree of the module (including the module itself).
+definition of a function named \llvm{@M\$op}. The parameter list of the
+functions thus produced contains one item for each input and output of the
+corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
+to pass the address of the structure encoding instance associated with that
+operation. Also the translation produces a special function \llvm{@M\$init\$}
+responsible for executing the initialization of \B{M}. The parameters of this
+function are the addresses of the instances found in the import tree of the
+module (including the module itself).
 
@@
 \item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
-  the data space of the module. In case the data space of a module is empty,
-  this section is empty and the type is not defined:
+  the data space of the module:
   \begin{tabbing}
@@
   \end{tabbing}
-\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data}, if
-  this has been defined. 
+  In case the data space of a module is empty, this section is empty and the
+  type is not defined.
+\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
+  defined):
   \begin{tabbing}
@@
   operation of the module. If the data space of the module is empty, no function
-  definition corresponding to the initialisation is emitted
+  definition corresponding to the initialisation is emitted:
   \begin{tabbing}
@@
   \end{tabbing}
-\item[implementation:] Defines the functions declared in the interface. This
-  section is produced only for the developed modules.
+\item[implementation:] Defines the functions declared in the interface:
   \begin{tabbing}
@@
   \end{tabbing}
+  This section is produced only for the developed modules.
 \end{description}
-Assuming these three sections are available, the \emph{COMP\/} translation
+Based on the definition of these sections, the \emph{COMP\/} translation
 assembles them as follows:
@@
     \textit{for each imported module} \B{Q} \textit{generate} \\
-    \> \textit{generate a type declaration for the state space of} \B{Q} \\
-    \> \llvm{\%Q\$state\$ = type opaque} \\
+    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
     \> \textit{include the interface section of }\B{Q} \\
@@
 \begin{description}
+\item[Notation] and some auxiliary definitions are presented in sec.~\ref{sec:notation}.
 \item[Modularity] constructs to combine B components are described in
@@
 
-\paragraph{Notations.} Code generation is defined with a set of rules defined by
+\section{Notations} 
+\label{sec:notation}
+
+Code generation is defined with a set of rules defined by
 structural recursion. For each grammatical construct, a code generation function
@@
 name of the component where it is defined and its own name.  Also, we define
-three auxiliary functions responsible to produce such identifiers:
+additional auxiliary functions responsible for producing such identifiers:
 \begin{align*}
 \state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
+\stateref(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$ref\$"} & \text{ (the type pointer to state)} \\
 \init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
@@
 We assume an unlimited supply of labels and local variables: we use $\lbl$ to
-return a fresh LLVM label and $\variable$ for a fresh local variable name.
-
-Finally, we need a function that, given a concrete variable, or an imported
-machine, returns the position of the corresponding element in the structure
-representing the state of the implementation. The position of the elements in
-the structure is the same as the position of corresponding B construct in the
-source code:
-\begin{align*}
-  \idx(\B{n}) & \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
-\B{n.root.vari})
-\end{align*}
-where $\mathit{index}(e, s)$ is a function that returns the position of $e$ in
-sequence $s$ and $\conc$ is the concatenation operator for sequences.
+return a fresh LLVM label and $\variable$ for a fresh local variable name. Also, let us define:
+\begin{itemize}
+\item $\developed(\B{n}) \isdef \B{n.impl} \neq \bot$ to test if a machine is
+  developed.
+\item $\stateless(\B{n}) \isdef \B{n.vars} =  \bot \land \forall i \in \B{n.impo}, \stateless(i)$ to test if a machine is stateless.
+\item $\importedmodules(\B{n}) \isdef \{ \B{i.mach} \mid \B{i} \in \B{n.impo}
+  \}$ to get the set of machines directly imported from implementation $\B{n}$.
+\item $\importedinstances(\B{n}) \isdef \B{n.impo}$ to get the set of machines
+  instances directly imported from implementation $\B{n}$.
+\item $\trimportedmodules(\B{n}) \isdef \ldots$ to get the set of machines
+transitively imported from implementation $\B{n}$.
+\item $\trimportedinstances(\B{n}) \isdef \ldots$ to get the set of machines
+instances transitively imported from implementation $\B{n}$.
+\item $\idx(\B{n}) \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
+  \B{n.root.vari})$ to get, from a given concrete variable $\B{n}$, the position
+  of the corresponding element in the structure representing the state of the
+  implementation. Here where $\mathit{index}(e, s)$ is a function that returns
+  the position of $e$ in sequence $s$ and $\conc$ is the concatenation operator
+  for sequences.
+\end{itemize}
 
@@
 
-[[[ THIS SECTION IS DEPRECATED ]]]
-
-The translation of a B implementation to LLVM, following the template sketched
-in Figure~\ref{fig:skel}, is specified by the following rule, where \B{n} is the
-root of the implementation representation and $\top^?$ a boolean indicating if
-\B{n} is the implementation of a top-level component or of a library machine:
-\begin{align*}
-  \lefteqn{\trad{\B{n}, $\top^?$}{\B{Impl}} \isdef } \\
-  & \llvm{"\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{State}}}_{def}}} & \text{defines the type of imported modules} \\
-  & \llvm{\state(\B{n}) = type \PH{\trad{n}{State}_{expr}}} & \text{defines the type of current module} \\
-  & \llvm{\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{Bhv}}}_{decl}}} & \text{declares of imported modules operations} \\
-  & \llvm{\PH{\trad{\B{n.init}}{\B{Init}}}} & \text{defines the initialization function} \\
-  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}}} & \text{defines functions implementing operations}\\
-  & \llvm{\PH{\trad{\B{n}, $\top^?$}{\B{Inst}}}"} & \text{creates instance (conditional)}
-\end{align*}
-
-To handle the importation of a machine, one needs to produce the definition of
-the type for the state of its implementation. Since this implementation may reference
-other modules, the definition is recursive (parameter $A$ is an accumulator to avoid
-duplicating such definitions):
-\begin{align*}
-  \trad{\B{n}, $A$}{\B{State}}_{def} \isdef 
-  & \IF \B{n} \not\in A \THEN \\
-  & \quad \LET \B{m} = \B{n.mach} \IN \\
-  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
-  & \quad \quad \quad \llvm{"\PH{\trad{\B{i.impo}, $A \cup \{ \B{n} \}$}{\ListOf{\B{State}}}_{def}}} \\
-  & \quad \quad \quad \llvm{\PH{\state(\B{m})} = type \PH{\trad{i}{State}_{expr}} \nl"}
-\end{align*}
-
-The following rules specify how the type representing the state of an
-implementation is generated. It is a structure type, with one element per
-imported machine, and one element per variable. The definition of such type is
-just the enumeration of the type of its elements (the individual elements are
-referenced, not by name, but by position inside the structure):
-\begin{align*}
-  \trad{\B{n}}{\B{State}}_{expr} \isdef & \llvm{"\{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{expr}},  \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{expr}} \}"} \\
-  \trad{\B{n}}{\B{Impo}}_{expr} \isdef & \state(\B{n.mach}) \\
-  \trad{\B{n}}{\B{Vari}}_{expr} \isdef & \trad{\B{n.type}}{\B{Type}}
-\end{align*}
-
-The code generation for the functions implementing the initialization and
-operations of imported machine is specified by the following rules. Not that
-this definition needs not be recursive, as only the operations of a directly
-imported machine may be called in an implementation. Again, parameter $A$ plays
-the role of an accumulator to avoid (harmless, but useless) duplicate
-declarations.
-\begin{align*}
-  \trad{\B{n}, $A$}{\B{Bhv}}_{decl} \isdef 
-  & \IF \B{n} \not\in A \THEN \\
-  & \quad \LET \B{m} = \B{n.mach} \IN \\
-  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
-  & \quad \quad \quad \llvm{declare void \PH{\init(\B{m})}(\PH{\state(\B{m})} *) \nl} \\
-  & \quad \quad \quad \llvm{\PH{\trad{\B{m.op}}{\ListOf{\B{Oper}}}_{decl}}"} \\
-  \trad{\B{n}}{\B{Oper}}_{decl} \isdef &
-  \llvm{"declare void \PH{\Global(\B{n})}(\PH{\state(\B{n.root})}, \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}}, \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}})"}\\
-  \trad{\B{n}}{\B{inp}}_{decl} \isdef & \trad{\B{n.type}}{\B{Type}} \\
-  \trad{\B{n}}{\B{out}}_{decl} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
-\end{align*}
-
-The last rule is responsible for producing the code creating the instance when the
-implementation is that of the top-level component of a design:
-\begin{align*}
-\lefteqn{  \trad{\B{n}, $\top^?$}{\B{Inst}} \isdef } \\
-  & \IF \top^? \THEN \llvm{"\PH{\self(\B{n})} = common global \PH{\state(\B{n})} zeroinitializer \nl"} \\
-  & \ELSE \llvm{""}
-\end{align*}
+\subsection{The typedef section}
+
+This section corresponds to the LLVM type definition for the state space of a
+given module. If the module is developed, then the state space is that of the
+corresponding implementation, otherwise the module is basic. If the basic
+machine is stateless, the state space is empty; otherwise, it is represented by
+an aggregate type, the elements of which correspond to the variables of the
+machine:
+\begin{align*}
+  \trad{\B{n}}{\B{Mach}}_{\text{typedef}} \isdef 
+  & \IF \developed(\B{n}) \THEN \trad{\B{n.impl}, \B{n}}{\B{Impl}}_{\text{typedef}} \\
+  & \ELSE \IF \stateless(\B{n}) \THEN \llvm{""} \\
+  & \ELSE \llvm{"\PH{\state(\B{n})} = type \{ \PH{\trad{\B{n.vars}}{\ListOf{\B{Vari}}}_{\text{typedef}}} \} "}
+\end{align*}
+
+For the case of a developed machine \B{m} with implementation \B{n}, if \B{n} is
+stateless, no type definition is emitted; otherwise, it is an aggregate type
+composed of one element for each import and variable:
+\begin{align*}
+  \trad{\B{n}, \B{m}}{\B{Impl}}_{\text{typedef}} \isdef 
+  & \IF \stateless(\B{n}) \THEN \llvm{""} \\
+  & \ELSE \llvm{"\PH{\state(\B{m})} = type \{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{\text{typedef}}}, \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{\text{typedef}}} \} "}
+\end{align*}
+
+In these aggregate types, the elements corresponding to an import \B{n} is
+defined in the following rule. If the imported machine is stateless, there is no
+element, otherwise it is the name of the type pointer to the state of the
+machine:
+\begin{align*}
+  \trad{\B{n}}{\B{Impo}}_{\text{typedef}} \isdef 
+  & \llvm{""} \triangleleft \stateless(\B{n.mach}) \triangleright \stateref(\B{n.mach})
+\end{align*}
+
+The elements corresponding to concrete variables are the LLVM data types
+corresponding to the B types of those variables (defined in
+sec.~\ref{sec:data}):
+\begin{align*}
+  \trad{\B{n}}{\B{Impl}}_{\text{typedef}} \isdef 
+  & \trad{\B{n.type}}{\B{Type}}
+\end{align*}
+
+\subsection{The interface section}
+
+The interface section of a module \B{n} defines a type $\stateref(\B{n})$,
+pointer to $\state(\B{n})$, the type to represent the data space of \B{n}, and
+declares functions for executing 1) executing the initialization of an instance
+of \B{n} and 2) each operation in \B{n}:
+\begin{align*}
+  \trad{\B{n}}{\B{Mach}}_{\text{interface}} \isdef 
+  & \llvm{"\PH{\stateref(\B{n})} = type \PH{\state(\B{n})}*\nl} \\
+  & \llvm{\PH{\trad{\B{n}}{\B{init}}_{\text{interface}}}}\\
+  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}_{\text{interface}}}"}
+\end{align*}
+
+The function responsible for initializing an instance of \B{n} takes as
+parameter a reference to such instance, as well as a reference for each
+transitively imported stateful instance and is as follows:
+\begin{align*}
+  \trad{\B{n}}{\B{init}}_{\text{interface}} \isdef 
+  & \llvm{"declare void \PH{\init(\B{n})} ( \PH{\stateref(\B{n})}, 
+    \PH{\trad{\trimportedinstances(\B{n})}{\ListOf{\B{inst}}}_{\text{decl}}} )\nl"} \\
+  \trad{\B{n}}{\B{inst}}_{\text{decl}} \isdef 
+  & \bot \triangleleft \stateless(\B{n.mach}) \triangleright \stateref(\B{n.mach}) 
+\end{align*}
+
+The function responsible for executing an operation \B{n} has as parameter a
+reference to the instance of the corresponding module, if it is stateful, and
+one parameter for each input and output of the corresponding operation:
+\begin{align*}
+  \trad{\B{n}}{\B{oper}}_{\text{interface}} \isdef 
+  & 
+\LET t = \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
+& \llvm{"declare void \PH{\Global(\B{n})} 
+( \PH{t}, 
+  \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}_{\text{decl}}}, 
+  \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}_{\text{decl}}}) \nl"} \\
+  \trad{\B{n}}{\B{inp}}_{\text{decl}} \isdef & \trad{\B{n.type}}{\B{Type}} \\
+  \trad{\B{n}}{\B{out}}_{\text{decl}} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
+\end{align*}
+It is noteworthy that the outputs of B operations are encoded as pointer
+parameters.
+\subsection{The implementation section}
+
+The implementation section is composed of the definitions for each function
+in the corresponding interface:
+\begin{align*}
+  \trad{\B{n}}{\B{Mach}}_{\text{implementation}} \isdef 
+  & \LET \B{i} = \B{n.impl} = \bot \IN\\
+  & \quad \IF \B{i} = \bot \THEN \llvm{""} \\
+  & \quad \ELSE \\
+  & \quad \quad \llvm{"\PH{\trad{\B{i}}{\B{init}}}} \\
+  & \quad \quad \llvm{\PH{\trad{\B{i.op}}{\ListOf{\B{Oper}}}}"}
+\end{align*}
+Code generation for operations and the initialization is described in
+sec.~\ref{sec:control}.
 
@@
 
-\paragraph{Arithmetic operations}\DD{Actually some arithmetic operations are not binary and different translation rules shall be written.} have a similar 
+\paragraph{Arithmetic operations} have a similar 
 structure and interpretation as comparisons. Here is the rule for binary
@@
 \end{align*}
-
-Next is the translation of the unary \B{succ} operator:
+Non-binary applications of associative operators are handled as nested
+applications of binary applications. Next is the translation of the unary
+\B{succ} operator:
 \begin{align*}

### 2013-10-16 translation.tex
Many conceptual and implementation changes to implement support for imports. The prototype is not yet functional.

@@
 corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
-to pass the address of the structure encoding instance associated with that
+to pass the address of the instance encoding structure associated with that
 operation. Also the translation produces a special function \llvm{@M\$init\$}
@@
 a given module \B{M}, the sections are the following:
-\begin{description}
-\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
+\begin{enumerate}
+\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
   the data space of the module:
@@
   type is not defined.
-\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
+\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
   defined):
@@
   \end{tabbing}
-  The interface also contains declarations of LLVM
-  functions: one implementing the initialisation of the module, and one for each
-  operation of the module. If the data space of the module is empty, no function
-  definition corresponding to the initialisation is emitted:
+  The interface also contains declarations of one LLVM functions for each
+  operation of the module. 
   \begin{tabbing}
-    foo \= \kill
-    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
+    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
+  \end{tabbing}
+  If the data space of the module is not empty, a function definition
+  corresponding to the initialisation is also emitted:
+  \begin{tabbing}
     \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
   \end{tabbing}
-\item[implementation:] Defines the functions declared in the interface:
+\item implementation: Defines the functions declared in the interface:
   \begin{tabbing}

### 2013-10-16 PROTO/b2llvm.py

@@ -2,15 +2,28 @@
 import argparse
 from b2llvm.translate import translate_bxml
 
-progdescription='Generates LLVM code for a B implementation. The code contains a type declaration, corresponding to the implementation state, and function declarations corresponding to the initialization as well as the operations. Optionally, the generator includes a declaration of a global variable holding the machine state (option --top).'
+progdescription='Generates LLVM code for a B module. In component mode, the generator produces type and function definitions corresponding to the state and operations of a given developed machine. In project mode, the generator instantiates the given module and all the instances of imported modules, and defines a function to link these modules and initialize their execution.'
 parser = argparse.ArgumentParser(description=progdescription)
-parser.add_argument('bxml_file', help='input BXML file name')
-parser.add_argument('llvm_file', help='output LLVM file name')
-parser.add_argument('-t', '--top', action='store_true', help='LLVM shall construct an instance of the component (do not set to generate code for library machines)')
+parser.add_argument('b_module', 
+                    help='name of the B module (machine)')
+parser.add_argument('llvm_file', 
+                    help='output LLVM file name')
+parser.add_argument('directory', 
+                    help='when set, the program will lookup files in that directory')
+parser.add_argument('settings', 
+                    help='project settings file')
+parser.add_argument('-m', '--mode', choices=['comp','proj'], default='comp', 
+                    help='Selects code generation mode.')
 
 args = parser.parse_args()
-translate_bxml(args.bxml, args.llvm, args.top)
-print("b2llvm code generation completed")
-print("- input: " + args.bxml_file)
-print("- output: " + args.llvm_file)
-print("- mode: " + ("top-level" if args.top else "library machine"))
+if args.comp and args.proj:
+    print("error: choose only one code generation mode.")
+elif not args.comp and not args.proj:
+    print("error: choose one code generation mode")
+else:
+    translate_bxml(b_module, args.llvm_file, mode=args.mode, 
+                   dirname=args.directory, settings=args.settings)
+    print("b2llvm code generation completed")
+    print("- input: " + args.bxml_file)
+    print("- output: " + args.llvm_file)
+    print("- mode: " + ("top-level" if args.top else "library machine"))

### 2013-10-16 PROTO/b2llvm/ast.py

@@ -159,13 +159,13 @@ def make_oper(id, inp, out, body):
 def make_import(mach, prefix = None):
     '''
     - Input: 
-    mach: a machine
+    mach: a machine or a library machine
     prefix: a string prefix (optional).
     - Output:
-    Creates an element of an import clause.
+    Creates an element of an import clause, i.e., a module.
     '''
-    assert(mach["kind"] in { "Mach" })
-    return { "kind": "Impo", "mach": mach, "pre": prefix }
+    assert(mach["kind"] in { "Machine" })
+    return { "kind": "Module", "mach": mach, "pre": prefix }
 
 def make_implementation(id, imports, consts, vars, init, ops):
     assert type(imports) is list
@@ -173,16 +173,48 @@ def make_implementation(id, imports, consts, vars, init, ops):
     assert type(vars) is list
     assert type(init) is list
     assert type(ops) is list
-    root = { "kind": "Impl", "id": id,
+    root = { "kind": "Impl", 
+             "id": id,
+             "machine": None,
              "imports": imports,
              "concrete_constants": consts,
-             "concrete_variables": vars, 
-             "initialisation": init, "operations": ops }
+             "variables": vars, 
+             "initialisation": init, "operations": ops,
+             "stateful": None}
     for node in imports + vars + ops:
         node["root"] = root
     return root
 
-def make_machine(id, impl):
-    root = { "kind": "Mach", "id": id, "impl": impl }
-    return root
+def make_base_machine(id, consts, vars, ops):
+    '''
+    Constructor for node that represent a machine interface, namely
+    the elements of a machine that may be accessed by a module depending
+    on that machine.
+    '''
+    assert type(consts) is list
+    assert type(vars) is list
+    assert type(ops) is list
+    return { "kind": "Machine", 
+             "concrete_constants": consts,
+             "variables": vars,
+             "operations": ops,
+             "implementation": None,
+             "stateful": None,
+             "comp_direct": None,
+             "comp_indirect": None}
+
+def make_developed_machine(id, impl):
+    '''
+    Constructor for node that represents a developed machine.
+
+    For now we are just interested in the implementation of that machine.
+    '''
+    return { "kind": "Machine", 
+             "id": id,
+             "variables": [],
+             "implementation": impl,
+             "stateful": None,
+             "comp_direct": None,
+             "comp_indirect": None
+             }
 

### 2013-10-16 PROTO/b2llvm/bproject.py

@@ -1,0 +1,58 @@
+from b2llvm.strutils import list_of
+import xml.etree.ElementTree as ET
+
+class BProject:
+    '''
+    Class representing B projects settings.
+    '''
+    def __init__(self, name):
+        '''
+        Constructor, takes as input the path of a bproject.xml file.
+        '''
+        root = ET.parse(name).getroot()
+        implements = root.findall("./developed")
+        self.implement = { e.find("./machine").text:e.find("./implementation").text 
+                           for e in implements }
+        foreign = root.findall("./base")
+        self.developed = self.implement.keys()
+        self.base = { e.find("./machine").text for e in foreign }
+
+    def is_developed(self, m):
+        ''' Tests if a machine is a developed machine.
+        
+        Inputs:
+        - m: a machine name
+        Returns:
+        True iff m is declared as developed.
+        '''
+        return m in self.developed
+
+    def is_base(self, m):
+        ''' Tests if a machine is a base machine.
+
+        Inputs:
+        - m: a machine name
+        Returns:
+        True iff m is declared as base.
+        '''
+        return m in self.base
+
+    def has(self, m):
+        ''' Tests if a machine belongs to a B project
+
+        Inputs:
+        - m: a machine name
+        Returns:
+        True if m is a developed or a base machine declared in project, 
+        False otherwise.
+        '''
+        return self.is_developed(m) or self.is_base(m)
+
+    def implementation(self, m):
+        assert self.is_developed(m)
+        return self.implement[m]
+
+    def __str__(self):
+        return ("developed = {" + list_of(self.developed) + "}\n" + 
+                "implement = {" + list_of([ m + ":" + i for m,i in self.implement.items()]) + "}\n" +
+                "base = {" + list_of(self.base) + "}\n")

### 2013-10-16 PROTO/b2llvm/cache.py

@@ -1,0 +1,16 @@
+class Cache:
+    '''
+    This class encapsulates a map with three operations: has,
+    get and set. It is just a wrapper around dict operations.
+    '''
+    def __init__(self):
+        self.store = dict()
+
+    def has(self, n) :
+        return n in self.store.keys()
+
+    def get(self, n) :
+        return self.store[n]
+
+    def set(self, n, v) :
+        self.store[n] = v

### 2013-10-16 PROTO/b2llvm/loadbxml.py

@@ -22,6 +22,14 @@
 # the input XML. Nevertheless, the inclusion in the symbol table is done
 # through function sym_table_add, which checks there is no such conflict.
 #
+# The elements stored into the symbol table are:
+#   - concrete variables
+#   - concrete constants/values
+#   - imports: name of import is mapped to ast import node
+#   - operations: name of operation is mapped to ast operation node
+#   - operation parameters
+#   - local variables
+#
 # When entering a new scope (e.g. an operation, or a VAR...IN substitution),
 # a copy of the symbol table is created, and the local symbols are added to
 # that copy.
@@ -34,142 +42,497 @@
 # For instance a + b + c would be represented as ((a + b) + c).
 #
 
+import os
 import xml.etree.ElementTree as ET
-import ast
+import b2llvm.ast as ast
+import b2llvm.cache as cache
+from b2llvm.bproject import BProject
 
 ###
 #
-# rudimentary error handling
+# global variables
 #
 ###
+loading = set()
 
-error_nb = 0
-def error(message):
+#
+# MODULE ENTRY POINTS
+#
+
+def load_project(dirname='bxml', filename='project.xml'):
     '''
-    Input:
-      - a string
-    Output: none
-    Prints the input string to stdoud and increments error_nb
+    Keywords:
+      - dirname: path to directory where bxml files are stored;
+      default value is 'bxml'.
+      - filename: the name of a bproject.bxml file. The default
+      value is 'bproject.bxml'. 
+    Loads the contents of the given file, if it exists, into global variable
+    project_db. If the file does not exist, then project_db is left empty, 
+    i.e. a pair formed by an empty dictionary and an empty set.
     '''
-    global error_nb
-    error_nb += 1
-    print("error: " + message)
+    return BProject(dirname+os.sep+filename)
 
-warn_nb = 0
-def warn(message):
-    '''
-    Input:
-      - a string
-    Output: none
-    Prints the input string to stdout and increments warn_nb
-    '''
-    global warn_nb
-    warn_nb += 1
-    print("warning: " + message)
-
-def display_report(id, filename):
-    global error_nb, warn_nb
-    print("B implementation " + id + " from file " + filename + " loaded.")
-    print(str(error_nb)+" error(s), "+str(warn_nb)+" warning(s) reported.")
-
-###
-#
-# symbol table stuff
-#
-###
-def sym_table_add(table, id, pyt):
-    if id in table:
-        error("name clash ("+id+")")
-    table[id] = pyt
-
-def sym_table_new():
-    table = dict()
[…]

### 2013-10-16 PROTO/b2llvm/names.py

@@ -1,0 +1,45 @@
+###
+#
+# b2llvm.names 
+#
+# Module for generating LLVM identifiers.
+#
+###
+
+llvm_local_var_counter = 0
+llvm_label_counter = 0
+
+def new_local():
+    '''
+    Input: None
+    Output: A string representing a LLVM local variable. This string
+    is composed by the prefix "%" followed by a number. Function
+    reset_llvm_names is responsible for zeroing that number.
+    '''
+    global llvm_local_var_counter
+    result = "%"+str(llvm_local_var_counter)
+    llvm_local_var_counter += 1
+    return result
+
+def new_label():
+    '''
+    Input: None
+    Output: A string representing a LLVM instruction label. This string
+    is composed by the prefix "label" followed by a number. Function
+    reset_llvm_names is responsible for zeroing that number.
+    '''
+    global llvm_label_counter
+    result = "label"+str(llvm_label_counter)
+    llvm_label_counter += 1
+    return result
+
+def reset():
+    '''
+    Output: None
+    The role of this function is to zero the counters used to build
+    label and local variable identifiers.
+    '''
+    global llvm_local_var_counter, llvm_label_counter
+    llvm_local_var_counter = 0
+    llvm_label_counter = 0
+

### 2013-10-16 PROTO/b2llvm/printer.py

@@ -12,9 +12,7 @@
 # - Some elements of the B representation are not present in the Python representation and
 # are therefore not output.
 #
-sp = " "
-tb = "  "
-nl = "\n"
+from b2llvm.strutils import commas, sp, tb, nl
 
 def term(n):
     '''
@@ -248,20 +246,19 @@ def oper(n):
 
 def implementation(n):
     global sp, tb, nl
-    imp, consts = n["imports"], n["concrete_constants"]
-    vars = n["concrete_variables"]
+    imp, consts, vars = n["imports"], n["concrete_constants"], n["variables"]
     init, ops = n["initialisation"], n["operations"]
     result = ""
     result += "IMPLEMENTATION" + sp + n["id"] + nl
     if imp != []:
         result += "IMPORTS" + nl
-        result += tb + ",".join([imports(e) for e in imp]) + nl
+        result += tb + commas([imports(e) for e in imp]) + nl
     if consts != []:
         result += "VALUES" + nl
-        result += tb + ",".join([value(e) for e in consts]) + nl
+        result += tb + commas([value(e) for e in consts]) + nl
     if vars != []:
         result += "VARIABLES" + nl
-        result += tb + (","+sp).join([e["id"] for e in vars]) + nl
+        result += tb + commas([e["id"] for e in vars]) + nl
         result += "INVARIANT" + nl
         result += tb + (sp+"&"+nl+tb).join([invar(e) for e in vars]) + nl
     if init != []:

### 2013-10-16 PROTO/b2llvm/strutils.py

@@ -1,0 +1,26 @@
+###
+#
+# b2llvm.strutils
+#
+# Module for string utilities
+#
+###
+
+
+def commas(li):
+    '''
+    Returns a comma-separated string of the given strings.
+    '''
+    return ", ".join(li)
+
+def nconc(li):
+    '''
+    Returns a comma-separated string of the given strings.
+    '''
+    return ", ".join(li)
+
+nl = "\n"
+sp = " "
+tb = "  "
+tb2 = tb*2
+

### 2013-10-16 PROTO/b2llvm/translate.py

@@ -8,48 +8,640 @@
 #
 ###
 
-import ast
+import b2llvm.ast as ast
+import b2llvm.loadbxml as loadbxml 
+import b2llvm.names as names
+from b2llvm.strutils import commas, nconc, sp, nl, tb, tb2
+from b2llvm.bproject import BProject
+
+#
+# Main entry point for this module
+#
+
+def translate_bxml(bmodule, outfile, mode='comp', dir='bxml', settings='project.xml'):
+    '''
+    Main function for applying the code generator to a B module
+
+    Keywords:
+      - mode: the code generation mode
+      - bmodule: the name of the B machine to have code generated for
+      - outfile: the name of the file where code shall be output
+      - dir: the name of the directory where files shall be read (default is 'bxml')
+      - settings: the name of the file in dir where the project settings
+      are stored (default: project.xml)
+    Result:
+      None
+    '''
+    project = loadbxml.load_project(dirname=dir, filename=settings)
+    ast = loadbxml.load_module(dir, project, bmodule)
+    if mode == 'comp':
+        filecontents = translate_mode_comp(ast)
+    else:
+        filecontents = translate_mode_proj(ast)
+    llvm = open(outfile, 'w')
+    llvm.write(";;; -*- mode: asm -*-"+nl) # emacs syntax highlight on
+    llvm.write(";;; file name: "+outfile+nl)
+    llvm.write(";;; B module: "+bmodule+nl)
+    llvm.write(";;; generation mode: "+mode+nl)
+    llvm.write(";;; generated with b2llvm."+nl)
+    llvm.write(filecontents)
+    llvm.close()
+
+#
+# TOP-LEVEL FUNCTION FOR EACH TRANSLATION MODE
+#
+
+def translate_mode_comp(m):
+    '''
+    Translation in component mode.
+
+    Inputs:
+      - c: cache
+      - n: root AST node for a B machine in proj
+    Output:
+      LLVM text corresponding to the implementation of n. The produced
+      LLVM module is a component to be linked against in the compilation
+      of a full project.
+    '''
+    check_kind(m, {"Machine"})
+    if is_base(m):
+        return section_typedef(m)
+    else:
+        assert is_developed(m)
+        res = str()
+        i = implementation(m)
+        for q in comp_indirect(m):
+            res += state_opaque_typedef(q.mach)
+            res += state_ref_typedef(q.mach)
+        for q in comp_direct(m):
[…]

### 2013-10-22 translation.tex
Compilation of translation.tex
* Put the required B-project.pdf (used convert B-project.tiff B-project.pdf)
* Fixed a very minor bug (description -> enumerate) in the tex file
Several bug fixes, including in the specification of the translation for operation calls. The translator now produces LLVM code for counter and wd, the latter importing the former. The produced LLVM code compiles without error with "llvm-as-mp-3.1".

@@
   This section is produced only for the developed modules.
-\end{description}
+\end{enumerate}
 Based on the definition of these sections, the \emph{COMP\/} translation
@@
   \trad{c.out}{\ListOf{Out}} \AND \\
-  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
+  & \quad id = {\Global(\B{c})} \AND ty = \llvm{"\PH{\stateref(\B{c.root})}} \AND \\
   & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
   & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
-  & \quad \quad \ELSE \LET v = \variable \IN \\
-  & \quad \quad \quad v \sep \\
-  & \quad \quad \quad \llvm{"\PH{v} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
+  & \quad \quad \ELSE \LET v_1, v_2 = \variable, \variable \IN \\
+  & \quad \quad \quad v_2 \sep \\
+  & \quad \quad \quad \llvm{"\PH{v_1} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl} \\
+  & \quad \quad \quad \quad \llvm{\PH{v_2} = load \PH{ty}* \PH{v_1} \nl"} \\
   & \IN \\
@@
   & \quad \llvm{\PH{p}} \\
-  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
+  & \quad \llvm{call void \PH{id}(\PH{ty} \PH{val}, \PH{il}, \PH{ol}) \nl"}
 \end{align*}

### 2013-10-22 PROTO/b2llvm/translate.py

@@ -64,7 +64,6 @@ def translate_mode_comp(m):
       of a full project.
     '''
     check_kind(m, {"Machine"})
-    print("translating machine "+m["id"])
     if is_base(m):
         return section_typedef(m)
     else:
@@ -128,10 +127,7 @@ def section_interface(m):
     Text of LLVM declarations (see section interface in translation definition).
     '''
     check_kind(m, {"Machine"})
-    print("building section interface of "+m["id"])
     res = str()
-    if is_stateful(m):
-        res += state_ref_typedef(m)
     res += section_interface_init(m)
     res += nconc([ section_interface_op(m, op) for op in operations(m) ])
     return res
@@ -204,7 +200,6 @@ def section_typedef(m):
     '''
     global nl
     check_kind(m, {"Machine"})
-    print("building section typede of "+m["id"])
     if is_developed(m):
         return section_typedef_impl(implementation(m), m)
     else:
@@ -250,7 +245,6 @@ def section_implementation(m):
     project proj.    
     '''
     check_kind(m, {"Machine"})
-    print("building section implementation of "+m["id"])
     res = str()
     if is_developed(m):
         i = implementation(m)
@@ -327,10 +321,10 @@ def x_init(m, i):
         mq = q.mach     # the imported machine
         arg_list2 = []  # to store parameters needed to initialize mq
         if is_stateful(mq):
-            arg_list += state_r_name(mq)+sp+lexicon[q]
+            arg_list2.append(state_r_name(mq)+sp+lexicon[q])
         n = len([x for x in comp_indirect(mq) if is_stateful(x.mach)])
         arg_list2.extend(arg_list[offset:offset+n])
-        res += tb+"call void "+init_name(mq)+"("+commas(arg_list)+")"+nl
+        res += tb+"call void "+init_name(mq)+"("+commas(arg_list2)+")"+nl
     # 2.4 translate initialisation instructions
     res += x_inst_list_label(i["initialisation"], "exit")
     res += "exit:"+nl
@@ -615,7 +609,6 @@ def x_lvalue(n):
 def x_call(n):
     global sp
     check_kind(n, {"Call"})
-    print("x_call, n.id="+n["op"]["id"])
     res = str()
     # pi, po: evaluate arguments - il, ol: get parameters types and names
     pi, il = x_inputs(n["inp"])
@@ -637,15 +630,17 @@ def x_call(n):
             v = "%self$"
         else:
             t = state_r_name(n["inst"]["mach"])
[…]

### 2013-10-24 translation.tex
Fixed root Python script. Fixed code generation so that LLVM code units generated for B multi-machines projects are linkable.

@@
 \begin{center}
-\includegraphics[width=.8\textwidth]{b-project.pdf}
+ \includegraphics[width=.8\textwidth]{B-project.pdf}
 \caption{Structure of the imports relation in a B project}
@@
 For this, we follow a pragmatic approach, based on testing and simulation of the
-generated LLVM artifacts, and using the existing LLVM tool set.
-
-First, we define a sample set of small B implementation modules, that contain
-all the B constructs handled in the translation. For each such module, the rules
-are applied manually, yielding a LLVM module. This LLVM module is then amended
-with extra code that have two purposes:
-\begin{itemize}
-\item a standalone executable may be generated from it;
-\item the executable creates an instance of the B implementation, applies the
-  initialization code, and offers to the user the choice to interactively invoke
-  all the operations of the implementation; after initialization and each
-  operation application the state of the implementation is printed. 
-\end{itemize}
+generated LLVM artifacts, and using the existing LLVM tool set. Consequently,
+the rules described in the preceding sections have been prototyped in the Python
+programming language. The input of the translation is in a XML-based format,
+named BXML, that is generated by the forthcoming version of Atelier-B. For each
+B component (machine or implementation), Atelier-B generates a BXML file that
+may then be used as input by our prototype. The output of the translation is in
+LLVM-IR format. Such output may then be processed by the LLVM tools either to
+generate binary in some compatible target platform or LLVM bitcode that may then
+be executed using the LLVM interpreter. We used the latter approach to validate
+our translation.
 
@@
 \begin{verbatim}
-llvm-as-mp-3.1 example-01.llvm -o example-01.bc
+llvm-as example-01.llvm -o example-01.bc
+\end{verbatim}
+\item[llvm-link] The LLVM bitcode linker, which may be invoked as follows:
+\begin{verbatim}
+llvm-link example-01.bc example-02.bc -o prog.bc
 \end{verbatim}
@@
 \begin{verbatim}
-lli-mp-3.1 example-01.bc
+lli-mp example-01.bc
 \end{verbatim}
@@
 
-Once a rule has been manually validated, we implement it in a prototype tool
-written in Python. Unfortunately the tool does not take as input the B0 source
-file, but instead a collection of Python definitions corresponding to the
-syntactic entities present in the parse tree. For instance, for an integer
-concrete variable named \B{value}, we would have the following Python entity:
+We describe below the different commands required to use the LLVM framework to
+animate a B implementation \B{wd} that uses an instance of a B module
+\B{counter}:
 \begin{verbatim}
-value = { "kind": "Vari",
-          "id": "value",
-          "type": "INT",
-          "scope": "impl",
-          "root": counter_i}
+b2llvm counter counter.llvm
+b2llvm wd counter.llvm
+b2llvm --mode proj wd init-wd.llvm
+llvm-as counter.llvm -o counter.bc
+llvm-as wd.llvm -o wd.bc
+llvm-as init-wd.llvm -o init-wd.bc
+llvm-as scaffold-wd.llvm -o scaffold-wd.bc
+llvm-link counter.bc wd.bc init-wd.bc scaffold-wd.bc -o anim-wd.bc
+lli anim-wd.bc
 \end{verbatim}
-where \verb'counter_i' would be the Python entity representing the parse tree
-root for the B implementation. Again the output generated by the prototype
-is compiled and simulated using off-the-shelf LLVM tools \verb'llvm-as' and
-\verb'lli'.
 

### 2013-10-24 PROTO/b2llvm.py

@@ -16,14 +16,11 @@ parser.add_argument('-m', '--mode', choices=['comp','proj'], default='comp',
                     help='Selects code generation mode.')
 
 args = parser.parse_args()
-if args.comp and args.proj:
-    print("error: choose only one code generation mode.")
-elif not args.comp and not args.proj:
-    print("error: choose one code generation mode")
-else:
-    translate_bxml(b_module, args.llvm_file, mode=args.mode, 
-                   dirname=args.directory, settings=args.settings)
-    print("b2llvm code generation completed")
-    print("- input: " + args.bxml_file)
-    print("- output: " + args.llvm_file)
-    print("- mode: " + ("top-level" if args.top else "library machine"))
+translate_bxml(args.b_module, args.llvm_file, mode=args.mode, 
+               dir=args.directory, settings=args.settings)
+print("b2llvm code generation completed")
+print("- BXML directory: " + args.directory)
+print("- B project settings file: " + args.settings)
+print("- B module: " + args.b_module)
+print("- LLVM output file: " + args.llvm_file)
+print("- code generation mode: " + args.mode)

### 2013-10-24 PROTO/b2llvm/translate.py

@@ -184,7 +184,7 @@ def section_interface_init(m):
     comp = list()
     if is_stateful(m):
         comp.append(m)
-    comp.extend([comp.mach for comp in comp_indirect(m)])
+    comp.extend([x.mach for x in comp_indirect(m)])
     res = str()
     res += "declare void"+sp+init_name(m)
     res += "("+commas(list_machine_refs(comp))+")"+nl
@@ -207,7 +207,7 @@ def section_interface_op(m, op):
         tl.append(state_r_name(m))
     tl.extend([ x_type(i["type"]) for i in op["inp"] ])
     tl.extend([ x_type(o["type"])+"*" for o in op["out"] ])
-    return "declare void"+sp+global_name(op)+"("+commas(tl)+")"+nl
+    return "declare void"+sp+op_name(op)+"("+commas(tl)+")"+nl
 
 def section_typedef(m):
     '''
@@ -363,7 +363,7 @@ def x_operation(n):
     check_kind(n, {"Oper"})
     names.reset()
     res = str()
-    res += "define void "+global_name(n)
+    res += "define void "+op_name(n)
     res += "("+commas([state_r_name(n["root"])+sp+"%self$"]+
                            [x_type(i["type"])+sp+"%"+i["id"] for i in n["inp"]]+
                            [x_type(o["type"])+"*"+sp+"%"+o["id"] for o in n["out"]])+")"
@@ -663,7 +663,7 @@ def x_call(n):
         args.append(t + sp + v2)
     args.extend(il)
     args.extend(ol)
-    id = global_name(operation)
+    id = op_name(operation)
     res += tb + "call void" + sp + id + "(" + commas(args) + ")" + nl
     return res
 
@@ -728,24 +728,30 @@ def state_r_name(n):
     else:
         return state_r_name(machine(n))
 
-def global_name(n):
+def op_name(n):
     '''
     - Input:
-      n: A node representing a B construct
+      n: A node representing a B operation
     - Output:
       A string for the name of the LLVM construct representing n.
     '''
-    return "@" + n["root"]["id"] + "$" + n["id"]
+    check_kind(n, "Operation")
+    root = n["root"]
+    check_kind(root, "Impl")
+    machine = root["machine"]
+    return "@" + machine["id"] + "$" + n["id"]
 
 def init_name(n):
     '''
     - Input:
-      n : a node representing a B implementation
+      n : a node representing a B machine or implementation
     - Output:
[…]

### 2013-11-18 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
A first draft about array and structure

\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsfonts}
\usepackage{amssymb}

\usepackage[fleqn]{amsmath}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage[colorlinks=false]{hyperref}
\usepackage{minted}   %Used to syntax highlight

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

\newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}
\newcommand{\VGM}[1]{\marginpar{\tiny{VGM: #1}}}

\newcommand{\stateindex}[0]{\ensuremath{\sf\tau}}

\author{David Déharbe, Valério Medeiros Jr.}
\title{Array}

\begin{document}

%TODO Review the text before present to David

\definecolor{bg}{rgb}{0.95,0.95,0.95}

\usemintedstyle{friendly}

\section{The abstract syntax tree with support for advanced types}

The supported advanced types are array and structures. The nodes Array and 
Structure can be used to declare a variable or to select an element in the array 
or the structure.

\begin{figure}[H]
  \begin{center}
    {\footnotesize
      \frame{
    \begin{tabular}[t]{ccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \B{impl} & : & opt \B{Impl} \\
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{impo} & : & seq \B{Import} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
        \hline
        \B{mach} & : & \B{Machine} \\
        \B{pre} & : & opt \B{Name}
        \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
	\hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
       \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\

       \end{tabular}

       \begin{tabular}[t]{rcl}

        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\   
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
	\hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{Vari} \\
        \B{rhs} & : & \B{Expr} \\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Oper} \\
        \B{inp} & : & seq \B{Expr} \\
        \B{out} & : & seq \B{Name} \\
        \B{inst} & : & opt \B{Impo} \\
        \hline
        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
        \hline
        \B{expr} & : & \B{Expr} \\
        \B{branches} & : & seq \B{CaseBr}\\
	\hline
        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
        \hline
        \B{val} & : & seq \B{Value} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
        \hline
        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Vari}} \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\

      \end{tabular}

      \begin{tabular}[t]{rcl}

        \hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
        \hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{Array}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{element} & : &  \B{Name}\\
        \B{id} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\	
        \hline
        \multicolumn{3}{|c|}{Data array : \B{Array}} \\
        \hline
	\B{dec} & : &  seq \B{Vector} \\
	\B{type} & : & seq \B{CSet}\\
	\B{indexes} & : & seq \B{STerm}\\
	\hline
  	\multicolumn{3}{|c|}{Vector : \B{Vector}} \\
        \hline
	\B{elements} & : & seq \B{Expr}\\ 
	\hline
	\multicolumn{3}{|c|}{Concrete set : \B{CSet}} \\
	\hline
	\multicolumn{3}{c}{BOOL \alt NAT \alt NAT$_{1}$ \alt } \\
	\multicolumn{3}{c}{INT \alt B0\_INTERVAL} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{B0\_INTERVAL}} \\
	\hline
	\B{start} & : & \B{Exp} \\
	\B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Simple term : \B{STerm}} \\
	\hline
	\B{index} & : &  \B{BOOL} \alt \B{INT} \alt \\ 
	& &  \B{Vari} \alt \B{Cons} \\
%	\multicolumn{3}{c}{ \B{BOOL} \alt \B{INT} \alt }\\
%	\multicolumn{3}{c}{ \B{Vari} \alt \B{Cons} }\\
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

\section{Translation of Array and Structures}

Two approaches are possibles to support arrays. An approach is importing
reusable libraries of AtelierB that are reported at official document of
ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
This approach is supported, but the generated code is more complex and it needs
more intermediate steps to manipulate the data of imported components. Not only,
this approach allows dynamic allocation for arrays, but it is extremely not
recommended to safety-critical systems. A second approach enjoys the direct 
correspondence between arrays native B0 and LLVM-IR and this approach is presented here.

A concrete array is a total function, where the domain indexes the range 
elements. The domain is composed by Cartesian product of sequence of concrete 
sets. The node Data array in AST can represent a declaration of array or a 
selection of element. The first element ``dec'' declares explicitly an array or does a 
reference to a variable array. The second element ``type'' defines a sequence of 
array types, where the last element defines the range of function and the others 
elements defines the domain. The third element ``indexes'' is optional, and it 
represents the selection of an element in array.

Some examples of array are:
\begin{itemize}
	\item $array1 \in (0..10) \to INT $
	\item $array2 \in (0..10) \times (0..10) \to INT $   (a multidimensional example)
	\item $array3 \in BOOL \times (0..10) \to INT$ (a domain with two index types)
\end{itemize}

The B0 language supports attributions between arrays, even though the attribution
must have the same domain type and range type. 
%TODO to add this restriction in become equals.

\subsection{Relation arrays B0 and LLVM-IR}
\VGM{It must be extended to support structure.}

The arrays in B0 are represented by functions where an index pointer to an 
element. By default, the element of index not initialized has the zero value.

This section shows the relation between arrays B0 and LLVM-IR. The sequence of
steps are: defining  an array, getting an element and setting an
element.

\subsubsection{Defining an array}

The essential elements of array are its name, size and type. The
following code declares an array global called ``arr'' with 100 integer
numbers with 32 bits and initialized with zero.

\begin{footnotesize}
\begin{minted}[frame=single,linenos=true,bgcolor=bg]{llvm}
@arr = internal global [100 x i32] zeroinitializer
\end{minted}
\end{footnotesize}

Some type examples are:
\begin{itemize}
	\item \verb|[25 x i32]| - Array with 25 elements of 32-bit integer values.
	\item \verb|[4 x [5 x [6 x i16]]] | - Multidimensional array with (4 x 5
x 6) elements of 16-bit integer values.
\end{itemize}

\subsubsection{Getting a element from array}
\VGM{Maybe it can be summarized with text of section 3}

The LLVM-IR instruction \textit{getelementptr} is used to select an element in
the array or structure. A simple example is:

\begin{footnotesize}
\begin{minted}[frame=single,linenos=true,bgcolor=bg]{llvm}
%4 = getelementptr  ([100 x i32]* @Array__arr, i64 0, i64 %3)
\end{minted}
\end{footnotesize}

The first parameter defines the type and name of variable, and the others are
indexes to select the element in the data type. The second parameter (i64 0)
selects the first element in the pointer (* @Array\_\_arr). The last element,
(i64 \%3) selects the value in position obtained from \%3 of the array of 32-bit
integer values. When the array is multidimensional, news index must be added
separated by comma to get the specific element.

\iffalse
\begin{verbatim}
<temporary variable> = getelementptr  
  [<number of elements> x <element type>]   * @<variable_name>, 
  <type and index of pointer> <index_value>,
  <type and index. ex:i64>  <index_value>
\end{verbatim}
\fi

\subsubsection{Setting an element to array}

The following blocks illustrate a B code setting and getting values with array
``arr''  and its translation to LLVM-IR. 
\begin{footnotesize}

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
  arr(2):=5;
  i = arr(2);
\end{minted}
\end{footnotesize}

\begin{footnotesize}
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}
  store i32 5, i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  %1 = load i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  store i32 %1, i32* @i, align 4
\end{minted}
\end{footnotesize}

\subsection{Draft of Rules}

\paragraph{An atomic expression}

\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a primitive state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr  \PH{\stateref(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t\\
& \ELSE (\B{n} \mbox{ is an array or structure state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$,
\PH{\stateindex(n)} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\
\end{align*}

Definition of a new function:

\begin{itemize}
\item $\stateindex(\B{n}) \isdef$ to get a sequence of index to select the 
element. If $\B{n} $ is integer vector ``arr'' with 100 elements selecting the 
element 1 then the result is this text ``, i32 0, i64 1''. It means to select 
the element in dimension zero and position one of array, where the index types are 
respectively ``i32'' and ``i64''. More detailed examples are presented in the reference 
manual of LLVM-IR\ref{}.

\end{itemize}

\subsubsection{Issues and problems}
\begin{itemize}
	\item Integrate it to report

	:promptfind
\begin{itemize}
	\item Identify the rules that must be updated:
	\begin{itemize}
		\item \textbf{Atomic expression} - getting an element to manipulate.
		\item %TODO - setting a element 
		\item Others involved rules ?

	\end{itemize}

	\end{itemize}

\end{itemize}

\newpage

\section*{Attachments}

\iffalse  % This content is not visible to reader.
A generic example available at repository ProB\footnote{
\url{https://github.com/bendisposto/probparsers/blob/develop/bparser/src/test/resources/parsable/Array.mch}}.

\subsection{A very simple and generic example}

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
MACHINE
   Array(maxindex, TYPE)
CONSTRAINTS maxindex : NAT
VARIABLES arr
INVARIANT arr : 1..maxindex +-> TYPE
INITIALISATION arr := {}
OPERATIONS

set(ix,tt)=
        PRE ix : 1..maxindex & tt : TYPE
        THEN arr := arr <+ {ix |-> tt}
        END ;

tt <-- read(ix) =
        PRE ix : dom (arr)
        THEN tt := arr(ix)
        END;

swap(ix, jx)=
        PRE ix : dom(arr) & jx : dom(arr)
        THEN arr := arr <+ {ix |-> arr(jx), jx|-> arr(ix)}
        END

END
\end{minted}
\fi
\subsection{A  simple example and its LLVM-IR}
\footnotesize{
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
MACHINE
   Array
VARIABLES arr

INVARIANT 
    arr : ((0..99) --> 0..1000)

INITIALISATION arr := {1|->100}

OPERATIONS

set(ix,tt)=
        PRE ix : (0..99) & tt : (0..1000)
        THEN arr := arr <+ {ix |-> tt}
        END ;

tt <-- read(ix) =
        PRE ix : (0..99)
        THEN tt := arr(ix)
        END;

swap(ix, jx)=
        PRE ix : (0..99) & jx : (0..99)
        THEN arr := arr <+ {ix |-> arr(jx), jx|-> arr(ix)}
        END

END
\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

      arr(ix) := tt 
   END;

   tt <-- read ( ix ) =
   IF
       ix >= 0 & ix <=99
   THEN
       tt := arr ( ix )
   ELSE
       tt := 0
   END;

   swap ( ix , jx ) =
   IF
        ix >= 0 & ix <=99 & jx >= 0 & jx <=99 
   THEN

       VAR temp IN
           temp := arr(jx);
           arr(jx) := arr(ix);
           arr(ix) := temp
       END
   END

END

\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}

@Array__arr = internal global [100 x i32] zeroinitializer, align 16

define void @Array__INITIALISATION() nounwind ssp {
  store i32 100, i32* getelementptr  ([100 x i32]* @Array__arr, i32 0, i64 1)
  ret void
}

define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %tt, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %19

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %19

; <label>:8                                       ; preds = %5
  %9 = load i32* %2, align 4
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %19

; <label>:11                                      ; preds = %8
  %12 = load i32* %2, align 4
  %13 = icmp sle i32 %12, 1000
  br i1 %13, label %14, label %19

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = load i32* %1, align 4
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}

define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32*, align 8
  store i32 %ix, i32* %1, align 4
  store i32* %tt, i32** %2, align 8
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %14

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %14

; <label>:8                                       ; preds = %5
  %9 = load i32* %1, align 4
  %10 = sext i32 %9 to i64
  %11 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %10
  %12 = load i32* %11
  %13 = load i32** %2, align 8
  store i32 %12, i32* %13
  br label %16

; <label>:14                                      ; preds = %5, %0
  %15 = load i32** %2, align 8
  store i32 0, i32* %15
  br label %16

; <label>:16                                      ; preds = %14, %8
  ret void
}

define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  %temp = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %jx, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %30

; <label>:5                                       ; preds = %0
  %6 = load i32* %1, align 4
  %7 = icmp sle i32 %6, 99
  br i1 %7, label %8, label %30

; <label>:8                                       ; preds = %5
  %9 = load i32* %2, align 4
  %10 = icmp sge i32 %9, 0
  br i1 %10, label %11, label %30

; <label>:11                                      ; preds = %8
  %12 = load i32* %2, align 4
  %13 = icmp sle i32 %12, 99
  br i1 %13, label %14, label %30

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = sext i32 %15 to i64
  %17 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %16
  %18 = load i32* %17
  store i32 %18, i32* %temp, align 4
  %19 = load i32* %1, align 4
  %20 = sext i32 %19 to i64
  %21 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %20
  %22 = load i32* %21
  %23 = load i32* %2, align 4
  %24 = sext i32 %23 to i64
  %25 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %24
  store i32 %22, i32* %25
  %26 = load i32* %temp, align 4
  %27 = load i32* %1, align 4
  %28 = sext i32 %27 to i64
  %29 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %28
  store i32 %26, i32* %29
  br label %30

; <label>:30                                      ; preds = %14, %11, %8, %5, %0
  ret void
}

\end{minted}

}

\subsection{Real example - Bubble Sort}

\begin{footnotesize}
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
MACHINE
   Bubble
VARIABLES
   vec,sort
INVARIANT
   vec : 0..99 --> 0..99 & sort : 0..1 &
   ( sort = 1 =>  !ii.(ii : 0..98 => vec(ii)<=vec(ii+1)))
INITIALISATION
   vec :: 0..99 --> 0..99 || sort := 0
OPERATIONS
   op_sort =
   ANY sorted_vector WHERE
    sorted_vector : 0..99 --> 0..99 &
    !ii.(ii : 0..98 => sorted_vector(ii)<=sorted_vector(ii+1))
   THEN
    vec := sorted_vector|| sort := 1
   END

END

IMPLEMENTATION
   Bubble_i
REFINES
   Bubble
CONCRETE_VARIABLES
   vec1,sort1
INVARIANT
   vec1 : 0..99 --> 0..99 & sort1 : 0..1 &
   vec = vec1 & sort1 = sort
INITIALISATION
   vec1(0) := 0  ;
   sort1 := 0
OPERATIONS
   op_sort =
   VAR nn, swapped, ii, tmp, vi, vi2 IN
	swapped := 1;
	nn := 100;
	ii:=0;
	WHILE swapped = 1 DO
		swapped := 0;
		ii:= 0;
		nn:= nn-1;

		WHILE ii<= nn DO
		vi :=  vec1(ii);
		vi2 := vec1(ii+1);
		IF vi > vi2 THEN
			tmp:= vec1(ii+1);
			vec1(ii+1):=vec1(ii);
			vec1(ii):=tmp;
			swapped:= 1
		END;
		ii:= ii+1

		INVARIANT 1>0
		VARIANT nn-ii
		END

	INVARIANT 1>0
	VARIANT nn-ii
	END
   END
END

\end{minted}
\end{footnotesize}

\end{document}

### 2013-11-19 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Fixing some grammar problems and adding a presentation produced last week.

@@
 
+\section{The LLVM grammar with support to array multidimensional}
+
+\begin{figure}
+  \begin{center}
+    \begin{tabular}{rcl}
+      \nt{module} & ::= & \ListOf{\nt{item}} \\
+      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def}
+      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
+      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
+      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
+      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
+      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
+      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
+      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
+      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
+      \nt{param} & ::= & \nt{type} \nt{name} \\
+      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
+      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
+      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
+      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
+      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
+      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
+      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
+      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
+      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
+      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
+%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
+      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
+      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
+      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
+      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
+      \nt{arg} & ::= & \nt{type} \nt{exp}
+    \end{tabular}
+  \end{center}
+  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
+    and \nt{name} correspond respectively to integer types, integer literals,
+    labels and names. Choices are separated by \alt and optionally delimited by
+    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
+    elements of the annotated entity.}
+  \label{fig:llvm-grammar}
+\end{figure}
+
 \section{Translation of Array and Structures}
 
-Two approaches are possibles to support arrays. An approach is importing
-reusable libraries of AtelierB that are reported at official document of
+Two approaches are possible to support arrays. An approach is importing
+reusable libraries of AtelierB that are reported in the official document of
 ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
 This approach is supported, but the generated code is more complex and it needs
-more intermediate steps to manipulate the data of imported components. Not only,
-this approach allows dynamic allocation for arrays, but it is extremely not
+more intermediate steps to manipulate the data of imported components. Moreover,
+this approach allows dynamic allocation for arrays, but it is strongly not
 recommended to safety-critical systems. A second approach enjoys the direct 
-correspondence between arrays native B0 and LLVM-IR and this approach is presented here.
+correspondence between B0 native arrays and LLVM-IR and this approach is presented here.
 
-A concrete array is a total function, where the domain indexes the range 
-elements. The domain is composed by Cartesian product of sequence of concrete 
+A concrete array is a total function where the domain indexes the range 
+elements. The domain is composed of Cartesian product of sequence of concrete 
 sets. The node Data array in AST can represent a declaration of array or a 
-selection of element. The first element ``dec'' declares explicitly an array or does a 
-reference to a variable array. The second element ``type'' defines a sequence of 
-array types, where the last element defines the range of function and the others 
-elements defines the domain. The third element ``indexes'' is optional, and it 
-represents the selection of an element in array.
+selection of element. The first element ``dec'' declares explicitly an array or 
+references a variable array. The second element ``type'' defines a sequence of 
+array types, where the last element defines the range of function and the other 
+elements define the domain. The third element ``indexes'' is optional, and it 
+represents the selection of an element in array, or the same type structure 
+when it is a structure.
 
@@
 
-The B0 language supports attributions between arrays, even though the attribution
-must have the same domain type and range type. 
+\paragraph{Attributions between array and structure}
+
+The B0 language supports attribution directly between arrays or structure, even though the attribution
+must have the same domain type and range type. \VGM{Project Decision: How translate a direct 
+attribution between advanced types, for example, arrayA := arrayB ? The Commenc 
+uses a function called ``memmove''. }
+
 %TODO to add this restriction in become equals.
@@
 
-The arrays in B0 are represented by functions where an index pointer to an 
-element. By default, the element of index not initialized has the zero value.
+The arrays in B0 are represented by functions where an index points to an 
+element. By default, the element of index not initialized has a zero value.
 
@@
 
-The first parameter defines the type and name of variable, and the others are
+The first parameter defines the type and name of the variable, and the others are
 indexes to select the element in the data type. The second parameter (i64 0)
@@
 (i64 \%3) selects the value in position obtained from \%3 of the array of 32-bit
-integer values. When the array is multidimensional, news index must be added
+integer values. When the array is multidimensional, the news index must be added
 separated by comma to get the specific element.

### 2013-11-19 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Fixing the grammar and the examples.

@@
         \hline
-	\B{element} & : &  \B{Name}\\
-        \B{id} & : & seq \B{Name} \\
+	\B{id} & : &  \B{Name}\\
+        \B{elements} & : & seq \B{Name} \\
 	\B{type} & : & seq \B{Type}\\	
@@
 	\hline
-	\multicolumn{3}{c}{BOOL \alt NAT \alt NAT$_{1}$ \alt } \\
+	\multicolumn{3}{c}{BOOL \alt NAT \alt NAT$_{1}$\alt } \\
 	\multicolumn{3}{c}{INT \alt B0\_INTERVAL} \\
@@
 	\hline
-	\B{index} & : &  \B{BOOL} \alt \B{INT} \alt \\ 
+	\B{index} & : &  \B{BOOL} \alt \B{INT}\alt \\ 
 	& &  \B{Vari} \alt \B{Cons} \\
-%	\multicolumn{3}{c}{ \B{BOOL} \alt \B{INT} \alt }\\
-%	\multicolumn{3}{c}{ \B{Vari} \alt \B{Cons} }\\
       \end{tabular}
@@
       & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
-      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
+      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} 
+\nt{exp}\llvm{,} \nt{index}\llvm{,} \ListOf{\nt{index}} \\
       & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
@@
       & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
-      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
+      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} 
+\llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \ListOf{\nt{index}} \llvm{)} \\
       \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
@@
     \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
-    elements of the annotated entity.}
+    elements of the annotated entity. }
   \label{fig:llvm-grammar}
 \end{figure}
+
+\VGM{To support arrays and structures, I
+
+added only + after index.}
 
@@
 attribution between advanced types, for example, arrayA := arrayB ? The Commenc 
-uses a function called ``memmove''. }
-
-%TODO to add this restriction in become equals.
+uses a function called ``memmove''.  %TODO to add this restriction in become equals.
+ }
 
@@
 \begin{itemize}
-	\item Integrate it to report
-
-	:promptfind
-\begin{itemize}
-	\item Identify the rules that must be updated:
-	\begin{itemize}
-		\item \textbf{Atomic expression} - getting an element to manipulate.
-		\item %TODO - setting a element 
-		\item Others involved rules ?
+	\item To develop support to structures (more easy)
+	\item To integrate it in translation.tex
+	\item To develop support to copy all  of an advanced types
+	\item To identify the rules that must be updated:
 
@@
 
-	\end{itemize}
-
-\end{itemize}
-
-\newpage
-
 \section*{Attachments}
 
-\iffalse  % This content is not visible to reader.
+%\iffalse  % This content is not visible to reader.
 A generic example available at repository ProB\footnote{
 \url{https://github.com/bendisposto/probparsers/blob/develop/bparser/src/test/resources/parsable/Array.mch}}. 
+This example can be used in approach using imported components.
 
@@
 \end{minted}
-\fi
+%\fi
 \subsection{A  simple example and its LLVM-IR}
@@
 }
+\end{minted}
+
+\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}
 

### 2013-12-02 PROTO/b2llvm.py
Fixed regression introduced to please pylint. Added a script to generate a coverage report. Updated .gitignore with files generated by coverage report script.

@@ -32,6 +32,6 @@ if (args.verbose):
     print("- code generation mode: " + args.mode)
     print("- emits traceability information: " + str(args.trace))
     print("- emits printing functions: " + str(args.emit_printer))
-translate_bxml(args.b_module, args.llvm_file, mode=args.mode,
+translate_bxml(args.b_module, args.llvm_file, trace, mode=args.mode,
                dir=args.directory, settings=args.settings,
                emit_printer=args.emit_printer)

### 2013-12-02 PROTO/b2llvm/ast.py

@@ -108,7 +108,7 @@ def make_lt(term1, term2):
 
 def make_ge(term1, term2):
     """Creates an AST node for a B "greater or equal" expression."""
-    return make_comp("<=", term1, term2)
+    return make_comp(">=", term1, term2)
 
 def make_gt(term1, term2):
     """Creates an AST node for a B "greater than" expression."""
@@ -229,7 +229,7 @@ def make_implementation(name, imports, consts, variables, init, ops):
     """Creates an AST node for a B implementation module."""
     assert type(imports) is list
     assert type(consts) is list
-    assert type(vars) is list
+    assert type(variables) is list
     assert type(init) is list
     assert type(ops) is list
     root = { "kind": "Impl",
@@ -240,7 +240,7 @@ def make_implementation(name, imports, consts, variables, init, ops):
              "variables": variables,
              "initialisation": init, "operations": ops,
              "stateful": None}
-    for node in imports + vars + ops:
+    for node in imports + variables + ops:
         node["root"] = root
     return root
 
@@ -251,7 +251,7 @@ def make_base_machine(name, consts, variables, ops):
     on that machine.
     '''
     assert type(consts) is list
-    assert type(vars) is list
+    assert type(variables) is list
     assert type(ops) is list
     root = { "kind": "Machine",
              "id": name,
@@ -263,7 +263,7 @@ def make_base_machine(name, consts, variables, ops):
              "stateful": None,
              "comp_direct": None,
              "comp_indirect": None}
-    for node in vars + ops:
+    for node in variables + ops:
         node["root"] = root
     return root
 

### 2013-12-02 PROTO/b2llvm/loadbxml.py

@@ -216,7 +216,7 @@ def load_developed_machine(dirname, project, loaded, machine):
     assert root.get("type") == "abstraction"
     assert name_attr(root) == machine
     impl_id = project.implementation(machine)
-    pyimpl = load_implementation(dir, project, loaded, impl_id)
+    pyimpl = load_implementation(dirname, project, loaded, impl_id)
     pymachine = ast.make_developed_machine(machine, pyimpl)
     pyimpl["machine"] = pymachine
     _LOADING.remove(machine)
@@ -550,9 +550,9 @@ def load_var_in(node, symast, symimp):
         pyvar = ast.make_loc_var(name, pytype)
         symast2.add(name, pyvar)
         pyvars.append(pyvar)
-        pybody = [ load_sub(xmlbody.find("./*"), symast2, symimp) ]
-        symast2.clear()
-        return ast.make_var_decl(pyvars, pybody)
+    pybody = [ load_sub(xmlbody.find("./*"), symast2, symimp) ]
+    symast2.clear()
+    return ast.make_var_decl(pyvars, pybody)
 
 def load_binary_substitution(node, symast, symimp):
     """Load an XML element for a binary substitution to an AST node."""
@@ -659,7 +659,7 @@ def load_exp(node, symast):
 
 def load_identifier(node, symast):
     """Load an XML element for an identifier to an AST node."""
-    return symast[value(node)]
+    return symast.get(value(node))
 
 def load_boolean_literal(node):
     """Load an XML element for an Boolean literal to an AST node."""
@@ -786,6 +786,9 @@ class SymbolTable(object):
         """Constructor. Prefills with MAXINT."""
         self._table = dict()
         self._table["MAXINT"] = ast.MAXINT
+    def __str__(self):
+        """Converts object to string."""
+        return "{"+",".join([key for key in self._table.keys()])+"}"
     def add(self, name, node):
         """Adds an element, checking if there is a conflict."""
         if name in self._table:
@@ -793,17 +796,25 @@ class SymbolTable(object):
         self._table[name] = node
     def rem(self, name):
         """Removes an element, checking if it is present."""
-        if name not in self._table:
+        if name not in self._table.keys():
             _LOG.error(name+"not found in table")
         self._table.pop(name)
     def get(self, name):
         """Gets an element (i.e. an AST node) from its name."""
-        if name not in self._table:
+        if name not in self._table.keys():
             _LOG.error(name+"not found in table")
         return self._table[name]
     def clear(self):
[…]

### 2013-12-02 PROTO/b2llvm/translate.py

@@ -42,9 +42,9 @@ def translate_bxml(bmodule, outfile, trace, mode='comp', dir='bxml', settings='p
     res = bytearray()
     res.extend(";; -*- mode: asm -*-"+NL) # emacs syntax highlight on
     trace.outu(res, "file generated with b2llvm")
+    trace.outu(res, "B project settings: "+settings)
     trace.outu(res, "B module: "+bmodule)
     trace.outu(res, "B project directory: "+dir)
-    trace.outu(res, "B project settings: "+settings)
     trace.outu(res, "code generation mode: " + ("component" if mode == "comp" else "project"))
     trace.outu(res, "output file: "+outfile)
     if mode == 'comp':
@@ -865,7 +865,7 @@ def x_if_br(text, lbr, lbl, trace):
             x_formula(text, br["cond"], lbl_1, lbl_2, trace)
             text.extend(lbl_1 + ":" + NL)
             x_inst_label(text, br["body"], lbl, trace)
-            text.extend(lbl_2 + ":" + NL, trace)
+            text.extend(lbl_2 + ":" + NL)
 
 ### TRANSLATION OF WHILE INSTRUCTIONS
 

### 2013-12-02 PROTO/coverage.sh

@@ -1,0 +1,37 @@
+#
+# This is a crude script to generate coverage data from the set of benchmarks
+# that we currently have.
+# It relies on the installation of the Python script "coverage"
+# https://bitbucket.org/ned/coveragepy
+#
+#!/bin/bash
+COV_RUN="coverage run --source=b2llvm.py,b2llvm -p"
+echo "Cleaning up existing coverage data."
+coverage erase
+echo "Generating new coverage data."
+$COV_RUN ./b2llvm.py -v -m comp counter counter.llvm bxml project.xml > /dev/null
+$COV_RUN ./b2llvm.py -m comp ioint ioint.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m comp counter counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m comp wd wd.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m comp timer timer.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj counter init-counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj wd init-wd.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj timer init-timer.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m comp ioint ioint.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m comp counter counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m comp wd wd.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m comp timer timer.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m proj counter init-counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m proj wd init-wd.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -t -m proj timer init-timer.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -p -t -m comp ioint ioint.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -p -t -m comp counter counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -p -t -m comp timer timer.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -p -t -m comp wd wd.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -p -t -m proj wd wd.llvm bxml project.xml
+echo "Combining coverage data."
[…]

### 2013-12-03 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Reworked introduction of arrays in AST.
Added the rules to support array and the Node Lvalue.

@@
         \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
-
-       \end{tabular}
-
-       \begin{tabular}[t]{rcl}
-
         \hline
@@
         \B{body} & : & seq \B{Inst} \\
+	\end{tabular}
+
+       \begin{tabular}[t]{rcl}   
         \hline
@@
         \hline
-        \B{lhs} & : & \B{Vari} \\
+        \B{lhs} & : & \B{LValue} \\
         \B{rhs} & : & \B{Expr} \\
+	\hline
+	\multicolumn{3}{|c|}{Label value: \B{LValue}} \\	
+	\hline
+	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} }\\
         \hline
@@
         \hline
-        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Vari}} \\
+        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Cons}}\\
+	\multicolumn{3}{c} {\B{Vari} \alt \B{Array} \alt \B{ArrayItem} } \\
         \hline
@@
         \B{args} & : & seq \B{Exp} \\
-
   \end{tabular}
@@
   \begin{tabular}[t]{rcl}
-
 	\hline
@@
         \hline
-        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
-	\multicolumn{3}{c}{\B{Struct} \alt \B{Array}}\\
+        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt \B{ArrayType}}\\
 	\hline	
-  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
+        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
         \hline 
-	\B{id} & : &  \B{Name}\\
-        \B{elements} & : & seq \B{Name} \\
-	\B{type} & : & seq \B{Type}\\	
+	\B{dom} & : & seq \B{Set}\\
+	\B{ran} & : & \B{Set}\\
         \hline
-        \multicolumn{3}{|c|}{Data array : \B{Array}} \\
+        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
         \hline
-	\B{dec} & : &  seq \B{Vector} \\
-	\B{type} & : & seq \B{CSet}\\
-	\B{indexes} & : & seq \B{STerm}\\
+	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} }\\
 	\hline
-  	\multicolumn{3}{|c|}{Vector : \B{Vector}} \\
+  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
         \hline
-	\B{elements} & : & seq \B{Expr}\\ 
+	\B{arg1} & : & seq \B{Expr}\\ 
+	\B{arg2} & : & \B{Expr}\\ 
 	\hline
-	\multicolumn{3}{|c|}{Concrete set : \B{CSet}} \\
+  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
         \hline
-	\multicolumn{3}{c}{BOOL \alt NAT \alt NAT$_{1}$\alt } \\
-	\multicolumn{3}{c}{INT \alt B0\_INTERVAL} \\
+	\B{arg1} & : & seq \B{Expr}\\ 
+	\B{arg2} & : & \B{Expr}\\ 	
 	\hline
-	\multicolumn{3}{|c|}{B0 interval : \B{B0\_INTERVAL}} \\
+	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
         \hline
-	\B{start} & : & \B{Exp} \\
-	\B{end} & : & \B{Exp} \\
+	\B{base} & : &  \B{Array} \\
+	\B{indices} & : & seq \B{Expr}\\
+	\B{type} & : & \B{ArrayType}\\
 	\hline
-	\multicolumn{3}{|c|}{Simple term : \B{STerm}} \\
+	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
 	\hline
-	\B{index} & : &  \B{BOOL} \alt \B{INT}\alt \\ 
-	& &  \B{Vari} \alt \B{Cons} \\
+	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
+	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
+	\hline
+	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
+	\hline
+	\B{start}, \B{end} & : & \B{Expr} \\
+	\hline
+	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
+	\hline
+	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
       \end{tabular}
@@
 
+\VGM{To add in text: The nodes \B{Int\_lit} and \B{Bool\_lit} are literals in its 
+respective set.}
 \section{The LLVM grammar with support to array multidimensional}
@@
 
-\VGM{To support arrays and structures, I
-
-added only + after index.}
+\VGM{To support arrays and structures, I added only + after index.}
 
@@
 
+\iffalse 
 \paragraph{An atomic expression}
@@
 & \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t\\
-& \ELSE (\B{n} \mbox{ is an array or structure state variable }) \LET p = \variable \AND v = \variable \IN \\
+& \ELSE (\B{n} \mbox{ is an array}) \LET p = \variable \AND v = \variable \IN \\
 & \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$,
@@
 & \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
-\
+\\
+\end{align*}
+\fi
+
+\paragraph{Becomes equal instructions} must evaluate the target and the source
+of the assignment, and then copy the result of the latter in the former, using
+the LLVM instruction \llvm{store}. The evaluation of the target is defined with
+$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
+variable $p$ containing the assigned location. The evaluation of the source
+yields the corresponding code $r$, the variable $v$ holding the result, and its
+type $t$. \VGM{Todo: update only the function LValue}
+\begin{align*}
+\begin{split}
+  \trad{a}{\B{Beq}} \isdef
+  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
+  & \llvm{"\PH{l}} \\
+  & \llvm{\PH{r}} \\
+  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
+%   & l \conc r \conc
+%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
+%   \conc t \conc \llvm{"* "} \conc p \conc \nl
+\end{split}
 \end{align*}
 
-Definition of a new function:
+The assignment target may be either local to the operation or a state
+variable. In the former case, it can be referenced by the corresponding LLVM
+identifier. In the latter case, it is represented as an element of the state
+structure $\Sigma$, and its address must be calculated and assigned to a new temporary
+variable.
+\begin{align*}
+\lefteqn{\trad{n}{lvalue} \isdef} \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
+& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
+& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,\stateindex(\B{n.index,n.type})  \nl"}\\
+& \quad \quad \sep v \sep t \\
+& \quad \ELSE \LET v = \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
+\end{align*}
 
-\begin{itemize}
-\item $\stateindex(\B{n}) \isdef$ to get a sequence of index to select the 
-element. If $\B{n} $ is integer vector ``arr'' with 100 elements selecting the 
-element 1 then the result is this text ``, i32 0, i64 1''. It means to select 
-the element in dimension zero and position one of array, where the index types are 
-respectively ``i32'' and ``i64''. More detailed examples are presented in the reference 
-manual of LLVM-IR\ref{}.
-
-\end{itemize}
+\paragraph{Definition of function:} $\stateindex(\B{index,type}) \isdef$ to get a 
+sequence of index to select the element. If $\B{n} $ is integer vector ``arr'' 
+with 100 elements selecting the element 1 then the result is this text ``, i32 
+0, i64 1''. It means to select the element in dimension zero and position one of 
+array, where the index types are respectively ``i32'' and ``i64''. More detailed 
+examples are presented in the reference manual of LLVM-IR. %TODO: put her a 
+% reference of manual
 

### 2013-12-03 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Added the rules to support array and the Node Lvalue.

@@
 
+\paragraph{Restriction on array and structure:} It is not yet supported the 
+translation of direct attribution between array expression, for example:
+
+\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+ array := array2;
+ structure1 := structure2;
+\end{minted}
+
 \subsubsection{Issues and problems}
@@
 	\item To integrate it in translation.tex
-	\item To develop support to copy all  of an advanced types
-	\item To identify the rules that must be updated:
-
 \end{itemize}
@@
 \section*{Attachments}
+
+\subsection{Structure examples}
+
+The following code example is defining a structure:
+
+\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+Year : struct(
+	Year_number : INT,
+	Leap_year : BOOL,
+	Number_of_days : (1..12) * (28..31),
+	Weather : struct( Average_Temperature : (1..12) *  INT,
+		Average_Rain : (1..12) * INT) )
+\end{minted}
+
+The following code declares a record example:
+
+\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+Year := rec(
+	Year_number : 2013,
+	Leap_year : TRUE,
+	Number_of_days : (12,28),
+	Weather : rec(  Average_Temperature : (5,8),
+			Average_Rain : (5,10)))
+\end{minted}
+
+\subsection{Arrays examples}
 

### 2013-12-03 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Adding the first notes about record/structure

@@
         \B{args} & : & seq \B{Exp} \\
-  \end{tabular}
-
-  \begin{tabular}[t]{rcl}
 	\hline
@@
         \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
+\end{tabular}
+
+  \begin{tabular}[t]{rcl}
 	\hline
@@
         \hline
-        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt \B{ArrayType}}\\
+        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
+	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
 	\hline
+  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
+        \hline
+	\B{id} & : &  \B{Name}\\
+        \B{elements} & : & seq \B{Name} \\
+	\B{type} & : & seq \B{Type}\\
+	\hline
+%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
+%        \hline
+%	\B{id} & : &  \B{Name}\\
+%        \B{elements} & : & seq \B{Name} \\
+%	\B{type} & : & \B{Struct}\\
+%	\hline
         \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
@@
 \begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
- array := array2;
+ array1 := array2;
  structure1 := structure2;
 \end{minted}
-
-\subsubsection{Issues and problems}
-\begin{itemize}
-	\item To develop support to structures (more easy)
-	\item To integrate it in translation.tex
-\end{itemize}
 
@@
 \end{minted}
+
+This code stores a value in the record and get the value of element.
+%TODO example getting
+\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+
+\end{minted}
+
+%TODO example setting
 

### 2013-12-17 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Fixing the rules to support Array multidimensional. It is not finished, but it is almost there.

@@
 \begin{document}
-
-%TODO Review the text before present to David
 
@@
 
-\subsection{Draft of Rules}
+%TODO update the text showing that to get the address is needed to get a 
+%sequence of steps.
+
+\subsection{Rules}
 
@@
 
+TODO: This function updates the Name function ?
+TODO: Integrate with EXPArray
+% Rule to get a element in array, it can be used in become equal or Expressions
+\begin{align*}
+\lefteqn{\trad{n}{lvalue} \isdef} \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
+& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
+& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \trad{n}{AccessArray}  \sep v \sep t \\
+& \quad \ELSE \LET v = \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
+\end{align*}
+
+%The new rule created 
+%It can appear in LHS or RHS in become equal 
+% We cannot use the tal function, because it is not able to solve any 
+% expressions.
+ - Não poderemos usar a função Tal, pois ela não resolverá expressões quaisquer!
+\begin{align*}
+\lefteqn{\trad{n}{AccessArray} \isdef} \\
+&  \LET p_1 \sep v_1 \sep t_1  = \trad{base}{Exp}  \IN \\
+&  \LET p_2 \sep v_2 \sep t_2  = \trad{indices}{Exp}  \IN \\
+&  \LET v =  \variable  \IN \\
+&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0, 
+i32  \PH{v_2}\nl"} \sep v\\
+&  \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,\stateindex(\B{n.index,n.type})  \nl"}\\
+\end{align*}
+
 The assignment target may be either local to the operation or a state
@@
 variable.
-\begin{align*}
-\lefteqn{\trad{n}{lvalue} \isdef} \\
-& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
-& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \\
-& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,\stateindex(\B{n.index,n.type})  \nl"}\\
-& \quad \quad \sep v \sep t \\
-& \quad \ELSE \LET v = \variable \IN \\
-& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
-& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
-\end{align*}
 
@@
 
+%TODO update this example
 \begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
@@
 This code stores a value in the record and get the value of element.
-%TODO example getting
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
-
-\end{minted}
-
-%TODO example setting
 
@@
 
-\subsection{A very simple and generic example}
-
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
-MACHINE
-   Array(maxindex, TYPE)
-CONSTRAINTS maxindex : NAT
-VARIABLES arr
-INVARIANT arr : 1..maxindex +-> TYPE
-INITIALISATION arr := {}
-OPERATIONS
-
-set(ix,tt)=
-        PRE ix : 1..maxindex & tt : TYPE
-        THEN arr := arr <+ {ix |-> tt}
-        END ;
-
-tt <-- read(ix) =
-        PRE ix : dom (arr)
-        THEN tt := arr(ix)
-        END;
-
-swap(ix, jx)=
-        PRE ix : dom(arr) & jx : dom(arr)
-        THEN arr := arr <+ {ix |-> arr(jx), jx|-> arr(ix)}
-        END
-
-END
-\end{minted}
-%\fi
 \subsection{A  simple example and its LLVM-IR}

### 2013-12-17 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
Fixing the rules to support array

@@
 TODO: This function updates the Name function ?
-TODO: Integrate with EXPArray
 % Rule to get a element in array, it can be used in become equal or Expressions
@@
 \lefteqn{\trad{n}{lvalue} \isdef} \\
-& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
 & \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \trad{n}{AccessArray}  \sep v \sep t \\
+& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \trad{n}{ArrayItem}  \sep v \sep t \\
 & \quad \ELSE \LET v = \variable \IN \\
@@
 
-%The new rule created 
-%It can appear in LHS or RHS in become equal 
 % We cannot use the tal function, because it is not able to solve any 
 % expressions.
- - Não poderemos usar a função Tal, pois ela não resolverá expressões quaisquer!
+
+The rule ArrayItem is applied in expressions as ``$a(n1,...,n_{n})$'', where $a$ is n.base 
+and $n1,...,n_{n}$ is a sequence of indices to select an item from array.
+
 \begin{align*}
-\lefteqn{\trad{n}{AccessArray} \isdef} \\
-&  \LET p_1 \sep v_1 \sep t_1  = \trad{base}{Exp}  \IN \\
-&  \LET p_2 \sep v_2 \sep t_2  = \trad{indices}{Exp}  \IN \\
-&  \LET v =  \variable  \IN \\
-&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0, 
-i32  \PH{v_2}\nl"} \sep v\\
-&  \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,\stateindex(\B{n.index,n.type})  \nl"}\\
+\lefteqn{\trad{n}{ArrayItem} \isdef} \\
+&  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
+&  \LET p_2 \sep v_2 \sep t_2  = \trad{n.indices}{\ListOf{Exp}}  \IN \\
+&  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
+&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0, i32  \PH{v_2}\nl"}  \sep v \sep t\\
 \end{align*}

### 2013-12-23 DOCUMENTATION/AdvancedTypes.tex
Intermediate commit (reorganizing repo).

--- 2013-12-17 PAPERS/Report/AdvancedTypes/AdvancedTypes.tex
@@
 \newcommand{\state}[0]{\ensuremath{\sf\Theta}}
-\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
 \newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
   arr(2):=5;
@@
 \begin{footnotesize}
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
   store i32 5, i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
@@
 
-%TODO update the text showing that to get the address is needed to get a 
-%sequence of steps.
+TODO: show a simple example getting or setting an element of array 
+multidimensinal using the rule $ArrayItem$. For example, show the translation of 
+this code: $ arr(2,3) :=4$.
 
 \subsection{Rules}
-
-\iffalse 
-\paragraph{An atomic expression}
-
-\begin{align*}
-\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
-& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
-& \ELSIF \B{n} \mbox{ is a local variable } \THEN
-\llvm{""} \sep \local(\B{n}) \sep t \\
-& \ELSE (\B{n} \mbox{ is a primitive state variable }) \LET p = \variable \AND v = \variable \IN \\
-& \quad \llvm{"\PH{p} = getelementptr  \PH{\stateref(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
-& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t\\
-& \ELSE (\B{n} \mbox{ is an array}) \LET p = \variable \AND v = \variable \IN \\
-& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$,
-\PH{\stateindex(n)} \nl} \\
-& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
-\\
-\end{align*}
-\fi
 
@@
 yields the corresponding code $r$, the variable $v$ holding the result, and its
-type $t$. \VGM{Todo: update only the function LValue}
+type $t$.
+
 \begin{align*}
@@
   & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
-%   & l \conc r \conc
-%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
-%   \conc t \conc \llvm{"* "} \conc p \conc \nl
 \end{split}
@@
 
-TODO: This function updates the Name function ?
 % Rule to get a element in array, it can be used in become equal or Expressions
@@
 & \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \trad{n}{ArrayItem}  \sep v \sep t \\
+& \quad \IF \mbox{\B{n} is an array} \THEN  \trad{n}{ArrayItem} \\
 & \quad \ELSE \LET v = \variable \IN \\
@@
 
-% We cannot use the tal function, because it is not able to solve any 
-% expressions.
-
-The rule ArrayItem is applied in expressions as ``$a(n1,...,n_{n})$'', where $a$ is n.base 
-and $n1,...,n_{n}$ is a sequence of indices to select an item from array.
+\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
+where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indices that 
+calculates the address of element.
 
@@
 
-The assignment target may be either local to the operation or a state
-variable. In the former case, it can be referenced by the corresponding LLVM
-identifier. In the latter case, it is represented as an element of the state
-structure $\Sigma$, and its address must be calculated and assigned to a new temporary
-variable.
-
-\paragraph{Definition of function:} $\stateindex(\B{index,type}) \isdef$ to get a 
-sequence of index to select the element. If $\B{n} $ is integer vector ``arr'' 
-with 100 elements selecting the element 1 then the result is this text ``, i32 
-0, i64 1''. It means to select the element in dimension zero and position one of 
-array, where the index types are respectively ``i32'' and ``i64''. More detailed 
-examples are presented in the reference manual of LLVM-IR. %TODO: put her a 
-% reference of manual
-
 \paragraph{Restriction on array and structure:} It is not yet supported the 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
  array1 := array2;
@@
 
-%TODO update this example
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
-Year : struct(
-	Year_number : INT,
-	Leap_year : BOOL,
-	Number_of_days : (1..12) * (28..31),
-	Weather : struct( Average_Temperature : (1..12) *  INT,
-		Average_Rain : (1..12) * INT) )
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+
+MACHINE Rec
+SETS
+ ID={aa,bb}
+VARIABLES xx, bank, total
+INVARIANT
+ xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
+ total:struct(money:NAT)
+INITIALISATION 
+ bank := rec(name:bb) ||
+ xx:= rec(name:aa, balance:10) ||
+ total := rec(money:20)
+OPERATIONS
+
+  res<--positive = PRE xx'balance > 0 THEN res:=TRUE END;
+
+  withrdaw(amt) = PRE amt:NAT1 & (xx'balance >= amt) THEN
+     xx := rec(name:xx'name, balance:(xx'balance - amt))
+  END;
+
+  unsafe_dec = BEGIN
+       xx'balance := xx'balance - 1
+  END
+END
+
 \end{minted}
 
-The following code declares a record example:
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+IMPLEMENTATION
+    Rec_i
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
-Year := rec(
-	Year_number : 2013,
-	Leap_year : TRUE,
-	Number_of_days : (12,28),
-	Weather : rec(  Average_Temperature : (5,8),
-			Average_Rain : (5,10)))
+REFINES
+    Rec
+
+CONCRETE_VARIABLES
+   xx ,
+   bank ,
+   total
+
+INITIALISATION
+
+   bank := rec(name:bb) ;
+   xx := rec(name:aa, balance:10) ;
+   total :=rec(money:20)
+OPERATIONS
+
+  res<--positive = 
+  IF xx'balance > 0 THEN
+       res:= TRUE 
+  ELSE res := FALSE
+  END;
+
+  withrdaw(amt) = IF   amt >=0 & amt <= MAXINT & (xx'balance >= amt) THEN 
+     xx := rec(name:xx'name, balance:(xx'balance - amt))
+  END;
+
+  unsafe_dec = BEGIN
+       xx'balance := xx'balance -1 
+  END
+END
 \end{minted}
-
-This code stores a value in the record and get the value of element.
 
@@
 \subsection{A  simple example and its LLVM-IR}
-\footnotesize{
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+
 MACHINE
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
 IMPLEMENTATION 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
 
@@
 
-}
-
 \subsection{Real example - Bubble Sort}
 
-\begin{footnotesize}
-\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
 MACHINE
@@
 \end{minted}
-\end{footnotesize}
 

### 2014-01-03 DOCUMENTATION/AdvancedTypes.tex
Update the example and rule to support array.

@@
 \usepackage[colorlinks=false]{hyperref}
+
 \usepackage{minted}   %Used to syntax highlight
+\definecolor{bg}{rgb}{0.95,0.95,0.95}
+\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
+\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
+\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
 
@@
 \begin{document}
-
-\definecolor{bg}{rgb}{0.95,0.95,0.95}
 
@@
 The B0 language supports attribution directly between arrays or structure, even though the attribution
-must have the same domain type and range type. \VGM{Project Decision: How translate a direct 
-attribution between advanced types, for example, arrayA := arrayB ? The Commenc 
-uses a function called ``memmove''.  %TODO to add this restriction in become equals.
- }
+must have the same domain type and range type.
 
 \subsection{Relation arrays B0 and LLVM-IR}
-\VGM{It must be extended to support structure.}
 
@@
 The essential elements of array are its name, size and type. The
-following code declares an array global called ``arr'' with 100 integer
-numbers with 32 bits and initialized with zero.
+following code declares an array global called ``Array3'' with 6000 (10*20*30)
+integer numbers with 32 bits and initialized with zero.
 
-\begin{footnotesize}
-\begin{minted}[frame=single,linenos=true,bgcolor=bg]{llvm}
-@arr = internal global [100 x i32] zeroinitializer
-\end{minted}
-\end{footnotesize}
+\begin{llvmcode}
+@Array3 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
+\end{llvmcode}
 
@@
 
-\begin{footnotesize}
-\begin{minted}[frame=single,linenos=true,bgcolor=bg]{llvm}
-%4 = getelementptr  ([100 x i32]* @Array__arr, i64 0, i64 %3)
-\end{minted}
-\end{footnotesize}
+\begin{llvmcode}
+  %7 = getelementptr [10 x [20 x [30 x i32]]]* @Array3, i32 0, i64 2, i64 3, i64 4
+\end{llvmcode}
 
 The first parameter defines the type and name of the variable, and the others are
-indexes to select the element in the data type. The second parameter (i64 0)
-selects the first element in the pointer (* @Array\_\_arr). The last element,
-(i64 \%3) selects the value in position obtained from \%3 of the array of 32-bit
-integer values. When the array is multidimensional, the news index must be added
-separated by comma to get the specific element.
+indexes to select the element in the data type. The second parameter (i32 0)
+selects the first element in the pointer (* @Array3). The last elements 
+(i64 2, i64 3, i64 4) selects the value from position (2,3,4) storing it in \%7.
+This array is multidimensional then several index must be declared separated
+by comma to get the specific element.
 
@@
 
-\subsubsection{Setting an element to array}
+\subsubsection{Setting and getting array}
 
-The following blocks illustrate a B code setting and getting values with array
-``arr''  and its translation to LLVM-IR. 
-\begin{footnotesize}
+The following blocks illustrate a C code setting a value to array
+and its translation to LLVM-IR.
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
-  arr(2):=5;
-  i = arr(2);
-\end{minted}
-\end{footnotesize}
+\begin{ccode} 
+  int x=2; int y=3 ; int z=4 ;
+  Array3[x][y][z]=7;
+\end{ccode}
 
-\begin{footnotesize}
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
-  store i32 5, i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
-  %1 = load i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
-  store i32 %1, i32* @i, align 4
-\end{minted}
-\end{footnotesize}
+This example is translated in four steps. First, the variables \%x, \%y, \%z 
+are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
+Third, the variables are loaded to local variables and converted to 64 bits. 
+Finally, the address of array is calculated in local variable \%7 and the value 
+7 is stored in it.
 
-TODO: show a simple example getting or setting an element of array 
-multidimensinal using the rule $ArrayItem$. For example, show the translation of 
-this code: $ arr(2,3) :=4$.
+\begin{llvmcode}
+%x = alloca i32, align 4
+%y = alloca i32, align 4
+%z = alloca i32, align 4
+store i32 2, i32* %x, align 4
+store i32 3, i32* %y, align 4
+store i32 4, i32* %z, align 4
+%1 = load i32* %z, align 4
+%2 = sext i32 %1 to i64
+%3 = load i32* %y, align 4
+%4 = sext i32 %3 to i64
+%5 = load i32* %x, align 4
+%6 = sext i32 %5 to i64
+%7 = getelementptr[10x[20x[30xi32]]]*@Array3,i32 0,i64 %6,i64 %4,i64 %2;getting an element
+store i32 7, i32* %7, align 4;setting an element
+\end{llvmcode}
 
@@
 where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indices that 
-calculates the address of element.
+calculates the address of element. This rule also supports to translate 
+expressions of arrays.
 
@@
 &  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
-&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0, i32  \PH{v_2}\nl"}  \sep v \sep t\\
+&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \ListOf{(,\PH{t_2} \PH{v_2})}\nl"}  \sep v \sep t\\
 \end{align*}
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+\begin{pascalcode}
  array1 := array2;
  structure1 := structure2;
-\end{minted}
+\end{pascalcode}
 

### 2014-01-04 DOCUMENTATION/AdvancedTypes.tex
Adding support/explanation about the different types of Array %- Declaring the type of array  a : (INT X INT) --> INT %- Declaring a array dynamically {1|->2|->3, 2|->3|->4}

@@
 respective set.}
-\section{The LLVM grammar with support to array multidimensional}
+\VGM{The LLVM grammar with support to array multidimensional.}
 
@@
 \begin{itemize}
-	\item $array1 \in (0..10) \to INT $
-	\item $array2 \in (0..10) \times (0..10) \to INT $   (a multidimensional example)
-	\item $array3 \in BOOL \times (0..10) \to INT$ (a domain with two index types)
+	\item $array1 \in (0..10) \to INT $	(a simple array)
+	\item $array2 \in (0..10) \times (0..10) \to INT $   (a multidimensional array)
+	\item $array3 \in BOOL \times (0..10) \to INT$ (an array with two index types)
+	\item $\{1\vdash2\vdash3, 2\vdash3\vdash4\}  \in (INT \times INT) \times INT$ (an explicit array)
 \end{itemize}
@@
 
+%TODO: Explain the other rules of Array 
+%- Declaring the type of array  a : (INT X INT) --> INT
+%- Declaring a array dynamically {1|->2|->3, 2|->3|->4}
+
 \subsubsection{Getting a element from array}
@@
 \begin{ccode} 
-  int x=2; int y=3 ; int z=4 ;
+  int x=2; int y=3; int z=4;
   Array3[x][y][z]=7;
@@
 
-\paragraph{Restriction on array and structure:} It is not yet supported the 
-translation of direct attribution between array expression, for example:
+\paragraph{Restriction on array and structure:} The translation of direct attribution
+ between array expression is not yet supported, for example:
 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
-
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
 MACHINE Rec
@@
 
-  res<--positive = PRE xx'balance > 0 THEN res:=TRUE END;
+  res<--positive = 
+  PRE xx'balance > 0 THEN
+  res:=TRUE
+  END;
 
-  withrdaw(amt) = PRE amt:NAT1 & (xx'balance >= amt) THEN
+  withrdaw(amt) = 
+  PRE amt:NAT1 & (xx'balance >= amt) THEN
      xx := rec(name:xx'name, balance:(xx'balance - amt))
@@
 
-  unsafe_dec = BEGIN
+  unsafe_dec = 
+  BEGIN
        xx'balance := xx'balance - 1
@@
 END
-
 \end{minted}
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
 IMPLEMENTATION
@@
 OPERATIONS
-
   res<--positive = 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
 IMPLEMENTATION 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
+\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
 MACHINE

### 2014-01-08 DOCUMENTATION/specification.tex
Added a file containing the specification of the code generation. It synthesizes information provided elsewhere before (translation.tex).

--- 2013-10-24 translation.tex
@@
 \pagestyle{headings}
-\title{Translation of B Implementations to the LLVM: Control and Modularity}
+\title{Translation of B Implementations to the LLVM: Specification}
 \author{David Déharbe \and Valério Medeiros Jr}
-\institute{Program of Graduate Studies in Systems and Computing (PPgSC) \\
-  Federal University of Rio Grande do Norte \\
+\institute{Federal University of Rio Grande do Norte \\
   Natal (Brazil)}
@@
 
-\begin{abstract}
-  The aim of this work is to lay the foundations of a multi-platform code
-  generator for the B method. In particular, this paper presents a translation
-  procedure from a large subset of the B language for implementations towards
-  LLVM source code. This translation is defined formally as a set of rules
-  defined recursively on the abstract syntax for B implementations. It handles
-  the following elements of the B language: simple data types, imperative
-  instructions and component compositions.
-\end{abstract}
-
-\section{Introduction}
-
-B is a formal refinement-based software design method~\cite{Abrial1996}. It has
-a single language encompassing abstract constructs suitable for specification,
-as well as classic imperative constructs for computer programming. The starting
-point of a B development is a specification, called a \emph{machine}, that is
-incrementally refined to an \emph{implementation}, where only imperative-like
-constructs may be employed~\cite{Clearsy}. Such implementation may then be
-translated to source code in a programming language, say C~\cite{ComenC} or
-Ada. All steps in the B method are formally verified using certified theorem
-proving technologies. However the translation to some programming language and
-its subsequent compilation to the target platform do not benefit from the same
-mathematical rigor. In practice, redundancy in the tool chains and execution
-platform are employed to increase the level of confidence to the desired levels.
-
-The goal of this work is to connect the B method to the LLVM compilation
-framework~\cite{Lattner04LLVM}.  LLVM is a free, open-source project upon which
-many compiling contemporary technologies and tools are based. LLVM provides an
-intermediate assembly language upon which may be applied techniques such as
-optimization, static analysis, code generation, debugging, etc. Providing a
-connection from the B method to this framework gives the users of the B method
-access to these technologies. Our approach is to define a translation from B0
-(the subset of the B language that is used to describe imperative programs) to
-the LLVM intermediate format.
-
-Considering the correctness of the proposed translation, we shall rely upon
-human inspection for the time being. The standard uniform semantic framework for
-B and LLVM that would be necessary to carry out a proof of this correctness does
-not exist (yet). A possible approach to mitigate the risk of having an error in
-the translation would be to include assertions in the target code. Indeed the
-artifacts produced in the B development have many such assertions and it would
-be possible to translate at least some of them (B has an assertion instruction
-that introduces additional verification conditions in the development). The
-produced code could then be (at least partially) verified by testing the absence
-of assertion violations.
-
-\paragraph{Overview.} Following this introduction, the paper is organized as
-follows. The implementation constructs of the B language, and selected elements
-of the LLVM are presented in section~\ref{sec:b-impl}
-and~\ref{sec:llvm}. Section~\ref{sec:overview} presents a bird's-eye view that
-is then detailed in sections~\ref{sec:module} (for component modularity)
-~\ref{sec:data} (for data aspects),~\ref{sec:expr} (expressions) and
-instructions~\ref{sec:control}. We conclude and discuss future work
-in~\ref{sec:conclusion}.
-
-\section{Implementation Language in the B-method}
-\label{sec:b-impl}
-
-This section describes the B implementation constructs that are considered in
-the compilation to the LLVM. This description introduces the notations used in
-the definition of the translation. The elements of the B language are denoted
-using a \B{sans serif} font.
+\section{Description of the input of the code generator}
+\label{sec:b-ast}
+
+We detail here the constructs that are considered in the compilation to the
+LLVM. This description introduces the notations used in the definition of the
+translation. The elements of the B language are denoted using a \B{sans serif}
+font.
 
@@
 \B{op}, a set of operations, each being an algorithmic description of the
-different functionalities provided by the implementation.\DD{Include the
-possibility that the import is a machine without implementation. Instead it
-has a list of typed, implementable, variables.}
+different functionalities provided by the implementation.
+
+We distinguish the base module machines from developed module machines by
+testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
+the machine is basic, otherwise it is the root of the corresponding
+implementation.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[t]
   \begin{center}
@@
 
-LLVM aims to be a general, language agnostic, intermediate representation for
-compilers. The compiler front-end translates the source programming language to
-LLVM source code, upon which optimization and other static analysis tasks may be
-performed, and the compiler back-end translates to the target platform assembly
-language. One key feature of LLVM is that it is a single-static assignment (SSA)
-language: a variable may only be assigned once. So for each assignment to a
-variable found in the source language, the front-end shall generate a new
-temporary variable in the LLVM program. For illustration,
-figure~\ref{fig:ex-llvm} presents a LLVM program (for clarity, the corresponding
-C code appears on the left) with two temporary variables \llvm{\%0} and
-\llvm{\%1}.
-\begin{figure}
-\vspace*{-3em}
-  \begin{minipage}[t]{.3\textwidth}
-    \begin{tabbing}
-      foo \= \kill  \\
-      void inc(int * pi) \{\\
-      \> *pi += 1; \\
-      \}\end{tabbing}
-  \end{minipage}
-  \hfill
-  \begin{minipage}[t]{.6\textwidth}
-    \begin{tabbing}
-      foo \= \kill  \\
-      \llvm{define void @inc(i32* \%pi) nounwind \{} \\
-      \llvm{entry:}                      \\
-      \> \llvm{\%0 = load i32* \%pi}     \\
-      \> \llvm{\%1 = add i32 \%0, 1}     \\
-      \> \llvm{store i32 \%1, i32* \%pi} \\
-      \> \llvm{ret void} \\
-      \llvm{\}}
-    \end{tabbing}
-  \end{minipage}
-  \caption{Simple example of a C function and its corresponding LLVM function.
-    The first line contains the signature: return type \llvm{void}, the name
-    \llvm{@inc} and one parameter named \llvm{\%pi} and typed \llvm{i32*}.  Next
-    is the body with a single block, labeled \llvm{entry}, and temporary
-    variables \llvm{\%0} and \llvm{\%1}, created in the conversion to SSA. The
-    block has four instructions: \llvm{load}, \llvm{add}, \llvm{store} and
-    \llvm{ret}. For instance, \llvm{\%1 = add i32 \%0, 1} is an addition
-    (\llvm{add}), has result type \llvm{i32} and assigns to
-    \llvm{\%1} the sum of variable \llvm{\%0} and integer
-    literal \llvm{1}.}
-  \label{fig:ex-llvm}
-\end{figure}
-
-The introduction to the LLVM that follows is focused in the features
-that are relevant for the representation of B
-implementations. Figure~\ref{fig:llvm-grammar} presents a grammar of this subset
-of the LLVM.
+Figure~\ref{fig:llvm-grammar} presents a grammar of this subset of the LLVM.
 
@@
 
-LLVM programs are organized into modules, one per translation unit. A module may
-contain declarations of external entities (functions, constants) and definitions
-of internal items (functions, variables, constants). All data is explicitly
-typed. The name and type of external entities must be declared.  All names,
-e.g. non-reserved identifiers, shall start with \llvm{@}, when they are global,
-or\llvm{\%}, when they are local. For instance, \llvm{@max = external constant
-  i32} declares \llvm{@max} as a 32-bit integer constant and \llvm{declare void
-  @inc(i32*)} declares \llvm{@inc} as a function having as parameter a pointer
-to an integer and \llvm{void} as a return type.
-
-The type system contains the empty type \llvm{void}, an infinite, countable
-number of integer types, one for each possible bit width (e.g. \llvm{i8} is the
-type for 8-bit integers), and type constructors pointer (monadic \llvm{$\cdot$
-  *}) and structure (polyadic \llvm{\{$\cdots$\}}). For instance \llvm{\{ i8*,
-  i8, i8 \}} is the type for structures with three fields, the first having as
-type pointer to \llvm{i8}. Grammar rule \nt{type\_def} states how types are
-named, e.g.: \llvm{\%T1 = type \{i32, i32\}} and \llvm{\%T2 = type \{\%T1 *,
-  \%T1 *\}}. Note that there is no \llvm{void *} type in the LLVM, instead
-pointer values are of integer types.
-
-Local entities are constants, variables or functions. An example of constant
-definition is \llvm{@secret = constant i32 42}, it is composed of a name, type
-and value. The elements of a variable definition are the name, type and code
-generation attributes, e.g \llvm{@count = common global i32 zeroinitializer}.
-Attributes \llvm{common}, \llvm{global} and \llvm{zeroinitializer} provide
-information for target code generation: linkage type (the variable is merged
-with other of the same name), the scope (global) and initialization of the
-variable (zeroed is mandatory for common linkage). LLVM defines other attributes
-for these aspects, but they are not needed in this work and their presentation
-is omitted. For each such definition, a memory block is statically allocated to
-stores the variable value. Function definitions are composed of the signature
-and body. The signature contains the return type, name, parameters, and
-attributes for target code generation. Attribute \llvm{nounwind} states that the
-function never generate exceptions. The body is a sequence of blocks of
-instructions in single-static assignment (SSA) form, i.e. are such that each
-variable may be assigned exactly once.
-
-The grammar rule \nt{inst} describes the different kind of instructions. All
-instructions producing a value assign it to a fresh variable.  Instruction
-\llvm{alloca} allocates a memory block the size of the given type on the stack
-segment. This memory is automatically freed when the current frame is popped
-from the stack. The LLVM does not include instructions for allocating from the
-heap. Arithmetic operations are binary, and include signed quotient and
-reminder. Arithmetic comparisons compare two values according to the first
-operand and return a 1-bit integer value.  Instruction \llvm{call} invokes the
-given function with the given arguments and assigns the result to a fresh
-variable, which is of the given type.  In general, \llvm{getelementptr} gets the
-address of an element in an aggregate object through indexing. In the target
-LLVM subset, aggregate objects are structures, the first index has type
-\llvm{i32} and value \llvm{0} to select the first structure at the given
-location \nt{exp}, and the second index selects a field in that structure.
-Instruction \llvm{load} assigns to a fresh variable \nt{name} the contents of a
-memory address of type \nt{type} specified by \nt{exp} (e.g. in
-fig.~\ref{fig:ex-llvm}).  Instruction \llvm{store} writes a value to memory
-address (again, see fig.~\ref{fig:ex-llvm}).  Instruction \llvm{br} has two
-forms.  In the first form, it directs conditionally the control flow to one of
-two blocks.  If the given expression evaluates to \llvm{1}, control goes to the
-first block, otherwise to the second. In the second form, the control jumps
-unconditionally to the given block.
-% Instruction \llvm{switch} directs the
-% control flow to one of several blocks. The given expression is evaluated and
-% control goes to the first branch with the matching value, or to the default
-% branch (appearing first in the instruction).
-Finally, instruction \llvm{ret}
-ends the current function call, optionally returning a value.
-
-The expression language is limited to names (local and global), integer literals
-and selection of an element in a structure.
-
-To conclude, we make no assumption on the existence of a library to obtain
-resources managed by the operating system, such as dynamic memory allocation. In
-the absence of such resource, all data must be allocated either statically, or
-on the current stack frame, by using the \llvm{alloca} instruction.
-
-\section{Translation: a global picture}
+\section{Code generation modes}
 \label{sec:overview}
 
-In addition to producing LLVM code that is faithful to the design developed in
-B, our code generator is designed to satisfy the following two requirements:
+Code generation modes are:
 \begin{enumerate}
-\item Memory is allocated statically, save for the function call frame stack: no
-  dynamic memory allocation library is required. Many safety critical systems
-  preclude use of this programming facility.
-\item The code generator features separate compilation: an internal change in a
-  component should only require generating new LLVM code for that component, and
-  not for the other components that depend on it. This condition is important
-  for large, distributed projects.
+\item Code generation aims at producing a LLVM module,
+  that will later be instantiated when generating the code for a project. We
+  call this the \emph{COMP} mode for code generation.
+\item Code generation is applied to build a full system from a project. Given
+  the root module of the project, the code generator must identify all the
+  transitively imported module instances. We call this the \emph{PROJ} mode for
+  code generation.
 \end{enumerate}
 
-In Atelier-B, software development is organized in libraries of modules. A
-module has a specification, called a machine. A machine may be:
-\begin{description}
-\item[basic] In that case its implementation is carried out outside the realm of
-  the B method. Even though no LLVM code needs to be generated to implement the
-  basic machines, the code generator still needs to produce LLVM declarations of
-  the entities (e.g., constants and operations) that are supplied by those
-  machines.
-\item[developed] In that case, this module also contains a series of
-  refinements, eventually reaching an implementable version, from which LLVM
-  code may be generated.
-\end{description}
-We distinguish basic machines from developed machines in the abstract syntax by
-testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
-the machine is basic, otherwise it is the root of the corresponding
-implementation.
-
-Amongst the several constructs of the B notation supporting modular design, the
-\emph{import} relation is related to separate compilation. At the implementation
-level, one module may import several instances of a machine to construct its
-internal data structures. Such imported machines may be either basic, or
-developed. In the latter case, the implementation of the imported machines may
-in turn import other machine instances and there is no pre-established limit to
-such chain of imports. Hence, the import relation between module instances forms
-a tree, where the root is an implementation and the descendants of a node are
-the instances imported from the module in that node.
-
-\begin{figure}
+\begin{figure}[t]
 \begin{center}
- \includegraphics[width=.8\textwidth]{B-project.pdf}
-\caption{Structure of the imports relation in a B project}
-\label{fig:structure}
+\newcommand{\myind}{\hspace*{2em}}
+\framebox{
+\begin{minipage}{.9\textwidth}
+\noindent\textbf{typedef section:}
+Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
+the data space of the module: \\
+\myind \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}} \\
+In case the data space of a module is empty, this section is empty and the
+type is not defined. \\
+\\
+\noindent\textbf{interface section:} 
+Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
+defined):\\
+\myind \llvm{\%M\$ref\$ = type \%M\$state\$*} \\
+The interface also contains declarations of one LLVM functions for each
+operation of the module. \\
+\myind \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
+If the data space of the module is not empty, a function definition
+corresponding to the initialisation is also emitted: \\
+\myind \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})} \\
+\\
+\noindent\textbf{implementation section:}
+Defines the functions declared in the interface (only for developed modules): \\
+\myind \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
+\myind \myind \llvm{\ListOf{\nt{block}}} \\
+\myind \myind \llvm{exit: ret void} \\
+\myind \llvm{\}}
+\end{minipage}
+}
 \end{center}
+\caption{Summary of the different sections and the pattern of LLVM code
+  composing them.}
+\label{fig:summary}
 \end{figure}
-Figure~\ref{fig:structure} illustrates this with the structure of a B project
-where an implementation \B{I0} of a specification \B{M0} imports one unnamed
-instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
-developed machine \B{M2}. Its implementation \B{I2} in turn imports two
-instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
-(which has no imports itself).  The generation of LLVM code for the project
-shall produce the representation for one instance of \B{I0}, two instances of
-\B{I2} and four instances of \B{I3}.
-
-At that point, let us remark that, when generating code for a given module, we
-need to consider the following two cases:
-\begin{enumerate}
-\item The code generation aims at producing a component, that will further be
-  instantiated by importations. We call this the \emph{COMP} mode for code
-  generation. This case is further split into two sub-cases:
-  \begin{enumerate}
-  \item The module is basic, and only declarations need to be emitted.
-  \item The module is developed, and declarations and definitions are produced.
-  \end{enumerate}
-\item The generated code corresponds to a full project. Given the root module,
-  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
-\end{enumerate}
-One key issue in the code generation is how memory shall be allocated to store
-the state of the modules. To satisfy our first initial requirement (no dynamic
-memory allocation), this allocation shall be static, that is to say the
-generator must include the code that creates the module instances when the
-project starts execution. Note that this cannot be done in \emph{COMP} mode as,
-at that point, we cannot assume any knowledge on the number of instances the
-module being compiled will there be at run time. Consequently, the
-responsibility for the static allocation of instances is assigned to the
-\emph{PROJ} mode, which shall also perform the proper bindings between the
-actual and formal instances in the generated code.
-
-To encode the data space of a module, the translation produces a structure type,
-named \llvm{\%M\$state\$}, such that the elements of the structure represent the
-variables and the instances of imported modules. To satisfy our second initial
-requirement (separate compilation), the instances cannot be part of the
-structure itself, instead references to those instances (encoded as pointers)
-are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
-To encode behavior of a module, each operation \B{op} is translated into the
-definition of a function named \llvm{@M\$op}. The parameter list of the
-functions thus produced contains one item for each input and output of the
-corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
-to pass the address of the instance encoding structure associated with that
-operation. Also the translation produces a special function \llvm{@M\$init\$}
-responsible for executing the initialization of \B{M}. The parameters of this
-function are the addresses of the instances found in the import tree of the
-module (including the module itself).
-
-Our approach consists in dividing the translation into three sections, and to
-combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
-a given module \B{M}, the sections are the following:
-\begin{enumerate}
-\item typedef: Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
-  the data space of the module:
-  \begin{tabbing}
-    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
-  \end{tabbing}
-  In case the data space of a module is empty, this section is empty and the
-  type is not defined.
-\item interface: Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
-  defined):
-  \begin{tabbing}
-    \llvm{\%M\$ref\$ = type \%M\$state\$*}
-  \end{tabbing}
-  The interface also contains declarations of one LLVM functions for each
-  operation of the module.
-  \begin{tabbing}
-    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}
-  \end{tabbing}
-  If the data space of the module is not empty, a function definition
-  corresponding to the initialisation is also emitted:
-  \begin{tabbing}
-    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
-  \end{tabbing}
-\item implementation: Defines the functions declared in the interface:
-  \begin{tabbing}
-    foo \= \kill
-    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
-    \> \llvm{\ListOf{\nt{block}}} \\
-    \> \llvm{exit: ret void} \\
-    \llvm{\}}
-  \end{tabbing}
-  This section is produced only for the developed modules.
-\end{enumerate}
-Based on the definition of these sections, the \emph{COMP\/} translation
-assembles them as follows:
+
+Figure~\ref{fig:summary} summarizes the three sections defined in our approach
+for the code generation. These sections are combined differently in
+\emph{PROJ\/} and \emph{COMP\/} mode. The \emph{COMP\/} translation assembles
+them as follows:
 \begin{center}
@@
 \end{center}
-
-The detailed presentation of the translation of the different elements of a B
-implementation is divided into:
-\begin{description}
-\item[Notation] and some auxiliary definitions are presented in sec.~\ref{sec:notation}.
-\item[Modularity] constructs to combine B components are described in
-  sec.~\ref{sec:module}.
-\item[Data] translation is briefly described sec.~\ref{sec:data}, including
-  the encoding of B data types.
-\item[Expressions] and conditions are codified as LLVM instruction blocks, as
-  described in section~\ref{sec:expr}.
-\item[Control] flow, instructions are translated to LLVM instructions as
-  discussed in sec.~\ref{sec:control}, which also presents the encoding
-  of B operations and initialization.
-\end{description}
 
@@
 
-\section{Validation}
-\label{sec:validation}
-
-Before implementing this translation, we want validate the rules that define it.
-For this, we follow a pragmatic approach, based on testing and simulation of the
-generated LLVM artifacts, and using the existing LLVM tool set. Consequently,
-the rules described in the preceding sections have been prototyped in the Python
-programming language. The input of the translation is in a XML-based format,
-named BXML, that is generated by the forthcoming version of Atelier-B. For each
-B component (machine or implementation), Atelier-B generates a BXML file that
-may then be used as input by our prototype. The output of the translation is in
-LLVM-IR format. Such output may then be processed by the LLVM tools either to
-generate binary in some compatible target platform or LLVM bitcode that may then
-be executed using the LLVM interpreter. We used the latter approach to validate
-our translation.
-
-The LLVM tools that are employed are:
-\begin{description}
-\item[llvm-as] The LLVM assembler generates LLVM bitcode, invoked, e.g., as
-  follows:
-\begin{verbatim}
-llvm-as example-01.llvm -o example-01.bc
-\end{verbatim}
-\item[llvm-link] The LLVM bitcode linker, which may be invoked as follows:
-\begin{verbatim}
-llvm-link example-01.bc example-02.bc -o prog.bc
-\end{verbatim}
-\item[lli] The LLVM bitcode interpreter, which may be invoked as follows:
-\begin{verbatim}
-lli-mp example-01.bc
-\end{verbatim}
-\end{description}
-
-We describe below the different commands required to use the LLVM framework to
-animate a B implementation \B{wd} that uses an instance of a B module
-\B{counter}:
-\begin{verbatim}
-b2llvm counter counter.llvm
-b2llvm wd counter.llvm
-b2llvm --mode proj wd init-wd.llvm
-llvm-as counter.llvm -o counter.bc
-llvm-as wd.llvm -o wd.bc
-llvm-as init-wd.llvm -o init-wd.bc
-llvm-as scaffold-wd.llvm -o scaffold-wd.bc
-llvm-link counter.bc wd.bc init-wd.bc scaffold-wd.bc -o anim-wd.bc
-lli anim-wd.bc
-\end{verbatim}
-
-\section{Conclusion and future work}
-\label{sec:conclusion}
-
-This paper presents the formal specification of a translation from a large
-subset of the B implementation language to LLVM, a modern compiler internal
-language.  Even though this is still a work in progress, the definition is
-self-contained and has a large enough scope to be applied to B implementations
-where the data belongs to basic types. This specification is a blueprint for a
-code synthesis tool for B that is currently being implemented. This tool will be
-distributed as an extension to Atelier-B under an open-source license.
-
-The next step is to extend the scope to the full B implementation language. This
-entails that the translation must be defined for aggregate data types and
-adapted to handle some syntactic sugar. We are also planning for producing a
-LLVM output with debugging information. Such output would be indeed very helpful
-to provide feedback to the user when applying testing to validate the produced
-code.
-
-To prove the verification of the correctness of the translation, we would have
-to define the semantics of both LLVM and B in a unified framework. One possible
-starting point is Vellvm~\cite{vellvm}, a framework to reason about the
-correctness of LLVM programs and transformations. We would then have to extend
-the framework to include B implementation language. Another possible approach
-would be to translate verification conditions from the B development artifacts
-as assertinons in the generated LLVM code. The compiled program would include
-checks that such assertions hold while executing.
-
-\bibliographystyle{plain}
-\bibliography{b2llvm}
-
-\newpage
-\appendix
-
-\section{Test summary}
-
-For each test file, indicates the B0 constructs found in the implementation.
-\begin{center}
-\newcommand{\no}[0]{--}
-\newcommand{\yes}[0]{$\circ$}
-\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
-\hline
-Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 & 08 \\
-\hline
-\hline
-\multicolumn{9}{|c|}{clauses} \\
-\hline
-import w.o prefix & \no & \no & \no & \no & \no & \no & \yes & \no \\
-import w prefix & \no & \no & \no & \no & \no & \no & \no & \yes \\
-variables & \yes & \yes & \yes & \yes & \yes & \yes & \no & \yes \\
-constants & \no & \yes & \no & \no & \yes & \no & \yes & \yes \\
-op. input & \yes & \no & \yes & \no & \no & \no & \no & \no \\
-op. output & \yes & \no & \yes & \no & \no & \no & \yes & \yes \\
-\hline
-\multicolumn{9}{|c|}{instructions} \\
-\hline
-block & \yes & \no & \yes & \yes & \no & \yes & \no & \yes \\
-local var & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
-identity & \no & \no & \yes & \no & \no & \no & \no & \no \\
-becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-operation call & \no & \no & \no & \no & \no & \no & \yes & \yes \\
-condition & \yes & \no & \no & \no & \yes & \no & \yes & \yes \\
-case & \no & \no & \no & \no & \no & \yes & \no & \no \\
-assertion & \no & \no & \no & \no & \no & \no & \no & \no \\
-sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes & \yes \\
-while & \no & \no & \no & \yes & \no & \no & \yes & \no \\
-\hline
-\multicolumn{9}{|c|}{types} \\
-\hline
-INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-BOOL & \yes & \yes & \no & \no & \yes & \no & \yes & \yes \\
-\hline
-\multicolumn{9}{|c|}{expressions} \\
-\hline
-int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
-bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes & \yes \\
-vars & \yes & \yes & \yes & \yes & \no & \yes & \no & \yes \\
-loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
-const & \no & \yes & \no & \no & \yes & \no & \yes & \yes\\
-succ & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
-pred & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
-+ & \yes & \no & \no & \yes & \no & \yes & \yes & \no \\
-- & \yes & \no & \no & \no & \no & \no & \yes & \no \\
-* & \no & \no & \no & \no & \no & \yes & \no & \no \\
-\hline
-\multicolumn{9}{|c|}{formulas} \\
-\hline
-and & \no & \no & \no & \no & \yes & \no & \no & \no \\
-or & \no & \no & \no & \no & \yes & \no & \no & \no \\
-not & \no & \no & \no & \no & \yes & \no & \no & \no \\
-$<$ & \yes & \yes & \no & \no & \no & \no & \yes & \yes \\
-$>$ & \no & \no & \no & \no & \yes & \no & \yes & \no \\
-\hline
-\end{tabular}
-\end{center}
-
 \end{document}

### 2014-01-13 DOCUMENTATION/AdvancedTypes.tex
Fixed the rule to support array and added an example in coverage.sh

@@
 &  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
-&  \LET p_2 \sep v_2 \sep t_2  = \trad{n.indices}{\ListOf{Exp}}  \IN \\
+&  \LET r \sep p_2  = \trad{n.indices}{LRExp}  \IN \\
 &  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
-&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \ListOf{(,\PH{t_2} \PH{v_2})}\nl"}  \sep v \sep t\\
+&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \PH{r}\nl"}  \sep v \sep t\\
+\end{align*}
+
+A list of type and variable is generated by the following rule.
+%TODO: check this rule with David
+
+\begin{align*}
+&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
+&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
+ \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
+& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
 \end{align*}

### 2014-01-13 coverage.sh

@@ -29,6 +29,7 @@ $COV_RUN ./b2llvm.py -p -t -m comp counter counter.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -p -t -m comp timer timer.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -p -t -m comp wd wd.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -p -t -m proj wd wd.llvm bxml project.xml
+#$COV_RUN ./b2llvm.py -p -t -m proj array array.llvm bxml project.xml
 echo "Combining coverage data."
 coverage combine
 echo "Generating coverage report."

### 2014-02-18 DOCUMENTATION/AdvancedTypes.tex
Created small introduction about shift/interval of array. Wrote about initialisation of array …

@@
 	\B{base} & : &  \B{Array} \\
-	\B{indices} & : & seq \B{Expr}\\
+	\B{indexes} & : & seq \B{Expr}\\
 	\B{type} & : & \B{ArrayType}\\
@@
 	\item $array1 \in (0..10) \to INT $	(a simple array)
-	\item $array2 \in (0..10) \times (0..10) \to INT $   (a multidimensional array)
-	\item $array3 \in BOOL \times (0..10) \to INT$ (an array with two index types)
-	\item $\{1\vdash2\vdash3, 2\vdash3\vdash4\}  \in (INT \times INT) \times INT$ (an explicit array)
+	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
+	\item $\{1\vdash1\vdash4, 1\vdash2\vdash5\}  \in (INT \times INT) \times INT$ (an initialised array)
 \end{itemize}
-
-\paragraph{Attributions between array and structure}
-
-The B0 language supports attribution directly between arrays or structure, even though the attribution
-must have the same domain type and range type.
 
@@
 
-\subsubsection{Defining an array}
+\subsubsection{Declaring an array}
 
-The essential elements of array are its name, size and type. The
-following code declares an array global called ``Array3'' with 6000 (10*20*30)
-integer numbers with 32 bits and initialized with zero.
+The essential elements of the array are its name, size and type. 
+The size of ``e'' array with the indexes (i1, ... , in) are defined according 
+to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
+e.in\_start+1)$.
+
+The following code declares a multidimensional global array called ``Array2'' 
+with 6000 (10*20*30) integer numbers with 32 bits and initialized with zero.
 
 \begin{llvmcode}
-@Array3 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
+@Array2 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
 \end{llvmcode}
 
-Some type examples are:
-\begin{itemize}
-	\item \verb|[25 x i32]| - Array with 25 elements of 32-bit integer values.
-	\item \verb|[4 x [5 x [6 x i16]]] | - Multidimensional array with (4 x 5
-x 6) elements of 16-bit integer values.
-\end{itemize}
+\VGM{It is not implemented yet, because it requires generating an XML example.}
+The following example initialises a new array in C:
 
-%TODO: Explain the other rules of Array 
-%- Declaring the type of array  a : (INT X INT) --> INT
-%- Declaring a array dynamically {1|->2|->3, 2|->3|->4}
+\begin{ccode}
+int32_t arr[1][2][1]= {{{4},{5}}};
+\end{ccode}
+
+The syntax rule used to initialize an array in LLVM is:\\
+$@<var\_name> = constant [<size> x <type>] [<type> <value>]$.\\
+The translation of the last example is:
+
+\begin{llvmcode}
+@arr=constant[1 x [2 x [1 x i32]]] [[2 x [1 x i32]] [[1 x i32] [i32 4],[1 x i32] [i32 5]]]
+\end{llvmcode}
 
 \subsubsection{Getting a element from array}
-\VGM{Maybe it can be summarized with text of section 3}
 
@@
 \begin{llvmcode}
-  %7 = getelementptr [10 x [20 x [30 x i32]]]* @Array3, i32 0, i64 2, i64 3, i64 4
+%7 = getelementptr [10 x [20 x [30 x i32]]]* @Array2, i32 0, i64 2, i64 3, i64 4
 \end{llvmcode}
@@
 indexes to select the element in the data type. The second parameter (i32 0)
-selects the first element in the pointer (* @Array3). The last elements 
-(i64 2, i64 3, i64 4) selects the value from position (2,3,4) storing it in \%7.
-This array is multidimensional then several index must be declared separated
+selects the first element in the pointer (* @Array2). The last elements 
+(i64 2, i64 3, i64 4) select the value from position (2,3,4) storing it in \%7.
+This array is multidimensional, so several indexes must be declared separated
 by comma to get the specific element.
-
-\iffalse
-\begin{verbatim}
-<temporary variable> = getelementptr  
-  [<number of elements> x <element type>]   * @<variable_name>, 
-  <type and index of pointer> <index_value>,
-  <type and index. ex:i64>  <index_value>
-\end{verbatim}
-\fi
 
@@
 int x=2; int y=3; int z=4;
-  Array3[x][y][z]=7;
+Array2[x][y][z]=7;
 \end{ccode}
@@
 %6 = sext i32 %5 to i64
-%7 = getelementptr[10x[20x[30xi32]]]*@Array3,i32 0,i64 %6,i64 %4,i64 %2;getting an element
+%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
 store i32 7, i32* %7, align 4;setting an element
@@
 
+The element access of array can need a shift in indexes. When the domain of 
+array is an interval and it is not started from zero, it must adjust the 
+indexes shifting the access, because the addressing  of array in intermediate 
+representation LLVM starts at 0.
+
 % Rule to get a element in array, it can be used in become equal or Expressions
@@
 \paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
-where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indices that 
+where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indexes that 
 calculates the address of element. This rule also supports to translate 
@@
 &  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
-&  \LET r \sep p_2  = \trad{n.indices}{LRExp}  \IN \\
+&  \LET r \sep p_2  = \trad{n.indexes}{LRExp}  \IN \\
 &  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
@@
 A list of type and variable is generated by the following rule.
-%TODO: check this rule with David
-
+%TODO: insert a rule to do the shift address with the first element of interval
 \begin{align*}
@@
 \end{align*}
+
+%TODO: support to shift between interval
+\begin{itemize}
+	\item $array2 \in (10..50) \times ((-10)..100) \to INT $   (a multidimensional array)
+\end{itemize}
+
+The array indexes can composed of only primitive types (INTEGER and BOOLEAN). To access an element the formula is $address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$.
+
+%TODO create a new rule based on the below rule.
+
+%it must change the declaration and the rule of ArrayItem
 
@@
 \end{pascalcode}
+
+The B0 language supports attribution directly between arrays or structures, even though the attribution
+must have the same domain type and range type.
 

### 2014-02-18 b2llvm/translate.py
Fixing the size of array.

@@ -396,7 +396,7 @@ def x_arrayType(t):
     tr =""
     domain = t.get("dom")
     for elem in domain:
-        size =int(elem.get("end")) - int(elem.get("start"))
+        size =int(elem.get("end")) - int(elem.get("start"))+1
         tl += "[ "+str(size)+ " x "
         tr += "]"
     return tl + ranType + tr
@@ -940,7 +940,6 @@ def x_lvalue(buf, n):
     if n["kind"] == "arrayItem": 
         buf.trace.tab()
         v1,t1 = x_expression(buf, n["base"])
-         
         buf.trace.out("Variable array (base) \""+n["base"]["id"]+"\" is stored at position "+v1+" of \"%self$\". (arrayItem)")
         buf.trace.untab()
         
@@ -948,7 +947,8 @@ def x_lvalue(buf, n):
         #TODO: Create the function LRExp to getting a sequence of selected elements.
         #commas([ term(x) for x in n["lhs"]])
         
-        buf.trace.tab()
+        buf.trace.tab() 
+        #for elem in n["index"]:
         vi,ti = x_expression(buf, n["index"])
         v = names.new_local()
         buf.code(LOADI,v,t1+"*",v1,vi)
@@ -956,7 +956,6 @@ def x_lvalue(buf, n):
         buf.trace.out("Variable array (index) \""+printer.term(n["index"])+"\" is stored at position "+vi+" of \"%self$\". (arrayItem)")
         buf.trace.untab()
         
-        #TODO: Considering the interval (a..b):
         #TODO: Adjust the size o vector to size=(b-a+1)
         #TODO: Add suport to interval position(p) = (p-a)
         return (v, t)  

### 2014-02-18 DOCUMENTATION/AdvancedTypes.tex
All details about the translation of Array, included multidimensional and interval (shift support and cast types).

@@
         \hline
-	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} }\\
+	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
 	\hline
@@
 
-The element access of array can need a shift in indexes. When the domain of 
-array is an interval and it is not started from zero, it must adjust the 
-indexes shifting the access, because the addressing  of array in intermediate 
-representation LLVM starts at 0.
-
-% Rule to get a element in array, it can be used in become equal or Expressions
+The rule lvalue is:
 \begin{align*}
@@
 \begin{align*}
-&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
+&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{ShiftExp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
 &\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
@@
 
-%TODO: support to shift between interval
-\begin{itemize}
-	\item $array2 \in (10..50) \times ((-10)..100) \to INT $   (a multidimensional array)
-\end{itemize}
+The element access of array can need a shift in indexes. When the domain of 
+array is an interval and it is not started from zero, it must adjust the 
+indexes shifting the access, because the addressing  of array in intermediate 
+representation LLVM starts at 0. To access an element from array the formula is 
+$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The following rule 
+\footnote{The function $TSymbol$ returns the "start", "end" and type of the one 
+index.} makes the shift for each index of array. \VGM{TODO: The function TSymbol
+returns the "start", "end" and type of the one index.}
 
-The array indexes can composed of only primitive types (INTEGER and BOOLEAN). To access an element the formula is $address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$.
+\begin{align*}
+&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
+&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
+&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
+&\quad \ELSE \LET  p \sep v \sep v' \sep t'  = \trad{p,t,v,itype,istart}{Cast} \IN \\
+&\quad \llvm{",\PH{t} \PH{l} "} \sep  \llvm{"\PH{p} \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
+\end{align*}
 
-%TODO create a new rule based on the below rule.
+The \trad{}{ShiftExp} rule does a subtraction and this operation must have operators 
+with the same data types so the following cast rule is used:
 
-%it must change the declaration and the rule of ArrayItem
+\begin{align*}
+&\trad{p,t,v,t',v'}{Cast} \isdef \LET  \ell = \lbl \IN \\
+&\quad \IF ( t == t') \THEN p \sep v \sep v' \sep t  \\
+&\quad \ELSE \IF (t<t') \THEN \\
+&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t} \PH{v} , \PH{t'}\nl"}  \sep l \sep v' \sep t'\\ 
+&\quad \ELSE \\ 
+&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t'} \PH{v'} , \PH{t}\nl"} \sep v \sep l \sep t \\ 
+\end{align*}
 

### 2014-02-27 DOCUMENTATION/AdvancedTypes.tex
Adding structure support details.

@@
 \newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
+\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
+\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}
 
@@
 \newcommand{\VGM}[1]{\marginpar{\tiny{VGM: #1}}}
-
-\newcommand{\stateindex}[0]{\ensuremath{\sf\tau}}
 
@@
         \B{body} & : & seq \B{Inst} \\
-	\end{tabular}
-
-       \begin{tabular}[t]{rcl}   
         \hline
@@
         \B{body} & : & seq \B{Inst} \\
+\end{tabular}
+\begin{tabular}[t]{rcl}   
   	\hline
@@
         \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
-\end{tabular}
-
-  \begin{tabular}[t]{rcl}
 	\hline
@@
         \B{args} & : & seq \B{Pred} \\
+\end{tabular}
+
+\begin{tabular}[t]{rcl}
         \hline
@@
 	\hline
-%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
-%        \hline
-%	\B{id} & : &  \B{Name}\\
-%        \B{elements} & : & seq \B{Name} \\
-%	\B{type} & : & \B{Struct}\\
-%	\hline
+	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
+        \hline
+	\B{base} & : &  \B{Struct} \\
+	\B{element} & : & \B{Name}\\
+	\B{type} & : & \B{Type}\\
+	\hline
         \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
@@
 
-Two approaches are possible to support arrays. An approach is importing
-reusable libraries of AtelierB that are reported in the official document of
-ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
-This approach is supported, but the generated code is more complex and it needs
-more intermediate steps to manipulate the data of imported components. Moreover,
-this approach allows dynamic allocation for arrays, but it is strongly not
-recommended to safety-critical systems. A second approach enjoys the direct 
-correspondence between B0 native arrays and LLVM-IR and this approach is presented here.
+Two approaches are possible to support arrays and structures.
+An approach is importing reusable libraries of AtelierB that are reported in the official document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
+This approach is supported, but the generated code is more complex and it needs more intermediate steps to manipulate the data of imported components.
+Moreover, this approach allows dynamic allocation for arrays, but it is strongly not recommended to safety-critical systems.
+A second approach enjoys the direct correspondence between B0 native arrays and LLVM-IR and this approach is presented here.
 
+\section{Structure}
+The struct declaration defines a list of variables to be placed under one name, 
+allowing several variables to be accessed by one base name. This example get andset one value of xx'balance and initializes a structure.
+
+\begin{ccode}
+xx := rec(name:aa, balance:10) ;
+xx'balance := xx'balance -1;
+\end{ccode}
+
+The following rule get an element value from the struct.
+The same rule is used in left side of attribution and right side of attribution.
+The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.
+\begin{align*}
+\begin{split}
+  \trad{n}{\B{StructItem}} \isdef \\
+   & \LET \ell = \lbl \AND \PH{t} = n.type \AND \PH{v} = n.element \IN\\
+   & \llvm{"\PH{\ell} = getelementptr \%\PH{t} @\PH{v}, i32 0, i32  \SBPos(\B{n})"} \\
+   & \sep  \ell  \sep \SBType(\B{n})  \\
+\end{split}
+\end{align*}
+
+AST
+
+record structure : Record
+elements : seq Name;
+values : seq Expr;
+
+Isto é uma expressão
+
+To remove the restrictions:
+( ) The direct attribution between arrays  is madded in similar way
+
+\section{Arrays}
+%TODO: update according the new names in AST, just change the names.
 A concrete array is a total function where the domain indexes the range 
 elements. The domain is composed of Cartesian product of sequence of concrete 
-sets. The node Data array in AST can represent a declaration of array or a 
+sets. The node Array in AST represents a declaration of array or a 
 selection of element. The first element ``dec'' declares explicitly an array or 
@@
 The essential elements of the array are its name, size and type. 
-The size of ``e'' array with the indexes (i1, ... , in) are defined according 
+The size of ``e'' array with the indexes (i1,...,in) are defined according 
 to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
@@
 A list of type and variable is generated by the following rule.
-%TODO: insert a rule to do the shift address with the first element of interval
 \begin{align*}
@@
 
-The \trad{}{ShiftExp} rule does a subtraction and this operation must have operators 
-with the same data types so the following cast rule is used:
+The \trad{}{ShiftExp} rule does a subtraction and this operation must have operators with the same data types so the following cast rule is used:
 
@@
 
-\paragraph{Restriction on array and structure:} The translation of direct attribution
+\paragraph{Restriction on array:} The translation of direct attribution
  between array expression is not yet supported, for example:
-
 \begin{pascalcode}
  array1 := array2;
- structure1 := structure2;
 \end{pascalcode}
-
-The B0 language supports attribution directly between arrays or structures, even though the attribution
-must have the same domain type and range type.
 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
+\begin{ccode}
 MACHINE Rec
@@
 END
-\end{minted}
+\end{ccode}
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
+\begin{ccode}
 IMPLEMENTATION
@@
 END
-\end{minted}
+\end{ccode}
 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
-
+\begin{ccode}
 MACHINE
@@
 END
-\end{minted}
+\end{ccode}
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
+\begin{ccode}
 IMPLEMENTATION 
@@
    END
+END
+\end{ccode}
 
-END
-
-\end{minted}
-
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
-
+\begin{ccode}
 @Array__arr = internal global [100 x i32] zeroinitializer, align 16
@@
 }
-\end{minted}
+\end{ccode}
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
-
+\begin{llvmcode}
 define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
@@
 }
-\end{minted}
+\end{llvmcode}
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
-
+\begin{llvmcode}
 define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
@@
 
-\end{minted}
+\end{llvmcode}
 
@@
 
-\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
+\begin{ccode}
 MACHINE
@@
 END
-
-\end{minted}
+\end{ccode}
 

### 2014-03-02 DOCUMENTATION/specification.tex
Adding to TODO in specification.tex

@@
 translation of scalar constants consists in substituting them with their value.
-\DD{The translation of non-scalar constants will be defined in a next version of
-  this document.}  The definition of the translation for expressions is
+%TODO Look the following text. 
+The translation of non-scalar constants will be defined in a next version of
+this document. The definition of the translation for expressions is
 described next in section~\ref{sec:expr}.
@@
 
-[[[ DEPRECATED ]]]
+%TODO It is deprecated 
+%[[[TODO: DEPRECATED ]]]
 

### 2014-03-02 DOCUMENTATION/AdvancedTypes.tex
Adding structure support details.
The next step is review the AdvancedTypes documentation

@@
 	\hline
-	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} }\\
+	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\
         \hline
@@
         \hline
-        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Cons}}\\
-	\multicolumn{3}{c} {\B{Vari} \alt \B{Array} \alt \B{ArrayItem} } \\
+        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Record}}\\
+	\multicolumn{3}{c}{\B{Vari}\alt\B{StructItem}\alt\B{ArrayItem} } \\
         \hline
@@
         \B{args} & : & seq \B{Pred} \\
-\end{tabular}
-
-\begin{tabular}[t]{rcl}
 	\hline
@@
         \B{arg1}, \B{arg2} & : & \B{Exp} \\
+
+\end{tabular}
+
+\begin{tabular}[t]{rcl}
         \hline
@@
 	\B{type} & : & \B{Type}\\
+	\hline
+	\multicolumn{3}{|c|}{Record Structure : \B{Record}} \\
+        \hline
+	\B{elements} & : & seq \B{Name}\\
+	\B{values} & : & seq \B{Expr}\\
 	\hline
@@
 
-\section{Translation of Array and Structures}
+\subsection{Rules to be updated in first specification}
+
+\paragraph{Becomes equal instructions} must evaluate the target and the source
+of the assignment, and then copy the result of the latter in the former, using
+the LLVM instruction \llvm{store}. The evaluation of the target is defined with
+$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
+variable $p$ containing the assigned location. The evaluation of the source
+yields the corresponding code $r$, the variable $v$ holding the result, and its
+type $t$.
+
+\begin{align*}
+\begin{split}
+  \trad{a}{\B{Beq}} \isdef
+  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
+  & \llvm{"\PH{l}} \\
+  & \llvm{\PH{r}} \\
+  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
+\end{split}
+\end{align*}
+
+The rule lvalue is:
+\begin{align*}
+\lefteqn{\trad{n}{lvalue} \isdef} \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
+& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
+& \quad \IF \mbox{\B{n} is an array item} \THEN  \trad{n}{ArrayItem} \\
+& \quad \IF \mbox{\B{n} is a struct item} \THEN  \trad{n}{StructItem} \\
+& \quad \ELSE \LET v = \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
+\end{align*}
+
+\section{Translation of Structures and Array}
 
@@
 
-\section{Structure}
+\subsection{Structure}
 The struct declaration defines a list of variables to be placed under one name, 
@@
 The following rule get an element value from the struct.
-The same rule is used in left side of attribution and right side of attribution.
+The same rule is used in left side of attribution as a $LValue$ and right side of attribution as an $Expression$.
 The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.
+
 \begin{align*}
@@
 
-AST
-
-record structure : Record
-elements : seq Name;
-values : seq Expr;
-
-Isto é uma expressão
-
-To remove the restrictions:
-( ) The direct attribution between arrays  is madded in similar way
+The attribution from record structure to a structure variable, line 1 from last code example, is translated by a sequence of application of rule $StructItem$.
 
 \section{Arrays}
-%TODO: update according the new names in AST, just change the names.
 A concrete array is a total function where the domain indexes the range 
@@
 sets. The node Array in AST represents a declaration of array or a 
-selection of element. The first element ``dec'' declares explicitly an array or 
-references a variable array. The second element ``type'' defines a sequence of 
-array types, where the last element defines the range of function and the other 
-elements define the domain. The third element ``indexes'' is optional, and it 
-represents the selection of an element in array, or the same type structure 
-when it is a structure.
+selection of element.
 
@@
 
-\subsection{Rules}
-
-\paragraph{Becomes equal instructions} must evaluate the target and the source
-of the assignment, and then copy the result of the latter in the former, using
-the LLVM instruction \llvm{store}. The evaluation of the target is defined with
-$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
-variable $p$ containing the assigned location. The evaluation of the source
-yields the corresponding code $r$, the variable $v$ holding the result, and its
-type $t$.
-
-\begin{align*}
-\begin{split}
-  \trad{a}{\B{Beq}} \isdef
-  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
-  & \llvm{"\PH{l}} \\
-  & \llvm{\PH{r}} \\
-  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
-\end{split}
-\end{align*}
-
-The rule lvalue is:
-\begin{align*}
-\lefteqn{\trad{n}{lvalue} \isdef} \\
-& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
-& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an array} \THEN  \trad{n}{ArrayItem} \\
-& \quad \ELSE \LET v = \variable \IN \\
-& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
-& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
-\end{align*}
+\subsection{Rules for support Array}
 
@@
  between array expression is not yet supported, for example:
+
 \begin{pascalcode}

### 2014-03-05 C/AdvancedTypes/Array_i.c
Adding C examples

@@ -1,0 +1,115 @@
+#include "stdio.h"
+#include "Array.h"
+
+/* Clause CONCRETE_CONSTANTS */
+/* Basic constants */
+
+/* Array and record constants */
+/* Clause CONCRETE_VARIABLES */
+
+static int32_t Array2[10][20][30];
+static int32_t Array__arr[100];
+static int32_t Array__arr_n[100][100];
+static int32_t Array__tmp[100];
+
+void Array__setZero(void);
+void Array__setZero2(void);
+void Array__setZeroSimp(void);
+void Array__setArray(void);
+
+int main(void){
+   Array__setZero();
+   Array__setZero2();
+   printf("%d\n", Array2[2][3][4]);
+  return 0;
+}
+
+/* Clause INITIALISATION */
+void Array__INITIALISATION(void)
+{
+    Array__arr[1] = 100;
+    Array__arr_n[0][0] = 10;
+    memmove(Array__tmp,Array__arr,100 * sizeof(int32_t));
+    Array__setZero();
+
+}
+
+/* Clause OPERATIONS */
+void Array__setZero(void)
+{   int32_t x=2; int32_t y=3 ; int32_t z=4 ;
+    Array2[x][y][z]=7;
+}
+
+void Array__setZero2(void)
+{
+    int x =4;
+    x= x-2;
+    Array2[x][3][4]=7;
+}
+void Array__setArray(void)
+{
+    int32_t cprogram[3][2][4]={
+        {{0, 022, 03, 43}, {23, 47, -9, 2}},
+        {{0, 36, 45, 4}, {12, 24, 022, -1}},
+        {{8, 32, 342, 01}, {21, 32, 43, -20}}
+    };
+    int32_t Array3[1][2][1]= {{{4},{5}}};
+    /*{ { {1}, {2} , {3}},
+        
+              { {4}, {5} , {6}},
+              { {7}, {8} , {9}}
+             } ;*/
+}
+void Array__setZeroSuggestedByDavid(void)
+{  /* David gives a similar sugestion wit intermediate steps */   
+
+	Array__arr_n[2][5]=4;
+    	printf("%d\n",Array__arr_n[2][5]);     
+	int * mid = Array__arr_n[2];
+	printf("%d\n",mid[5]);
+}
+
+
+
+
+void Array__set(int32_t ix, int32_t tt)
+{
+    if(((((ix) >= (0)) &&
+            ((ix) <= (99))) &&
[…]

### 2014-03-05 C/AdvancedTypes/Rec_i_changed.c

@@ -1,0 +1,80 @@
+#include <stdio.h>
+#include "Rec.h"
+
+/* Clause CONCRETE_CONSTANTS */
+/* Basic constants */
+
+/* Array and record constants */
+/* Clause CONCRETE_VARIABLES */
+
+static struct R_1 { int32_t number;int32_t balance; double dd; } Rec__account;
+void Rec__set(void);
+void Rec__unsafe_dec(void);
+void Rec__get_and_set(void);
+typedef struct point point;
+/* Declare the struct with integer members x, y */
+struct point {
+   int    x;
+   int    y;
+};
+
+int main (){
+    
+    Rec__set();
+    printf("%d\n",Rec__account.balance);
+    Rec__get_and_set();
+    printf("%d\n",Rec__account.balance);
+    Rec__INITIALISATION();
+    printf("%lf\n",Rec__account.dd);
+}
+/* Clause INITIALISATION */
+void Rec__INITIALISATION(void)
+{
+    point p = {.y = 2, .x = 1};
+    point pi;// = {700,900};
+    pi.y=700;
+    pi.x=900;
+    Rec__account.number = 0;
+    Rec__account.balance = 10;
+    Rec__account.dd=7;
+
+}
+
+/* Clause OPERATIONS */
+
+void Rec__positive(bool *res)
+{
+    if((Rec__account.balance) > (0))
+    {
+        (*res) = true;
+    }
+    else
+    {
+        (*res) = false;
+    }
+}
+
+void Rec__withrdaw(int32_t amt)
+{
+    if((((amt) >= (0)) &&
+        ((amt) <= (2147483647))) &&
+    (((Rec__account.balance) >= (amt))))
+    {
+        Rec__account.number = Rec__account.number;
+        Rec__account.balance = (Rec__account.balance-amt);
+    }
+}
+void Rec__set(void)
+{
+    Rec__account.balance = 50;
+}
+void Rec__get_and_set(void)
+{
+    Rec__account.balance = Rec__account.balance + 50;
+}
+
+void Rec__unsafe_dec(void)
+{
+    Rec__account.balance = Rec__account.balance-1;
+}
+

### 2014-03-05 DOCUMENTATION/AdvancedTypes.tex
Doing the first review by me.

@@
 \author{David Déharbe, Valério Medeiros Jr.}
-\title{Array}
+\title{Specification of Advanced Types Translation}
 
@@
 The supported advanced types are array and structures. The nodes Array and 
-Structure can be used to declare a variable or to select an element in the array 
-or the structure.
+Structure can be used to declare a variable, or  declare an element, or to 
+select an element in the array or the structure.
 
@@
   \begin{center}
-    {\footnotesize
+    {\scriptsize
       \frame{
@@
         \hline
-        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Record}}\\
-	\multicolumn{3}{c}{\B{Vari}\alt\B{StructItem}\alt\B{ArrayItem} } \\
+        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
+	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
+        \hline
+        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
+        \hline
+	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\
         \hline
@@
         \B{args} & : & seq \B{Pred} \\
+\end{tabular}
+
+\begin{tabular}[t]{rcl}
 	\hline
@@
         \B{arg1}, \B{arg2} & : & \B{Exp} \\
-
-\end{tabular}
-
-\begin{tabular}[t]{rcl}
         \hline
@@
 	\hline
-  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
+  	\multicolumn{3}{|c|}{Data struct : \B{Struct}} \\
         \hline
@@
 	\hline
-	\multicolumn{3}{|c|}{Record Structure : \B{Record}} \\
+	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
         \hline
@@
 \end{figure}
-
-\VGM{To add in text: The nodes \B{Int\_lit} and \B{Bool\_lit} are literals in its 
-respective set.}
-\VGM{The LLVM grammar with support to array multidimensional.}
 
@@
 & \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an array item} \THEN  \trad{n}{ArrayItem} \\
-& \quad \IF \mbox{\B{n} is a struct item} \THEN  \trad{n}{StructItem} \\
+& \quad \IF \mbox{\B{n} is an ArrayItem} \THEN  \trad{n}{ArrayItem} \\
+& \quad \IF \mbox{\B{n} is a StructItem} \THEN  \trad{n}{StructItem} \\
 & \quad \ELSE \LET v = \variable \IN \\
@@
 
+The expressions have elements represented by pointers that need a special 
+care. The attributions between these elements are made  directly by the memory 
+address.  However, the expression resolution are made by parts, where the 
+partial results are stored to temporary variables. Therefore, the expression 
+resolution with pointers needs an additional instruction $LOAD$ for store the 
+partial results. This translation is represented by $ValueItem$.
+
+\begin{align*}
+\begin{split}
+  \trad{n}{\B{ValueItem}} \isdef \\
+   & \LET \ell = \lbl  \IN\\
+   & \IF \B{n} \mbox{ is \B{StructItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{StructItem}} \\
+   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
+   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
+   & \sep  \ell  \sep t'\\
+\end{split}
+\end{align*}
+
 \section{Translation of Structures and Array}
 
-Two approaches are possible to support arrays and structures.
-An approach is importing reusable libraries of AtelierB that are reported in the official document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
-This approach is supported, but the generated code is more complex and it needs more intermediate steps to manipulate the data of imported components.
-Moreover, this approach allows dynamic allocation for arrays, but it is strongly not recommended to safety-critical systems.
-A second approach enjoys the direct correspondence between B0 native arrays and LLVM-IR and this approach is presented here.
+Two approaches are possible to support arrays and structures. An approach is 
+importing reusable libraries of AtelierB that are reported in the official 
+document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
+This approach is supported, but the generated code is more complex and it needs 
+more intermediate steps to manipulate the data of imported components. 
+Moreover, this approach allows dynamic allocation for arrays, but it is 
+strongly not recommended for safety-critical systems. A second approach enjoys 
+the direct correspondence between B0 native types and LLVM and this approach 
+is presented here.
 
-\subsection{Structure}
-The struct declaration defines a list of variables to be placed under one name, 
-allowing several variables to be accessed by one base name. This example get andset one value of xx'balance and initializes a structure.
+\subsection{Structure} \VGM{It is not implemented yet, because it requires generating an XML example.}
+The structure declaration defines a list of variables to be placed under one 
+name, allowing several variables to be accessed by one base name. This example  
+initializes a structure, and gets and sets one value of xx'balance.
 
 \begin{ccode}
-xx := rec(name:aa, balance:10) ;
+xx := rec(id:1, balance:10);
 xx'balance := xx'balance -1;
 \end{ccode}
+The lines 1 and 2 are represented in translation in LLVM code:
 
-The following rule get an element value from the struct.
-The same rule is used in left side of attribution as a $LValue$ and right side of attribution as an $Expression$.
+\begin{llvmcode}
+%1 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 0
+store i32 1, i32 * %1
+%2 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+store i32 10, i32 * %2
+
+%3 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+%4 = load i32* %3
+%5 = sub nsw i32 %4, 1
+%6 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+store i32 %5, i32* %6
+\end{llvmcode}
+
+The $StructItem$ rule gets an element value from the struct.
+The same rule is used on the left side of attribution as a $LValue$ and right side of attribution as an $Expression$.
 The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.
@@
    & \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
-   & \llvm{"\PH{\ell} = getelementptr \%\PH{t} @\PH{v}, i32 0, i32  \SBPos(\B{n})"} \\
-   & \sep  \ell  \sep \SBType(\B{n})  \\
+   & \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
+%   & \llvm{\PH{\ell2}  = load  \PH{\SBType(\B{n})}* \PH{\ell1} \nl"} \\
+   & \sep  \ell  \sep \SBType(\B{n})\\
 \end{split}
@@
 
-The attribution from record structure to a structure variable, line 1 from last code example, is translated by a sequence of application of rule $StructItem$.
+The attribution from record structure to a structure variable, line 1 from last 
+B code example, is translated by a sequence of application of rule $StructItem$.
 
 \section{Arrays}
-A concrete array is a total function where the domain indexes the range 
-elements. The domain is composed of Cartesian product of sequence of concrete 
-sets. The node Array in AST represents a declaration of array or a 
-selection of element.
 
-Some examples of array are:
+The arrays store a fixed-size sequential collection of elements with the same type. 
+They consist of contiguous memory locations, where the lowest address (0) 
+corresponds to the first element and the highest address to the last element.
+In B, a concrete array is a total function where the domain indexes the range 
+elements. Declaration examples of arrays are:
 \begin{itemize}
@@
 
-\subsection{Relation arrays B0 and LLVM-IR}
+\subsection{Relation arrays B0 and LLVM}
 
@@
 
-This section shows the relation between arrays B0 and LLVM-IR. The sequence of
-steps are: defining  an array, getting an element and setting an
+This section shows the relation between arrays B0 and LLVM. The sequence of
+elements are defining an array, getting an element and setting an
 element.
@@
 The size of ``e'' array with the indexes (i1,...,in) are defined according 
-to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
-e.in\_start+1)$.
+to the formula:\\
+ $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - e.in\_start+1)$.
 
@@
 \VGM{It is not implemented yet, because it requires generating an XML example.}
-The following example initialises a new array in C:
+The following example initialises an new array in C:
 
@@
 
-The syntax rule used to initialize an array in LLVM is:\\
-$@<var\_name> = constant [<size> x <type>] [<type> <value>]$.\\
+The syntax rule\footnote{The underlined element type can be expanded in others types definitions.} used to initialize an array in LLVM is:\\
+$@<var\_name> = constant [<size> x \underline{<type>}] [<type> <value>]$.\\
 The translation of the last example is:
@@
 
-The LLVM-IR instruction \textit{getelementptr} is used to select an element in
+The LLVM instruction \textit{getelementptr} is used to select an element in
 the array or structure. A simple example is:
@@
 The following blocks illustrate a C code setting a value to array
-and its translation to LLVM-IR.
+and its translation to LLVM.
 
@@
 7 is stored in it.  
-
+\begin{figure}[h]
 \begin{llvmcode}
-%x = alloca i32, align 4
-%y = alloca i32, align 4
-%z = alloca i32, align 4
-store i32 2, i32* %x, align 4
-store i32 3, i32* %y, align 4
-store i32 4, i32* %z, align 4
-%1 = load i32* %z, align 4
+%x = alloca i32
+%y = alloca i32
+%z = alloca i32
+store i32 2, i32* %x
+store i32 3, i32* %y
+store i32 4, i32* %z
+%1 = load i32* %z
 %2 = sext i32 %1 to i64
-%3 = load i32* %y, align 4
+%3 = load i32* %y
 %4 = sext i32 %3 to i64
-%5 = load i32* %x, align 4
+%5 = load i32* %x
 %6 = sext i32 %5 to i64
 %7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
-store i32 7, i32* %7, align 4;setting an element
+store i32 7, i32* %7         ;setting an element
 \end{llvmcode}
+\end{figure}
 
@@
 
-\subsection{A  simple example and its LLVM-IR}
+\subsection{A  simple example and its LLVM}
 
@@
 define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
-  %1 = alloca i32, align 4
-  %2 = alloca i32, align 4
-  store i32 %ix, i32* %1, align 4
-  store i32 %tt, i32* %2, align 4
-  %3 = load i32* %1, align 4
+  %1 = alloca i32
+  %2 = alloca i32
+  store i32 %ix, i32* %1
+  store i32 %tt, i32* %2
+  %3 = load i32* %1
   %4 = icmp sge i32 %3, 0
@@
 ; <label>:5                                       ; preds = %0
-  %6 = load i32* %1, align 4
+  %6 = load i32* %1
   %7 = icmp sle i32 %6, 99
@@
 ; <label>:8                                       ; preds = %5
-  %9 = load i32* %2, align 4
+  %9 = load i32* %2
   %10 = icmp sge i32 %9, 0
@@
 ; <label>:11                                      ; preds = %8
-  %12 = load i32* %2, align 4
+  %12 = load i32* %2
   %13 = icmp sle i32 %12, 1000
@@
 ; <label>:14                                      ; preds = %11
-  %15 = load i32* %2, align 4
-  %16 = load i32* %1, align 4
+  %15 = load i32* %2
+  %16 = load i32* %1
   %17 = sext i32 %16 to i64

### 2014-03-06 C/DerivedTypes/Array.h
renaming folder

@@ -1,33 +1,0 @@
-#ifndef _Array_h
-#define _Array_h
-
-#include <stdint.h>
-#include <stdbool.h>
-#include <string.h>
-#ifdef __cplusplus
-extern "C" {
-#endif /* __cplusplus */
-
-
-/* Clause SETS */
-
-/* Clause CONCRETE_VARIABLES */
-
-
-/* Clause CONCRETE_CONSTANTS */
-/* Basic constants */
-/* Array and record constants */
-extern void Array__INITIALISATION(void);
-
-/* Clause OPERATIONS */
-
-extern void Array__set(int32_t ix, int32_t tt);
-extern void Array__read(int32_t ix, int32_t *tt);
-extern void Array__swap(int32_t ix, int32_t jx);
-
-#ifdef __cplusplus
-}
-#endif /* __cplusplus */
-
-
-#endif /* _Array_h */

### 2014-03-06 C/DerivedTypes/Rec_i.c

@@ -1,48 +1,0 @@
-
-#include "Rec.h"
-
-/* Clause CONCRETE_CONSTANTS */
-/* Basic constants */
-
-/* Array and record constants */
-/* Clause CONCRETE_VARIABLES */
-
-static struct R_1 { int32_t number;int32_t balance; } Rec__account;
-/* Clause INITIALISATION */
-void Rec__INITIALISATION(void)
-{
-    
-    Rec__account.number = 0;
-    Rec__account.balance = 10;
-}
-
-/* Clause OPERATIONS */
-
-void Rec__positive(bool *res)
-{
-    if((Rec__account.balance) > (0))
-    {
-        (*res) = true;
-    }
-    else
-    {
-        (*res) = false;
-    }
-}
-
-void Rec__withrdaw(int32_t amt)
-{
-    if((((amt) >= (0)) &&
-        ((amt) <= (2147483647))) &&
-    (((Rec__account.balance) >= (amt))))
-    {
-        Rec__account.number = Rec__account.number;
-        Rec__account.balance = (Rec__account.balance-amt);
-    }
-}
-
-void Rec__unsafe_dec(void)
-{
-    Rec__account.balance = Rec__account.balance-1;
-}
-

### 2014-03-06 C/DerivedTypes/Rec_i_changed.c

@@ -1,80 +1,0 @@
-#include <stdio.h>
-#include "Rec.h"
-
-/* Clause CONCRETE_CONSTANTS */
-/* Basic constants */
-
-/* Array and record constants */
-/* Clause CONCRETE_VARIABLES */
-
-static struct R_1 { int32_t number;int32_t balance; double dd; } Rec__account;
-void Rec__set(void);
-void Rec__unsafe_dec(void);
-void Rec__get_and_set(void);
-typedef struct point point;
-/* Declare the struct with integer members x, y */
-struct point {
-   int    x;
-   int    y;
-};
-
-int main (){
-    
-    Rec__set();
-    printf("%d\n",Rec__account.balance);
-    Rec__get_and_set();
-    printf("%d\n",Rec__account.balance);
-    Rec__INITIALISATION();
-    printf("%lf\n",Rec__account.dd);
-}
-/* Clause INITIALISATION */
-void Rec__INITIALISATION(void)
-{
-    point p = {.y = 2, .x = 1};
-    point pi;// = {700,900};
-    pi.y=700;
-    pi.x=900;
-    Rec__account.number = 0;
-    Rec__account.balance = 10;
-    Rec__account.dd=7;
-
-}
-
-/* Clause OPERATIONS */
-
-void Rec__positive(bool *res)
-{
-    if((Rec__account.balance) > (0))
-    {
-        (*res) = true;
-    }
-    else
-    {
-        (*res) = false;
-    }
-}
-
-void Rec__withrdaw(int32_t amt)
-{
-    if((((amt) >= (0)) &&
-        ((amt) <= (2147483647))) &&
-    (((Rec__account.balance) >= (amt))))
-    {
-        Rec__account.number = Rec__account.number;
-        Rec__account.balance = (Rec__account.balance-amt);
-    }
-}
-void Rec__set(void)
-{
-    Rec__account.balance = 50;
-}
-void Rec__get_and_set(void)
-{
-    Rec__account.balance = Rec__account.balance + 50;
-}
-
-void Rec__unsafe_dec(void)
-{
-    Rec__account.balance = Rec__account.balance-1;
-}
-

### 2014-03-06 DOCUMENTATION/AdvancedTypes.tex
Small grammar problems fixed.

@@
 
-\subsubsection{Getting a element from array}
+\subsubsection{Getting an element from array}
 
@@
 $address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The following rule 
-\footnote{The function $TSymbol$ returns the "start", "end" and type of the one 
+\footnote{The function $TSymbol$ returns the ``start'', ``end'' and type of the one 
 index.} makes the shift for each index of array. \VGM{TODO: The function TSymbol
-returns the "start", "end" and type of the one index.}
+returns the ``start'', ``end'' and type of the one index.}
 
@@
 \paragraph{Restriction on array:} The translation of direct attribution
- between array expression is not yet supported, for example:
+ between arrays is not yet supported, for example:
 

### 2014-03-08 DOCUMENTATION/AdvancedTypes.tex
Removed unneeded identifier and other small fix.

@@
         \B{lhs} & : & \B{LValue} \\
-        \B{rhs} & : & \B{Expr} \\
+        \B{rhs} & : & \B{Exp} \\
 	\hline
-	\multicolumn{3}{|c|}{Label value: \B{LValue}} \\	
+	\multicolumn{3}{|c|}{Left value: \B{LValue}} \\	
 	\hline
@@
         \B{op} & : & \B{Oper} \\
-        \B{inp} & : & seq \B{Expr} \\
+        \B{inp} & : & seq \B{Exp} \\
         \B{out} & : & seq \B{Name} \\
@@
         \hline
-        \B{expr} & : & \B{Expr} \\
+        \B{expr} & : & \B{Exp} \\
         \B{branches} & : & seq \B{CaseBr}\\
@@
         \hline
-        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
+        \multicolumn{3}{|c|}{Expression: \B{Exp}} \\
         \hline
@@
         \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
-	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
+	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
 	\hline
-  	\multicolumn{3}{|c|}{Data struct : \B{Struct}} \\
+  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
         \hline
-	\B{id} & : &  \B{Name}\\
         \B{elements} & : & seq \B{Name} \\
@@
         \hline
-	\B{base} & : &  \B{Struct} \\
-	\B{element} & : & \B{Name}\\
+	\B{base} & : &  \B{Name} \\
+	\B{element} & : & seq \B{Name}\\
 	\B{type} & : & \B{Type}\\
@@
 	\B{elements} & : & seq \B{Name}\\
-	\B{values} & : & seq \B{Expr}\\
+	\B{values} & : & seq \B{Exp}\\
 	\hline
@@
         \hline
-	\B{arg1} & : & seq \B{Expr}\\ 
-	\B{arg2} & : & \B{Expr}\\ 
+	\B{arg1} & : & seq \B{Exp}\\ 
+	\B{arg2} & : & \B{Exp}\\ 
 	\hline
@@
         \hline
-	\B{arg1} & : & seq \B{Expr}\\ 
-	\B{arg2} & : & \B{Expr}\\ 	
+	\B{arg1} & : & seq \B{Exp}\\ 
+	\B{arg2} & : & \B{Exp}\\ 	
 	\hline
@@
 	\B{base} & : &  \B{Array} \\
-	\B{indexes} & : & seq \B{Expr}\\
+	\B{indexes} & : & seq \B{Exp}\\
 	\B{type} & : & \B{ArrayType}\\
@@
 	\hline
-	\B{start}, \B{end} & : & \B{Expr} \\
+	\B{start}, \B{end} & : & \B{Exp} \\
 	\hline
@@
       & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
+      & \alt & \nt{name} \llvm{=} \llvm{bitcast} \nt{type} \nt{name}, \nt{type} \\
       & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
@@
   \trad{a}{\B{Beq}} \isdef
-  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
+  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
   & \llvm{"\PH{l}} \\
@@
   \trad{n}{\B{StructItem}} \isdef \\
-   & \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
-   & \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
-%   & \llvm{\PH{\ell2}  = load  \PH{\SBType(\B{n})}* \PH{\ell1} \nl"} \\
-   & \sep  \ell  \sep \SBType(\B{n})\\
+   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
+   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
+   &\quad  \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
+   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
 \end{split}

### 2014-03-08 DOCUMENTATION/AdvancedTypes.tex
Integrating the AdvancedTypes.tex and specfication.tex

@@
-\documentclass[10pt,a4paper]{article}
-\usepackage[utf8]{inputenc}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-
-\usepackage[fleqn]{amsmath}
-\usepackage{xspace,framed,here,keystroke}
-\usepackage{graphicx}
-\usepackage[colorlinks=false]{hyperref}
-
-\usepackage{minted}   %Used to syntax highlight
-\definecolor{bg}{rgb}{0.95,0.95,0.95}
-\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
-\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
-\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
-
-\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
-\newcommand{\nl}[0]{\ensuremath{\downarrow}}
-\newcommand{\mty}[0]{\texttt{""}}
-\DeclareMathOperator{\conc}{\diamond}
-\DeclareMathOperator{\isdef}{\equiv}
-\DeclareMathOperator{\dom}{\mbox{dom}}
-\DeclareMathOperator{\lbl}{\mathcal{L}()}
-\DeclareMathOperator{\variable}{\mathcal{V}()}
-\newcommand{\llvm}[1]{\texttt{#1}}
-\newcommand{\B}[1]{\textsf{#1}}
-\newcommand{\lalt}[0]{$\langle$\xspace}
-\newcommand{\ralt}[0]{$\rangle$\xspace}
-\newcommand{\alt}[0]{$\mid\,$}
-\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
-\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
-\newcommand{\Dict}[0]{\mathbb{D}}
-\newcommand{\Text}[0]{\mathbb{T}}
-\newcommand{\IF}[0]{\textbf{ if }}
-\newcommand{\ELSIF}[0]{\textbf{ else if }}
-\newcommand{\ELSE}[0]{\textbf{ else }}
-\newcommand{\THEN}[0]{\textbf{ then }}
-\newcommand{\LET}[0]{\textbf{ let }}
-\newcommand{\IN}[0]{\textbf{ in }}
-\newcommand{\AND}[0]{\textbf{ and }}
-\newcommand{\PH}[1]{\framebox{$#1$}}
-\newcommand{\sep}[0]{\otimes}
-\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
-\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
-\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
-\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
-\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
-\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
-\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
-\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
-\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
-\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
-\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
-\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
-\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
-\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
-\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
-\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
-\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}
-
-\newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}
-\newcommand{\VGM}[1]{\marginpar{\tiny{VGM: #1}}}
-
-\author{David Déharbe, Valério Medeiros Jr.}
-\title{Specification of Advanced Types Translation}
-
-\begin{document}
-
-\usemintedstyle{friendly}
-
-\section{The abstract syntax tree with support for advanced types}
-
-The supported advanced types are array and structures. The nodes Array and 
-Structure can be used to declare a variable, or  declare an element, or to 
-select an element in the array or the structure.
-
-\begin{figure}[H]
-  \begin{center}
-    {\scriptsize
-      \frame{
-    \begin{tabular}[t]{ccc}
-      \begin{tabular}[t]{rcl}
-        \hline
-        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{const} & : & seq \B{Cons} \\
-        \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} \\
-        \B{impl} & : & opt \B{Impl} \\
-        \hline
-        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{impo} & : & seq \B{Import} \\
-        \B{const} & : & seq \B{Cons} \\
-        \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} \\
-        \hline
-        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
-        \hline
-        \B{mach} & : & \B{Machine} \\
-        \B{pre} & : & opt \B{Name}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{type} & : & \B{Type} \\
-        \B{val} & : & \B{Value} \\
-        \hline
-        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{type} & : & \B{Type} \\
-        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
-	\hline
-        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{inp} & : & seq \B{Vari} \\
-        \B{out} & : & seq \B{Vari} \\
-        \B{body} & : & \B{Inst}
-        \\
-       \hline
-        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
-        \hline
-        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
-        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
-        \hline
-        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
-        \hline
-        \B{body} & : & seq \B{Inst} \\
-        \hline
-        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
-        \hline
-        \B{vars} & : & seq \B{Name} \\
-        \B{body} & : & seq \B{Inst} \\
-\end{tabular}
-\begin{tabular}[t]{rcl}   
-  	\hline
-        \multicolumn{3}{|c|}{If: \B{If}} \\
-        \hline
-        \B{branches} & : & seq \B{IfBr} \\
-        \hline
-        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
-        \hline
-        \B{lhs} & : & \B{LValue} \\
-        \B{rhs} & : & \B{Exp} \\
-	\hline
-	\multicolumn{3}{|c|}{Left value: \B{LValue}} \\	
-	\hline
-	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\
-        \hline
-        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
-        \hline
-        \B{op} & : & \B{Oper} \\
-        \B{inp} & : & seq \B{Exp} \\
-        \B{out} & : & seq \B{Name} \\
-        \B{inst} & : & opt \B{Impo} \\
-        \hline
-        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
-        \hline
-        \B{expr} & : & \B{Exp} \\
-        \B{branches} & : & seq \B{CaseBr}\\
-	\hline
-        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
-        \hline
-        \B{val} & : & seq \B{Value} \\
-        \B{body} & : & \B{Inst}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{While loop: \B{While}} \\
-        \hline
-        \B{cond} & : & \B{Pred} \\
-        \B{body} & : & \B{Inst} \\
-        \hline
-        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
-        \hline
-        \B{cond} & : & opt \B{Pred} \\
-        \B{body} & : & \B{Inst} \\
-        \hline
-        \multicolumn{3}{|c|}{Expression: \B{Exp}} \\
-        \hline
-        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
-	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
-        \hline
-        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
-        \hline
-	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\
-        \hline
-        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
-        \hline
-        \B{op} & : & \B{ArithOp} \\
-        \B{args} & : & seq \B{Exp} \\
-	\hline
-        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
-        \hline
-        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
-	\hline
-        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
-        \hline
-        \B{op} & : & \B{BoolOp} \\
-        \B{args} & : & seq \B{Pred} \\
-\end{tabular}
-
-\begin{tabular}[t]{rcl}
-	\hline
-        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
-        \hline
-        \B{op} & : & \B{RelOp} \\
-        \B{arg1}, \B{arg2} & : & \B{Exp} \\
-        \hline
-        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
-        \hline
-        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
-	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
-	\hline
-  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
-        \hline
-        \B{elements} & : & seq \B{Name} \\
-	\B{type} & : & seq \B{Type}\\
-	\hline
-	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
-        \hline
-	\B{base} & : &  \B{Name} \\
-	\B{element} & : & seq \B{Name}\\
-	\B{type} & : & \B{Type}\\
-	\hline
-	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
-        \hline
-	\B{elements} & : & seq \B{Name}\\
-	\B{values} & : & seq \B{Exp}\\
-	\hline
-        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
-        \hline 
-	\B{dom} & : & seq \B{Set}\\
-	\B{ran} & : & \B{Set}\\
-        \hline
-        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
-        \hline
-	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
-	\hline
-  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
-        \hline
-	\B{arg1} & : & seq \B{Exp}\\ 
-	\B{arg2} & : & \B{Exp}\\ 
-	\hline
-  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
-        \hline
-	\B{arg1} & : & seq \B{Exp}\\ 
-	\B{arg2} & : & \B{Exp}\\ 	
-	\hline
-	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
-        \hline
-	\B{base} & : &  \B{Array} \\
-	\B{indexes} & : & seq \B{Exp}\\
-	\B{type} & : & \B{ArrayType}\\
-	\hline
-	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
-	\hline
-	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
-	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
-	\hline
-	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
-	\hline
-	\B{start}, \B{end} & : & \B{Exp} \\
-	\hline
-	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
-	\hline
-	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
-      \end{tabular}
-    \end{tabular}
-  }
-  }
-    \caption{Abstract syntax structure of a B implementation.  For each abstract
-      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
-      branch in a case instruction, and the different attributes of the class),
-      or a list of the possible sub-classes (e.g. \B{Inst} for the different
-      kinds of instructions).
-      Sequence and optional attributes are denoted ``seq'' and ``opt'',
-      respectively.}
-    \label{tab:node-attr}
-  \end{center}
-\end{figure}
-
-\begin{figure}
-  \begin{center}
-    \begin{tabular}{rcl}
-      \nt{module} & ::= & \ListOf{\nt{item}} \\
-      \nt{item} & ::= & \nt{const\_decl} \alt \nt{function\_decl} \alt \nt{type\_def}
-      \alt \nt{const\_def} \alt \nt{var\_def} \alt \nt{function\_def} \\
-      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
-      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
-      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
-      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
-      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
-      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
-      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
-      \nt{param} & ::= & \nt{type} \nt{name} \\
-      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}} \\
-      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
-      & \alt & \nt{name} \llvm{=} \lalt \llvm{add} \alt \llvm{sub} \alt \llvm{mul} \alt \llvm{sdiv} \alt \llvm{srem} \ralt \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
-      & \alt & \nt{name} \llvm{=} \llvm{icmp} \lalt \llvm{eq} \alt \llvm{ne} \alt \llvm{sgt} \alt \llvm{sge} \alt \llvm{slt} \alt \llvm{sle} \ralt \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
-      & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
-      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} 
-\nt{exp}\llvm{,} \nt{index}\llvm{,} \ListOf{\nt{index}} \\
-      & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
-      & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
-      & \alt & \nt{name} \llvm{=} \llvm{bitcast} \nt{type} \nt{name}, \nt{type} \\
-      & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
-      & \alt & \llvm{br} \llvm{label} \nt{lbl} \\
-%       & \alt & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
-      & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
-      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} 
-\llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \ListOf{\nt{index}} \llvm{)} \\
-      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
-      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
-      \nt{arg} & ::= & \nt{type} \nt{exp}
-    \end{tabular}
-  \end{center}
-  \caption{Grammar of the target LLVM subset: \nt{itype}, \nt{iliteral}, \nt{lbl}
-    and \nt{name} correspond respectively to integer types, integer literals,
-    labels and names. Choices are separated by \alt and optionally delimited by
-    \lalt and \ralt.  The \ListOf{} superscript denotes a comma-separated list of
-    elements of the annotated entity. }
-  \label{fig:llvm-grammar}
-\end{figure}
-
-\VGM{To support arrays and structures, I added only + after index.}
-
 \subsection{Rules to be updated in first specification}
 
-\paragraph{Becomes equal instructions} must evaluate the target and the source
-of the assignment, and then copy the result of the latter in the former, using
-the LLVM instruction \llvm{store}. The evaluation of the target is defined with
-$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
-variable $p$ containing the assigned location. The evaluation of the source
-yields the corresponding code $r$, the variable $v$ holding the result, and its
-type $t$.
+\paragraph{Becomes equal instructions} \textcolor{blue}{ must evaluate the
+ target and the source of the assignment, and then copy the result of the 
+latter in the former, using the LLVM instruction \llvm{store}. The evaluation 
+of the target is defined with $\trad{}{lvalue}$, which yields the corresponding 
+LLVM source code $l$ and the variable $p$ containing the assigned location. The 
+evaluation of the source yields the corresponding code $r$, the variable $v$ 
+holding the result, and its type $t$. }
 
@@
 
-The expressions have elements represented by pointers that need a special 
+\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
 care. The attributions between these elements are made  directly by the memory 
@@
 resolution with pointers needs an additional instruction $LOAD$ for store the 
-partial results. This translation is represented by $ValueItem$.
+partial results. This translation is represented by $ValueItem$.}
 
@@
    &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
-   &\quad  \llvm{"\PH{\ell} = getelementptr \%\PH{t}* @\PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
+   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
    &\quad  \sep  \ell  \sep \SBType(\B{n})\\
@@
 B code example, is translated by a sequence of application of rule $StructItem$.
+The nested structures will be designed in the next project phase.
 
@@
 \begin{align*}
-&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
+&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\ \\
 &\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
@@
 \end{pascalcode}
-
-\section*{Attachments}
-
-\subsection{Structure examples}
-
-The following code example is defining a structure:
-
-\begin{ccode}
-MACHINE Rec
-SETS
- ID={aa,bb}
-VARIABLES xx, bank, total
-INVARIANT
- xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
- total:struct(money:NAT)
-INITIALISATION 
- bank := rec(name:bb) ||
- xx:= rec(name:aa, balance:10) ||
- total := rec(money:20)
-OPERATIONS
-
-  res<--positive = 
-  PRE xx'balance > 0 THEN
-  res:=TRUE
-  END;
-
-  withrdaw(amt) = 
-  PRE amt:NAT1 & (xx'balance >= amt) THEN
-     xx := rec(name:xx'name, balance:(xx'balance - amt))
-  END;
-
-  unsafe_dec = 
-  BEGIN
-       xx'balance := xx'balance - 1
-  END
-END
-\end{ccode}
-
-\begin{ccode}
-IMPLEMENTATION
-    Rec_i
-
-REFINES
-    Rec
-
-CONCRETE_VARIABLES
-   xx ,
-   bank ,
-   total
-
-INITIALISATION
-
-   bank := rec(name:bb) ;
-   xx := rec(name:aa, balance:10) ;
-   total :=rec(money:20)
-OPERATIONS
-  res<--positive = 
-  IF xx'balance > 0 THEN
-       res:= TRUE 
-  ELSE res := FALSE
-  END;
-
-  withrdaw(amt) = IF   amt >=0 & amt <= MAXINT & (xx'balance >= amt) THEN 
-     xx := rec(name:xx'name, balance:(xx'balance - amt))
-  END;
-
-  unsafe_dec = BEGIN
-       xx'balance := xx'balance -1 
-  END
-END
-\end{ccode}
-
-\subsection{Arrays examples}
-
-%\iffalse  % This content is not visible to reader.
-A generic example available at repository ProB\footnote{
-\url{https://github.com/bendisposto/probparsers/blob/develop/bparser/src/test/resources/parsable/Array.mch}}. 
-This example can be used in approach using imported components.
-
-\subsection{A  simple example and its LLVM}
-
-\begin{ccode}
-MACHINE
-   Array
-VARIABLES arr
-
-INVARIANT 
-    arr : ((0..99) --> 0..1000)
-
-INITIALISATION arr := {1|->100}
-
-OPERATIONS
-
-set(ix,tt)=
-        PRE ix : (0..99) & tt : (0..1000)
-        THEN arr := arr <+ {ix |-> tt}
-        END ;
-
-tt <-- read(ix) =
-        PRE ix : (0..99)
-        THEN tt := arr(ix)
-        END;
-
-swap(ix, jx)=
-        PRE ix : (0..99) & jx : (0..99)
-        THEN arr := arr <+ {ix |-> arr(jx), jx|-> arr(ix)}
-        END
-
-END
-\end{ccode}
-
-\begin{ccode}
-IMPLEMENTATION 
-   Array_i
-REFINES
-   Array
-CONCRETE_VARIABLES
-   arr
-INVARIANT 
-    arr :  ((0..99) --> 0..1000)
-INITIALISATION
-   arr(1) := 100
-OPERATIONS
-   set ( ix , tt ) =
-   IF
-       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
-   THEN
-
-      arr(ix) := tt 
-   END;
-
-   tt <-- read ( ix ) =
-   IF
-       ix >= 0 & ix <=99
-   THEN
-       tt := arr ( ix )
-   ELSE
-       tt := 0
-   END;
-
-   swap ( ix , jx ) =
-   IF
-        ix >= 0 & ix <=99 & jx >= 0 & jx <=99 
-   THEN
-
-       VAR temp IN
-           temp := arr(jx);
-           arr(jx) := arr(ix);
-           arr(ix) := temp
-       END
-   END
-END
-\end{ccode}
-
-\begin{ccode}
-@Array__arr = internal global [100 x i32] zeroinitializer, align 16
-
-define void @Array__INITIALISATION() nounwind ssp {
-  store i32 100, i32* getelementptr  ([100 x i32]* @Array__arr, i32 0, i64 1)
-  ret void
-}
-
-define void @Array__set(i32 %ix, i32 %tt) nounwind ssp {
-  %1 = alloca i32
-  %2 = alloca i32
-  store i32 %ix, i32* %1
-  store i32 %tt, i32* %2
-  %3 = load i32* %1
-  %4 = icmp sge i32 %3, 0
-  br i1 %4, label %5, label %19
-
-; <label>:5                                       ; preds = %0
-  %6 = load i32* %1
-  %7 = icmp sle i32 %6, 99
-  br i1 %7, label %8, label %19
-
-; <label>:8                                       ; preds = %5
-  %9 = load i32* %2
-  %10 = icmp sge i32 %9, 0
-  br i1 %10, label %11, label %19
-
-; <label>:11                                      ; preds = %8
-  %12 = load i32* %2
-  %13 = icmp sle i32 %12, 1000
-  br i1 %13, label %14, label %19
-
-; <label>:14                                      ; preds = %11
-  %15 = load i32* %2
-  %16 = load i32* %1
-  %17 = sext i32 %16 to i64
-  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
-  store i32 %15, i32* %18
-  br label %19
-; <label>:19                                      ; preds = %14, %11, %8, %5, %0
-  ret void
-}
-\end{ccode}
-
-\begin{llvmcode}
-define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
-  %1 = alloca i32, align 4
-  %2 = alloca i32*, align 8
-  store i32 %ix, i32* %1, align 4
-  store i32* %tt, i32** %2, align 8
-  %3 = load i32* %1, align 4
-  %4 = icmp sge i32 %3, 0
-  br i1 %4, label %5, label %14
-
-; <label>:5                                       ; preds = %0
-  %6 = load i32* %1, align 4
-  %7 = icmp sle i32 %6, 99
-  br i1 %7, label %8, label %14
-
-; <label>:8                                       ; preds = %5
-  %9 = load i32* %1, align 4
-  %10 = sext i32 %9 to i64
-  %11 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %10
-  %12 = load i32* %11
-  %13 = load i32** %2, align 8
-  store i32 %12, i32* %13
-  br label %16
-
-; <label>:14                                      ; preds = %5, %0
-  %15 = load i32** %2, align 8
-  store i32 0, i32* %15
-  br label %16
-
-; <label>:16                                      ; preds = %14, %8
-  ret void
-}
-\end{llvmcode}
-
-\begin{llvmcode}
-define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
-  %1 = alloca i32, align 4
-  %2 = alloca i32, align 4
-  %temp = alloca i32, align 4
-  store i32 %ix, i32* %1, align 4
-  store i32 %jx, i32* %2, align 4
-  %3 = load i32* %1, align 4
-  %4 = icmp sge i32 %3, 0
-  br i1 %4, label %5, label %30
-
-; <label>:5                                       ; preds = %0
-  %6 = load i32* %1, align 4
-  %7 = icmp sle i32 %6, 99
-  br i1 %7, label %8, label %30
-
-; <label>:8                                       ; preds = %5
-  %9 = load i32* %2, align 4
-  %10 = icmp sge i32 %9, 0
-  br i1 %10, label %11, label %30
-
-; <label>:11                                      ; preds = %8
-  %12 = load i32* %2, align 4
-  %13 = icmp sle i32 %12, 99
-  br i1 %13, label %14, label %30
-
-; <label>:14                                      ; preds = %11
-  %15 = load i32* %2, align 4
-  %16 = sext i32 %15 to i64
-  %17 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %16
-  %18 = load i32* %17
-  store i32 %18, i32* %temp, align 4
-  %19 = load i32* %1, align 4
-  %20 = sext i32 %19 to i64
-  %21 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %20
-  %22 = load i32* %21
-  %23 = load i32* %2, align 4
-  %24 = sext i32 %23 to i64
-  %25 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %24
-  store i32 %22, i32* %25
-  %26 = load i32* %temp, align 4
-  %27 = load i32* %1, align 4
-  %28 = sext i32 %27 to i64
-  %29 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %28
-  store i32 %26, i32* %29
-  br label %30
-
-; <label>:30                                      ; preds = %14, %11, %8, %5, %0
-  ret void
-}
-
-\end{llvmcode}
-
-\subsection{Real example - Bubble Sort}
-
-\begin{ccode}
-MACHINE
-   Bubble
-VARIABLES
-   vec,sort
-INVARIANT
-   vec : 0..99 --> 0..99 & sort : 0..1 &
-   ( sort = 1 =>  !ii.(ii : 0..98 => vec(ii)<=vec(ii+1)))
-INITIALISATION
-   vec :: 0..99 --> 0..99 || sort := 0
-OPERATIONS
-   op_sort =
-   ANY sorted_vector WHERE
-    sorted_vector : 0..99 --> 0..99 &
-    !ii.(ii : 0..98 => sorted_vector(ii)<=sorted_vector(ii+1))
-   THEN
-    vec := sorted_vector|| sort := 1
-   END
-
-END
-
-IMPLEMENTATION
-   Bubble_i
-REFINES
-   Bubble
-CONCRETE_VARIABLES
-   vec1,sort1
-INVARIANT
-   vec1 : 0..99 --> 0..99 & sort1 : 0..1 &
-   vec = vec1 & sort1 = sort
-INITIALISATION
-   vec1(0) := 0  ;
-   sort1 := 0
-OPERATIONS
-   op_sort =
-   VAR nn, swapped, ii, tmp, vi, vi2 IN
-	swapped := 1;
-	nn := 100;
-	ii:=0;
-	WHILE swapped = 1 DO
-		swapped := 0;
-		ii:= 0;
-		nn:= nn-1;
-
-		WHILE ii<= nn DO
-		vi :=  vec1(ii);
-		vi2 := vec1(ii+1);
-		IF vi > vi2 THEN
-			tmp:= vec1(ii+1);
-			vec1(ii+1):=vec1(ii);
-			vec1(ii):=tmp;
-			swapped:= 1
-		END;
-		ii:= ii+1
-
-		INVARIANT 1>0
-		VARIANT nn-ii
-		END
-
-	INVARIANT 1>0
-	VARIANT nn-ii
-	END
-   END
-END
-\end{ccode}
-
-\end{document}

### 2014-03-08 DOCUMENTATION/specification.tex
Integrating the AdvancedTypes.tex and specfication.tex

@@
 \usepackage{graphicx}
+\usepackage{color}
+\usepackage{minted}
+
 \pagestyle{headings}
+
 \title{Translation of B Implementations to the LLVM: Specification}
-\author{David Déharbe \and Valério Medeiros Jr}
-\institute{Federal University of Rio Grande do Norte \\
+\author{David Déharbe and Valério Medeiros Jr}
+\institute{Federal University of Rio Grande do Norte\\
+Federal Institute of Education Science and Technology\\
  Natal (Brazil)}
-\date{May 2013}
+
+\date{March 2014}
+\definecolor{bg}{rgb}{0.95,0.95,0.95}
+\newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
+\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
+\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
 
@@
 \newcommand{\IN}[0]{\textbf{ in }}
-\newcommand{\AND}[0]{\textbf{ and }}
+\newcommand{\AND}[1]{\textbf{ and }}
 \newcommand{\PH}[1]{\framebox{$#1$}}
@@
 \newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
+\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
+\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}
 
 \newcommand{\DD}[1]{\marginpar{\tiny{DD: #1}}}
+\newcommand{\VGM}[1]{\marginpar{\tiny{VGM: #1}}}
 
 \begin{document}
+
 \maketitle
@@
       respectively}
+    \label{tab:node-attr}
+  \end{center}
+\end{figure}
+
+\section{The abstract syntax tree with support for advanced types}
+
+\textcolor{blue}{
+The supported advanced types are array and structures. The nodes Array and 
+Structure can be used to declare a variable, or  declare an element, or to 
+select an element in the array or the structure. 
+}
+
+\begin{figure}[H]
+  \begin{center}
+    {\footnotesize
+      \frame{
+    \begin{tabular}[t]{ccc}
+      \begin{tabular}[t]{rcl}
+        \hline
+        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{const} & : & seq \B{Cons} \\
+        \B{var} & : & seq \B{Vari} \\
+        \B{init} & : & seq \B{Inst} \\
+        \B{op} & : & seq \B{Oper} \\
+        \B{impl} & : & opt \B{Impl} \\
+        \hline
+        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{impo} & : & seq \B{Import} \\
+        \B{const} & : & seq \B{Cons} \\
+        \B{var} & : & seq \B{Vari} \\
+        \B{init} & : & seq \B{Inst} \\
+        \B{op} & : & seq \B{Oper} \\
+        \hline
+        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
+        \hline
+        \B{mach} & : & \B{Machine} \\
+        \B{pre} & : & opt \B{Name}
+        \\
+        \hline
+        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{type} & : & \B{Type} \\
+        \B{val} & : & \B{Value} \\
+        \hline
+        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{type} & : & \B{Type} \\
+        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
+	\hline
+        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
+        \B{inp} & : & seq \B{Vari} \\
+        \B{out} & : & seq \B{Vari} \\
+        \B{body} & : & \B{Inst}
+        \\
+       \hline
+        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
+        \hline
+        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
+        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
+        \hline
+        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
+        \hline
+        \B{body} & : & seq \B{Inst} \\
+        \hline
+        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
+        \hline
+        \B{vars} & : & seq \B{Name} \\
+        \B{body} & : & seq \B{Inst} \\
+\end{tabular}
+\begin{tabular}[t]{rcl}   
+  	\hline
+        \multicolumn{3}{|c|}{If: \B{If}} \\
+        \hline
+        \B{branches} & : & seq \B{IfBr} \\
+        \hline
+        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
+        \hline
+        \B{lhs} & : & \B{LValue} \\
+        \B{rhs} & : & \B{Exp} \\
+	\hline
+	\multicolumn{3}{|c|}{Left value: \B{LValue}} \\	
+	\hline
+	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\
+        \hline
+        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
+        \hline
+        \B{op} & : & \B{Oper} \\
+        \B{inp} & : & seq \B{Exp} \\
+        \B{out} & : & seq \B{Name} \\
+        \B{inst} & : & opt \B{Impo} \\
+        \hline
+        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
+        \hline
+        \B{expr} & : & \B{Exp} \\
+        \B{branches} & : & seq \B{CaseBr}\\
+	\hline
+        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
+        \hline
+        \B{val} & : & seq \B{Value} \\
+        \B{body} & : & \B{Inst}
+        \\
+        \hline
+        \multicolumn{3}{|c|}{While loop: \B{While}} \\
+        \hline
+        \B{cond} & : & \B{Pred} \\
+        \B{body} & : & \B{Inst} \\
+        \hline
+        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
+        \hline
+        \B{cond} & : & opt \B{Pred} \\
+        \B{body} & : & \B{Inst} \\
+        \hline
+        \multicolumn{3}{|c|}{Expression: \B{Exp}} \\
+        \hline
+        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
+	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
+        \hline
+        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
+        \hline
+	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\
+        \hline
+        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
+        \hline
+        \B{op} & : & \B{ArithOp} \\
+        \B{args} & : & seq \B{Exp} \\
+	\hline
+        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
+        \hline
+        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
+	\hline
+        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
+        \hline
+        \B{op} & : & \B{BoolOp} \\
+        \B{args} & : & seq \B{Pred} \\
+\end{tabular}
+
+\begin{tabular}[t]{rcl}
+	\hline
+        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
+        \hline
+        \B{op} & : & \B{RelOp} \\
+        \B{arg1}, \B{arg2} & : & \B{Exp} \\
+        \hline
+        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
+        \hline
+        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
+	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
+	\hline
+  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
+        \hline
+        \B{elements} & : & seq \B{Name} \\
+	\B{type} & : & seq \B{Type}\\
+	\hline
+	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
+        \hline
+	\B{base} & : &  \B{Name} \\
+	\B{element} & : & seq \B{Name}\\
+	\B{type} & : & \B{Type}\\
+	\hline
+	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
+        \hline
+	\B{elements} & : & seq \B{Name}\\
+	\B{values} & : & seq \B{Exp}\\
+	\hline
+        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
+        \hline 
+	\B{dom} & : & seq \B{Set}\\
+	\B{ran} & : & \B{Set}\\
+        \hline
+        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
+        \hline
+	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
+	\hline
+  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
+        \hline
+	\B{arg1} & : & seq \B{Exp}\\ 
+	\B{arg2} & : & \B{Exp}\\ 
+	\hline
+  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
+        \hline
+	\B{arg1} & : & seq \B{Exp}\\ 
+	\B{arg2} & : & \B{Exp}\\ 	
+	\hline
+	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
+        \hline
+	\B{base} & : &  \B{Array} \\
+	\B{indexes} & : & seq \B{Exp}\\
+	\B{type} & : & \B{ArrayType}\\
+	\hline
+	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
+	\hline
+	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
+	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
+	\hline
+	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
+	\hline
+	\B{start}, \B{end} & : & \B{Exp} \\
+	\hline
+	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
+	\hline
+	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
+      \end{tabular}
+    \end{tabular}
+  }
+  }
+    \caption{Abstract syntax structure of a B implementation.  For each abstract
+      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
+      branch in a case instruction, and the different attributes of the class),
+      or a list of the possible sub-classes (e.g. \B{Inst} for the different
+      kinds of instructions).  Sequence and optional attributes are denoted 
+      ``seq'' and ``opt'',respectively.}
     \label{tab:node-attr}
@@
       & \alt & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
-      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index} \\
+      & \alt & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \textcolor{blue}{\ListOf{ \nt{index}}}  \\
       & \alt & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
       & \alt & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
+      & \alt & \textcolor{blue}{\nt{name} \llvm{=} \llvm{bitcast} \nt{type} \nt{name}, \nt{type} }\\
       & \alt & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
@@
       & \alt & \llvm{ret} \lalt \nt{type} \nt{exp} \alt \llvm{void} \ralt \\
-      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
+      \nt{exp} & ::= & \nt{name} \alt \nt{iliteral} \alt \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \textcolor{blue}{\ListOf{ \nt{index}}}  \llvm{)} \\
       \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
@@
 
+\include{AdvancedTypes}
+
+\include{Attachments}
+
 \end{document}

### 2014-03-17 DOCUMENTATION/specification.tex
Added spec rules for initialization (still incomplete).

--- 2014-03-02 DOCUMENTATION/specification.tex
@@
 
-Initialization is done is two steps: first all imported modules are
-initialized,  second the initialization clause is executed.
-\begin{align*}
-  \trad{i}{Init} \isdef
-  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
+Initialization of a module is implemented by a LLVM function that takes as
+parameters: the address of the aggregate representing the state of the module,
+which we name \llvm{\%self}, and the addressses of the aggregates representing
+the state of each component of the module, which we name \llvm{\%argN}, where
+\llvm{N} is a numeric identifier. The body of this function is composed as
+follows: first frame stack memory is allocated, to store the representation of
+the local variables in the initialization; second, all imported modules are
+bound to the component's elements and initialized; last the instructions
+composing the initialization clause are executed:
+
+\begin{align*}
+  \trad{\B{n}}{\B{init}}_{\text{interface}} \isdef
+  & \llvm{"define void \PH{\init(\B{n})} ( \PH{\stateref(\B{n})} \%self,
+    \PH{\trad{\trimportedinstances(\B{n})}{\B{args}}} \{ \nl"} \\
+  & \quad\llvm{entry: \nl} \\
+  & \quad\llvm{\PH{\trad{i.init}{alloc}}} \\
   & \quad\llvm{\PH{\trad{i, i.impl}{\ListOf{InitImp}}}} \\
@@
 \end{align*}
+
+Next, we specify how to generate the code to enumerate the list of arguments
+representing the components. Essentially, for each stateful component
+\B{inst}, the argument is a reference to the aggregate representing the
+state and named \llvm{\%argN}, where \llvm{N} is a number:
+\begin{align*}
+  \trad{}{args} \isdef & \llvm{""} \\
+  \trad{\B{inst} L}{args} \isdef &
+  \llvm{"\trad{\B{inst}, \textit{index}(\B{inst}, L)}{arg}, \trad{L}{args}"} \\
+  \trad{\B{inst}, $n$}{arg} \isdef &
+  \llvm{""} \triangleleft \stateless(\B{inst.mach}) \triangleright 
+  \llvm{"\PH{\stateref(\B{inst.mach})} \%arg\PH{n}"}
+\end{align*}
+
 The initialization of an imported module is performed by a call to the
@@
   & \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{impl.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{impo})}\nl}\\
+  & \quad \llvm{"store \PH{?ref} \%arg\PH{?X}, \PH{?ref} * \PH{\idx(\B{impo})}\nl}\\
   & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m.impl})}*
@@
 \end{align*}
++ add binding!
 

### 2014-03-18 DOCUMENTATION/specification.tex
Completed a first draft for full specification of initialization.

--- 2014-03-02 DOCUMENTATION/specification.tex
@@
 
-%TODO It is deprecated 
-%[[[TODO: DEPRECATED ]]]
-
-Initialization is done is two steps: first all imported modules are
-initialized,  second the initialization clause is executed.
-\begin{align*}
-  \trad{i}{Init} \isdef
-  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
-  & \quad\llvm{\PH{\trad{i, i.impl}{\ListOf{InitImp}}}} \\
-  & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
+\newcommand{\argname}[0]{\ensuremath{\mathsf{\nabla}}}
+\newcommand{\components}[0]{\ensuremath{\textsf{st-comp*}}}
+
+Initialization of a module is implemented by a LLVM function that takes as
+parameters: the address of the aggregate representing the state of the module,
+which we name \llvm{\%self}, and the addressses of the aggregates representing
+the state of each component of the module. 
+We first define a naming function $\argname$ having as domain the list of
+\emph{stateful} module components:
+\begin{align*}
+  \argname(l) \isdef
+  & \lambda c \cdot \llvm{"\%arg\PH{\mathit{index}(c, l)}"}
+\end{align*}
+So \llvm{\%arg0}, \llvm{\%arg1}, \ldots are the identifiers for the parameters
+representing the components of the initialized module.
+
+More precisely, the specification needs to make a distinction between stateful
+and stateless components, as only the former are represented by a LLVM
+aggregate. We use the expression $\components(\B{n}) \isdef
+\mathit{remove}(\trimportedinstances(\B{n}), \stateless)$ to denote the list of
+stateful components of $\B{n}$, so that $\argname(\components(\B{n}))$ is the
+function that, given a stateful component $c$ of \B{n}, returns the parameter of
+the initialization function of $c$. We are now ready to specify the 
+signature of this initialization function, using an auxiliary translation
+function to produce the LLVM parameter type and name for each stateful
+component $\B{c}$:
+\begin{align*}
+\trad{\B{n}}{init-sig} \isdef &
+\llvm{"\PH{\init(\B{n})}(\PH{\stateref(\B{n})} \%self, 
+  \PH{\trad{\components(\B{n}), \argname(\components(\B{n}))}{\ListOf{arg}}})"} \\
+\trad{\B{c}, \argname}{arg} \isdef & \llvm{"\stateref(\B{c.root}) \argname(\B{c})"}
+\end{align*}
+
+The body of this function is composed as follows: firstly, frame stack memory is
+allocated, to store the representation of the local variables in the
+initialization; second, all the directly imported, stateful, component modules
+are bound to the main module's elements; third, the directly imported components
+are initialized, by calling the corresponding initialization function; last, the
+instructions composing the initialization clause are executed:
+
+\begin{align*}
+  \trad{\B{n}}{\B{init}}_{\text{interface}} \isdef
+  & \llvm{"define void \PH{\trad{\B{n}}{init-sig}} \{ \nl"} \\
+  & \quad\llvm{entry: \nl} \\
+  & \quad\llvm{\PH{\trad{n.init}{alloc}}} \\
+  & \quad\llvm{\PH{\trad{n, \importedinstances(n), \argname(\components(\B{n}))}{\ListOf{BindImp}}}} \\
+  & \quad\llvm{\PH{\trad{\importedinstances(n), \argname(\components(\B{n}))}{\ListOf{InitImp}}}} \\
+  & \quad\llvm{\PH{\trad{n.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
   & \quad\llvm{exit: ret void \nl} \\
@@
 \end{align*}
-The initialization of an imported module is performed by a call to the
-corresponding initialization function:
-\begin{align*}
-  \lefteqn{\trad{\B{impl}, \B{impo}}{InitImp} \isdef} \\
-  & \LET v = \variable \AND \B{m} = \B{impo.mach} \IN \\
-  & \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{impl.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{impo})}\nl}\\
-  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m.impl})}*
-    \PH{\ell}) \nl"}
+
+The translation for the memory allocation and the instructions are specified in
+section~\ref{sec:trad-instr}.
+
+The aggregate representing the state of the root module has one element for
+each directly imported stateful module. The initialization function is
+responsible for binding this element with the corresponding argument.
+\begin{align*}
+  \lefteqn{\trad{\B{n}, \B{c}, \argname}{BindImp} \isdef} \\
+  & \LET v = \variable \IN \\
+  & \quad \llvm{"\PH{v} = getelementptr \PH{\stateref(\B{n})} \%self\$, i32 0, i32 \PH{\idx(\B{c})}\nl}\\
+  & \quad \llvm{store \PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}, 
+    \PH{\stateref(\B{c.root})}* \PH{v}\nl"}
+\end{align*}
+
+In general, the initialization function of a component module $c$ has as
+argument the address of the aggregate representing the module, as well as the
+addresses of the aggregates representing the modules in the importation tree of
+$c$.
+\begin{align*}
+  \trad{\B{c}, \argname}{InitImp} \isdef&
+  \llvm{call void \PH{\init(\B{c})}(\PH{\trad{c}{arg}}, \trad{\components(\B{c}), \argname}{\ListOf{arg}}) \nl"} \\
+\trad{\B{c}, \argname}{arg} \isdef& 
+  \llvm{"\PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}"}
 \end{align*}

### 2014-03-19 DOCUMENTATION/specification.tex
Specification: refactorings, small corrections, reworked section on data.

@@
 \newcommand{\LET}[0]{\textbf{ let }}
+\DeclareMathOperator{\BE}{\hat{=}}
 \newcommand{\IN}[0]{\textbf{ in }}
@@
 \newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
-\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
+\newcommand{\self}[0]{\llvm{\%self\$}}
 \newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
@@
 	\hline
-	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{StructItem} }\\
+	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} \alt \B{RecField} }\\
         \hline
@@
         \hline
-	\multicolumn{3}{c}{\B{StructItem}\alt\B{ArrayItem} } \\
+	\multicolumn{3}{c}{\B{RecField}\alt\B{ArrayItem} } \\
         \hline
@@
         \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
-	\multicolumn{3}{c}{\B{StructType} \alt \B{ArrayType}}\\
-	\hline
-  	\multicolumn{3}{|c|}{Data struct : \B{StructType}} \\
-        \hline
-        \B{elements} & : & seq \B{Name} \\
-	\B{type} & : & seq \B{Type}\\
-	\hline
-	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
-        \hline
-	\B{base} & : &  \B{Name} \\
-	\B{element} & : & seq \B{Name}\\
+	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType}}\\
+	\hline
+  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
+        \hline
+        \B{elements} & : & seq \B{Field} \\
+	\hline
+  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
+        \hline
+        \B{id} & : & \B{Name} \\
 	\B{type} & : & \B{Type}\\
 	\hline
-	\multicolumn{3}{|c|}{Record struct : \B{Record}} \\
-        \hline
-	\B{elements} & : & seq \B{Name}\\
+	\multicolumn{3}{|c|}{Record: \B{Record}} \\
+        \hline
+	\B{type} & : & \B{Structure}\\
 	\B{values} & : & seq \B{Exp}\\
+	\hline
+	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
+        \hline
+	\B{base} & : &  \B{name} \\
+	\B{element} & : & seq \B{Name} \\
 	\hline
@@
 Defines the functions declared in the interface (only for developed modules): \\
-\myind \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
+\myind \llvm{define void @M\$op(\%M\$state\$* \self, \ListOf{\nt{param}}) \{} \\
 \myind \myind \llvm{\ListOf{\nt{block}}} \\
@@
 \stateref(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$ref\$"} & \text{ (the type pointer to state)} \\
-\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
-\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
+\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)}
 \end{align*}
@@
   &
-\LET t = \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
+\LET t \BE \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
 & \llvm{"declare void \PH{\Global(\B{n})}
@@
   \trad{\B{n}}{\B{Mach}}_{\text{implementation}} \isdef
-  & \LET \B{i} = \B{n.impl} = \bot \IN\\
+  & \LET \B{i} \BE \B{n.impl} = \bot \IN\\
   & \quad \IF \B{i} = \bot \THEN \llvm{""} \\
@@
 
-The translation presented in this paper is currently limited to integers and
-booleans. In the case of integers, the translation is based on the supposition
-that the B0 definition of \B{INT} is the range $-2^{31}.. 2^{31}-1$. We have
-$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}$ and $\trad{INT}{\B{Type}} \isdef
-\llvm{i32}$.
-
-B constants shall be translated to constants in the LLVM program. The
-translation of scalar constants consists in substituting them with their value.
-%TODO Look the following text. 
-The translation of non-scalar constants will be defined in a next version of
-this document. The definition of the translation for expressions is
-described next in section~\ref{sec:expr}.
+The translation presented in this paper handles the following concrete data:
+\begin{description}
+\item[Booleans] There is no Boolean type in LLVM, so we encode Booleans with
+  one-bit integer:
+$$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}.$$
+\item[integers] Atelier-B offers three possible definitions for integers,
+  corresponding to 16-bit, 32-bit (default) and 64-bit integers. The code
+  generator has an option to specify which definition is employed, and we have:
+$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
+\item[enumerated sets] Again, enumerations are encoded as integer types, and the
+  width of the type depends on the size of the enumeration.
+$$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$
+\item[arrays] In LLVM, an array type is defined by its size and the type of its
+  elements, the main issues in encoding a B array type into LLVM being to
+  determine the size of the array and multi-dimensionality.
+
+  [[[ TODO ]]]
+
+\item[structures] The structure types of B are naturally encoded as structure
+  types in LLVM.
+
+  [[[ TODO ]]]
+\end{description}
+
+The current specification of the code generator does not handle deferred sets.
 
@@
 \begin{align*}
-\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
+\lefteqn{\trad{n}{\B{Name}} \isdef \LET t \BE \trad{n.type}{\B{Type}} \IN} \\
 & \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
@@
 \llvm{""} \sep \local(\B{n}) \sep t \\
-& \ELSE (\B{n} \mbox{ is a state variable }) \LET p = \variable \AND v = \variable \IN \\
-& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
+& \ELSE (\B{n} \mbox{ is a state variable }) \LET p \BE \variable \AND v = \variable \IN \\
+& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \self, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
 & \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
@@
   \LET
-  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
-  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
-& \IN \LET v = \variable \IN \\
+  p_1 \sep v_1 \sep t_1 \BE \trad{r.arg1}{\B{Expr}} \AND
+  p_2 \sep v_2 \sep t_2 \BE \trad{r.arg2}{\B{Expr}}} \\
+& \IN \LET v \BE \variable \IN \\
 & \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = icmp \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v
@@
   \LET
-  p_1 \sep v_1 \sep t_1 = \trad{r.arg1}{\B{Expr}} \AND
-  p_2 \sep v_2 \sep t_2 = \trad{r.arg2}{\B{Expr}}} \\
-& \IN \LET v = \variable \IN \\
+  p_1 \sep v_1 \sep t_1 \BE \trad{r.arg1}{\B{Expr}} \AND
+  p_2 \sep v_2 \sep t_2 \BE \trad{r.arg2}{\B{Expr}}} \\
+& \IN \LET v \BE \variable \IN \\
 & \llvm{"\PH{p_1}\quad\PH{p_2}\quad\PH{v} = i32 \PH{\opmap(\B{r.op})} \PH{t_1} \PH{v_1} \PH{v_2} \nl"} \sep v \sep \llvm{i32}
@@
   \trad{n}{\B{succ}} \isdef & \textbf{ let } p \sep v \sep t = \trad{n.arg}{\B{expr}} \IN  \\
-  & \quad \LET w = \variable \IN \\
+  & \quad \LET w \BE \variable \IN \\
   & \quad \quad \llvm{"\PH{p}} \\
@@
 \begin{split}
-  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell = \lbl \IN} \\
-  & \LET p_1 = \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
-  \AND p_2 = \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
+  \lefteqn{\trad{n, $\ell_t$, $\ell_f$}{\B{and}} \isdef \LET \ell \BE \lbl \IN} \\
+  & \LET p_1 \BE \trad{n.arg1, $\ell$, $\ell_f$}{\B{Form}}
+  \AND p_2 \BE \trad{n.arg2, $\ell_t$, $\ell_f$}{\B{Form}} \IN \\
   & \quad \llvm{"\PH{p_1} \PH{\ell} :  \nl \PH{p_2}"}
@@
 The definition of the translation for disjunctions is similar and is omitted.
+
+\paragraph{$\lambda$-expressions}
+
+[[[ TODO ]]]
 
@@
   \lefteqn{\trad{\B{op}}{\B{oper}} \isdef} \\
-  & \LET h = \trad{\B{op}}{sig} \AND   a = \trad{\B{op.body}}{alloc} \AND i = \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
+  & \LET h \BE \trad{\B{op}}{sig} \AND   a \BE \trad{\B{op.body}}{alloc} \AND i \BE \trad{\B{op.body}, \llvm{exit}}{\ListOf{\B{Inst}}}_L \IN  \\
   & \llvm{"define void \PH{h} \{\nl} \\
@@
 header: \\
-\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \%self\$, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
+\noindent$\trad{\B{o}}{sig} \isdef \llvm{"\PH{\Global(\B{o})}(\PH{\state(\B{o.root})}* \self, \PH{\trad{\B{o.inp}}{\ListOf{inp}}}, \PH{\trad{\B{o.out}}{\ListOf{out}}})"}$
 
@@
 & \ELSIF \B{i} \mbox{ is \B{Blk}} \THEN \trad{\B{i.body}, $\ell$}{\ListOf{\B{Inst}}}_L \\
-& \ELSE \LET p = \trad{\B{i}}{\B{Inst}} \IN \\
+& \ELSE \LET p \BE \trad{\B{i}}{\B{Inst}} \IN \\
 & \quad \llvm{"\PH{p}} \\
@@
 & \ELSE \\
-& \LET \B{i}, \B{il'} = \B{il}  \IN \\
+& \LET \B{i}, \B{il'} \BE \B{il}  \IN \\
 & \quad \quad \LET p_1 =
@@
   \IF \B{i} \mbox{ is \B{If} or \B{While}} \THEN \\
-  \quad \LET \ell' = \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
+  \quad \LET \ell' \BE \lbl \IN \llvm{"\PH{\trad{i, $\ell'$}{\B{Inst}}_L} \PH{\ell'}: \PH{p}"} \\
   \ELSE \llvm{"\PH{\trad{i}{\B{Inst}}} \PH{p}"} \IN
@@
   \trad{a}{\B{Beq}} \isdef
-  & \LET l \sep p \sep t'= \trad{i.lhs}{lv} \AND r \sep v \sep t = \trad{i.rhs}{\B{Expr}} \IN \\
+  & \LET l \sep p \sep t'= \trad{i.lhs}{lv} \AND r \sep v \sep t \BE \trad{i.rhs}{\B{Expr}} \IN \\
   & \llvm{"\PH{l}} \\
@@
 \lefteqn{\trad{n}{lv} \isdef} \\
-& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
+& \LET t \BE \llvm{"\PH{\trad{\B{n.type}}{Type}} *"} \IN \\
 & \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \ELSE \LET v = \variable \IN \\
-& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \ELSE \LET v \BE \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \self,} \\
 & \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
@@
   \trad{c}{\B{Call}} \isdef
-  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol =
+  & \LET p_i \sep il \BE \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol =
   \trad{c.out}{\ListOf{Out}} \AND \\
-  & \quad id = {\Global(\B{c})} \AND ty = \llvm{"\PH{\stateref(\B{c.root})}} \AND \\
-  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
-  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
-  & \quad \quad \ELSE \LET v_1, v_2 = \variable, \variable \IN \\
+  & \quad id \BE {\Global(\B{c})} \AND ty \BE \llvm{"\PH{\stateref(\B{c.root})}} \AND \\
+  & \quad val \sep p \BE \IF \B{c.inst} = \text{None} \\
+  & \quad \quad \THEN \llvm{"\self"} \sep \llvm{""} \\
+  & \quad \quad \ELSE \LET v_1, v_2 \BE \variable, \variable \IN \\
   & \quad \quad \quad v_2 \sep \\
-  & \quad \quad \quad \llvm{"\PH{v_1} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl} \\
+  & \quad \quad \quad \llvm{"\PH{v_1} = getelementptr \PH{ \state({\B{c.inst.root}})} \self, i32 0, i32 \PH{\idx(\B{c.inst})} \nl} \\
   & \quad \quad \quad \quad \llvm{\PH{v_2} = load \PH{ty}* \PH{v_1} \nl"} \\
@@
   \trad{i}{Inp} \isdef
-  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
+  & \LET p \sep v \sep t \BE \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
   \trad{o}{Out} \isdef
-  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
+  & \LET p \sep v \sep t \BE \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
 \end{align*}
@@
 \begin{split}
-  \trad{\B{b bl}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \IN \\
+  \trad{\B{b bl}, $\ell$}{\ListOf{\B{IfBr}}} \isdef & \LET \ell_1 \BE \lbl \AND \ell_2 \BE \lbl \IN \\
   & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell_2$}{\B{Pred}}} } \\
@@
   \trad{b.body, $\ell$}{\ListOf{Inst}}_L\\
-  & \ELSE \LET \ell_1 = \lbl \IN \\
+  & \ELSE \LET \ell_1 \BE \lbl \IN \\
   & \llvm{"\PH{\trad{b.cond, $\ell_1$, $\ell$}{\B{Pred}}} } \\
@@
 \begin{align*}
-  \trad{v, $\ell$}{\B{Case}} \isdef & \LET p \sep v \sep t = \trad{v.expr}{\B{Expr}} \IN \\
-  & \LET \ell_o = \lbl \IN \\
-  & \LET j, b = \trad{v.branches, $\ell_o$, $\ell$}{\ListOf{CaseBr}} \IN \\
+  \trad{v, $\ell$}{\B{Case}} \isdef & \LET p \sep v \sep t \BE \trad{v.expr}{\B{Expr}} \IN \\
+  & \LET \ell_o \BE \lbl \IN \\
+  & \LET j, b \BE \trad{v.branches, $\ell_o$, $\ell$}{\ListOf{CaseBr}} \IN \\
   & \quad \llvm{"\PH{p}} \\
@@
   \trad{b bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \isdef
-  & \LET \ell = \lbl \IN \\
-  & \LET j, b = \trad{bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \IN \\
+  & \LET \ell \BE \lbl \IN \\
+  & \LET j, b \BE \trad{bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \IN \\
   & \quad \llvm{"\PH{\trad{b.values, $\ell$}{\ListOf{\B{CaseVal}}}} \PH{j}"}
@@
   & \quad \llvm{""} \sep \llvm{"\PH{\ell_o}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L}"} \\
-  & \ELSE \LET \ell_b = \lbl \IN \\
+  & \ELSE \LET \ell_b \BE \lbl \IN \\
   & \quad \trad{b.values, $\ell_b$}{\ListOf{\B{CaseVal}}} \sep \\
@@
 \begin{align*}
-  \trad{v, $\ell$}{\B{CaseVal}} \isdef & \LET p \sep v \sep t = \trad{v}{\B{Expr}} \IN \\
+  \trad{v, $\ell$}{\B{CaseVal}} \isdef & \LET p \sep v \sep t \BE \trad{v}{\B{Expr}} \IN \\
   & \quad \llvm{"\PH{t} \PH{v}, label \%\PH{\ell} \nl"}
@@
 \trad{w, $\ell$}{\B{While}} \isdef
-  & \LET \ell_1 = \lbl \AND \ell_2 = \lbl \AND c \sep v = \trad{w.cond}{\B{Pred}} \IN \\
+  & \LET \ell_1 \BE \lbl \AND \ell_2 \BE \lbl \AND c \sep v \BE \trad{w.cond}{\B{Pred}} \IN \\
   & \llvm{"br label \%\PH{\ell_1} \nl} \\
@@
 
+We first define $\argname$, a function that given a list, yields a naming
+function for the elements of the list:
+\begin{align*}
+  \argname(l) \isdef
+  & \lambda c \cdot \llvm{"\%arg\PH{\mathit{index}(c, l)}"}
+\end{align*}
+So $\argname(l)$ is a function that returns \llvm{\%arg0}, \llvm{\%arg1}, \ldots
+for the first, second, \ldots elements of $l$.
+
 Initialization of a module is implemented by a LLVM function that takes as
 parameters: the address of the aggregate representing the state of the module,
-which we name \llvm{\%self}, and the addressses of the aggregates representing
-the state of each component of the module. 
-We first define a naming function $\argname$ having as domain the list of
-\emph{stateful} module components:
-\begin{align*}
-  \argname(l) \isdef
-  & \lambda c \cdot \llvm{"\%arg\PH{\mathit{index}(c, l)}"}
-\end{align*}
-So \llvm{\%arg0}, \llvm{\%arg1}, \ldots are the identifiers for the parameters
-representing the components of the initialized module.
-
-More precisely, the specification needs to make a distinction between stateful
-and stateless components, as only the former are represented by a LLVM
-aggregate. We use the expression $\components(\B{n}) \isdef
-\mathit{remove}(\trimportedinstances(\B{n}), \stateless)$ to denote the list of
-stateful components of $\B{n}$, so that $\argname(\components(\B{n}))$ is the
-function that, given a stateful component $c$ of \B{n}, returns the parameter of
-the initialization function of $c$. We are now ready to specify the 
-signature of this initialization function, using an auxiliary translation
-function to produce the LLVM parameter type and name for each stateful
-component $\B{c}$:
+which we name \self, and the addressses of the aggregates representing
+the state of each component of the module.  More precisely, the specification
+needs to make a distinction between stateful and stateless components, as only
+the former are represented by a LLVM aggregate. We use the expression
+$\components(\B{n}) \isdef \mathit{remove}(\trimportedinstances(\B{n}),
+\stateless)$ to denote the list of stateful components of $\B{n}$, so that
+$\argname(\components(\B{n}))$ is the function that, given a stateful component
+$c$ of \B{n}, returns the parameter of the initialization function of $c$. We
+are now ready to specify the signature of this initialization function, using an
+auxiliary translation function to produce the LLVM parameter type and name for
+each stateful component $\B{c}$:
 \begin{align*}
 \trad{\B{n}}{init-sig} \isdef &
-\llvm{"\PH{\init(\B{n})}(\PH{\stateref(\B{n})} \%self, 
+\llvm{"\PH{\init(\B{n})}(\PH{\stateref(\B{n})} \self, 
   \PH{\trad{\components(\B{n}), \argname(\components(\B{n}))}{\ListOf{arg}}})"} \\
-\trad{\B{c}, \argname}{arg} \isdef & \llvm{"\stateref(\B{c.root}) \argname(\B{c})"}
+\trad{\B{c}, \argname}{arg} \isdef & \llvm{"\stateref(\B{c.mach}) \argname(\B{c})"}
 \end{align*}
@@
   \lefteqn{\trad{\B{n}, \B{c}, \argname}{BindImp} \isdef} \\
-  & \LET v = \variable \IN \\
-  & \quad \llvm{"\PH{v} = getelementptr \PH{\stateref(\B{n})} \%self\$, i32 0, i32 \PH{\idx(\B{c})}\nl}\\
-  & \quad \llvm{store \PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}, 
-    \PH{\stateref(\B{c.root})}* \PH{v}\nl"}
+  & \LET v \BE \variable \IN \\
+  & \quad \llvm{"\PH{v} = getelementptr \PH{\stateref(\B{n})} \self, i32 0, i32 \PH{\idx(\B{c})}\nl}\\
+  & \quad \llvm{store \PH{\stateref(\B{c.mach})} \PH{\argname(\B{c})}, 
+    \PH{\stateref(\B{c.mach})}* \PH{v}\nl"}
 \end{align*}
@@
 \trad{\B{c}, \argname}{arg} \isdef& 
-  \llvm{"\PH{\stateref(\B{c.root})} \PH{\argname(\B{c})}"}
+  \llvm{"\PH{\stateref(\B{c.mach})} \PH{\argname(\B{c})}"}
 \end{align*}

### 2014-03-27 DOCUMENTATION/AdvancedTypes.tex
Fixing the rule CAST

@@
 &\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\ \\
-&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \ell = \lbl \IN \\
+&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \PH{\ell} = \lbl \IN \\
 &\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
-&\quad \ELSE \LET  p \sep v \sep v' \sep t'  = \trad{p,t,v,itype,istart}{Cast} \IN \\
-&\quad \llvm{",\PH{t} \PH{l} "} \sep  \llvm{"\PH{p} \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
+&\quad \ELSE \LET  p' \sep v \sep v' \sep t'  = \trad{t,v,itype,istart}{Cast} \IN \\
+&\quad \llvm{",\PH{t} \PH{\ell} "} \sep  \llvm{"\PH{p} \PH{p'}  \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
 \end{align*}
@@
 \begin{align*}
-&\trad{p,t,v,t',v'}{Cast} \isdef \LET  \ell = \lbl \IN \\
+&\trad{t,v,t',v'}{Cast} \isdef \LET  \ell = \lbl \IN \\
 &\quad \IF ( t == t') \THEN p \sep v \sep v' \sep t  \\
 &\quad \ELSE \IF (t<t') \THEN \\
-&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t} \PH{v} , \PH{t'}\nl"}  \sep l \sep v' \sep t'\\ 
+&\quad \quad \llvm{"\PH{\ell}= bitcast \PH{t} \PH{v} , \PH{t'}\nl"}  \sep l \sep v' \sep t'\\ 
 &\quad \ELSE \\ 
-&\quad \quad \llvm{"\PH{p} \PH{l}= bitcast \PH{t'} \PH{v'} , \PH{t}\nl"} \sep v \sep l \sep t \\ 
+&\quad \quad \llvm{"\PH{\ell}= bitcast \PH{t'} \PH{v'} , \PH{t}\nl"} \sep v \sep l \sep t \\ 
 \end{align*}

### 2014-03-31 DOCUMENTATION/specification.tex
Adding in AST LLVM array types
Added support for enumerated sets and enumerated values.

@@
 \newcommand{\IN}[0]{\textbf{ in }}
-\newcommand{\AND}[1]{\textbf{ and }}
+\newcommand{\AND}[0]{\textbf{ and }}
 \newcommand{\PH}[1]{\framebox{$#1$}}
@@
         \B{args} & : & seq \B{Pred} \\
+	\hline
+        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
+        \hline
+        \B{op} & : & \B{RelOp} \\
+        \B{arg1}, \B{arg2} & : & \B{Exp}
 \end{tabular}
@@
         \hline
-        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
-        \hline
-        \B{op} & : & \B{RelOp} \\
-        \B{arg1}, \B{arg2} & : & \B{Exp} \\
-        \hline
         \multicolumn{3}{|c|}{Data type: \B{Type}} \\
@@
         \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
-	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType}}\\
+	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType} \alt} \\
+	\multicolumn{3}{c}{\B{Enumeration}} \\
+	\hline
+  	\multicolumn{3}{|c|}{Enumerated set: \B{Enumeration}} \\
+        \hline
+        \B{elements} & : & seq \B{Enumerated} \\
+	\hline
+  	\multicolumn{3}{|c|}{Enumerated value: \B{Enumerated}} \\
+        \hline
+        \B{id} & : & \B{name} \\
+        \B{type} & : & \B{Enumeration} \\
 	\hline
@@
       \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
-      \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
+       \nt{type} & ::= & \llvm{void} \alt \nt{itype} \alt \textcolor{blue}{$\left[ iliteral\ x\ type  \right]$} \alt \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \alt \nt{type}\llvm{*} \\
       \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
@@
   one-bit integer:
-$$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}.$$
+$$\trad{BOOL}{\B{Type}} \isdef \llvm{"i1"}.$$
 \item[integers] Atelier-B offers three possible definitions for integers,
@@
   generator has an option to specify which definition is employed, and we have:
-$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
+$$\trad{INT}{\B{Type}} \isdef \llvm{"i16"} \mid \llvm{"i32"} \mid \llvm{"i64"}.$$
 \item[enumerated sets] Again, enumerations are encoded as integer types, and the
   width of the type depends on the size of the enumeration.
-$$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$
+$$\trad{\B{n}}{\B{Type}} \isdef \LET w \BE \log_2 \mathit{size}(\B{n.elements}) \IN \llvm{i\PH{w}}.$$
 \item[arrays] In LLVM, an array type is defined by its size and the type of its
@@
 
-For identifiers, three cases are possible: global constants, local variables,
-and module state variables. For state variables, the translation issues
-statements to assign the address of the corresponding element in the state
-structure to pointer variable $p$ and then dereference $p$ to assign $v$:
+Enumerated values are encoded as integers, using as code their position in the
+corresponding set declaration: $\trad{\B{n}}{\B{Enum}} \isdef \LET t =
+\trad{\B{n.type}}{\B{Type}} \AND v = \mathit{index}(\B{n}, \B{n.type.elements})
+\IN \llvm{""} \sep \PH{v} \sep \PH{t}$.
+
+For identifiers, the following cases are possible: global constants, local
+variables, and module state variables. For state variables, the translation
+issues statements to assign the address of the corresponding element in the
+state structure to pointer variable $p$ and then dereference $p$ to assign $v$:
 \begin{align*}

### 2014-03-31 b2llvm/printer.py

@@ -30,6 +30,8 @@ def term(n):
         return (SP + term(n["base"])+"("+ term(n["index"])+ ")")
     elif kind == "Vari":
         return n["id"]
+    elif kind == "Enumerated":
+        return n["id"]
     elif n["op"] in { "succ", "pred" }:
         return n["op"] + "(" + term(n["args"][0]) + ")"
     elif n["op"] in { "+", "*", "-" }:
@@ -67,7 +69,7 @@ def condition(n):
         return "<<UNRECOGNIZED >>"
 
 def type(n):
-    return n
+    return n["id"]
 
 def imports(n):
     result = ""
@@ -75,6 +77,11 @@ def imports(n):
         result += n["pre"] + "."
     result += n["mach"]["id"]
     return result
+
+def sets(n):
+    result = ""
+    return (n["id"] + " = " + 
+            "{" + commas([e["name"] for e in n["elements"]]) + "}")
 
 def value(n):
     return n["id"] + " = " + term(n["value"])
@@ -259,13 +266,17 @@ def oper(n):
 
 def implementation(n):
     global SP, TB, NL
-    imp, consts, vars = n["imports"], n["concrete_constants"], n["variables"]
+    imp, sets = n["imports"], n["sets"]
+    consts, vars = n["concrete_constants"], n["variables"]
     init, ops = n["initialisation"], n["operations"]
     result = ""
     result += "IMPLEMENTATION" + SP + n["id"] + NL
     if imp != []:
         result += "IMPORTS" + NL
         result += TB + commas([imports(e) for e in imp]) + NL
+    if sets != []:
+        result += "SETS" + NL
+        result += TB + commas([sets(e) for e in sets]) + NL        
     if consts != []:
         result += "VALUES" + NL
         result += TB + commas([value(e) for e in consts]) + NL

### 2014-03-31 b2llvm/translate.py

@@ -11,6 +11,7 @@
 # tries to proceed processing the input.
 #
 ###
+import math
 
 import b2llvm.ast as ast
 import b2llvm.loadbxml as loadbxml
@@ -134,7 +135,7 @@ def translate_mode_proj(buf, m, emit_printer):
     Appends to text the LLVM code corresponding to the LLVM code generation for m
     in project mode.
     '''
-    check_kind(m, "Machine")
+    check_kind(m, { "Machine" })
     assert is_developed(m)
     buf.trace.outu("Preamble")
     buf.trace.outu("")
@@ -374,16 +375,21 @@ def section_implementation(buf, m, emit_printer):
 
 ### TYPE TRANSLATION ###
 
+def bit_width(n):
+    """Returns the number of bits to represent n different values."""
+    return str(max(1, math.log(n, 2)))
 #
 # This function is responsible for translation B0 type names to LLVM types
 #
 def x_type(t):
     """Returns the type for declaration."""
-    assert(t == ast.INTEGER or t == ast.BOOL or (t.get("kind")== "arrayType") )
+    check_kind(t, {"Integer", "Bool", "Enumeration", "arrayType"})
     if (t == ast.INTEGER):
         return "i32" 
     if (t == ast.BOOL):
         return "i1"
+    if (t["kind"] == "Enumeration"):
+        return "i"+str(bit_width(len(t["elements"])))
     if (t.get("kind")== "arrayType"):
         return x_arrayType(t)
 
@@ -1177,13 +1183,16 @@ def x_expression(buf, n):
       A pair containing, the identifier of the LLVM temporary variable storing the value
       of the expression, and the LLVM type of this temporary variable.
     '''
-    check_kind(n, {"IntegerLit", "BooleanLit", "Vari", "Term", "Cons"})
+    check_kind(n, {"IntegerLit", "BooleanLit", "Enumerated", 
+                   "Vari", "Term", "Cons"})
     buf.trace.out("Evaluate expression \""+ellipse(printer.term(n))+"\".")
     buf.trace.tab()
     if n["kind"] == "IntegerLit":
         res = x_integerlit(buf, n), "i32"
     elif n["kind"] == "BooleanLit":
         res = x_booleanlit(buf, n), "i1"
+    elif n["kind"] == "Enumerated":
+        res = x_enumerated(buf, n)
     elif n["kind"] == "Vari":
         res = x_name(buf, n)
     elif n["kind"] == "Term":
@@ -1223,6 +1232,23 @@ def x_booleanlit(buf, n):
[…]

### 2014-03-31 coverage.sh

@@ -10,6 +10,7 @@ echo "Cleaning up existing coverage data."
 coverage erase
 echo "Generating new coverage data."
 $COV_RUN ./b2llvm.py -v -m comp counter counter.llvm bxml project.xml > /dev/null
+$COV_RUN ./b2llvm.py -m comp enumeration enumeration.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -m comp array array.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -m comp ioint ioint.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -m comp counter counter.llvm bxml project.xml

### 2014-04-01 DOCUMENTATION/specification.tex
Adding information about the AST relating each Derived Type

--- 2014-03-19 DOCUMENTATION/specification.tex
@@
 \usepackage{minted}
+\usepackage{array}
 
@@
 
-\section{The abstract syntax tree with support for advanced types}
+\section{The abstract syntax tree with support for derived types}
 
 \textcolor{blue}{
-The supported advanced types are array and structures. The nodes Array and 
+The supported derived types are array and structures. The nodes Array and 
 Structure can be used to declare a variable, or  declare an element, or to 
@@
         \hline
-	\B{base} & : &  \B{name} \\
+	\B{base} & : &  \B{Name} \\
 	\B{element} & : & seq \B{Name} \\
@@
 $$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$
+
+\end{description}
+
+\subsection{Derived Data Types}
+
+The data types \textbf{structures} and \textbf{arrays} require detailed 
+representation in the abstract syntax structure from the table \ref{tab:node-attr}.
+The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.
+
+The following examples relate the B code and their abstract syntax representation. The \textcolor{gray}{gray elements} are not related to their abstract syntax tree to simplify the examples.
+
+\begin{description}
+
+\item[structures] The structure types of B are naturally encoded as structure
+  types in LLVM.
+
+Declaring an structure type:
+\[
+  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
+\]
+Initializing a record with a sequence of expressions:
+\[
+  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
+\]
+
+Setting a record field to a variable:
+\[
+  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
+\]
+
 \item[arrays] In LLVM, an array type is defined by its size and the type of its
@@
   determine the size of the array and multi-dimensionality.
-
-  [[[ TODO ]]]
-
-\item[structures] The structure types of B are naturally encoded as structure
-  types in LLVM.
-
-  [[[ TODO ]]]
+Declaring an array type:
+\[
+\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
+\]
+Initializing an array:
+\[
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{1 \mapsto  100,2 \mapsto 101}_{seq\ Map}\}}_{Array}
+\]
+\[
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{0,1\}}_{Exp} }_{Product}\}}_{Array}
+\]
+
+Updating a selected element from array by a value of variable:
+\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
+\]
+
+Setting a selected element from array to a variable:
+\[ 
+\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \]
 \end{description}
@@
 The current specification of the code generator does not handle deferred sets.
-
-\section{Translation: Expressions}
+These data types and logic/math operators compose expressions that is in 
+the next section.
+
+\section{Translation Expressions}
 \label{sec:expr}
@@
 
-\include{AdvancedTypes}
+\include{DerivedTypes}
 

### 2014-04-01 DOCUMENTATION/specification.tex

@@
   generator has an option to specify which definition is employed, and we have:
-$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
+$$\trad{INT}{\B{Type}} \isdef \llvm{"i16"} \mid \llvm{"i32"} \mid \llvm{"i64"}.$$
 \item[enumerated sets] Again, enumerations are encoded as integer types, and the

### 2014-04-01 DOCUMENTATION/DerivedTypes.tex
Removing C examples

--- 2014-03-27 DOCUMENTATION/AdvancedTypes.tex
@@
 \section{Translation of Structures and Array}
+\label{sec:StructArray}
 
@@
 
-\section{Arrays}
+\subsection{Arrays}
 
@@
 	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
-	\item $\{1\vdash1\vdash4, 1\vdash2\vdash5\}  \in (INT \times INT) \times INT$ (an initialised array)
+	\item $\{1\mapsto1\mapsto4, 1\mapsto2\mapsto5\}  \in (INT \times INT) \times INT$ (an explicit array)
 \end{itemize}
 
-\subsection{Relation arrays B0 and LLVM}
+\subsubsection{Relation arrays B0 and LLVM}
 
@@
 
-The following blocks illustrate a C code setting a value to array
+The following blocks illustrate a B code setting a value to array
 and its translation to LLVM.
 
-\begin{ccode} 
-int x=2; int y=3; int z=4;
-Array2[x][y][z]=7;
-\end{ccode}
+\begin{pascalcode} 
+VAR xx, yy, zz IN
+	xx=2;
+	yy=3;
+	zz=4;
+	Array2(x,y,z):=7
+END
+\end{pascalcode}
 
@@
 \begin{llvmcode}
-%x = alloca i32
-%y = alloca i32
-%z = alloca i32
-store i32 2, i32* %x
-store i32 3, i32* %y
-store i32 4, i32* %z
-%1 = load i32* %z
+%xx = alloca i32
+%yy = alloca i32
+%zz = alloca i32
+store i32 2, i32* %xx
+store i32 3, i32* %yy
+store i32 4, i32* %zz
+%1 = load i32* %zz
 %2 = sext i32 %1 to i64
-%3 = load i32* %y
+%3 = load i32* %yy
 %4 = sext i32 %3 to i64
-%5 = load i32* %x
+%5 = load i32* %xx
 %6 = sext i32 %5 to i64

### 2014-04-01 b2llvm/ast.py
Merge branch 'master' of https://bitbucket.org/ddeharbe/b2llvm
Conflicts:
DOCUMENTATION/specification.tex

@@ -15,8 +15,34 @@ oriented towards the implementation of the translation to LLVM.
 
 ### TYPES ###
 
-INTEGER = "INTEGER"
-BOOL = "BOOL"
+# We are only concerned with representing concrete data.
+# The following types of concrete data are supported
+# - INTEGERS
+# - BOOL
+# - Enumerations
+# The following types of concrete data are partially supported 
+# and are being handled by Valerio, the corresponding code is
+# present elsewhere and shall eventually be merged in this section:
+# - Arrays (=> Valerio)
+# - Intervals (=> Valerio)
+#
+
+INTEGER = { "kind": "Integer", "id" : "INTEGER" }
+BOOL = { "kind": "Bool", "id": "BOOL" }
+
+def make_enumerated(name):
+    """Creates an AST node for an element of an enumerated set.
+    The created node does not have the type attribute set."""
+    return { "kind": "Enumerated", "id": name }
+
+def make_enumeration(name, elements):
+    """Creates an AST node for an enumerated set. Sets the
+    type nodes for all the elements to the created node."""
+    assert type(elements) is list
+    n = { "kind": "Enumeration", "id": name, "elements": elements }
+    for e in elements:
+        e["type"] = n
+    return n
 
 ### TERMINALS ###
 
@@ -84,18 +110,6 @@ def become_list(elem):
     if (type( elem ) == list ): 
         return elem
     else: return  [elem]
-
-def make_sset_bool():
-    """Creates an AST node for a B simple bool set."""
-    return { "kind" : "set_BOOL" }
-
-def make_sset_nat():
-    """Creates an AST node for a B simple natural set."""
-    return { "kind" : "set_NAT" }
-
-def make_sset_int():
-    """Creates an AST node for a B simple integer set."""
-    return { "kind" : "set_INT" }
 
 def make_interval(start,end):
     """Creates an AST node for a B simple interval set."""
@@ -272,10 +286,11 @@ def make_import(mach, prefix = None):
     assert(mach["kind"] in { "Machine" })
     return { "kind": "Module", "mach": mach, "pre": prefix }
 
-def make_implementation(name, imports, consts, variables, init, ops):
+def make_implementation(name, imports, sets, consts, variables, init, ops):
     """Creates an AST node for a B implementation module."""
     assert type(imports) is list
     assert type(consts) is list
+    assert type(sets) is list
     assert type(variables) is list
     assert type(init) is list
     assert type(ops) is list
[…]

### 2014-04-01 b2llvm/loadbxml.py

@@ -170,10 +170,12 @@ def load_base_machine(dirname, loaded, machine):
     root = tree.getroot()
     assert root.tag == 'Machine'
     assert root.get("type") == "abstraction"
+    sets = []
     constants = load_values(root, symast)
     variables = load_concrete_variables(root, symast)
     operations = load_operations(root, symast, symimp)
-    pymachine = ast.make_base_machine(machine, constants, variables, operations)
+    pymachine = ast.make_base_machine(machine, sets, constants, variables, 
+                                      operations)
     symast.clear()
     symimp.clear()
     _LOADING.remove(machine)
@@ -215,9 +217,10 @@ def load_developed_machine(dirname, project, loaded, machine):
     assert root.tag == 'Machine'
     assert root.get("type") == "abstraction"
     assert name_attr(root) == machine
+    sets = []
     impl_id = project.implementation(machine)
     pyimpl = load_implementation(dirname, project, loaded, impl_id)
-    pymachine = ast.make_developed_machine(machine, pyimpl)
+    pymachine = ast.make_developed_machine(machine, sets, pyimpl)
     pyimpl["machine"] = pymachine
     _LOADING.remove(machine)
     loaded.set(machine, pymachine)
@@ -246,6 +249,8 @@ def load_implementation(dirname, project, loaded, module):
     if module in _LOADING:
         _LOG.error("there seems to be a recursive dependency in imports")
     symast = SymbolTable()
+    symast.add("BOOL", ast.BOOL)
+    symast.add("INTEGER", ast.INTEGER)
     symimp = SymbolTable()
     _LOADING.add(module)
     tree = ET.parse(path(dirname, module))
@@ -255,14 +260,15 @@ def load_implementation(dirname, project, loaded, module):
     impl_name = name_attr(root)
     assert impl_name == module
     imports = load_imports(root, symast, symimp, dirname, project, loaded)
+    sets = load_sets(root, symast)
     constants = load_values(root, symast)
     variables = load_concrete_variables(root, symast)
     initialisation = load_initialisation(root, symast, symimp)
     operations = load_operations(root, symast, symimp)
     symast.clear()
     symimp.clear()
-    pyimpl = ast.make_implementation(impl_name, imports, constants, variables,
-                                initialisation, operations)
+    pyimpl = ast.make_implementation(impl_name, imports, sets, constants, 
[…]

### 2014-04-04 DOCUMENTATION/specification.tex
Organising the position of derived types section.

@@
 
+\include{derivedTypes}
+
 \section{Translation: Control Flow and Instructions
@@
 
+\paragraph{Becomes equal instructions} \textcolor{blue}{ must evaluate the
+ target and the source of the assignment, and then copy the result of the 
+latter in the former, using the LLVM instruction \llvm{store}. The evaluation 
+of the target is defined with $\trad{}{lvalue}$, which yields the corresponding 
+LLVM source code $l$ and the variable $p$ containing the assigned location. The 
+evaluation of the source yields the corresponding code $r$, the variable $v$ 
+holding the result, and its type $t$. }
+
+\begin{align*}
+\begin{split}
+  \trad{a}{\B{Beq}} \isdef
+  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
+  & \llvm{"\PH{l}} \\
+  & \llvm{\PH{r}} \\
+  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
+\end{split}
+\end{align*}
+
+The rule lvalue is:
+\begin{align*}
+\lefteqn{\trad{n}{lvalue} \isdef} \\
+& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
+& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
+& \quad \IF \mbox{\B{n} is an ArrayItem} \THEN  \trad{n}{ArrayItem} \\
+& \quad \IF \mbox{\B{n} is a RecField} \THEN  \trad{n}{RecField} \\
+& \quad \ELSE \LET v = \variable \IN \\
+& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
+& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
+\end{align*}
+
+\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
+care. The attributions between these elements are made  directly by the memory 
+address.  However, the expression resolution are made by parts, where the 
+partial results are stored to temporary variables. Therefore, the expression 
+resolution with pointers needs an additional instruction $LOAD$ for store the 
+partial results. This translation is represented by $ValueItem$.}
+
+\begin{align*}
+\begin{split}
+  \trad{n}{\B{ValueItem}} \isdef \\
+   & \LET \ell = \lbl  \IN\\
+   & \IF \B{n} \mbox{ is \B{RecField}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{RecField}} \\
+   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
+   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
+   & \sep  \ell  \sep t'\\
+\end{split}
+\end{align*}
+
 \paragraph{Call up} instructions apply B operations with some values and obtain
@@
 
-\include{DerivedTypes}
-
 \include{Attachments}

### 2014-04-04 C/AdvancedTypes/Array_i.c
Renaming folder (AdvancedTypes to derivedTypes)

@@ -1,115 +1,0 @@
-#include "stdio.h"
-#include "Array.h"
-
-/* Clause CONCRETE_CONSTANTS */
-/* Basic constants */
-
-/* Array and record constants */
-/* Clause CONCRETE_VARIABLES */
-
-static int32_t Array2[10][20][30];
-static int32_t Array__arr[100];
-static int32_t Array__arr_n[100][100];
-static int32_t Array__tmp[100];
-
-void Array__setZero(void);
-void Array__setZero2(void);
-void Array__setZeroSimp(void);
-void Array__setArray(void);
-
-int main(void){
-   Array__setZero();
-   Array__setZero2();
-   printf("%d\n", Array2[2][3][4]);
-  return 0;
-}
-
-/* Clause INITIALISATION */
-void Array__INITIALISATION(void)
-{
-    Array__arr[1] = 100;
-    Array__arr_n[0][0] = 10;
-    memmove(Array__tmp,Array__arr,100 * sizeof(int32_t));
-    Array__setZero();
-
-}
-
-/* Clause OPERATIONS */
-void Array__setZero(void)
-{   int32_t x=2; int32_t y=3 ; int32_t z=4 ;
-    Array2[x][y][z]=7;
-}
-
-void Array__setZero2(void)
-{
-    int x =4;
-    x= x-2;
-    Array2[x][3][4]=7;
-}
-void Array__setArray(void)
-{
-    int32_t cprogram[3][2][4]={
-        {{0, 022, 03, 43}, {23, 47, -9, 2}},
-        {{0, 36, 45, 4}, {12, 24, 022, -1}},
-        {{8, 32, 342, 01}, {21, 32, 43, -20}}
-    };
-    int32_t Array3[1][2][1]= {{{4},{5}}};
-    /*{ { {1}, {2} , {3}},
-        
-              { {4}, {5} , {6}},
-              { {7}, {8} , {9}}
-             } ;*/
-}
-void Array__setZeroSuggestedByDavid(void)
-{  /* David gives a similar sugestion wit intermediate steps */   
-
-	Array__arr_n[2][5]=4;
-    	printf("%d\n",Array__arr_n[2][5]);     
-	int * mid = Array__arr_n[2];
-	printf("%d\n",mid[5]);
-}
-
-
-
-
-void Array__set(int32_t ix, int32_t tt)
-{
-    if(((((ix) >= (0)) &&
-            ((ix) <= (99))) &&
[…]

### 2014-04-04 C/AdvancedTypes/Rec.h

@@ -1,32 +1,0 @@
-#ifndef _Rec_h
-#define _Rec_h
-
-#include <stdint.h>
-#include <stdbool.h>
-#ifdef __cplusplus
-extern "C" {
-#endif /* __cplusplus */
-
-
-/* Clause SETS */
-
-/* Clause CONCRETE_VARIABLES */
-
-
-/* Clause CONCRETE_CONSTANTS */
-/* Basic constants */
-/* Array and record constants */
-extern void Rec__INITIALISATION(void);
-
-/* Clause OPERATIONS */
-
-extern void Rec__positive(bool *res);
-extern void Rec__withrdaw(int32_t amt);
-extern void Rec__unsafe_dec(void);
-
-#ifdef __cplusplus
-}
-#endif /* __cplusplus */
-
-
-#endif /* _Rec_h */

### 2014-04-04 C/AdvancedTypes/array_example.c

@@ -1,23 +1,0 @@
-#include <stdio.h> 
-	//int arr[99];
-        struct tstruct { int arr[99]; int anyVar1; int anyVar2;  } ref;
-	const int start_arr =1  ;
-
-void set();
-void get();	
-
-int main (){
-	set();
-
-	printf("%d",ref.arr[4]);
-}
-
-void set(){
-	int ind = 7;
-	int tmp = 4 - start_arr;
-	ref.arr[tmp]=5;
-}
-
-void get(){
-	int a= ref.arr[4];
-}

### 2014-04-22 DOCUMENTATION/specification.tex
Removing old AST and small changes in paper.

@@
 \institute{Federal University of Rio Grande do Norte\\
-Federal Institute of Education Science and Technology\\
+Federal Institute of Education Science and Technology of Rio Grande do Norte\\
  Natal (Brazil)}
@@
 Note that the translation presented in this paper only considers types \B{INT}
-(integer) and \B{BOOL} (Boolean). Accordingly, the expression language consists
-in arithmetic operations \B{-} (unary and binary), \B{+}, \B{*}, \B{/}, \B{mod},
-% \B{**} (exponentiation),
-\B{succ} and \B{pred} and Boolean operations
-\B{$\land$}, \B{$\lor$}, \B{$\neg$} as well as relations \B{=}, \B{$\neq$},
-\B{$>$}, \B{$<$}, \B{$\leq$} and \B{$\neq$}. Atomic expressions are identifiers,
-Boolean constants \B{FALSE} and \B{TRUE}, integer literals, and integer
-constants \B{MAXINT} and \B{MININT}. The development through the application of
-the B method ensures the absence of overflows and underflows.
-
-\begin{figure}[t]
+(integer), \B{BOOL} (Boolean), enumerated set, arrays and structures.
+All elements and data types from B code are represented in abstract syntax tree.
+It is illustrated in Figure~\ref{tab:node-attr} and used as an intermediate representation of B code.
+
+\begin{figure}[H]
   \begin{center}
@@
       \frame{
-    \begin{tabular}[t]{ccccc}
+    \begin{tabular}[t]{ccc}
       \begin{tabular}[t]{rcl}
@@
         \B{type} & : & \B{Type} \\
-        \B{scope} & : & \B{Impl} \alt \B{Oper}
-        \\
-      \end{tabular}
-      & &
-      \begin{tabular}[t]{rcl}
+        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
 	\hline
@@
         \B{branches} & : & seq \B{IfBr} \\
-        \hline
-        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
-        \hline
-        \B{lhs} & : & \B{Vari} \\
-        \B{rhs} & : & \B{Expr} \\
-        \hline
-        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
-        \hline
-        \B{op} & : & \B{Oper} \\
-        \B{inp} & : & seq \B{Expr} \\
-        \B{out} & : & seq \B{Name} \\
-        \B{inst} & : & opt \B{Impo}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
-        \hline
-        \B{expr} & : & \B{Expr} \\
-        \B{branches} & : & seq \B{CaseBr}
-      \end{tabular}
-      & &
-      \begin{tabular}[t]{rcl}
-        \hline
-        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
-        \hline
-        \B{val} & : & seq \B{Value} \\
-        \B{body} & : & \B{Inst}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{While loop: \B{While}} \\
-        \hline
-        \B{cond} & : & \B{Pred} \\
-        \B{body} & : & \B{Inst}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
-        \hline
-        \B{cond} & : & opt \B{Pred} \\
-        \B{body} & : & \B{Inst} \\
-        \hline
-        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
-        \hline
-        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Vari}} \\
-        \hline
-        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
-        \hline
-        \B{op} & : & \B{ArithOp} \\
-        \B{args} & : & seq \B{Exp} \\
-        \hline
-        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
-        \hline
-        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
-        \hline
-        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
-        \hline
-        \B{op} & : & \B{BoolOp} \\
-        \B{args} & : & seq \B{Pred} \\
-        \hline
-        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
-        \hline
-        \B{op} & : & \B{RelOp} \\
-        \B{arg1}, \B{arg2} & : & \B{Exp} \\
-        \hline
-        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
-        \hline
-        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}}
-      \end{tabular}
-    \end{tabular}
-  }
-  }
-    \caption{Abstract syntax structure of a B implementation.  For each abstract
-      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
-      branch in a case instruction, and the different attributes of the class),
-      or a list of the possible sub-classes (e.g. \B{Inst} for the different
-      kinds of instructions).
-      Sequence and optional attributes are denoted ``seq'' and ``opt'',
-      respectively}
-    \label{tab:node-attr}
-  \end{center}
-\end{figure}
-
-\section{The abstract syntax tree with support for derived types}
-
-\textcolor{blue}{
-The supported derived types are array and structures. The nodes Array and 
-Structure can be used to declare a variable, or  declare an element, or to 
-select an element in the array or the structure. 
-}
-
-\begin{figure}[H]
-  \begin{center}
-    {\footnotesize
-      \frame{
-    \begin{tabular}[t]{ccc}
-      \begin{tabular}[t]{rcl}
-        \hline
-        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{const} & : & seq \B{Cons} \\
-        \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} \\
-        \B{impl} & : & opt \B{Impl} \\
-        \hline
-        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{impo} & : & seq \B{Import} \\
-        \B{const} & : & seq \B{Cons} \\
-        \B{var} & : & seq \B{Vari} \\
-        \B{init} & : & seq \B{Inst} \\
-        \B{op} & : & seq \B{Oper} \\
-        \hline
-        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
-        \hline
-        \B{mach} & : & \B{Machine} \\
-        \B{pre} & : & opt \B{Name}
-        \\
-        \hline
-        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{type} & : & \B{Type} \\
-        \B{val} & : & \B{Value} \\
-        \hline
-        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{type} & : & \B{Type} \\
-        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
-	\hline
-        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
-        \hline
-        \B{id} & : & \B{Name} \\
-        \B{inp} & : & seq \B{Vari} \\
-        \B{out} & : & seq \B{Vari} \\
-        \B{body} & : & \B{Inst}
-        \\
-       \hline
-        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
-        \hline
-        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
-        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
-        \hline
-        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
-        \hline
-        \B{body} & : & seq \B{Inst} \\
-        \hline
-        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
-        \hline
-        \B{vars} & : & seq \B{Name} \\
-        \B{body} & : & seq \B{Inst} \\
+
 \end{tabular}
 \begin{tabular}[t]{rcl}   
-  	\hline
-        \multicolumn{3}{|c|}{If: \B{If}} \\
-        \hline
-        \B{branches} & : & seq \B{IfBr} \\
         \hline
@@
         \hline
-        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt\B{Vari}}\\
-	\multicolumn{3}{c}{\B{Record}\alt\B{Array}\alt\B{ValueItem} } \\
-        \hline
-        \multicolumn{3}{|c|}{Value item : \B{ValueItem}} \\
-        \hline
+        \multicolumn{3}{c}{\B{Lit}\alt\B{Term}\alt\B{Pred}\alt\B{Cons}\alt}\\
+	\multicolumn{3}{c}{\B{Vari}\alt\B{Record}\alt\B{Array}\alt} \\
 	\multicolumn{3}{c}{\B{RecField}\alt\B{ArrayItem} } \\
@@
         \B{op} & : & \B{RelOp} \\
-        \B{arg1}, \B{arg2} & : & \B{Exp}
-\end{tabular}
-
-\begin{tabular}[t]{rcl}
+        \B{arg1}, \B{arg2} & : & \B{Exp}\\
         \hline
@@
 	\multicolumn{3}{c}{\B{Enumeration}} \\
+
+\end{tabular}
+
+\begin{tabular}[t]{rcl}
 	\hline
@@
 \end{figure}
+
+ Accordingly, the expression language consists
+in arithmetic operations \B{-} (unary and binary), \B{+}, \B{*}, \B{/}, \B{mod},
+% \B{**} (exponentiation),
+\B{succ} and \B{pred} and Boolean operations
+\B{$\land$}, \B{$\lor$}, \B{$\neg$} as well as relations \B{=}, \B{$\neq$},
+\B{$>$}, \B{$<$}, \B{$\leq$} and \B{$\neq$}. Atomic expressions are identifiers,
+Boolean constants \B{FALSE} and \B{TRUE}, integer literals, and integer
+constants \B{MAXINT} and \B{MININT}. The development through the application of
+the B method ensures the absence of overflows and underflows.
 
@@
 \[
-\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{0,1\}}_{Exp} }_{Product}\}}_{Array}
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
 \]

### 2014-04-25 DOCUMENTATION/DerivedTypes.tex
rewriting the initialisation

@@
-\subsection{Rules to be updated in first specification}
-
-\paragraph{Becomes equal instructions} \textcolor{blue}{ must evaluate the
- target and the source of the assignment, and then copy the result of the 
-latter in the former, using the LLVM instruction \llvm{store}. The evaluation 
-of the target is defined with $\trad{}{lvalue}$, which yields the corresponding 
-LLVM source code $l$ and the variable $p$ containing the assigned location. The 
-evaluation of the source yields the corresponding code $r$, the variable $v$ 
-holding the result, and its type $t$. }
-
-\begin{align*}
-\begin{split}
-  \trad{a}{\B{Beq}} \isdef
-  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
-  & \llvm{"\PH{l}} \\
-  & \llvm{\PH{r}} \\
-  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
-\end{split}
-\end{align*}
-
-The rule lvalue is:
-\begin{align*}
-\lefteqn{\trad{n}{lvalue} \isdef} \\
-& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
-& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
-& \quad \IF \mbox{\B{n} is an ArrayItem} \THEN  \trad{n}{ArrayItem} \\
-& \quad \IF \mbox{\B{n} is a StructItem} \THEN  \trad{n}{StructItem} \\
-& \quad \ELSE \LET v = \variable \IN \\
-& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
-& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
-\end{align*}
-
-\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
-care. The attributions between these elements are made  directly by the memory 
-address.  However, the expression resolution are made by parts, where the 
-partial results are stored to temporary variables. Therefore, the expression 
-resolution with pointers needs an additional instruction $LOAD$ for store the 
-partial results. This translation is represented by $ValueItem$.}
-
-\begin{align*}
-\begin{split}
-  \trad{n}{\B{ValueItem}} \isdef \\
-   & \LET \ell = \lbl  \IN\\
-   & \IF \B{n} \mbox{ is \B{StructItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{StructItem}} \\
-   & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
-   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
-   & \sep  \ell  \sep t'\\
-\end{split}
-\end{align*}
-
-\section{Translation of Structures and Array}
+\subsection{Translation of structures and arrays}
 \label{sec:StructArray}
@@
 the direct correspondence between B0 native types and LLVM and this approach 
-is presented here.
+is presented here. This section describes the translation of derived data 
+types showed in section~\ref{sec:data}.
 
-\subsection{Structure} \VGM{It is not implemented yet, because it requires generating an XML example.}
+\subsubsection{Structure} 
 The structure declaration defines a list of variables to be placed under one 
@@
 
-\begin{ccode}
+\begin{pascalcode}
 xx := rec(id:1, balance:10);
 xx'balance := xx'balance -1;
-\end{ccode}
+\end{pascalcode}
+
 The lines 1 and 2 are represented in translation in LLVM code:
@@
 
-The $StructItem$ rule gets an element value from the struct.
-The same rule is used on the left side of attribution as a $LValue$ and right side of attribution as an $Expression$.
-The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure.
+The $RecField$ rule gets an element value from the record. The same rule is 
+used on the left side of attribution as a $LValue$ and right side of 
+attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
+respectively the position and type of $n$ element in the structure.
 
@@
 \begin{split}
-  \trad{n}{\B{StructItem}} \isdef \\
+  \trad{n}{\B{RecField}} \isdef \\
    & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
@@
 
-The attribution from record structure to a structure variable, line 1 from last 
-B code example, is translated by a sequence of application of rule $StructItem$.
-The nested structures will be designed in the next project phase.
+The attribution from record structure to a record structure variable, line 1 
+from last B code example, is translated by a sequence of application of rule 
+$RecField$. The nested structures will be designed in the next project phase.
+%TODO after test the first examples.
 
-\subsection{Arrays}
+\subsubsection{Arrays}
 
@@
 In B, a concrete array is a total function where the domain indexes the range 
-elements. Declaration examples of arrays are:
-\begin{itemize}
-	\item $array1 \in (0..10) \to INT $	(a simple array)
-	\item $array2 \in (1..10) \times (1..20) \times (1..30) \to INT $   (a multidimensional array)
-	\item $\{1\mapsto1\mapsto4, 1\mapsto2\mapsto5\}  \in (INT \times INT) \times INT$ (an explicit array)
-\end{itemize}
+elements.
+
+%TODO Attributions and initializations
 
@@
 elements are defining an array, getting an element and setting an
-element.
+element. The white code block and gray code block are respectively B code and LLVM code.
 
@@
 The following code declares a multidimensional global array called ``Array2'' 
-with 6000 (10*20*30) integer numbers with 32 bits and initialized with zero.
+with 16 (2*2*4) integer numbers with 32 bits.
 
 \begin{llvmcode}
-@Array2 = internal global [10 x [20 x [30 x i32]]] zeroinitializer, align 16
+@array2 = internal global [2 x [2 x [4 x i32]]]
 \end{llvmcode}
 
-\VGM{It is not implemented yet, because it requires generating an XML example.}
-The following example initialises an new array in C:
+\subsubsection{Initialing an array}
+The following two attribution commands have the same semantic effect.
 
-\begin{ccode}
-int32_t arr[1][2][1]= {{{4},{5}}};
-\end{ccode}
+\begin{pascalcode}
+array2 := ( 0..1 * 0..1 * 0..3 * {5});
 
-The syntax rule\footnote{The underlined element type can be expanded in others types definitions.} used to initialize an array in LLVM is:\\
-$@<var\_name> = constant [<size> x \underline{<type>}] [<type> <value>]$.\\
-The translation of the last example is:
+array2 := { 0|->0|->0|->5 , 0|->0|->1|->5 , 
+            0|->1|->0|->5 , 0|->1|->1|->5 , 
+            1|->0|->0|->5 , 1|->0|->1|->5 , 
+            1|->1|->0|->5 , 1|->1|->1|->5 ,
+            0|->0|->2|->5 , 0|->0|->3|->5 ,
+            0|->1|->2|->5 , 0|->1|->3|->5 , 
+            1|->0|->2|->5 , 1|->0|->3|->5 , 
+            1|->1|->2|->5 , 1|->1|->3|->5 };
+\end{pascalcode}
 
-\begin{llvmcode}
-@arr=constant[1 x [2 x [1 x i32]]] [[2 x [1 x i32]] [[1 x i32] [i32 4],[1 x i32] [i32 5]]]
-\end{llvmcode}
+An alternative way to represent these attributions cannot be using sequential attributions as follow:
 
-\subsubsection{Getting an element from array}
+\begin{pascalcode}
+array2(0,0,0) := 5;
+// A sequence of attributions ...
+array2(1,1,3) := 5;
+\end{pascalcode}
+
+It is not possible, because B array attribution are atomic. The B array 
+attribution are made in two steps. First, the new value is calculated and
+copied to a temporal place. Second, the sequence of new values are setted 
+in theirs address. Finally, considering these steps and the execution of
+following code then the value of ``array2(0,0,1)'' is ``2''.
+
+\begin{pascalcode}
+ array2 := ( (0..1) * (0..1) * (0..3) * {1});
+ array2 := { 0|->0|->0|->2 , 0|->0|->1|->(a3d(0,0,0)+1), ...  };
+\end{pascalcode}
 
@@
 \begin{llvmcode}
-%7 = getelementptr [10 x [20 x [30 x i32]]]* @Array2, i32 0, i64 2, i64 3, i64 4
+%7 = getelementptr [2 x [2 x [4 x i32]]]* @Array2, i32 0, i64 1, i64 1, i64 3
 \end{llvmcode}
 
-The first parameter defines the type and name of the variable, and the others are
-indexes to select the element in the data type. The second parameter (i32 0)
-selects the first element in the pointer (* @Array2). The last elements 
-(i64 2, i64 3, i64 4) select the value from position (2,3,4) storing it in \%7.
+The first two parameters define the type and name of the variable, and the
+others are indexes to select the element in the data type. The third parameter (i32 0)
+selects the first element in the pointer (* @array2). The last elements 
+(i64 1, i64 1, i64 3) select the value from position (1,1,3) storing it in \%7.
 This array is multidimensional, so several indexes must be declared separated
@@
 VAR xx, yy, zz IN
-	xx=2;
-	yy=3;
-	zz=4;
-	Array2(x,y,z):=7
+	xx=0;
+	yy=1;
+	zz=3;
+	array2(x,y,z):=7
 END
@@
 %zz = alloca i32
-store i32 2, i32* %xx
-store i32 3, i32* %yy
-store i32 4, i32* %zz
+store i32 0, i32* %xx
+store i32 1, i32* %yy
+store i32 3, i32* %zz
 %1 = load i32* %zz
@@
 %6 = sext i32 %5 to i64
-%7 = getelementptr[10x[20x[30xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
+%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
 store i32 7, i32* %7         ;setting an element
@@
 
-\subsection{Rules for support Array}
+\subsubsection{Rules for support array}
 
@@
 \end{pascalcode}
+
+%TODO Describe here the restrictions of declarations arrays
+\paragraph{Relevant restrictions of work:}
+\begin{itemize}
+\item  B0 definition - An array must be initialized declaring all values 
+to avoid no determinism.
+\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:
+
+\begin{pascalcode}
+array := array <+ {1|->2|->0|->4,  1|->2|->1|->5}
+\end{pascalcode}
+
+\end{itemize}

### 2014-04-25 DOCUMENTATION/specification.tex
rewriting the initialisation

@@
 \date{March 2014}
-\definecolor{bg}{rgb}{0.95,0.95,0.95}
+\definecolor{bg}{rgb}{0.93,0.93,0.93}
+\definecolor{white}{rgb}{1,1,1}
 \newminted{llvm}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
-\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
-\newminted{c}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=bg}
+\newminted{pascal}{frame=single,firstnumber=1,linenos=true,fontsize=\scriptsize,bgcolor=white}
 
@@
 \]
-Initializing an array:
+
+There are three attributions cases related to array.
+\subitem\textbf{Case 1.1} - Initializing an unique value to an array:
 \[
-\textcolor{gray}{arr :=} \underbrace{\{\underbrace{1 \mapsto  100,2 \mapsto 101}_{seq\ Map}\}}_{Array}
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2  }_{seq\ Exp} \times \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
 \]
+\subitem\textbf{Case 1.2} - Initializing different values to an array:
 \[
-\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
+\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(0 \mapsto 0) \mapsto  100,(0 \mapsto 1) \mapsto  101}_{seq\ Map}\}}_{Array}
 \]
 
-Updating a selected element from array by a value of variable:
-\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
+\subitem\textbf{Case 2.1} - Updating a selected element from array by a value of variable:
+\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
 \]
 
-Setting a selected element from array to a variable:
+\subitem\textbf{Case 2.2} - Setting a selected element from array to a variable:
 \[ 
-\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \]
+\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem}
+\]
+
+\subitem\textbf{Case 3} - Attribution between arrays:
+\[
+\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
+\]
 \end{description}

### 2014-04-25 beta_tests/Counter/Counter_using_CUTest 1.5/AllTests.c
Adding a set of beta tests for LLVM code (counter example is working).

@@ -1,0 +1,80 @@
+#include "CuTest.h"
+#include <stdio.h>
+#include <string.h>
+#include "Counter.h"
+
+    
+    CuSuite* StrUtilGetSuite();
+    
+    counter$state$ counter;
+
+
+  void Test_oper_inc_testCase1(CuTest *tc) {
+	    // setting manually
+	counter.value = 0 ;
+	counter.error = false;
+	    // or setting using the initialization, you can chose one option
+	counter$INITIALISATION(&counter);
+	    // calling the function
+	counter$inc(&counter);
+    
+	bool condition = !( counter.error == true) || (counter.value == 2147483647);
+	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
+
+    }
+
+
+    void Test_oper_inc_testCase2(CuTest *tc) {
+	    // setting manually
+	counter.value = 3 ;
+	counter.error = true;
+    	
+	// calling the function
+	counter$inc(&counter);
+    	
+	// Neste caso, o invariante está sendo quebrado então a formula não deveria ser negada ?
+	bool condition =  !( counter.error == true) || (counter.value == 2147483647) ;
+	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
+
+    }
+
+
+    void Test_oper_zero_testCase1(CuTest *tc) {
+	    // setting manually
+	counter.value = 3 ;
+	counter.error = true;
+    	
+	// calling the function
+	counter$zero(&counter);
+    
+	bool condition = !( counter.error == true) || (counter.value == 2147483647);
+	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
+
+    }
+
+   
+    CuSuite* testsSuite() {
+        CuSuite* suite = CuSuiteNew();
+	SUITE_ADD_TEST(suite, Test_oper_inc_testCase1);
+        SUITE_ADD_TEST(suite, Test_oper_inc_testCase2);
+	SUITE_ADD_TEST(suite, Test_oper_zero_testCase1);
+
+        return suite;
+    }
+
+    
+    void RunAllTests(void) {
+        CuString *output = CuStringNew();
+        CuSuite* suite = CuSuiteNew();
+        
+        CuSuiteAddSuite(suite, testsSuite());
+    
+        CuSuiteRun(suite);
+        CuSuiteSummary(suite, output);
+        CuSuiteDetails(suite, output);
[…]

### 2014-04-25 beta_tests/Counter/Counter_using_CUTest 1.5/Exemplo_gerado/AllTests.c

@@ -1,0 +1,112 @@
+#include <stdio.h>
+#include <string.h>
+#include <stdint.h>
+#include "CuTest.h"
+#include "counter.h"
+
+counter$state$ counter;
+
+void check_invariant(CuTest* tc, counter$state$ counter) {
+	
+	if(!((!((counter.error == true)) || (counter.value == INT32_MAX)))){
+		CuFail(tc, "The invariant '((error = TRUE) => (value = MAXINT))' was unsatisfied");
+	}
+	
+	if(!(0 <= counter.value)){
+		CuFail(tc, "The invariant '0 <= value' was unsatisfied");
+	}
+	
+	if(!(counter.value <= INT32_MAX)){
+		CuFail(tc, "The invariant 'value <= MAXINT' was unsatisfied");
+	}
+	
+}
+		 
+
+/**
+* Test Case 1
+* Formula: value <= MAXINT & 0 <= value & value < MAXINT & error : BOOL & value : INT & ((error = TRUE) => (value = MAXINT))
+*/
+void counter_inc_test_case_1(CuTest* tc)
+{
+	//counter$INITIALISATION(&counter);
+	
+	bool error = false; 
+	counter.error = error; 
+	int32_t value = 0; 
+	counter.value = value; 
+	
+	
+	counter$inc(&counter);
+	
+	 
+	bool errorExpected; // Add expected value here.
+	CuAssertTrue(tc, counter.error == false);
+	
+	int32_t valueExpected; // Add expected value here.
+	CuAssertTrue(tc, counter.value == 1);
+	 
+	check_invariant(tc, counter);
+}
+		 
+/**
+* Test Case 2
+* Formula: value <= MAXINT & 0 <= value & not(value < MAXINT) & error : BOOL & value : INT & ((error = TRUE) => (value = MAXINT))
+*/
+void counter_inc_test_case_2(CuTest* tc)
+{
+	//counter$INITIALISATION(&counter);
+	
+	bool error = true; 
+	counter.error = error; 
+	int32_t value = INT32_MAX; 
+	counter.value = value; 
+	
+	
+	counter$inc(&counter);
+	
+	 
+	bool errorExpected; // Add expected value here.
+	CuAssertTrue(tc, counter.error == true);
+	
+	int32_t valueExpected; // Add expected value here.
+	CuAssertTrue(tc, counter.value == INT32_MAX);
+	 
+	check_invariant(tc, counter);
+}
+		 
+/**
[…]

### 2014-04-25 beta_tests/Counter/Counter_using_CUTest 1.5/Exemplo_gerado/comp_exec.sh

@@ -1,0 +1,8 @@
+# Before you must generate the counter.llvm by this command:
+# ./b2llvm.py -v -m comp counter counter.llvm bxml project.xml
+clang -S -emit-llvm -c CuTest.c AllTests.c
+llc-mp-3.5  counter.llvm
+llc-mp-3.5  CuTest.ll 
+llc-mp-3.5  AllTests.ll
+clang -o executable counter.llvm.s  CuTest.s AllTests.s
+./executable 

### 2014-04-25 beta_tests/Counter/Counter_using_CUTest 1.5/Exemplo_gerado/counter.h

@@ -1,0 +1,45 @@
+#ifndef _counter_h
+#define _counter_h
+
+#include <stdint.h>
+#include <stdbool.h>
+#ifdef __cplusplus
+extern "C" {
+#endif /* __cplusplus */
+    
+    
+    /* Clause SETS */
+    
+    /* Clause CONCRETE_VARIABLES */
+    
+    
+    /* Clause CONCRETE_CONSTANTS */
+    /* Basic constants */
+    /* Array and record constants */
+
+   // Passo 2 - Declarando os typedef 
+   typedef struct  { int32_t value; bool error; } counter$state$;
+   typedef counter$state$ * counter$ref$;
+
+
+
+   // Passo 1 - Adicionando a renomeiação das variáveis e usando o parâmetro de contexto.
+    extern void counter$INITIALISATION(counter$ref$ self);
+   // Passo 3 - Remove as declarações das variáveis antigas no arquivo counter.c 
+   // Passo 4 - Atualiza as variáveis no arquivo de teste para usar a referencia a estrutura de estado. Esta mesma sequência de passos segui no arquivo de teste.
+   // Passo 5 - Faz a reomeação das funções nos arquivos .c e .h. Exemplo: counter__zero -> counter$zero. Precisa somente counter__-> counter$
+   // Passo 6 - Comparar a compatilibilidade entre o arquivo .c baseado no atelierb e o llvm criando com B2llvm
+ 
+
+    /* Clause OPERATIONS */
+    
+    extern void counter$zero(counter$ref$ self);
+    extern void counter$inc(counter$ref$ self);
+    extern void counter$get(counter$ref$ self, int32_t *res);
+    
+#ifdef __cplusplus
+}
+#endif /* __cplusplus */
+
+
+#endif /* _counter_h */

### 2014-05-13 DOCUMENTATION/DerivedTypes.tex
Adding declaration for Structures and Arrays

@@
 
+Initializing a record with a sequence of expressions:
+\[
+  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
+\]
+
+Setting a record field to a variable:
+\[
+  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
+\]
+
+A B code example:
+
 \begin{pascalcode}
 xx := rec(id:1, balance:10);
-xx'balance := xx'balance -1;
+xx'balance := xx'balance-1;
 \end{pascalcode}
@@
   \trad{n}{\B{RecField}} \isdef \\
-   & \IF \B{n.type} \mbox{ is \B{BOOL or NAT or INT} } \THEN \\
-   &\quad  \LET \ell = \lbl  \AND  \PH{t} = n.type \AND \PH{v} = n.element \IN\\
-   &\quad  \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \\
-   &\quad  \sep  \ell  \sep \SBType(\B{n})\\
+   &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
+   &code \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
+   &\IF \B{n.base.type} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
+   &\quad \THEN  code \sep  \ell  \sep \SBType(\B{n})\\
+   &\IF \B{n.base.type} \mbox{ is \B{RecField} }  \THEN \\
+   &\LET \ell2 = \lbl  \AND  t = n.type \AND v = \B{tail}(n.element) \AND \\
+   &\quad n2.base  \BE ?? \AND n2.element = \B{head}(n.element) \AND a, b, c = \trad{n2}{RecField}
+\IN\\
+   &\quad  code \conc  \llvm{"\PH{\ell2} = getelementptr \PH{t}* \PH{\ell}, i32 0, i32  \SBPos(\B{n})\nl"} \\
+   &\quad \sep  \ell2  \sep \SBType(\B{n})\\
 \end{split}
@@
 
-%TODO Attributions and initializations
+%[[ Write a text here to define the rule for it]]
+
+There are three attributions cases related to array.
+\begin{description}
+
+\item[Case 1.1] - Initializing an unique value to an array:
+\[
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2 \times 0..3  }_{seq\ Exp} \times \underbrace{\{5\}}_{Exp} }_{Product}\}}_{Array}
+\]
+\item[Case 1.2] - Initializing different values to an array\footnote{B0 definition requires the declaration of all values.}:
+\[
+\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{5}_{Exp} , ... ,
+(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{5}_{Exp} }_{seq\ Map}\}}_{Array}
+\]
+
+Theses cases use a declared constant that is attributed to the variable:
+\begin{llvmcode}
+@array = global [3 x [3 x [4 x i32]]] 
+[[3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
+ 	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
+\end{llvmcode}
+
+The rule for declare array constants:
+
+\begin{align*}
+\begin{split}
+  \trad{array}{\B{Type}} \isdef \\
+   & \LET \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
+   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}\\
+   & \quad \ELSIF array \mbox{ is \B{Map} } \THEN\\
+   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
+   & \quad \IN \llvm{" global \PH{dec}"}\\
+\end{split}
+\end{align*}
+\VGM{Alguma sugestão com relação as funções:head, tail.}
+
+\begin{align*}
+\begin{split}
+  \trad{type, seqIntl, value}{\B{Product}} \isdef \\
+   &\LET \B{h = head(seqIntl)} \AND\\
+   & \B{times = h.end - h.start + 1} \AND\\
+   & values = \trad{times,type,value}{\B{TV}} \IN\\
+   & \IF \B{seqIntl.length} == 1 \THEN  \\
+   & \llvm{"[\PH{times} x \PH{type}] [\PH{values}]"}  \\
+   & \ELSE \LET \\
+   &  tvalues = \trad{type,tail(seqIntl),values}{\B{Product}} \\
+   & \IN \\
+   & \llvm{"[\PH{times} x \PH{type}]  \PH{tvalues}"}  \\
+\end{split}
+\end{align*}
+
+\begin{align*}
+\begin{split}
+  \trad{times, type, value}{\B{TV}} \isdef \\
+   &\LET res1 = \llvm{" \PH{type} \PH{value} "} \IN\\
+   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
+   & \ELSE \LET   res2 =   \trad{times-1,type,value}{\B{TV}}  \IN  \\
+   & \llvm{"\PH{res1} , \PH{res2}"} \\
+\end{split}
+\end{align*}
+
+The ``head'' function returns the first element from sequence.
+The ``tail'' function returns the given sequence without its first element.
+
+\begin{comment}
+\$VarName = global [size X type] X [(type value,)^+]
+
+|array|Type = LET
+				  IF array is Product THEN dec = |array.arg1, array.arg2 |CProduct  ( Quebrando o padrão?)
+				  ELSEIF array is seq Map THEN dec = |array.Map |Map
+				IN	  
+					" global "^ dec
+				END
+|sIntl, value |CProduct = 
+    	IF sIntl is  empty THEN  ""
+	ELSE THEN
+		LET sizeType = |sIntl, "i32"|PrintSizeType &
+		 tailSizeType=|sIntl.tail,"i32"|PrintSizeType & 
+		 values= |tailSizeType,values|CProduct IN
+		"[sizeType]  [ tailSizeType  values ]"		
+		END
+	END
+
+|sIntl,t| PrintSizeType =
+    	IF sIntl.size == 1  THEN  "t"
+	ELSE THEN
+		LET size = sIntl.head.size()  &
+			type = |sIntl.tail,t| PrintSizeType    &
+		"[size X type ]"		
+		END
+	END
+
+|array|Type = IF t is Map THEN ...
+\end{comment}
+
+\item[Case 2.1] - Updating a selected element from array by a value of variable:
+\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
+\]
+\item[Case 2.2] - Setting a selected element from array to a variable:
+\[ 
+\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem}
+\]
+\item[Case 3] - Attribution between arrays\footnote{B0 definition requires the same type.}:
+\[
+\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
+\]
+\end{description}
 
@@
 element. By default, the element of index not initialized has a zero value.
-
-This section shows the relation between arrays B0 and LLVM. The sequence of
-elements are defining an array, getting an element and setting an
-element. The white code block and gray code block are respectively B code and LLVM code.
-
-\subsubsection{Declaring an array}
-
-The essential elements of the array are its name, size and type. 
-The size of ``e'' array with the indexes (i1,...,in) are defined according 
-to the formula:\\
- $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - e.in\_start+1)$.
-
-The following code declares a multidimensional global array called ``Array2'' 
-with 16 (2*2*4) integer numbers with 32 bits.
-
-\begin{llvmcode}
-@array2 = internal global [2 x [2 x [4 x i32]]]
-\end{llvmcode}
+The different ways to evaluate an array are described as follow:
 
@@
 array2(0,0,0) := 5;
-// A sequence of attributions ...
+array2(0,0,1) := 5;
+// ... - A sequence of attributions
 array2(1,1,3) := 5;
@@
  array2 := ( (0..1) * (0..1) * (0..3) * {1});
- array2 := { 0|->0|->0|->2 , 0|->0|->1|->(a3d(0,0,0)+1), ...  };
+ array2 := { 0|->0|->0|->2 , 0|->0|->1|->(array2(0,0,0)+1), ...  };
 \end{pascalcode}
@@
 VAR xx, yy, zz IN
-	xx=0;
+	xx=1;
 	yy=1;
@@
 This example is translated in four steps. First, the variables \%x, \%y, \%z 
-are allocated. Second, the respective values 2, 3, 4 are stored in variables. 
+are allocated. Second, the respective values 0, 1, 3 are stored in variables. 
 Third, the variables are loaded to local variables and converted to 64 bits. 
@@
 %zz = alloca i32
-store i32 0, i32* %xx
+store i32 1, i32* %xx
 store i32 1, i32* %yy
@@
 
+\paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.
+
+\paragraph{seq Map} declares every address and values from array, it is represented by \textbf{case 1.2}.
+
 \paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
@@
 calculates the address of element. This rule also supports to translate 
-expressions of arrays.
+expressions of arrays. This rule is used to support array attributions 
+from \textbf{cases} \textbf{2.1} and \textbf{2.2}.
 
@@
 \begin{align*}
-&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\ \\
+&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
 &\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \PH{\ell} = \lbl \IN \\
@@
 
-\paragraph{Restriction on array:} The translation of direct attribution
- between arrays is not yet supported, for example:
-
-\begin{pascalcode}
- array1 := array2;
-\end{pascalcode}
-
-%TODO Describe here the restrictions of declarations arrays
-\paragraph{Relevant restrictions of work:}
+\paragraph{Relevant translation restrictions:}
 \begin{itemize}

### 2014-05-13 DOCUMENTATION/specification.tex
Adding declaration for Structures and Arrays

@@
 \usepackage{color}
+\usepackage{verbatim}
 \usepackage{minted}
@@
 
-\date{March 2014}
+\date{May 2014}
 \definecolor{bg}{rgb}{0.93,0.93,0.93}
@@
 translation. The elements of the B language are denoted using a \B{sans serif}
-font.
+font, when used the code block so white code block and gray code block means
+respectively B code and LLVM code.
 
@@
         \hline
-	\B{arg1} & : & seq \B{Exp}\\ 
+	\B{arg1} & : & seq \B{Interval}\\ 
 	\B{arg2} & : & \B{Exp}\\ 	
@@
 	\hline
-	\B{start}, \B{end} & : & \B{Exp} \\
+	\multicolumn{3}{c}{ \B{start}, \B{end} : \B{Exp} } \\
 	\hline
@@
 representation in the abstract syntax structure from the table \ref{tab:node-attr}.
+
 The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.
 
-The following examples relate the B code and their abstract syntax representation. The \textcolor{gray}{gray elements} are not related to their abstract syntax tree to simplify the examples.
+The following examples relate the B code and their abstract syntax representation. 
+The \textcolor{gray}{gray elements} are not related to their abstract syntax tree to simplify the examples.
 
 \begin{description}
-
 \item[structures] The structure types of B are naturally encoded as structure
@@
 
-Declaring an structure type:
+Declaring a structure variable and nested structure type:
+%\[
+%  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
+%\] 
 \[
-  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
+\textcolor{gray}{T_1 =} \underbrace{struct( \underbrace{num1 \in \mathbb{Z}, num2 \in  \mathbb{Z} }_{seq\ Field} )}_{Struct};\\
+  \textcolor{gray}{name1 \in} \ T_1;
 \]
-Initializing a record with a sequence of expressions:
+
 \[  
-  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
+  \textcolor{gray}{name2 \in} \underbrace{struct( \underbrace{num1 \in \mathbb{Z}, elem \in  T_1 }_{seq\ Field} )}_{Struct}
 \]
-
-Setting a record field to a variable:
-\[
-  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
-\]
+The simplified LLVM translation is:
+
+\begin{llvmcode}
+%struct.T_1 = type { i32, i32 }
+%struct.T_2 = type { i32, %struct.T_1 }
+%struct.state = type { %struct.T_1, %struct.T_2 }
+
+@name2 =  global %struct.T_2 zeroinitializer
+@name1 =  global %struct.T_1 zeroinitializer
+\end{llvmcode}
+
+The rules to declare the structure type are:
+\begin{align*}
+\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{TField}}} \AND\\
+ l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
+\end{align*}
+$$\trad{t}{\B{TField}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$
+
+The records types are initialized in the initializing function of the module.\\
+
+\begin{comment}
+The following rule is invoked in the initializing function of the module.
+\begin{align*}
+\trad{instanceModule,Structure}{\B{InitDerivedType}} \isdef \\
+\LET name \BE \Global(\B{instanceModule,Structure.elements}) \AND\\
+tlabel \BE \local(Structure.elements) \IN\\
+\llvm{"\PH{name} = global \PH{tlabel} zeroinitializer\nl"}\\
+\end{align*}
+\end{comment}
 
@@
 \[
-\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
-\]
-
-There are three attributions cases related to array.
-\subitem\textbf{Case 1.1} - Initializing an unique value to an array:
-\[
-\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2  }_{seq\ Exp} \times \underbrace{\{1\}}_{Exp} }_{Product}\}}_{Array}
-\]
-\subitem\textbf{Case 1.2} - Initializing different values to an array:
-\[
-\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(0 \mapsto 0) \mapsto  100,(0 \mapsto 1) \mapsto  101}_{seq\ Map}\}}_{Array}
-\]
-
-\subitem\textbf{Case 2.1} - Updating a selected element from array by a value of variable:
-\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
-\]
-
-\subitem\textbf{Case 2.2} - Setting a selected element from array to a variable:
-\[ 
-\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+ii}_{seq\ Exp})}_{ArrayItem}
-\]
-
-\subitem\textbf{Case 3} - Attribution between arrays:
-\[
-\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
+\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..1) \times (0..1) \times (0..3)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
 \]
 \end{description}
+
+Declaring an array: the essential elements of the array are its name, size and type. 
+The following code declares a multidimensional global array called ``Arr'' 
+with 16 (2*2*4) integer numbers with 32 bits.
+
+\begin{llvmcode}
+%struct.state = type { ... , [2 x [2 x [4 x i32]]],.. }
+\end{llvmcode}
+
+The function ArrayType returns the type for a declared variable array in 
+structure state, for example:
+
+\begin{align*}
+\trad{Array}{\B{Type}} \isdef \\
+& \trad{Array.dom , Array.ran.type}{ArrayType}\\
+\trad{seqSet,type}{\B{ArrayType}} \isdef \\
+& \LET h \BE \B{head}(seqSet) \AND\\
+& times \BE h.end - h.start + 1 \IN\\
+& \IF seqSet.length == 1 \THEN\\
+& \llvm{"[\PH{times} x \PH{type}] \nl"} \\
+& \ELSE \\
+& \LET res \BE \trad{tail(seqSet),type}{ArrayType} \IN \\
+& \llvm{"[\PH{times} x \PH{res}] \nl"} \\
+\end{align*}
+
+[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]
 
@@
 
-To facilitate the defintiion of rules that apply to similar cases, we introduce
+To facilitate the definition of rules that apply to similar cases, we introduce
 mapping $\opmap$, that associates B operators to corresponding LLVM keywords:

### 2014-05-13 b2llvm/loadbxml.py
Updating the load XML (renaming some elements from BXML)

@@ -462,7 +462,7 @@ def load_sub(elem, symast, symimp):
         astnode = load_assert_substitution(elem)
     elif elem.tag == "If_Substitution":
         astnode = load_if_substitution(elem, symast, symimp)
-    elif elem.tag == "Affectation_Substitution":
+    elif elem.tag == "Assignement_Substitution":
         astnode = load_becomes_eq(elem, symast)
     elif elem.tag == "Case_Substitution":
         astnode = load_case_substitution(elem, symast, symimp)
@@ -510,9 +510,9 @@ def load_if_substitution(elem, symast, symimp):
     """Load an XML element for a IF substitution to an AST node."""
     global _LOG
     assert elem.tag == "If_Substitution"
-    if elem.get("elseif") != None:
-        _LOG.error("unrecognized elseif attribute in IF substitution")
-        return ast.make_skip()
+    #if elem.get("elseif") != None:
+    #    _LOG.error("unrecognized elseif attribute in IF substitution")
+    #    return ast.make_skip()
     xmlcond = elem.find("./Condition")
     xmlthen = elem.find("./Then")
     xmlelse = elem.find("./Else")
@@ -529,7 +529,7 @@ def load_if_substitution(elem, symast, symimp):
 def load_becomes_eq(elem, symast):
     """Load an XML element for a becomes equal substitution to an AST node."""
     global _LOG
-    assert elem.tag == "Affectation_Substitution"
+    assert elem.tag == "Assignement_Substitution"
     lhs = elem.findall("./Variables/*")
     rhs = elem.findall("./Values/*")
     if len(lhs) != 1 or len(rhs) != 1:
@@ -672,9 +672,9 @@ def load_exp(node, symast):
         pynode = load_nary_expression(node, symast)
     elif node.tag == "Unary_Expression":
         pynode = load_unary_expression(node, symast)
-    elif node.tag == "Boolean_Litteral":
+    elif node.tag == "Boolean_Literal":
         pynode = load_boolean_literal(node)
-    elif node.tag == "Integer_Litteral":
+    elif node.tag == "Integer_Literal":
         pynode = load_integer_literal(node)
     elif node.tag == "Identifier":
         pynode = load_identifier(node, symast)
@@ -696,7 +696,7 @@ def load_identifier(node, symast):
 def load_boolean_literal(node):
     """Load an XML element for an Boolean literal to an AST node."""
     global _LOG
-    assert node.tag == "Boolean_Litteral"
[…]

### 2014-07-07 DOCUMENTATION/DerivedTypes.tex
Small updates about array

@@
 
-Two approaches are possible to support arrays and structures. An approach is 
+Two approaches are possible to support arrays and structures. One approach is 
 importing reusable libraries of AtelierB that are reported in the official 
@@
 is presented here. This section describes the translation of derived data 
-types showed in section~\ref{sec:data}.
+types showed in Section~\ref{sec:data}.
 
@@
 
-Initializing a record with a sequence of expressions:
+\begin{description}
+\item[Case 1.1] Initializing a record with a sequence of expressions:
 \[
@@
 
-Setting a record field to a variable:
+\item[Case 1.2] Setting a record field to a variable:
 \[
@@
 \]
+
+\item[Case 2]  Attribution between structures:
+\[
+  \underbrace{newRecord}_{Record} :=  \underbrace{accountRecord}_{Record} 
+\]
+\end{description}
 
@@
 xx := rec(id:1, balance:10);
-xx'balance := xx'balance-1;
+xx`balance := xx`balance-1;
 \end{pascalcode}
 
-The lines 1 and 2 are represented in translation in LLVM code:
+Lines 1 and 2 are represented in translation in LLVM code:
 
@@
 The $RecField$ rule gets an element value from the record. The same rule is 
-used on the left side of attribution as a $LValue$ and right side of 
+used on the left side of attribution as an $LValue$ and right side of 
 attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
@@
    &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
-   &code \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
-   &\IF \B{n.base.type} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
-   &\quad \THEN  code \sep  \ell  \sep \SBType(\B{n})\\
-   &\IF \B{n.base.type} \mbox{ is \B{RecField} }  \THEN \\
-   &\LET \ell2 = \lbl  \AND  t = n.type \AND v = \B{tail}(n.element) \AND \\
-   &\quad n2.base  \BE ?? \AND n2.element = \B{head}(n.element) \AND a, b, c = \trad{n2}{RecField}
-\IN\\
-   &\quad  code \conc  \llvm{"\PH{\ell2} = getelementptr \PH{t}* \PH{\ell}, i32 0, i32  \SBPos(\B{n})\nl"} \\
-   &\quad \sep  \ell2  \sep \SBType(\B{n})\\
+   &code1 \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
+   &\IF \B{n.element} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
+   &\quad \THEN  code1 \sep  \ell  \sep \SBType(\B{n})\\
+   &\IF \B{head(n.element)} \mbox{ is \B{RecField} }  \THEN \\
+   &\LET n2 \BE new \B{RecField}() \AND  \B{n2.base}  \BE \B{head}(n.element)\\
+   &\AND \B{n2.element} \BE \B{tail}(n.element) \AND \\
+   &\quad code2, label, pos = \trad{n2}{RecField} \IN\\
+   &\quad \llvm{"\PH{code1} \PH{code2}"} \sep  label  \sep pos \\
 \end{split}
@@
 
-The attribution from record structure to a record structure variable, line 1 
-from last B code example, is translated by a sequence of application of rule 
-$RecField$. The nested structures will be designed in the next project phase.
-%TODO after test the first examples.
+The attribution from record structure to a record structure variable, Line 1 
+from the last B code example, is translated by a sequence of applications of
+the  rule $RecField$. This rule was also designed to support nested structures.
+
+The structure attribution are made in atomic way and it is simple example:
+
+\begin{pascalcode}
+simpleRec := rec(elem1 : 2, elem2:3);
+simpleRec := rec(elem1 : simpleRec`elem1+1, elem2: simpleRec`elem1+1);
+\end{pascalcode}
+
+This code is translated to the following code and solved in three steps: solve
+the expression, store it to a temporal variable and  store the result in the
+final place.
+
+\begin{pascalcode}
+VAR tmp1, tmp2 IN
+tmp1 := 1;
+tmp2 := 2;
+simpleRec`elem1 := tmp1;
+simpleRec`elem2 := tmp2;
+END;
+VAR tmp3, tmp4 IN
+tmp3 := simpleRec`elem1 + 1;
+tmp4 := simpleRec`elem1 + 1;
+simpleRec`elem1 := tmp3;
+simpleRec`elem2 := tmp4;
+END
+\end{pascalcode}
+
+In this example, note that the final value of $simpleRec'elem2$ is 2. This 
+happens due to atomic assignment. This translation is represented by the 
+$Record$ rule and it is omitted, because it is meta rule very simple.
+
+The \textbf{case 2} is the atrribution between records:
+
+\begin{pascalcode}
+  newSimpleRec = simpleRec;
+\end{pascalcode}
+
+The following code is its translation to LLVM:
+
+\begin{llvmcode}
+  %4 = bitcast %struct.database* %newSimpleRec to i8*
+  %5 = bitcast %struct.database* %simpleRec to i8*
+  call void @llvm.memcpy.p0i8.p0i8.i64(i8* %4, i8* %5, i64 12, i32 0, i1 false)
+....
+; Function Attrs: nounwind
+declare void @llvm.memcpy.p0i8.p0i8.i64(i8* nocapture, i8* nocapture readonly, i64, i32, i1) #1
+\end{llvmcode}
+
+The instruction bitcast is used to get the reference pointer from source address 
+and the destination address. The function \textit{memcpy} get a sequence of argument.
+The first argument is a pointer to the destination, the second is a pointer to the source.
+The third argument is an integer argument specifying the number of bytes to copy, the
+fourth argument is the alignment of the source and destination locations, and the fifth
+is a boolean indicating a volatile access. By default, the last two arguments are 0 and false.
 
@@
 
-%[[ Write a text here to define the rule for it]]
-
-There are three attributions cases related to array.
+There are three attribution cases related to array.
 \begin{description}
 
-\item[Case 1.1] - Initializing an unique value to an array:
+\item[Case 1.1] - Initializing a unique value to an array using Cartesian product:
 \[
@@
 \]
-\item[Case 1.2] - Initializing different values to an array\footnote{B0 definition requires the declaration of all values.}:
+
+\item[Case 1.2] - Initializing different values to an 
+array\footnote{B0 definition requires the declaration of all values.} using a sequence of map:
 \[
-\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{5}_{Exp} , ... ,
-(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{5}_{Exp} }_{seq\ Map}\}}_{Array}
+\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{1}_{Exp} , ... ,
+(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{27}_{Exp} }_{seq\ Map}\}}_{Array}
 \]
-
-Theses cases use a declared constant that is attributed to the variable:
-\begin{llvmcode}
-@array = global [3 x [3 x [4 x i32]]] 
-[[3 x [4 x i32]] 
-	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
-	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
-	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
- [3 x [4 x i32]] 
-	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
-	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
- 	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
- [3 x [4 x i32]] 
-	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
-	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
-	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
-\end{llvmcode}
-
-The rule for declare array constants:
-
-\begin{align*}
-\begin{split}
-  \trad{array}{\B{Type}} \isdef \\
-   & \LET \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
-   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}\\
-   & \quad \ELSIF array \mbox{ is \B{Map} } \THEN\\
-   & \quad \quad  dec = \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
-   & \quad \IN \llvm{" global \PH{dec}"}\\
-\end{split}
-\end{align*}
-\VGM{Alguma sugestão com relação as funções:head, tail.}
-
-\begin{align*}
-\begin{split}
-  \trad{type, seqIntl, value}{\B{Product}} \isdef \\
-   &\LET \B{h = head(seqIntl)} \AND\\
-   & \B{times = h.end - h.start + 1} \AND\\
-   & values = \trad{times,type,value}{\B{TV}} \IN\\
-   & \IF \B{seqIntl.length} == 1 \THEN  \\
-   & \llvm{"[\PH{times} x \PH{type}] [\PH{values}]"}  \\
-   & \ELSE \LET \\
-   &  tvalues = \trad{type,tail(seqIntl),values}{\B{Product}} \\
-   & \IN \\
-   & \llvm{"[\PH{times} x \PH{type}]  \PH{tvalues}"}  \\
-\end{split}
-\end{align*}
-
-\begin{align*}
-\begin{split}
-  \trad{times, type, value}{\B{TV}} \isdef \\
-   &\LET res1 = \llvm{" \PH{type} \PH{value} "} \IN\\
-   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
-   & \ELSE \LET   res2 =   \trad{times-1,type,value}{\B{TV}}  \IN  \\
-   & \llvm{"\PH{res1} , \PH{res2}"} \\
-\end{split}
-\end{align*}
-
-The ``head'' function returns the first element from sequence.
-The ``tail'' function returns the given sequence without its first element.
 
@@
 
-\item[Case 2.1] - Updating a selected element from array by a value of variable:
+\item[Case 2.1] - Updating a selected element from array by a variable value :
 \[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
@@
 The arrays in B0 are represented by functions where an index points to an 
-element. By default, the element of index not initialized has a zero value.
-The different ways to evaluate an array are described as follow:
+element. The different ways to evaluate an array are described as follows:
 
@@
 
-An alternative way to represent these attributions cannot be using sequential attributions as follow:
+An alternative way to represent these attributions cannot be by
+using sequential attributions as follow:
 
@@
 
-It is not possible, because B array attribution are atomic. The B array 
-attribution are made in two steps. First, the new value is calculated and
-copied to a temporal place. Second, the sequence of new values are setted 
-in theirs address. Finally, considering these steps and the execution of
+It is not possible because B array attributions are atomic. The B array 
+attribution is made in two steps. First, the sequence of new value is calculated and
+copied to its temporal place. Second, the sequence of new values is setted 
+in its address. Finally, considering these steps and the execution of
 following code then the value of ``array2(0,0,1)'' is ``2''.
@@
 \begin{llvmcode}
-%7 = getelementptr [2 x [2 x [4 x i32]]]* @Array2, i32 0, i64 1, i64 1, i64 3
+%7 = getelementptr [2 x [2 x [4 x i32]]]* @array2, i32 0, i64 1, i64 1, i64 3
 \end{llvmcode}
@@
 7 is stored in it.
-\begin{figure}[h]
+
 \begin{llvmcode}
@@
 %6 = sext i32 %5 to i64
-%7 = getelementptr[2x[2x[4xi32]]]*@Array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
+%7 = getelementptr[2x[2x[4xi32]]]*@array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
 store i32 7, i32* %7         ;setting an element
 \end{llvmcode}
-\end{figure}
 
@@
 \paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.
+The first case uses a declared constant that is attributed to the variable:
+\begin{llvmcode}
+@array = global [3 x [3 x [4 x i32]]] 
+[[3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
+\end{llvmcode}
+
+The rule for declare array constants:
+
+\begin{align*}
+\begin{split}
+  \trad{array}{\B{Type}} \isdef \\
+   & \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
+   & \quad \quad \llvm{" global \PH{\trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}}"}\\
+%   & \quad \ELSIF array \mbox{ is \B{seq Map} } \THEN\\
+%   & \quad \quad  \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
+\end{split}
+\end{align*}
+
+\begin{align*}
+\begin{split}
+  \trad{type, seqSet, value}{\B{Product}} \isdef \\
+   & \LET\\
+   & \quad \B{times} \BE \B{head}(seqSet).end - \B{head}(seqSet).start + 1\\
+   & \quad left \BE \trad{seqSet,type}{ArrayType} \IN \\
+   & \IF \B{seqSet.length} == 1 \THEN  \\
+   & \quad \LET right = \trad{times,type,value}{\B{TV}} \IN\\
+   & \quad \llvm{"\PH{left} [\PH{right}]"}  \\
+   & \ELSE \\
+   & \quad \llvm{"\PH{left} \PH{\trad{type,tail(seqSet),value}{\B{Product}}}"}\\
+\end{split}
+\end{align*}
+
+\begin{align*}
+\begin{split}
+  \trad{times, type, value}{\B{TV}} \isdef \\
+   &\LET res1 \BE \llvm{" \PH{type} \PH{value} "} \IN\\
+   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
+   & \ELSE\\
+   & \quad \llvm{"\PH{res1} , \PH{\trad{times-1,type,value}{\B{TV}}}"} \\
+\end{split}
+\end{align*}
 
 \paragraph{seq Map} declares every address and values from array, it is represented by \textbf{case 1.2}.
+
+The attribution using an sequence of Map are made in atomic way and it is simple example.
+
+\begin{pascalcode}
+ array2 := ( (0..1) * (0..1) * (0..3) * {1}); // Cartesian product
+ array2 := { 0|->0|->0|->2, 0|->0|->1|->(array2(0,0,0)+1), ...}; // sequence of map
+\end{pascalcode}
+
+This code is translated in the following code and solved in three steps: solve 
+the expression, store it to a temporal variable and  store the result in the 
+final place.
+
+\begin{pascalcode}
+VAR tmp1, tmp2 IN
+tmp1 := 2;
+tmp2 := array(0,0,0)+1;
+array(0,0,0) := tmp1;
+array(0,0,1) := tmp2;
+END;
+\end{pascalcode}
+
+In this example, note that the final value of $array(0,0,1)$ is 2. This 
+happens due to atomic assignment. This translation is represented by the 
+$Map$ rule and it is omitted, because it is meta rule very simple.
 
@@
 representation LLVM starts at 0. To access an element from array the formula is 
-$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The following rule 
-\footnote{The function $TSymbol$ returns the ``start'', ``end'' and type of the one 
-index.} makes the shift for each index of array. \VGM{TODO: The function TSymbol
-returns the ``start'', ``end'' and type of the one index.}
+$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The function $TSymbol$
+returns the ``start'', ``end'' and type of the one index. The following rule 
+makes the shift for each index of array.
 
@@
 
-\paragraph{Relevant translation restrictions:}
+The \textbf{case 3} is the atrribution between arrays. This translation is 
+very similar to atribution between records.
+
+\begin{pascalcode}
+	newArray = simpleArray;
+\end{pascalcode}
+
+The following code is its translation to LLVM:
+
+\begin{llvmcode}
+%x4  = bitcast [10 x i32]* @newArray to i8*
+%x5  = bitcast [10 x i32]* @simpleArray to i8*
+call void @llvm.memcpy.p0i8.p0i8.i64(i8* %x4, i8* %x5, i64 40, i32 0, i1 false)
+\end{llvmcode}
+
+\paragraph{Relevant translation restrictions in B2LLVM:}
 \begin{itemize}

### 2014-07-07 DOCUMENTATION/specification.tex
Small updates about array

@@
 
+There are also functions to manipulate the sequences: the $\B{``head''}$ function 
+returns the first element from sequence; the $\B{``tail''}$ function returns 
+the given sequence without its first element.
+
 \section{Translation: Modularity Aspects}
@@
 The data types \textbf{structures} and \textbf{arrays} require detailed 
-representation in the abstract syntax structure from the table \ref{tab:node-attr}.
-
-The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.
+representation in the abstract syntax structure from the Table \ref{tab:node-attr}.
+
+Section \ref{sec:StructArray} also presents other examples and rules for these derived data types.
 
@@
 \begin{align*}
-\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{TField}}} \AND\\
- l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
-\end{align*}
-$$\trad{t}{\B{TField}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$
+\trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{Field}}}\\
+%\AND l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
+  \\IN  \llvm{" type \{\PH{tps}\}\nl"}
+\end{align*}
+$$\trad{t}{\B{Field}} \isdef \LET res \BE  \trad{t.type}{\B{Type}} \IN \llvm{"\PH{res},"}$$
 
@@
 
-[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]
+%TODO
+\VGM{[[ TODO: Write about the initialization (Structure and Array) unsing the concept of module/instance]]}
 
 The current specification of the code generator does not handle deferred sets.
-These data types and logic/math operators compose expressions that is in 
-the next section.
+These data types and logic/math operators compose expressions that are explained
+in the next section.
 
@@
   \trad{a}{\B{Beq}} \isdef
-  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Exp}} \IN \\
+  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{ValueItem}} \IN \\
   & \llvm{"\PH{l}} \\
@@
 
-\textcolor{blue}{ The expressions have elements represented by pointers that need a special 
+\textcolor{blue}{The expressions have elements represented by pointers that need a special 
 care. The attributions between these elements are made  directly by the memory 
@@
   \trad{n}{\B{ValueItem}} \isdef \\
-   & \LET \ell = \lbl  \IN\\
-   & \IF \B{n} \mbox{ is \B{RecField}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{RecField}} \\
+   & \LET \ell \BE \lbl  \IN\\
+   & \IF \B{n} \mbox{ is \B{RecField}} \THEN \LET p' \sep v' \sep t' = \trad{\B{n}}{\B{RecField}} \IN \\
+   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \sep  \ell  \sep t'\\
+   & \ELSIF \B{n} \mbox{ is \B{Record}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{Record}} \\
    & \ELSIF \B{n} \mbox{ is \B{ArrayItem}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{ArrayItem}} \\
-   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \\
-   & \sep  \ell  \sep t'\\
+   & \llvm{"\PH{p'} \PH{\ell}  = load  \PH{t'}* \PH{v'} \nl"} \sep  \ell  \sep t'\\
+   & \ELSIF \B{n} \mbox{ is \B{Array}} \THEN p' \sep v' \sep t' = \trad{\B{n}}{\B{Array}} \\
+   & code \sep  \ell  \sep t'\\
+   & \ELSE  p' \sep v' \sep t' = \trad{\B{n}}{\B{Exp}}
 \end{split}

### 2014-07-07 b2llvm/translate.py
Update the clause While to support conditions as formula.

@@ -363,7 +363,7 @@ def section_implementation(buf, m, emit_printer):
         i = implementation(m)
         x_init(buf, m, i)
         for op in i["operations"]:
-            x_operation(buf, op)
+            x_operation(buf, m, op)
         if emit_printer:
             x_printer(buf, m, i)
 
@@ -436,7 +436,9 @@ def x_init(buf, m, i):
             lexicon[q] = "%arg"+str(count)+"$"
             count += 1
     # 1.2 generate parameter type, name list
-    arg_list = [ tm+SP+"%self$" ]
+    arg_list = []
+    if is_stateful(m) : 
+                arg_list = [ tm+SP+"%self$" ]
     for q in comp_indirect(m):
         if is_stateful(q.mach):
             arg_list.append(state_r_name(q.mach)+SP+lexicon[q])
@@ -562,7 +564,7 @@ def x_printer(buf, m, i):
 
 ### TRANSLATION OF OPERATIONS
 
-def x_operation(buf, n):
+def x_operation(buf, mod, n):
     '''
     Code generation for B operations.
 
@@ -574,7 +576,10 @@ def x_operation(buf, n):
     check_kind(n, {"Oper"})
     names.reset()
     buf.trace.out("The LLVM function implementing B operation \""+n["id"]+"\" in \""+n["root"]["id"]+"\" follows.")
-    args = commas([state_r_name(n["root"])+SP+"%self$"]+
+    arg1 = []
+    if is_stateful(mod) : 
+                arg1 = [state_r_name(n["root"])+SP+"%self$"]
+    args = commas( arg1+
                   [x_type(i["type"])+SP+"%"+i["id"] for i in n["inp"]]+
                   [x_type(o["type"])+"*"+SP+"%"+o["id"] for o in n["out"]])
     buf.code(FNDEF, op_name(n), args)
@@ -896,9 +901,10 @@ def x_while(buf, n, lbl):
     lbl1 = names.new_label()
     buf.code(GOTO, lbl1)
     buf.code(LABEL, lbl1)
-    v = x_pred(buf, n["cond"])
     lbl2 = names.new_label()
-    buf.code(CGOTO, v, lbl2, lbl)
+        #v = x_pred(buf, n["cond"])
+    x_formula(buf,n["cond"],lbl2, lbl) 
+        #buf.code(CGOTO, v, lbl2, lbl)
[…]

### 2014-07-07 b2llvm/c_translate.py
Fixing the skeleton generator.

@@ -53,8 +53,8 @@ def generate_header_skeleton(ast,bmodule):
     #if transLLVM.is_stateful(ast):
     comp.append(ast)
     comp.extend([x.mach for x in transLLVM.comp_indirect(ast)])
-    #args = commas([c_state_r_name(x)+" self" for x in comp if transLLVM.is_stateful(x)])
-    args = commas([c_state_r_name(x)+" self" for x in comp ])
+    args = commas([c_state_r_name(x)+" self" for x in comp if transLLVM.is_stateful(x)])
+    #args = commas([c_state_r_name(x)+" self" for x in comp ])
     buf_header.code(EXTFNDEC, transLLVM.init_name(ast)[1:], args)
     
     
@@ -63,8 +63,9 @@ def generate_header_skeleton(ast,bmodule):
         # compute in tl the list of arguments types
         buf_header.trace.out("The declaration of the function implementing operation \""+op["id"]+"\" is:")
         tl = list()
-        # if transLLVM.is_stateful(ast):
-        tl.append(c_state_r_name(ast)+" self")
+        #TODO: 
+        if transLLVM.is_stateful(ast):
+            tl.append(c_state_r_name(ast)+" self")
         tl.extend([ cx_type(i["type"])+" "+i["id"] for i in op["inp"] ])
         tl.extend([ cx_type(o["type"])+"*"+" "+o["id"] for o in op["out"] ])
         buf_header.code(EXTFNDEC, transLLVM.op_name(op)[1:], commas(tl))
@@ -90,8 +91,8 @@ def section_typedef_impl(buf, i, m):
     import b2llvm.translate as transLLVM
     transLLVM.check_kind(i, {"Impl"})
     transLLVM.check_kind(m, {"Machine"})
-    #if transLLVM.is_stateful(i):
-    if True:
+    if transLLVM.is_stateful(i):
+    #if True:
         buf.trace.out("The type encoding the state of \""+m["id"] + "\" is an aggregate and is defined as")
         buf.trace.outu("(using implementation \""+i["id"]+"\"):")
         buf.trace.tab()

### 2014-07-07 coverage.sh
Adding new examples

@@ -9,32 +9,49 @@ COV_RUN="coverage run --source=b2llvm.py,b2llvm -p"
 echo "Cleaning up existing coverage data."
 coverage erase
 echo "Generating new coverage data."
+#$COV_RUN ./b2llvm.py -m comp Calculator Calculator_i.llvm bxml project.xml
+#$COV_RUN ./b2llvm.py -m proj Calculator init-Calculator.llvm bxml project.xml
+
+# Simple example BETA
+$COV_RUN ./b2llvm.py --trace --verbose  -m comp Division Division.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj Division init-Division.llvm bxml project.xml
+
+$COV_RUN ./b2llvm.py -m comp  Prime Prime.llvm bxml project.xml
+$COV_RUN ./b2llvm.py --trace --verbose  -m proj Prime init-Prime.llvm bxml project.xml
+
+#$COV_RUN ./b2llvm.py -m comp Rooms  Rooms.llvm bxml project.xml
+#$COV_RUN ./b2llvm.py -m proj Rooms init-Rooms.llvm bxml project.xml
+
+$COV_RUN ./b2llvm.py -m comp swap swap.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj swap init-swap.llvm bxml project.xml
+
 $COV_RUN ./b2llvm.py -v -m comp counter counter.llvm bxml project.xml > /dev/null
-$COV_RUN ./b2llvm.py -m comp enumeration enumeration.llvm bxml project.xml
-$COV_RUN ./b2llvm.py -m comp array array.llvm bxml project.xml
-$COV_RUN ./b2llvm.py -m comp ioint ioint.llvm bxml project.xml
-$COV_RUN ./b2llvm.py -m comp counter counter.llvm bxml project.xml
-$COV_RUN ./b2llvm.py -m comp wd wd.llvm bxml project.xml
-$COV_RUN ./b2llvm.py -m comp timer timer.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -m proj counter init-counter.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m comp enumeration enumeration.llvm bxml project.xml
+$COV_RUN ./b2llvm.py -m proj enumeration init-enumeration.llvm bxml project.xml
+#$COV_RUN ./b2llvm.py -m comp array array.llvm bxml project.xml
[…]

### 2014-07-07 coverage.sh
Adding a set of tests to diff the generated code and expected code.

@@ -53,5 +53,5 @@ echo "Coverage report is now available."
 
 
 echo "Generating diff report "
-for file in *.h *.llvm; do diff "$file" "expected_code/${file##*/}">expected_code/REPORT; echo "Diff for ${file##*/}:"; cat expected_code/REPORT;   done
+for file in *.h *.llvm; do diff "$file" "expected_code/${file##*/}">expected_code/REPORT_"$file".diff; echo "Diff for ${file##*/}:"; cat expected_code/REPORT_"$file".diff;   done
 echo "Report generated"

### 2014-07-07 expected_code/enumeration.h

@@ -1,0 +1,12 @@
+#ifndef _enumeration_h
+#define _enumeration_h
+
+#include <stdint.h>
+#include <stdbool.h>
+
+typedef struct {i1 current }   enumeration$state$;
+typedef enumeration$state$ * enumeration$ref$;
+extern void enumeration$init$(enumeration$ref$ self);
+extern void enumeration$tick(enumeration$ref$ self);
+extern void enumeration$read(enumeration$ref$ self, i1* res);
+#endif /* _enumeration_h */

### 2014-07-07 expected_code/swap.h

@@ -1,0 +1,13 @@
+#ifndef _swap_h
+#define _swap_h
+
+#include <stdint.h>
+#include <stdbool.h>
+
+typedef struct {int32_t v1; int32_t v2; }   swap$state$;
+typedef swap$state$ * swap$ref$;
+extern void swap$init$(swap$ref$ self);
+extern void swap$step(swap$ref$ self);
+extern void swap$set(swap$ref$ self, int32_t av1, int32_t av2);
+extern void swap$get(swap$ref$ self, int32_t* r1, int32_t* r2);
+#endif /* _swap_h */

### 2014-07-07 expected_code/timer.h

@@ -1,0 +1,18 @@
+#ifndef _timer_h
+#define _timer_h
+
+#include <stdint.h>
+#include <stdbool.h>
+
+typedef struct {int32_t value; bool error; }   counter$state$;
+typedef counter$state$ * counter$ref$;
+typedef struct {counter$ref$; counter$ref$; counter$ref$; bool is_running; bool overflow; }   timer$state$;
+typedef timer$state$ * timer$ref$;
+extern void timer$init$(timer$ref$ self, counter$ref$ self, counter$ref$ self, counter$ref$ self);
+extern void timer$tick(timer$ref$ self);
+extern void timer$reset(timer$ref$ self);
+extern void timer$stop(timer$ref$ self);
+extern void timer$start(timer$ref$ self);
+extern void timer$elapsed(timer$ref$ self, int32_t* hours, int32_t* minutes, int32_t* seconds);
+extern void timer$has_overflown(timer$ref$ self, bool* answer);
+#endif /* _timer_h */

### 2014-07-07 expected_code/wd.h

@@ -1,0 +1,15 @@
+#ifndef _wd_h
+#define _wd_h
+
+#include <stdint.h>
+#include <stdbool.h>
+
+typedef struct {int32_t value; bool error; }   counter$state$;
+typedef counter$state$ * counter$ref$;
+typedef struct {counter$ref$ }   wd$state$;
+typedef wd$state$ * wd$ref$;
+extern void wd$init$(wd$ref$ self, counter$ref$ self);
+extern void wd$start(wd$ref$ self);
+extern void wd$tick(wd$ref$ self);
+extern void wd$expired(wd$ref$ self, bool* res);
+#endif /* _wd_h */

### 2014-07-13 DOCUMENTATION/specification.tex
Adding new notes about array

@@
   call this the \emph{COMP} mode for code generation.
+\item Code generation for integration between to C/C++ code.
+  The code generated is a skeleton with types and functions declaration 
+  that defines a C header file. This produced skeleton file can be integrated to
+  C/C++ unit tests. This mode for code generation is called \emph{SKELETON}.
+
 \item Code generation is applied to build a full system from a project. Given
@@
 
+This Section presents only the rules related to declaration of derived type data types.
 Section \ref{sec:StructArray} also presents other examples and rules for these derived data types.
@@
 
+%\AND l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
 The rules to declare the structure type are:
@@
 \trad{Structure}{\B{Type}} \isdef \LET tps \BE \trad{Structure.elements}{\ListOf{\B{Field}}}\\
-%\AND l \BE \local(\B{Structure.elements})  \IN  \llvm{"\PH{l}= type \{\PH{tps}\}\nl"}
-  \\IN  \llvm{" type \{\PH{tps}\}\nl"}
+   \IN  \llvm{" type \{\PH{tps}\}\nl"} 
 \end{align*}
@@
 
-The function ArrayType returns the type for a declared variable array in 
-structure state, for example:
+The function ArrayType returns the type for a declared variable.
 
@@
 
-The current specification of the code generator does not handle deferred sets.
+This current specification of the code generator supports also multi dimensional array.
 These data types and logic/math operators compose expressions that are explained
@@
 
-\include{derivedTypes}
+\subsection{Translation of structures and arrays}
+\label{sec:StructArray}
+
+Two approaches are possible to support arrays and structures. One approach is 
+importing reusable libraries of AtelierB that are reported in the official 
+document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. 
+This approach is supported, but the generated code is more complex and it needs 
+more intermediate steps to manipulate the data of imported components. 
+Moreover, this approach allows dynamic allocation for arrays, but it is 
+strongly not recommended for safety-critical systems. A second approach enjoys 
+the direct correspondence between B0 native types and LLVM and this approach 
+is presented here. This section describes the translation of derived data 
+types showed in Section~\ref{sec:data}.
+
+\subsubsection{Structure} 
+The structure declaration defines a list of variables to be placed under one 
+name, allowing several variables to be accessed by one base name. This example  
+initializes a structure, and gets and sets one value of xx'balance.
+
+\begin{description}
+\item[Case 1.1] Initializing a record with a sequence of expressions:
+\[
+  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
+\]
+
+\item[Case 1.2] Setting a record field to a variable:
+\[
+  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
+\]
+
+\item[Case 2]  Attribution between structures:
+\[
+  \underbrace{newRecord}_{Record} :=  \underbrace{accountRecord}_{Record} 
+\]
+\end{description}
+
+The B code example illustrates the translation of \textbf{Case 1.1} and \textbf{Case 1.2}:
+
+\begin{pascalcode}
+xx := rec(id:1, balance:10);
+xx`balance := xx`balance-1;
+\end{pascalcode}
+
+Lines 1 and 2 are represented in translation in LLVM code:
+
+\begin{llvmcode}
+%1 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 0
+store i32 1, i32 * %1
+%2 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+store i32 10, i32 * %2
+
+%3 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+%4 = load i32* %3
+%5 = sub nsw i32 %4, 1
+%6 = getelementptr %struct.R_1* @Rec__account, i32 0, i32 1
+store i32 %5, i32* %6
+\end{llvmcode}
+
+The $RecField$ rule gets an element value from the record. The same rule is 
+used on the left side of attribution as an $LValue$ and right side of 
+attribution as an $Expression$. The functions \SBPos(n) and \SBType(n)  get 
+respectively the position and type of $n$ element in the structure.
+
+\begin{align*}
+\begin{split}
+  \trad{n}{\B{RecField}} \isdef \\
+   &\LET \ell = \lbl  \AND  t = n.type \AND v = \B{head}(n.element) \AND\\
+   &code1 \BE \llvm{"\PH{\ell} = getelementptr \PH{t}* \PH{v}, i32 0, i32  \SBPos(\B{n})\nl"} \IN \\
+   &\IF \B{n.element} \mbox{ is \B{BOOL or NAT or INT} \AND n.element.length=1}\\
+   &\quad \THEN  code1 \sep  \ell  \sep \SBType(\B{n})\\
+   &\ELSIF \B{head(n.element)} \mbox{ is \B{RecField} }  \THEN \\
+   &\LET n2 \BE new \B{RecField}() \AND  \B{n2.base}  \BE \B{head}(n.element)\\
+   &\AND \B{n2.element} \BE \B{tail}(n.element) \AND \\
+   &\quad code2, label, pos = \trad{n2}{RecField} \IN\\
+   &\quad \llvm{"\PH{code1} \PH{code2}"} \sep  label  \sep pos \\
+\end{split}
+\end{align*}
+
+\subsection{Atomic Multiple Attribution for Structure }
+The attribution from record structure to a record structure variable, Line 1 
+from the next B code example, is translated by a sequence of applications of
+the  rule $RecField$. This rule was also designed to support nested structures.
+
+The \textbf{Case 1} is a structure attribution made in atomic way, for example:
+
+\begin{pascalcode}
+simpleRecA := rec(elem1 : 1, elem2:2);
+simpleRecB := rec(elem1 : simpleRecB`elem1+1, elem2: simpleRecB`elem1+1);
+\end{pascalcode}
+
+This code is translated to the following code and solved in three steps: solve
+the expression, store it to a temporal variable and store the result in the
+final place.
+
+\begin{pascalcode}
+VAR tmp1, tmp2 IN
+tmp1 := 1;
+tmp2 := 2;
+simpleRecA`elem1 := tmp1;
+simpleRecA`elem2 := tmp2;
+END;
+VAR tmp3, tmp4 IN
+tmp3 := simpleRecB`elem1 + 1;
+tmp4 := simpleRecB`elem1 + 1;
+simpleRecB`elem1 := tmp3;
+simpleRecB`elem2 := tmp4;
+END
+\end{pascalcode}
+
+In this example, note that the final value of $simpleRecB'elem2$ is 2. This 
+happens due to atomic assignment. This translation is represented by the 
+$Record$ rule and it is omitted, because it is a meta rule simple.
+
+The \textbf{Case 2} is the atrribution between records:
+
+\begin{pascalcode}
+  newSimpleRec = simpleRec;
+\end{pascalcode}
+
+The following code is its translation to LLVM:
+
+\begin{llvmcode}
+%4 = bitcast %struct.database* %newSimpleRec to i8*
+%5 = bitcast %struct.database* %simpleRec to i8*
+call void @llvm.memcpy.p0i8.p0i8.i64(i8* %4, i8* %5, i64 12, i32 0, i1 false)
+....
+; Function Attrs: nounwind
+declare void @llvm.memcpy.p0i8.p0i8.i64(i8* nocapture, i8* nocapture readonly,i64,i32,i1)
+\end{llvmcode}
+
+The instruction bitcast is used to get the reference pointer from source address 
+and the destination address. The function \textit{memcpy} get a sequence of argument.
+The first argument is a pointer to the destination, the second is a pointer to the source.
+The third argument is an integer argument specifying the number of bytes to copy, the
+fourth argument is the alignment of the source and destination locations, and the fifth
+is a boolean indicating a volatile access. By default, the last two arguments are 0 and false.
+
+%[No futuro colocar a regra aqui ou apenas descrever como ela seria implementada]
+
+\subsubsection{Arrays}
+
+The arrays store a fixed-size sequential collection of elements with the same type. 
+They consist of contiguous memory locations, where the lowest address (0) 
+corresponds to the first element and the highest address to the last element.
+In B, a concrete array is a total function where the domain indexes the range 
+elements.
+
+There are three main attribution cases related to array.
+\begin{description}
+
+\item[Case 1.1] - Initializing a unique value to an array using Cartesian product:
+\[
+\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..2 \times 0..2 \times 0..3  }_{seq\ Exp} \times \underbrace{\{5\}}_{Exp} }_{Product}\}}_{Array}
+\]
+
+\item[Case 1.2] - Initializing different values to an 
+array\footnote{B0 definition requires the declaration of all values.} using a sequence of map:
+\[
+\textcolor{gray}{arr:=} \underbrace{\{\underbrace{(\underbrace{(0 \mapsto 0) \mapsto  0)}_{seq Exp} \mapsto \underbrace{1}_{Exp} , ... ,
+(\underbrace{(2 \mapsto 2) \mapsto  3)}_{seq\ Exp} \mapsto \underbrace{27}_{Exp} }_{seq\ Map}\}}_{Array}
+\]
+
+\begin{comment}
+\$VarName = global [size X type] X [(type value,)^+]
+
+|array|Type = LET
+				  IF array is Product THEN dec = |array.arg1, array.arg2 |CProduct  ( Quebrando o padrão?)
+				  ELSEIF array is seq Map THEN dec = |array.Map |Map
+				IN	  
+					" global "^ dec
+				END
+|sIntl, value |CProduct = 
+    	IF sIntl is  empty THEN  ""
+	ELSE THEN
+		LET sizeType = |sIntl, "i32"|PrintSizeType &
+		 tailSizeType=|sIntl.tail,"i32"|PrintSizeType & 
+		 values= |tailSizeType,values|CProduct IN
+		"[sizeType]  [ tailSizeType  values ]"		
+		END
+	END
+
+|sIntl,t| PrintSizeType =
+    	IF sIntl.size == 1  THEN  "t"
+	ELSE THEN
+		LET size = sIntl.head.size()  &
+			type = |sIntl.tail,t| PrintSizeType    &
+		"[size X type ]"		
+		END
+	END
+
+|array|Type = IF t is Map THEN ...
+\end{comment}
+
+\item[Case 2.1] - Updating a selected element from array by a variable value :
+\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
+\]
+\item[Case 2.2] - Setting a selected element from array to a variable:
+\[ 
+\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii,1+jj,1+xx}_{seq\ Exp})}_{ArrayItem}
+\]
+\item[Case 3] - Attribution between arrays\footnote{B0 definition requires the same type.}:
+\[
+\underbrace{arr1}_{Array} := \underbrace{arr2}_{Array} 
+\]
+\end{description}
+
+\subsection{Relation arrays B0 and LLVM}
+
+The arrays in B0 are represented by functions where an index points to an 
+element. The different ways to evaluate an array are described as follows:
+
+\subsubsection{Initializing a unique value to an array (Case 1.1)}
+The following two attribution commands have the same semantic effect.
+
+\begin{pascalcode}
+array2 := ( 0..1 * 0..1 * 0..3 * {5});
+
+array2 := { 0|->0|->0|->5 , 0|->0|->1|->5 , 
+            0|->1|->0|->5 , 0|->1|->1|->5 , 
+            1|->0|->0|->5 , 1|->0|->1|->5 , 
+            1|->1|->0|->5 , 1|->1|->1|->5 ,
+            0|->0|->2|->5 , 0|->0|->3|->5 ,
+            0|->1|->2|->5 , 0|->1|->3|->5 , 
+            1|->0|->2|->5 , 1|->0|->3|->5 , 
+            1|->1|->2|->5 , 1|->1|->3|->5 };
+\end{pascalcode}
+
+An alternative way to represent these attributions cannot be by
+using sequential attributions as follow:
+
+\begin{pascalcode}
+array2(0,0,0) := 5;
+array2(0,0,1) := 5;
+// ... - A sequence of attributions
+array2(1,1,3) := 5;
+\end{pascalcode}
+
+\subsubsection{Initializing different values to an array (Case 1.2) }
+It is not possible because B array attributions are atomic. The B array 
+attribution is made in two steps. First, the sequence of new value is calculated and
+copied to its temporal place. Second, the sequence of new values is setted 
+in its address. Finally, considering these steps and the execution of
+following code then the value of ``array2(0,0,1)'' is ``2''.
+
+\begin{pascalcode}
+ array2 := ( (0..1) * (0..1) * (0..3) * {1});
+ array2 := { 0|->0|->0|->2 , 0|->0|->1|->(array2(0,0,0)+1), ...  };
+\end{pascalcode}
+
+The LLVM instruction \textit{getelementptr} is used to select an element in
+the array or structure. A simple example is:
+
+\begin{llvmcode}
+%7 = getelementptr [2 x [2 x [4 x i32]]]* @array2, i32 0, i64 1, i64 1, i64 3
+\end{llvmcode}
+
+The first two parameters define the type and name of the variable, and the
+others are indexes to select the element in the data type. The third parameter (i32 0)
+selects the first element in the pointer (* @array2). The last elements 
+(i64 1, i64 1, i64 3) select the value from position (1,1,3) storing it in \%7.
+This array is multidimensional, so several indexes must be declared separated
+by comma to get the specific element.
+
+\subsubsection{Setting and getting array}
+
+The following blocks illustrate a B code setting a value to array
+and its translation to LLVM.
+
+\begin{pascalcode} 
+VAR xx, yy, zz IN
+	xx=1;
+	yy=1;
+	zz=3;
+	array2(x,y,z):=7
+END
+\end{pascalcode}
+
+This example is translated in four steps. First, the variables \%x, \%y, \%z 
+are allocated. Second, the respective values 0, 1, 3 are stored in variables. 
+Third, the variables are loaded to local variables and converted to 64 bits. 
+Finally, the address of array is calculated in local variable \%7 and the value 
+7 is stored in it.
+
+\begin{llvmcode}
+%xx = alloca i32
+%yy = alloca i32
+%zz = alloca i32
+store i32 1, i32* %xx
+store i32 1, i32* %yy
+store i32 3, i32* %zz
+%1 = load i32* %zz
+%2 = sext i32 %1 to i64
+%3 = load i32* %yy
+%4 = sext i32 %3 to i64
+%5 = load i32* %xx
+%6 = sext i32 %5 to i64
+%7 = getelementptr[2x[2x[4xi32]]]*@array2,i32 0,i64 %6,i64 %4,i64 %2;getting an element
+store i32 7, i32* %7         ;setting an element
+\end{llvmcode}
+
+\subsubsection{Rules for support array}
+
+\paragraph{Product} declares all address and a unique value from array, it is represented by \textbf{case 1.1}.
+The first case uses a declared constant that is attributed to the variable:
+\begin{llvmcode}
+@array = global [3 x [3 x [4 x i32]]] 
+[[3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]], 
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]],
+ [3 x [4 x i32]] 
+	[[4 x i32] [i32 5, i32 5, i32 5, i32 5],
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5], 
+	 [4 x i32] [i32 5, i32 5, i32 5, i32 5]]], align 16
+\end{llvmcode}
+
+The rule for declare array constants:
+
+\begin{align*}
+\begin{split}
+  \trad{array}{\B{Type}} \isdef \\
+   & \quad \IF \B{array} \mbox{ is \B{Product} } \THEN \\
+   & \quad \quad \llvm{" global \PH{\trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Product}}}"}\\
+%   & \quad \ELSIF array \mbox{ is \B{seq Map} } \THEN\\
+%   & \quad \quad  \trad{\llvm{"i32"}, array.arg1, array.arg2}{\B{Map}} \\
+\end{split}
+\end{align*}
+
+\begin{align*}
+\begin{split}
+  \trad{type, seqSet, value}{\B{Product}} \isdef \\
+   & \LET\\
+   & \quad \B{times} \BE \B{head}(seqSet).end - \B{head}(seqSet).start + 1\\
+   & \quad left \BE \trad{seqSet,type}{ArrayType} \IN \\
+   & \IF \B{seqSet.length} == 1 \THEN  \\
+   & \quad \LET right = \trad{times,type,value}{\B{TV}} \IN\\
+   & \quad \llvm{"\PH{left} [\PH{right}]"}  \\
+   & \ELSE \\
+   & \quad \llvm{"\PH{left} \PH{\trad{type,tail(seqSet),value}{\B{Product}}}"}\\
+\end{split}
+\end{align*}
+
+\begin{align*}
+\begin{split}
+  \trad{times, type, value}{\B{TV}} \isdef \\
+   &\LET res1 \BE \llvm{" \PH{type} \PH{value} "} \IN\\
+   & \IF times == 1 \THEN \llvm{"\PH{res1}"} \\
+   & \ELSE\\
+   & \quad \llvm{"\PH{res1} , \PH{\trad{times-1,type,value}{\B{TV}}}"} \\
+\end{split}
+\end{align*}
+
+\paragraph{seq Map} declares every address and values from array, it is represented by \textbf{case 1.2}.
+
+The attribution using an sequence of Map are made in atomic way and it is simple example.
+
+\begin{pascalcode}
+ array2 := ( (0..1) * (0..1) * (0..3) * {1}); // Cartesian product
+ array2 := { 0|->0|->0|->2, 0|->0|->1|->(array2(0,0,0)+1), ...}; // sequence of map
+\end{pascalcode}
+
+This code is translated in the following code and solved in three steps: solve 
+the expression, store it to a temporal variable and  store the result in the 
+final place.
+
+\begin{pascalcode}
+VAR tmp1, tmp2 IN
+tmp1 := 2;
+tmp2 := array(0,0,0)+1;
+array(0,0,0) := tmp1;
+array(0,0,1) := tmp2;
+END;
+\end{pascalcode}
+
+In this example, note that the final value of $array(0,0,1)$ is 2. This 
+happens due to atomic assignment. This translation is represented by the 
+$Map$ rule and it is omitted, because it is meta rule very simple.
+
+\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
+where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indexes that 
+calculates the address of element. This rule also supports to translate 
+expressions of arrays. This rule is used to support array attributions 
+from \textbf{cases} \textbf{2.1} and \textbf{2.2}.
+
+\begin{align*}
+\lefteqn{\trad{n}{ArrayItem} \isdef} \\
+&  \LET p_1 \sep v_1 \sep t_1  = \trad{n.base}{Exp}  \IN \\
+&  \LET r \sep p_2  = \trad{n.indexes}{LRExp}  \IN \\
+&  \LET v =  \variable \AND  t = \trad{\B{n.type}}{Type}  \IN \\
+&  \llvm{"\PH{p_1} \PH{p_2} \PH{v} =  getelementptr \PH{t_1}, \PH{v_1}, i32 0 \PH{r}\nl"}  \sep v \sep t\\
+\end{align*}
+
+A list of type and variable is generated by the following rule.
+\begin{align*}
+&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{ShiftExp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
+&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND
+ \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
+& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
+\end{align*}
+
+The element access of array can need a shift in indexes. When the domain of 
+array is an interval and it is not started from zero, it must adjust the 
+indexes shifting the access, because the addressing  of array in intermediate 
+representation LLVM starts at 0. To access an element from array the formula is 
+$address(e(i1,..,in))  = i1-i1\_start, ... , in-in\_start$. The function $TSymbol$
+returns the ``start'', ``end'' and type of the one index. The following rule 
+makes the shift for each index of array.
+
+\begin{align*}
+&\trad{n}{ShiftExp} \isdef \LET p \sep v \sep t  = \trad{n}{Exp} \AND \\
+&\quad iStart \sep iEnd \sep iType = \trad{n}{TSymbol}  \AND  \PH{\ell} = \lbl \IN \\
+&\quad \IF ( iStart == 0 ) \THEN  \llvm{",\PH{t} \PH{v} "} \sep p \\
+&\quad \ELSE \LET  p' \sep v \sep v' \sep t'  = \trad{t,v,itype,istart}{Cast} \IN \\
+&\quad \llvm{",\PH{t} \PH{\ell} "} \sep  \llvm{"\PH{p} \PH{p'}  \PH{\ell} = sub \PH{t'} \PH{v}, \PH{v'}\nl"}  \\
+\end{align*}
+
+The \trad{}{ShiftExp} rule does a subtraction and this operation must have operators with the same data types so the following cast rule is used:
+
+\begin{align*}
+&\trad{t,v,t',v'}{Cast} \isdef \LET  \ell = \lbl \IN \\
+&\quad \IF ( t == t') \THEN p \sep v \sep v' \sep t  \\
+&\quad \ELSE \IF (t<t') \THEN \\
+&\quad \quad \llvm{"\PH{\ell}= bitcast \PH{t} \PH{v} , \PH{t'}\nl"}  \sep l \sep v' \sep t'\\ 
+&\quad \ELSE \\ 
+&\quad \quad \llvm{"\PH{\ell}= bitcast \PH{t'} \PH{v'} , \PH{t}\nl"} \sep v \sep l \sep t \\ 
+\end{align*}
+
+The \textbf{case 3} is the atrribution between arrays. This translation is 
+very similar to atribution between records.
+
+\begin{pascalcode}
+	newArray = simpleArray;
+\end{pascalcode}
+
+The following code is its translation to LLVM:
+
+\begin{llvmcode}
+%x4  = bitcast [10 x i32]* @newArray to i8*
+%x5  = bitcast [10 x i32]* @simpleArray to i8*
+call void @llvm.memcpy.p0i8.p0i8.i64(i8* %x4, i8* %x5, i64 40, i32 0, i1 false)
+\end{llvmcode}
+
+\paragraph{Relevant translation restrictions in B2LLVM:}
+\begin{itemize}
+\item  B0 definition - An array must be initialized declaring all values 
+to avoid no determinism.
+\item B0 definition from CommenC - The operator overwrite from arrays is not supported, for example:
+
+\begin{pascalcode}
+array := array <+ {1|->2|->0|->4,  1|->2|->1|->5}
+\end{pascalcode}
+
+\end{itemize}
 

### 2014-07-13 C/derivedTypes/comp_exec_Test_bubble.sh
Added support to unidimensional array

@@ -1,0 +1,9 @@
+# Before you must generate the bubble.llvm by this command:
+# ./b2llvm.py -v -m comp counter bubble.llvm bxml project.xml
+cp ../../bubble.llvm .
+cp ../../bubble.h .
+clang -S -emit-llvm -c initc-bublle.c
+llc-mp-3.5  bubble.llvm
+llc-mp-3.5  initc-bublle.ll
+clang -o executable bubble.llvm.s  initc-bublle.s
+./executable 

### 2014-07-13 C/derivedTypes/initc-bublle.c

@@ -1,0 +1,32 @@
+#include "bubble.h"
+#include <stdio.h>
+
+bubble$state$ bubble;
+
+int main(){
+	int i=0;
+        bubble.vec1[0]=5;
+        for(int i=0;i<100;i++){
+        	bubble.vec1[i]=100-i; 
+        }
+
+        for(int i=0;i<=10;i++){
+            bubble.vec1[i]=100-i;
+            printf(" vec1[%d]= %d\n",i, bubble.vec1[i]  );
+        }
+	bubble$init$(&bubble);
+
+	for(int i=0;i<=10;i++)
+	{
+        printf(" vec1[%d]= %d\n",i, bubble.vec1[i]  );
+	}
+
+	bubble$op_sort(&bubble);
+        printf("Ordenado\n");
+        for(int i=0;i<100;i++)
+        {
+	         printf(" vec1[%d]= %d\n",i, bubble.vec1[i]  );
+        }
+
+} 
+

### 2014-07-13 b2llvm/ast.py

@@ -94,11 +94,10 @@ ZERO = make_intlit(0)
 ONE = make_intlit(1)
 MAXINT = make_intlit(2147483647)
 
-#VGM - TODO: move this functions ?
 def make_sset(node):
     """Creates an AST node for a B simple set."""
     if (node.get("operator") == ".."):
-        res = make_interval(node[0].get("value"),node[1].get("value"))
+        res = make_interval(node[1].get("value"),node[2].get("value"))
     elif (node.get("operator") == "*"):
         res = become_list(make_sset(node[0]))
         res += become_list(make_sset(node[1]))
@@ -124,9 +123,9 @@ def make_arrayType(domxml, ranxml):
     assert dom != None and ran != None
     return { "kind" : "arrayType", "dom": dom, "ran" : ran}
 
-def make_arrayItem(base,index):
+def make_arrayItem(base,indexes):
     """Creates an AST node for a B arrayType."""
-    return { "base": base, "index": index, 
+    return { "base": base, "indexes": indexes, 
             "kind":"arrayItem", "type": INTEGER} #TODO: change it support other typer different of INTEGER 
 
 ### COMPOSED EXPRESSIONS ###

### 2014-07-13 b2llvm/c_translate.py

@@ -66,8 +66,8 @@ def generate_header_skeleton(ast,bmodule):
         #TODO: 
         if transLLVM.is_stateful(ast):
             tl.append(c_state_r_name(ast)+" self")
-        tl.extend([ cx_type(i["type"])+" "+i["id"] for i in op["inp"] ])
-        tl.extend([ cx_type(o["type"])+"*"+" "+o["id"] for o in op["out"] ])
+        tl.extend([ cx_type_name_dec(False, i["type"], i["id"]) for i in op["inp"] ])
+        tl.extend([ cx_type_name_dec(True, o["type"], o["id"]) for o in op["out"] ])
         buf_header.code(EXTFNDEC, transLLVM.op_name(op)[1:], commas(tl))
 
     buf_header.code(ENDIFNDEF, bmodule)
@@ -110,7 +110,7 @@ def section_typedef_impl(buf, i, m):
         buf.trace.untab()
         imports = [c_state_r_name(imp["mach"]) for imp in i["imports"] 
                    if transLLVM.is_stateful(imp["mach"])]
-        variables = [ cx_type(var["type"])+" "+var["id"] for var in i["variables"]]
+        variables = [ cx_type_name_dec(False,var["type"], var["id"]) for var in i["variables"]]
         buf.code(TYPEDEFR,  "struct {" + semicolon(imports + variables) +" }  ", c_state_t_name(m))
         buf.code(TYPEDEFR,   c_state_t_name(m)+" *", c_state_r_name(m))
 
@@ -146,21 +146,35 @@ def c_state_r_name(n):
     else:
         return c_state_r_name(transLLVM.machine(n))
 
-def cx_type(t):
+
+def cx_type_name_dec(is_pointer, t, name):
     '''
     This function is responsible for translation B0 type names to C types
     Input:
+     - is_pointer : a boolean that indicates when it is a pointer 
      - t : a node object that represent the type
+     -name : a node object that represent the id 
     
     '''
     import b2llvm.translate as transLLVM
     transLLVM.check_kind(t, {"Integer", "Bool", "Enumeration", "arrayType"})
+    p=" "
+    if (is_pointer):
+        p = "* "
     if (t == transLLVM.ast.INTEGER):
-        return "int32_t" 
+        return "int32_t" +p+ name
     if (t == transLLVM.ast.BOOL):
-        return "bool"
+        return "bool" +p+ name
     if (t["kind"] == "Enumeration"):
[…]

### 2014-07-13 b2llvm/loadbxml.py

@@ -251,6 +251,7 @@ def load_implementation(dirname, project, loaded, module):
     symast = SymbolTable()
     symast.add("BOOL", ast.BOOL)
     symast.add("INTEGER", ast.INTEGER)
+    #symast.add("ARRAY", ast.ARRAY)
     symimp = SymbolTable()
     _LOADING.add(module)
     tree = ET.parse(path(dirname, module))
@@ -758,7 +759,7 @@ def load_binary_expression(node, symast):
     return load_binary(node, symast, "Binary_Expression",
                        {"+":ast.make_sum, "-":ast.make_diff,
                         "*":ast.make_prod,"(":ast.make_arrayItem,
-                        "mod":ast.make_mod})
+                        "mod":ast.make_mod,"..":ast.make_interval})
 
 def load_nary_expression(node, symast):
     """Load an XML element for a nary expression to an AST node."""
@@ -933,8 +934,8 @@ def get_inv_type(xmlid, xmlinv):
     function = elem[1]
     
     if (function.get("operator")=="-->"):
-        domxml = function[0] 
-        ranxml = function[1]
+        domxml = function[1] 
+        ranxml = function[2]
         res = ast.make_arrayType(domxml,ranxml)
     assert res != None
     return res   
@@ -950,7 +951,19 @@ def load_type(xmlid, symast):
     - Assumes that the XML attribute TypeInfo exists and
     is an identifier with the name of the type.
     '''
-    return symast.get(value(xmlid.find("./Attributes/TypeInfo/Identifier")))
+    node = xmlid.find("./Attributes/TypeInfo/Identifier")
+    if node != None : 
+        return symast.get(value(node))
+    else : 
+        return None
+
+def load_derivedTypes(node):
+    assert node != None 
+    if (node.get("operator") == "-->") :
+        return  ast.make_arrayType(node[0], node[1])
+    else :
+        return
+    
 
 def discard_attributes(exp):
     '''

### 2014-07-13 b2llvm/printer.py

@@ -27,7 +27,7 @@ def term(n):
     elif kind == "Cons":
         return n["id"]
     elif kind == "arrayItem":
-        return (SP + term(n["base"])+"("+ term(n["index"])+ ")")
+        return (SP + term(n["base"])+"("+ term(n["indexes"])+ ")")
     elif kind == "Vari":
         return n["id"]
     elif kind == "Enumerated":

### 2014-07-13 b2llvm/translate.py

@@ -951,28 +951,7 @@ def x_lvalue(buf, n):
     buf.trace.out("Evaluate address for \""+printer.term(n)+"\".")
     t = x_type(n["type"]) + "*"
     if n["kind"] == "arrayItem": 
-        buf.trace.tab()
-        v1,t1 = x_expression(buf, n["base"])
-        buf.trace.out("Variable array (base) \""+n["base"]["id"]+"\" is stored at position "+v1+" of \"%self$\". (arrayItem)")
-        buf.trace.untab()
-        
-        #TODO: Add the index to be loaded
-        #TODO: Create the function LRExp to getting a sequence of selected elements.
-        #commas([ term(x) for x in n["lhs"]])
-        
-        buf.trace.tab() 
-        #for elem in n["index"]:
-        vi,ti = x_expression(buf, n["index"])
-        v = names.new_local()
-        buf.code(LOADI,v,t1+"*",v1,vi)
-        t  = "i32"
-        buf.trace.out("Variable array (index) \""+printer.term(n["index"])+"\" is stored at position "+vi+" of \"%self$\". (arrayItem)")
-        buf.trace.untab()
-        
-        #TODO: Adjust the size o vector to size=(b-a+1)
-        #TODO: Add suport to interval position(p) = (p-a)
-        return (v, t)  
-
+        return x_arrayItem(buf, n)
     elif n["scope"] == "Impl":
         pos=state_position(n)
         v = names.new_local()
@@ -990,6 +969,7 @@ def x_lvalue(buf, n):
     else:
         print("error: unknown scope for variable " + v["id"])
         return ("UNKNOWN", "UNKNOWN")
+
 
 ### TRANSLATION OF CALL INSTRUCTIONS
 
@@ -1210,7 +1190,15 @@ def x_expression(buf, n):
         res = x_term(buf, n)
     elif n["kind"] == "Cons":
         res = x_expression(buf, n["value"])
+    elif n["kind"] == "arrayItem": 
+        p,_ = x_arrayItem(buf, n)
+        v1 = names.new_local()
+        buf.code(LOADD, v1, _ , p)
+        res = (v1,_)
     else:
+        print("error: unknown expression kind")
+        buf.code("<error inserted by b2llvm>")
+        assert False, "error: unknown expression kind"
         res = ("","")
     buf.trace.untab()
     buf.trace.outu("The evaluation of \""+ellipse(printer.term(n))+"\" is \""+res[0]+"\".")
[…]

### 2014-07-13 beta_tests/Counter/Counter_using_CUTest 1.5/AllTests.c

@@ -1,80 +1,0 @@
-#include "CuTest.h"
-#include <stdio.h>
-#include <string.h>
-#include "Counter.h"
-
-    
-    CuSuite* StrUtilGetSuite();
-    
-    counter$state$ counter;
-
-
-  void Test_oper_inc_testCase1(CuTest *tc) {
-	    // setting manually
-	counter.value = 0 ;
-	counter.error = false;
-	    // or setting using the initialization, you can chose one option
-	counter$INITIALISATION(&counter);
-	    // calling the function
-	counter$inc(&counter);
-    
-	bool condition = !( counter.error == true) || (counter.value == 2147483647);
-	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
-
-    }
-
-
-    void Test_oper_inc_testCase2(CuTest *tc) {
-	    // setting manually
-	counter.value = 3 ;
-	counter.error = true;
-    	
-	// calling the function
-	counter$inc(&counter);
-    	
-	// Neste caso, o invariante está sendo quebrado então a formula não deveria ser negada ?
-	bool condition =  !( counter.error == true) || (counter.value == 2147483647) ;
-	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
-
-    }
-
-
-    void Test_oper_zero_testCase1(CuTest *tc) {
-	    // setting manually
-	counter.value = 3 ;
-	counter.error = true;
-    	
-	// calling the function
-	counter$zero(&counter);
-    
-	bool condition = !( counter.error == true) || (counter.value == 2147483647);
-	CuAssert_Line(tc, __FILE__, __LINE__ , "Invariant invalid", condition);
-
-    }
-
-   
-    CuSuite* testsSuite() {
-        CuSuite* suite = CuSuiteNew();
-	SUITE_ADD_TEST(suite, Test_oper_inc_testCase1);
-        SUITE_ADD_TEST(suite, Test_oper_inc_testCase2);
-	SUITE_ADD_TEST(suite, Test_oper_zero_testCase1);
-
-        return suite;
-    }
-
-    
-    void RunAllTests(void) {
-        CuString *output = CuStringNew();
-        CuSuite* suite = CuSuiteNew();
-        
-        CuSuiteAddSuite(suite, testsSuite());
-    
-        CuSuiteRun(suite);
-        CuSuiteSummary(suite, output);
-        CuSuiteDetails(suite, output);
[…]

### 2014-07-13 beta_tests/Counter/Counter_using_CUTest 1.5/Exemplo_gerado/comp_exec.sh

@@ -1,8 +1,0 @@
-# Before you must generate the counter.llvm by this command:
-# ./b2llvm.py -v -m comp counter counter.llvm bxml project.xml
-clang -S -emit-llvm -c CuTest.c AllTests.c
-llc-mp-3.5  counter.llvm
-llc-mp-3.5  CuTest.ll 
-llc-mp-3.5  AllTests.ll
-clang -o executable counter.llvm.s  CuTest.s AllTests.s
-./executable 

### 2014-07-13 beta_tests/Counter/Counter_using_CUTest 1.5/counter.h

@@ -1,45 +1,0 @@
-#ifndef _counter_h
-#define _counter_h
-
-#include <stdint.h>
-#include <stdbool.h>
-#ifdef __cplusplus
-extern "C" {
-#endif /* __cplusplus */
-    
-    
-    /* Clause SETS */
-    
-    /* Clause CONCRETE_VARIABLES */
-    
-    
-    /* Clause CONCRETE_CONSTANTS */
-    /* Basic constants */
-    /* Array and record constants */
-
-   // Passo 2 - Declarando os typedef 
-   typedef struct  { int32_t value; bool error; } counter$state$;
-   typedef counter$state$ * counter$ref$;
-
-
-
-   // Passo 1 - Adicionando a renomeiação das variáveis e usando o parâmetro de contexto.
-    extern void counter$INITIALISATION(counter$ref$ self);
-   // Passo 3 - Remove as declarações das variáveis antigas no arquivo counter.c 
-   // Passo 4 - Atualiza as variáveis no arquivo de teste para usar a referencia a estrutura de estado. Esta mesma sequência de passos segui no arquivo de teste.
-   // Passo 5 - Faz a reomeação das funções nos arquivos .c e .h. Exemplo: counter__zero -> counter$zero. Precisa somente counter__-> counter$
-   // Passo 6 - Comparar a compatilibilidade entre o arquivo .c baseado no atelierb e o llvm criando com B2llvm
- 
-
-    /* Clause OPERATIONS */
-    
-    extern void counter$zero(counter$ref$ self);
-    extern void counter$inc(counter$ref$ self);
-    extern void counter$get(counter$ref$ self, int32_t *res);
-    
-#ifdef __cplusplus
-}
-#endif /* __cplusplus */
-
-
-#endif /* _counter_h */

### 2014-07-13 coverage.sh

@@ -19,6 +19,11 @@ $COV_RUN ./b2llvm.py -m proj Division init-Division.llvm bxml project.xml
 $COV_RUN ./b2llvm.py -m comp  Prime Prime.llvm bxml project.xml
 $COV_RUN ./b2llvm.py --trace --verbose  -m proj Prime init-Prime.llvm bxml project.xml
 
+echo "Generating bubble example"
+$COV_RUN ./b2llvm.py -m comp  bubble bubble.llvm bxml project.xml
+$COV_RUN ./b2llvm.py --trace --verbose  -m proj bubble init-bubble.llvm bxml project.xml
+echo "Generated code!"
+
 #$COV_RUN ./b2llvm.py -m comp Rooms  Rooms.llvm bxml project.xml
 #$COV_RUN ./b2llvm.py -m proj Rooms init-Rooms.llvm bxml project.xml
 
@@ -54,4 +59,8 @@ echo "Coverage report is now available."
 
 echo "Generating diff report "
 for file in *.h *.llvm; do diff "$file" "expected_code/${file##*/}">expected_code/REPORT_"$file".diff; echo "Diff for ${file##*/}:"; cat expected_code/REPORT_"$file".diff;   done
+find expected_code -size -1c -name "*.diff" -exec rm -f {} \;
 echo "Report generated"
+
+
+

### 2014-08-04 DOCUMENTATION/SLIDES/20140422-LORIA/ex.c
LORIA FM seminar

@@ -1,0 +1,17 @@
+extern void gcd(int *, int, int);
+typedef struct {
+  int s;
+  int n;
+  int d;
+} * Trat;
+void norm(Trat P)
+{
+  int g;
+  if (P->d == 0)
+    goto end;
+  gcd(&g, P->n, P->d);
+  P->n /= g;
+  P->d /= g;
+end:
+  return;
+}

### 2014-08-04 DOCUMENTATION/SLIDES/20140422-LORIA/figures_minted/gen-figures.sh

@@ -1,0 +1,137 @@
+# This script crops the generated PDFs # It uses cpdf from http://community.coherentpdf.com #!/bin/bash
+set -x
+alias CPDF=$HOME/bin/cpdf-binaries-master/OSX-Intel/cpdf  
+
+dx="4mm"
+dy="6mm"
+width="60mm"
+
+file="inc_llvm.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 32mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file generated"
+
+width="40mm"
+file="inc_c.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 20mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file generated"
+
+dx="0mm"
+width="100mm"
+
+file="count_i.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 33mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file generated"
+
+file="count_i_llvm_typedef.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 7mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file.pdf generated"
+
+file="count_i_llvm_interface.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 16mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file generated"
+
+file="count_i_llvm_implementation.pdf"
+make -s $file
+CPDF -crop "$dx $dy $width 114mm" $file -o tmp.pdf 2> /dev/null
+CPDF -frombox /CropBox -tobox /MediaBox tmp.pdf -o x_$file 2> /dev/null
+echo "x_$file generated"
+
+file="wd_i.pdf"
+make -s $file
[…]

### 2014-08-22 DOCUMENTATION/SLIDES/20140422-LORIA/figures_minted/with_trace.tex

\documentclass[10pt,a4paper]{article}
\include{codeFormat}

\begin{document}

\begin{llvmcode}
;;1 The type for the state of "counter" is defined in "counter_i",
;; it is an aggregate such that:
;;1.1 Position "0" represents variable "value".
;;1.2 Position "1" represents variable "error".
%counter$state$ = type {i32, i1}
;;2 The type for references to state encodings of "counter" is:
%counter$ref$ = type %counter$state$*
;;3 The function implementing initialisation for "counter" is
;;  named "@counter$init$" and has the following parameters:
;;3.1 "%self$": address of LLVM aggregate storing state of "counter";
define void @counter$init$(%counter$ref$ %self$) {
\end{llvmcode}

\end{document}

### 2014-08-22 DOCUMENTATION/SLIDES/20140422-LORIA/presentation.tex

\documentclass[12pt]{beamer}
\usecolortheme{beaver}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{thumbpdf}
\usepackage{wasysym}
\usepackage{ucs}
\usepackage[utf8]{inputenc}
\usepackage{pgf,pgfarrows,pgfnodes,pgfautomata,pgfheaps,pgfshade}
\usepackage{verbatim}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
%\usepackage{minted} % Added highlighting in source code

% Enable/Disable notes - It creates a white complementary slide to help the presenter save the spech
%\setbeameroption{show notes}
%\setbeamertemplate{note page}[plain]

\setbeamertemplate {navigation symbols}{} % Remove buttons in footnote

\pdfinfo
{
  /Title       (b2llvm: B developments onto the LLVM)
  /Creator     (David Déharbe)
  /Author      (David Déharbe)
}

\title{\bllvm: B developments onto the LLVM}
\author{David Déharbe\footnote{j.w.w. R. Bonichon, T. Lecomte, V. Medeiros}}
\institute[UFRN]{Federal University of Rio Grande do Norte, Natal, Brazil}
\date{May $20^\text{\scriptsize th}$, 2014}

\input{commands}

\begin{document}

% To enable numbered captions
\setbeamertemplate{caption}[numbered]

\frame{\titlepage}

\newcommand<>{\highlighton}[1]{%
  \alt#2{\structure{#1}}{{#1}}
}

\newcommand{\icon}[1]{\pgfimage[height=1em]{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% Content starts here %%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\begin{frame}

  \frametitle{Introduction}

\begin{minipage}{.4\textwidth}
\begin{center}
  \includegraphics[width=\textwidth]{figures/b-classic.pdf}
\end{center}
\end{minipage}
\begin{minipage}{.4\textwidth}
  \begin{itemize}
%   \item B-method: design of software \emph{components}
%     \begin{itemize}
%       \item From requirements: first-order logic, sets, integer arithmetics
%       \item To implementation: simple imperative programming language (no
%         recursion, no dynamic memory allocation).
%       \item Through: stepwise refinements and proof.
%       \end{itemize}
%   \item Atelier-B is an IDE that implements all these aspects of the B method.
%   \item There is no compiler for the B implementation language.
%   \item Instead, code in languages such as C and Ada is synthesized from
%     B implementations.
%   \item Source code synthesis is not subject to proof.
  \item \emph{Problem:\/} certify source code synthesis in Atelier-B
  \item \emph{Approach:\/} redundancy, by developing a new code generator to
    compare with.
  \end{itemize}
\end{minipage}
\end{frame}

\begin{frame}

  \frametitle{Introduction}

\begin{center}
  \includegraphics[width=\textwidth]{figures_minted/x_timer_i.pdf}
\vspace*{2mm}
  \includegraphics[width=\textwidth]{figures/ab-errors.pdf}
\end{center}

\begin{itemize}
\item \emph{Problem:\/} arbitrary limitations in current code synthesis
\item \emph{Approach:\/} remove limitations, keeping in mind application
  context.
\end{itemize}
\end{frame}

\begin{frame}

  \frametitle{The \bllvm project}

\begin{minipage}{.34\textwidth}
  \includegraphics[width=\textwidth]{figures/b-llvm.pdf}
\end{minipage}
\begin{minipage}{.56\textwidth}
  \includegraphics[width=\textwidth]{figures/b-certify.pdf}
\end{minipage}

  \begin{itemize}
  \item Code synthesis for B targeting LLVM.
  \item Working prototype.
  \end{itemize}

\end{frame}

\section{LLVM}

\begin{frame}
\frametitle{LLVM}

\begin{quote}
``The LLVM Project is a collection of modular and reusable compiler and toolchain technologies.'' (\url{http://www.llvm.org})
\end{quote}

\begin{itemize}
\item Used in many commercial and open-source projects
\begin{itemize}
\item Adobe, Apple, Cray, Intel, NVIDIA 
\end{itemize}
\item Used in academic research
\begin{itemize}
\item \emph{Formalizing the LLVM Intermediate Representation for Verified Program Transformations}. J. Zhao, S. Nagarakatte, M. Martin, S. Zdancewic. POPL 2012.
\end{itemize}
\item backbone of LLVM: an \emph{assembly language}
\begin{itemize}
\item an intermediate representation (IR)
\item generated by compiler front-ends (parsing, type analysis) 
\item consumed, transformed by back-ends (optimization, target code
  generation).
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{LLVM-IR}

\begin{itemize}
\item Example
\begin{center}
\begin{tabular}[t]{cc}
  C & LLVM \\
  \includegraphics{figures_minted/x_inc_c.pdf}&
  \includegraphics{figures_minted/x_inc_llvm.pdf}
\end{tabular}
\end{center}
\item Single-static assignment
\item Strongly typed
\item Modular
\end{itemize}

\end{frame}

\begin{frame}{LLVM module items}

\begin{itemize}
\item type declaration

\noindent\includegraphics{figures_minted/x_typedecl.pdf}
\item Type definitions

\noindent\includegraphics{figures_minted/x_typedef.pdf}
\item constant declaration

\noindent\includegraphics{figures_minted/x_constdecl.pdf}
\item constant definition

\noindent\includegraphics{figures_minted/x_constdef.pdf}
\item function declaration

\noindent\includegraphics{figures_minted/x_fundecl.pdf}

\item function definition...
\end{itemize}

\end{frame}

\begin{frame}{LLVM function definitions}

\noindent\includegraphics[width=\textwidth]{figures_minted/x_fundef1.pdf}

\end{frame}

\begin{frame}

  \frametitle{Some remarks on the B method}

  \begin{itemize}
  \item Construction of software \emph{modules\/}.
  \item A module may be
    \begin{itemize}
    \item a \emph{developed} module: implementation derived in B.
    \item a \emph{base} module: implementation derived outside of B.
    \end{itemize}
  \item A developed module has a B specification (machine) and a B
    implementation.
  \item A base module has a B specification (machine).
  \item A B module may \emph{import} other modules.
  \item A root B module may (transitively) contain several instances of other
    B modules.
  \end{itemize}

\end{frame}

\begin{frame}

  \frametitle{Some general requirements}

  \begin{itemize}
  \item No dynamic memory allocation
    \begin{itemize}
    \item This is a requirement in some applications (often
      safety-critical systems).
    \item No need for a third-party library.
    \end{itemize}
  \item Separate code generation
    \begin{itemize}
    \item This is a requirement for large projects.
    \item Internal changes in a B module shall only require generating code
      for that module.
    \item Changes in the interface of a B module requires generating code
      for that module and those other modules that use it.
    \end{itemize}
  \end{itemize}

\end{frame}

\begin{frame}{Code generation}

  \begin{itemize}
  \item Specified as a set of code generation rules\footnote{You will be spared a detailed presentation of the rules.}

    \includegraphics[width=.75\textwidth]{rule.pdf}
  \item Validated via 
    \begin{itemize}
    \item trial encoding
    \item rapid prototyping
      \begin{itemize}
      \item read XML files produced by a third-party tool
      \item Python
      \end{itemize}
    \end{itemize}
  \end{itemize}

\end{frame}

\begin{frame}{B constructs}

  \begin{itemize}
    \item Constants, variables
    \item Module instantiation (imports) \only<2>{\alert{$\leftarrow$ More on this}}
    \item Imperative constructs: conditionals, loop, local variables, assignments, operation calls.
    \item Arithmetic and relational expressions, conditionals
    \item Integers, Booleans, Enumerations.
    \item \textcolor{red}{-} lambda expressions, arrays, records,
    \item \textcolor{red}{- -} machine parameters.
  \end{itemize}
\end{frame}

\begin{frame}

  \frametitle{B projects and code generation}

  \includegraphics[height=.4\textheight]{B-project.pdf}

  To build the project LLVM code:
  \begin{description}
  \item[b2llvm] \alert{component} mode:
    \begin{itemize}
      \item \B{M0} \B{M2} \B{M3} $\Longrightarrow$ \llvm{m0.ll} \llvm{m2.ll} \llvm{m3.ll} 
    \end{itemize}
  \item[b2llvm] \alert{project} mode:
    \begin{itemize}
      \item \B{M0} $\Longrightarrow$ \llvm{init-m0.ll}
    \end{itemize}
  \item[LLVM] optimize, generate assembly, link
    \begin{itemize}
      \item \llvm{init-m0.ll} \llvm{m0.ll} \llvm{m2.ll} \llvm{m3.ll} \textcolor{red}{\llvm{m1.o}} $\Longrightarrow$ \llvm{a.out}
      \end{itemize}
  \end{description}
\end{frame}

\begin{frame}

  \frametitle{Bricks of LLVM code}

{
\newcommand{\myind}{\hspace*{2em}}
\small
\begin{minipage}{.9\textwidth}
\noindent\textbf{type def:}
A structure type (one item per state variable and import): 
\textcolor{blue}{\llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}} \\
\\
\noindent\textbf{type decl:}
\textcolor{blue}{\llvm{\%M\$state\$ = type opaque}} \\
\\
\noindent\textbf{type ref:}
A pointer type: 
\textcolor{blue}{\llvm{\%M\$ref\$ = type \%M\$state\$*}} \\
\\
\noindent\textbf{function decl:} 
An initialization function:\\
\textcolor{blue}{\llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}} \only<2>{\alert{$\leftarrow$ address of module instance}}\\
One function for each operation in the module: \\
\textcolor{blue}{\llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})}}\\
\\
\noindent\textbf{function def:}
(for developed modules): \\
\textcolor{blue}{\llvm{define void @M\$op(\%M\$ref\$ \%self\$, \ListOf{\nt{param}}) \{}} \\
\myind \textcolor{blue}{\llvm{\ListOf{\nt{block}}}} \\
\myind \textcolor{blue}{\llvm{exit: ret void}} \\
\textcolor{blue}{\llvm{\}}}
\end{minipage}
}

\end{frame}

\begin{frame}

  \frametitle{Example 1}

  \begin{center}
    \includegraphics{figures_minted/x_count_i.pdf}
  \end{center}

  \begin{itemize}
    \item type definition

{\small \textcolor{blue}{\llvm{\%counter\$state\$ = type \{ i32, i1 \}}}}
    \item reference type

{\small \textcolor{blue}{\llvm{\%counter\$ref\$ = type \%counter\$state\$*}}}
    \item function declarations:

{\small
      \textcolor{blue}{\llvm{declare void @counter\$init\$(\%counter\$ref\$)}}

      \textcolor{blue}{\llvm{declare void @counter\$inc(\%counter\$ref\$)}}

      \textcolor{blue}{\llvm{declare void @counter\$get(\%counter\$ref\$, i32*)}}
}
    \end{itemize}
\end{frame}

\begin{frame}

  \frametitle{Example 1 (a function definition)}

  \includegraphics{figures_minted/count_i-inc.pdf}

  \vspace*{-10mm}
  \includegraphics{figures_minted/count_i_llvm_inc.pdf}

\end{frame}

\begin{frame}

  \frametitle{Example 1 (another function definition)}

  \includegraphics{figures_minted/count_i-get.pdf}

  \vspace*{-10mm}
  \includegraphics{figures_minted/count_i_llvm-get.pdf}

\end{frame}

\begin{frame}

  \frametitle{Example 2}

  \begin{center}
    \includegraphics[height=.5\textheight]{figures_minted/x_wd_i.pdf}
  \end{center}

  \begin{itemize}
    \item type definition

{\small \textcolor{blue}{\llvm{\%wd\$state\$ = type \{ \%count\$ref\$ \}}}}
    \item function declarations:

    \includegraphics{figures_minted/x_wd_i_llvm-fun-decl.pdf}

    \end{itemize}
\end{frame}

\begin{frame}

  \frametitle{Example 2 (definition of function for initialization)}

  \includegraphics{figures_minted/x_wd_i_llvm-init-def.pdf}

\end{frame}

\begin{frame}

  \frametitle{Putting it all together: component mode}

  \begin{itemize}
  \item type definition
    \begin{itemize}
    \item need type references of imported modules
    \end{itemize}
  \item function definitions
    \begin{itemize}
    \item need type references of transitively imported
      modules (initialization)
    \item need function declarations of imported modules
      (operation calls, initialization)
    \end{itemize}
  \end{itemize}

\end{frame}

\begin{frame}

  \frametitle{Putting it all together: project mode}

  \begin{itemize}
    \item to create instances of all the instances of all the
      modules:
      \begin{itemize}
        \item type definition of module
        \item type reference of module
        \item type definitions of transitively imported modules
        \item type references of transitively imported modules
      \end{itemize}
      For each instance \B{Q} of a module \B{M}

      {\small 
        \textcolor{blue}{\llvm{@\$Q[path] = common global @\$Q\$state\$ zeroinitializer}}}

    \item to initialize the system:
      \begin{itemize}
      \item function declaration (initialization)
        of the module
      \end{itemize}

      {\small 
        \textcolor{blue}{\llvm{define void @\$init\$(void) \{}}

        \textcolor{blue}{\llvm{\hspace*{.5cm} call void @M\$init\$(\$M, ...)}}

        \textcolor{blue}{\llvm{\hspace*{.5cm} ret void}}

        \textcolor{blue}{\llvm{\}}}
        }

  \end{itemize}

\end{frame}

\begin{frame}{Putting it all together: project for example 2}

  \includegraphics{figures_minted/x_wd_llvm_main.pdf}

\end{frame}

% \begin{frame}{extra}
%   For each B module \B{M}, generate code for
%   \begin{description}
%   \item[typedef] An LLVM record type \llvm{@M\$state\$} to represent the state
%     of the module, with
%     \begin{itemize}
%     \item one record field for each state variable;
%     \item one record field for each imported module.
%     \end{itemize}
%   \item An LLVM pointer type \llvm{@M\$ref\$} to represent pointers to 
%     \llvm{@M\$state\$}.
%   \item A LLVM function \llvm{@M\$init\$} to initialize module instances,
%     \begin{itemize}
%     \item parameter: the (representation) of the instance to be initialized
%     \item extra parameters: the (representation) of the component instances.
%     \end{itemize}
%   \item one LLVM function \llvm{@M\$op\$} for each operation \B{op} in \B{M}.
%     \begin{itemize}
%     \item parameter: the (representation) of the called instance.
%     \item extra parameters: the inputs and the outputs of the operation.
%     \end{itemize}
%   \end{itemize}

% \end{frame}

\begin{frame}{Verification, validation}

\begin{itemize}
  \item Formal verification:
    \begin{itemize}
    \item Formalize LLVM-IR (available)
    \item Formalize B (partly available)
    \item Formalize and verify code generation rules (to be done)
    \end{itemize}
  \item Assertions in generated code.
    \begin{itemize}
    \item Implement assert in LLVM.
    \item Define code generation rules for full conditions (not only
      implementable conditions).
    \end{itemize}
  \item Test generated code (ongoing work)
  \item Generate debugging aids (some done)
  \item Traceability (done)
\end{itemize}

\end{frame}

\begin{frame}{Traceability: an example}

\includegraphics{figures_minted/x_with_trace.pdf}

\end{frame}

\section{Conclusion}

\begin{frame}{Conclusion}

\begin{itemize}
\item \bllvm is open-source (\url{http://www.b2llvm.org/b2llvm}).
\item \bllvm is compatible with Atelier-B 4.2 (availability: June 2014).
\item Available validation strategies: 
\begin{itemize}
\item output code may be annotated.
\item debugging code may be generated.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Ongoing work}
\begin{itemize}
\item lambda terms;
\item support to type constructors: arrays, records;
\item integration with test-generation tool.
\end{itemize}

\end{frame}

\begin{frame}

\begin{center}
\huge ?
\end{center}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Syntax of the target LLVM-IR: module elements}

  \begin{center}
    \begin{tabular}{rcl}
      \nt{module} & ::= & \ListOf{\nt{item}} \\
      \nt{item} & ::= &  \nt{type\_def} \ALT \nt{const\_decl} \ALT \nt{const\_def} \ALT \nt{var\_def}\\
      & & \ALT \nt{function\_decl} \ALT \nt{function\_def} \\
      \nt{type\_def} & ::= & \nt{name} \llvm{=} \llvm{type} \nt{type} \\
      \nt{type} & ::= & \llvm{void} \ALT \nt{itype} \ALT \llvm{\{} \ListOf{\nt{type}} \llvm{\}} \ALT \nt{type}\llvm{*} \ALT \llvm{opaque} \\
      \nt{const\_decl} & ::= & \nt{name} \llvm{=} \llvm{external} \llvm{constant} \nt{type} \\
      \nt{const\_def} & ::= & \nt{name} \llvm{=} \llvm{constant} \nt{type} \nt{iliteral} \\
      \nt{var\_def} & ::= & \nt{name} \llvm{=} \llvm{common} \llvm{global} \nt{type} \llvm{zeroinitializer} \\
      \nt{function\_decl} & ::= & \llvm{declare} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{type}} \llvm{)}\\
      \nt{function\_def} & ::= & \llvm{define} \nt{type} \nt{name} \llvm{(} \ListOf{\nt{param}} \llvm{)} \llvm{\{} \ListOf{\nt{block}} \llvm{\}} \\
      \nt{param} & ::= & \nt{type} \nt{name} \\
      \nt{block} & ::= & \nt{lbl} \llvm{:} \ListOf{\nt{inst}}
    \end{tabular}
  \end{center}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Syntax of the target LLVM-IR: instructions}

  \begin{center}
    \begin{tabular}{rcl}
      \nt{inst} & ::=  & \nt{name} \llvm{=} \llvm{alloca} \nt{type} \\
      & \ALT & \nt{name} \llvm{=} \nt{arith} \nt{itype} \nt{exp} \llvm{,} \nt{exp} \\
      & \ALT & \nt{name} \llvm{=} \llvm{icmp} \nt{rel} \llvm{i1} \nt{exp} \llvm{,} \nt{exp}\\
      & \ALT & \nt{name} \llvm{=} \llvm{call} \nt{type} \llvm{(} \ListOf{\nt{arg}} \llvm{)} \\
      & \ALT & \nt{name} \llvm{=} \llvm{getelementptr} \nt{type} \llvm{*} \nt{exp}\llvm{,} \nt{index}\llvm{,} \nt{index}  \\
      & \ALT & \nt{name} \llvm{=} \llvm{load} \nt{type} \nt{exp} \\
      & \ALT & \llvm{store} \nt{type} \nt{exp}, \nt{type} \llvm{*} \nt{exp} \\
      & \ALT & \llvm{br} \llvm{i1} \nt{exp} \llvm{,} \llvm{label} \nt{lbl} \llvm{,} \llvm{label} \nt{lbl} \\
      & \ALT & \llvm{br} \llvm{label} \nt{lbl} \\
      & \ALT & \llvm{switch} \nt{type} \nt{exp} \llvm{,} \llvm{branch} \nt{lbl} \llvm{[} \ListOf{\nt{branch}} \llvm{]} \\
      & \ALT & \llvm{ret} \nt{type} \nt{exp}\\
      & \ALT & \llvm{ret} \llvm{void} \\
    \end{tabular}
  \end{center}

\end{frame}

\begin{frame}[fragile]
  \frametitle{Syntax of the target LLVM-IR: expressions, misc.}

  \begin{center}
    \begin{tabular}{rcl}
      \nt{arith} & ::= & \llvm{add} \ALT \llvm{sub} \ALT \llvm{mul} \ALT \llvm{sdiv} \ALT \llvm{srem} \\
      \nt{rel} & ::= & \llvm{eq} \ALT \llvm{ne} \ALT \llvm{sgt} \ALT \llvm{sge} \ALT \llvm{slt} \ALT \llvm{sle} \\
      \nt{exp} & ::= & \nt{name} \ALT \nt{iliteral} \ALT \\
& & \llvm{getelementptr} \llvm{(} \nt{type} \nt{exp} \llvm{,} \nt{index} \llvm{,} \nt{index} \llvm{)} \\
      \nt{index} & ::= & \nt{itype} \nt{iliteral} \\
      \nt{branch} & ::= & \nt{iliteral} \nt{iliteral} \nt{lbl} \\
      \nt{arg} & ::= & \nt{type} \nt{exp}
    \end{tabular}
  \end{center}

Main missing construct: arrays.

\end{frame}

\begin{frame}{Example 1 (standalone module)}

  \begin{center}
    \includegraphics{figures_minted/x_count_i.pdf}
  \end{center}
\end{frame}

\begin{frame}{Example 2 (composed module)}

  \begin{center}
    \includegraphics{figures_minted/x_wd_i.pdf}
  \end{center}
\end{frame}

\begin{frame}

  \frametitle{B projects and code generation}

  \includegraphics[height=.4\textheight]{B-project.pdf}
  \begin{enumerate}
  \item modules access other modules (base or developed) through their interface
    $\Longrightarrow$ need to generate LLVM for module \emph{interface}.
  \item Memory for each instance of each \B{Mi} must be allocated statically
    $\Longrightarrow$ global variables.
  \item Instances should be created only when the full project is compiled
    $\Longrightarrow$ two code generation modes: \emph{component} and 
    \emph{project}.
  \end{enumerate}
\end{frame}

\end{document}
